### 2022-11-04 Docs/manuscript.Rnw
start sweaving

%%% start preambling . . .  %%%
\documentclass{article}

% required 
\usepackage{Sweave}
\usepackage{graphicx}

% https://tex.stackexchange.com/questions/60209/how-to-add-an-extra-level-of-sections-with-headings-below-subsubsection
\usepackage{titlesec}
\setcounter{secnumdepth}{4}

\titleformat{\paragraph}
{\normalfont\normalsize\bfseries}{\theparagraph}{1em}{}
\titlespacing*{\paragraph}
{0pt}{3.25ex plus 1ex minus .2ex}{1.5ex plus .2ex}

% recommended! Uncomment the below line and change the path for your computer!
\SweaveOpts{prefix.string=/Users/alina/Documents/git/localadaptclim/Docs/figure_ms/demoFig, eps=FALSE} 
%put your figures in one place! Also, note that here 'figures' is the folder and 'demoFig' is what each 
% figure produced will be titled plus its number or label (e.g., demoFig-nqpbetter.pdf')

% make your captioning look better
\usepackage[small]{caption}
\setlength{\captionmargin}{30pt}
\setlength{\abovecaptionskip}{0pt}
\setlength{\belowcaptionskip}{10pt}

% optional: muck with spacing
\topmargin -1.5cm        
\oddsidemargin 0.5cm   
\evensidemargin 0.5cm  % same as oddsidemargin but for left-hand pages
\textwidth 15.59cm
\textheight 21.94cm 
% \renewcommand{\baselinestretch}{1.5} % 1.5 lines between lines
\parindent 0pt		  % sets leading space for paragraphs
% optional: cute, fancy headers
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhead[LO]{November 2022}
\fancyhead[RO]{Manuscript}
% more optionals! %
\usepackage[hyphens]{url} % this wraps my URL versus letting it spill across the page, a bad habit LaTeX has

%%% end preambling. %%%

\begin{document}
\SweaveOpts{concordance=TRUE} % For RStudio hiccups
\title{{\huge Changes and trends in budburst and leaf flush across Europe and North America} \\A meta-analysis of local adaptation in spring phenology studies}
\author{Ziyun Zeng \& E. M. Wolkovich}
\date{November 2022}
\maketitle

\newpage

\section*{Abstract}
\section{Introduction}

\begin{itemize}
\item Define spring event, fall event
\item Background
\item Research question (links to results)
\end{itemize}

\section{Methods}
\begin{itemize}
\item Literature search
\item Provenance location and spring/fall DOY data gathering using ImageJ
% note that we primarily focused on papers with spring data
  \begin{itemize}
\item Read AB 2015 and A 2013, took paper with spring doy and provenance coordinates from them
\item Searched online for more papers (explain criteria here)
\item Out of 59 papers related to common garden/spring phenology that I read, only 22 can supply data that we need; Out of the 22, 4 studies have provenance and gardens on different continents, which were subsequently archived (can document in supplements)).
\end{itemize}
\item Present how many gardens, species, provenance we have using a table, which also documents the type of spring event they looked at, year, etc. 
\item Insert mapped locations of studies
\item Climate data gathering
  \begin{itemize}
  \item Simple climate matrix (MAT \& MSP/MAP) gathering
  \begin{itemize}
  \item ClimateNA by Dr. Tongli Wang
  \item Climate Information Tool by FAO
  \end{itemize}
  \item Daily climate data gathering: Gridded daily climate data downloaded for all provenances and gardens: We hope that coarse metrics such as latitude and MAT ultimately represent how similar the climates are between the provenances and gardens in times that matter for the events. If climates are very similar, then we would expect similar timings [add more here].
To this end, we estimated climate overlap in relevant months: For spring events, we considered overlap across March to May.
  \begin{itemize}
  \item North American daily climate data from Daymet R package
  \item Europe daily climate data (gridded: NetCDF-4) retrieved from E-OBS 
  \end{itemize}
  \end{itemize}
  \item Estimate climate overlap using daily climate data using Overlap R package
  \item Calculated GDDs, spherical distances
  \item Corrected for DOY (taking the difference), will put results in supplement
  \item Mixed effects modelling: We used hierarchical Bayesian models where we partially pooled by species. We estimated effects of continent and species type (angiosperm VS gymnosperm) from posterior estimates, in relation to single predictors (including provenance latitude, MAT) and two predictors (climate overlap percentage and standard deviation).
\end{itemize}

\section{Results}
%AZZNov04: all subsection titles are placeholders

\subsection{Provenance latitude and MAT do not affect spring event timing}

Overall, our models show that spring events are not related to provenance latitude or mean annual temperature in North America, and only weakly related in Europe (Figure \#). 
\\
\verb@Alina Says:@ Need to elaborate here. Note that all figures are subject to update for better resolution. I might add climate overlap plots here too.

\begin{figure}[!h] 
    \centering
 \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/springfall_latmat.png}
    \caption{Spring event Day of Year (DOY) in relation to provenance latitude and MAT, coded by symbol for species and color for garden with linear fits from hierarchical Bayesian models. Spring events shown on top and fall event at the bottom.}
    \label{figure:springfall_latmat}
\end{figure}

\newpage

\subsubsection {Similar results across provenance latitude, latitude difference, and distance}
\verb@Alina says: need to find a name for each metric and stick to it.@
Along with (a) provenance latitude, we also looked at how spring event timing is related to (b) the absolute difference between provenance latitude and garden latitude , as well as (c) the spherical distance between the garden and each provenance. All three metrics depict little to no relationship between spring event timing and the geographical location of the provenance \verb@(Supplement Figure #)@.
% can show that figure in supplement

\subsubsection {Climate overlap does not predict event dates much better than provenance latitude or MAT}

While comparing how similarity in climate relates to event dates, we observed very weak effects of climate overlap on spring events, nearly identical across angiosperms and gymnosperms. Fall events diverge as climate overlap declines for both angiosperms and gymnosperms, but slightly more strongly for gymnosperms.
% add figure here

\subsubsection {Strong relationship between provenance latitude, MAT, and GDDs}

I am unsure about how to commment on what we are seeing in this plot. There is a strong relationship between GDDs of each event day recorded and the latitude and MAT, but isn't that self-apparent? also I don't know how to link (Fig. \#) in text with the corresponding figure being displayed.

% all figures will be updated (color, font size, resolution)
\begin{figure}[!h] 
    \centering
 \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/gdd_ms.png}
    \caption{Growing Degree Days (GDD)on each day of spring event in relation to provenance latitude and MAT, coded by symbol for species and color for garden with linear fits from hierarchical Bayesian models.}
    \label{figure:gdd}
\end{figure}

\newpage

\subsection{Stronger clines in fall events observed in North America than Europe}

% old graphs can be found in C:\Users\alina\Documents\git\localadaptclim\Output\plotMay15_two predictors_experiment
% refer to analyses\script_model_continent_spp_type_effect
% https://github.com/lizzieinvancouver/localadaptclim/issues/16

% need to replot ofc

%  extract the mean estimate of the slope and report that instead of ~4 or such
% hmmm i forgot the code :)))

We find that fall events (budset, leaf senescence, leaf abcission) advance strongly with provenance latitude and mean annual temperature, meaning fall events are earlier where provenance mean annual temperature is lower (higher, more northern latitudes). This relationship, however, is observed mostly in North America where fall events advance 4.2 days per degree we move north, or 6.4 days when the MAT decreases by one $^{\circ}$C. In Europe, such relationship is weak: advance 0.5 days per degree we move north, or 0.6 days when the MAT decreases by one $^{\circ}$C.
% add figure here

\subsection{Effects of MAT on spring events diverge across angiosperms and gymnosperms}

Effects of provenance latitude on both spring and fall events are similar across angiosperms and gymnosperms. However, effects of MAT on spring events weakly diverge: spring events get earlier as MAT increases in angiosperms and delay as MAT increases in gymnosperms, except for \emph{Pseudotsuga menziesii}. Fall events delay in warmer locations for both species types, but slightly more so for gymnosperms (3.7 days VS. 6.2 days).

\section{Discussion}

The weak relationship between spring event dates and provenance latitude and MAT that we find in European studies might be explained by the higher extent of climate overlap in those studies. The more similar the climate is between provenances and gardens, the less difference between spring event dates.
\\

The inconsistent and weak clines in spring events that we found suggest high plasticity in spring phenology across continents and species. Fall events, on the other hand, exhibit stronger clines which suggest more local adaptation, especially in North America. Overall, our results predict that warming springs will continue to be tracked more closely phenologically by trees than warming fall temperatures.
\\

In contrast to spring events, we found strong latitudinal clines in fall events across both continents, with local adaptation appearing much stronger in North America than in Europe. Our results show that spring events are highly plastic, and thus may shift with warming, but data on more species and greater information on important factors, such as their geographic location in relation to their origins and elevation, are needed for forecasting.

\bibliography{..//..//refs/ranges}
\section{Figures}

\end{document}

% formatting code resources
% https://github.com/lizzieinvancouver/ospree/blob/master/docs/ranges/ranges_outline.tex
% https://github.com/lizzieinvancouver/ospree/blob/master/docs/traits/Traitors_Manuscript_supp.Rnw

### 2022-11-04 Docs/manuscript.tex
start sweaving

--- 2022-11-04 Docs/manuscript.Rnw
@@
 % recommended! Uncomment the below line and change the path for your computer!
-\SweaveOpts{prefix.string=/Users/alina/Documents/git/localadaptclim/Docs/figure_ms/demoFig, eps=FALSE} 
+
 %put your figures in one place! Also, note that here 'figures' is the folder and 'demoFig' is what each 
@@
 \begin{document}
-\SweaveOpts{concordance=TRUE} % For RStudio hiccups
+\input{/Users/alina/Documents/git/localadaptclim/Docs/figure_ms/demoFig-concordance} % For RStudio hiccups
 \title{{\huge Changes and trends in budburst and leaf flush across Europe and North America} \\A meta-analysis of local adaptation in spring phenology studies}
@@
   \begin{itemize}
-\item Read AB 2015 and A 2013, took paper with spring doy and provenance coordinates from them
+\item Read SB 2015 and A 2013, took paper with spring doy and provenance coordinates from them
 \item Searched online for more papers (explain criteria here)
@@
 \\
-\verb@Alina Says:@ Need to elaborate here. Note that all figures are subject to update for better resolution. I might add climate overlap plots here too.
+\verb@Alina Says: Need to elaborate here. Note that all figures are subject to update for better resolution. I might add climate overlap plots here too.@
 

### 2022-11-04 Docs/supplement.Rnw
start sweaving

--- 2022-11-04 Docs/manuscript.Rnw
@@
 {0pt}{3.25ex plus 1ex minus .2ex}{1.5ex plus .2ex}
-
 % recommended! Uncomment the below line and change the path for your computer!
-\SweaveOpts{prefix.string=/Users/alina/Documents/git/localadaptclim/Docs/figure_ms/demoFig, eps=FALSE} 
+\SweaveOpts{prefix.string=/Users/alina/Documents/git/localadaptclim/Docs/figure_supp/demoFig, eps=FALSE} 
 %put your figures in one place! Also, note that here 'figures' is the folder and 'demoFig' is what each 
@@
 \fancyhead[LO]{November 2022}
-\fancyhead[RO]{Manuscript}
+\fancyhead[RO]{Supplement}
 % more optionals! %
@@
 \SweaveOpts{concordance=TRUE} % For RStudio hiccups
-\title{{\huge Changes and trends in budburst and leaf flush across Europe and North America} \\A meta-analysis of local adaptation in spring phenology studies}
+\title{{\huge Supplements for: Changes and trends in budburst and leaf flush across Europe and North America} \\A meta-analysis of local adaptation in spring phenology studies}
 \author{Ziyun Zeng \& E. M. Wolkovich}
@@
 
-\newpage
+\section*{Methods}
+\subsection*{Table for all studies}
+\subsection*{Mapped locations for all studies}
 
-\section*{Abstract}
-\section{Introduction}
+\section*{Results}
 
-\begin{itemize}
-\item Define spring event, fall event
-\item Background
-\item Research question (links to results)
-\end{itemize}
-
-\section{Methods}
-\begin{itemize}
-\item Literature search
-\item Provenance location and spring/fall DOY data gathering using ImageJ
-% note that we primarily focused on papers with spring data
-  \begin{itemize}
-\item Read AB 2015 and A 2013, took paper with spring doy and provenance coordinates from them
-\item Searched online for more papers (explain criteria here)
-\item Out of 59 papers related to common garden/spring phenology that I read, only 22 can supply data that we need; Out of the 22, 4 studies have provenance and gardens on different continents, which were subsequently archived (can document in supplements)).
-\end{itemize}
-\item Present how many gardens, species, provenance we have using a table, which also documents the type of spring event they looked at, year, etc. 
-\item Insert mapped locations of studies
-\item Climate data gathering
-  \begin{itemize}
-  \item Simple climate matrix (MAT \& MSP/MAP) gathering
-  \begin{itemize}
-  \item ClimateNA by Dr. Tongli Wang
-  \item Climate Information Tool by FAO
-  \end{itemize}
-  \item Daily climate data gathering: Gridded daily climate data downloaded for all provenances and gardens: We hope that coarse metrics such as latitude and MAT ultimately represent how similar the climates are between the provenances and gardens in times that matter for the events.
-If climates are very similar, then we would expect similar timings [add more here].
-To this end, we estimated climate overlap in relevant months: For spring events, we considered overlap across March to May.
-  \begin{itemize}
-  \item North American daily climate data from Daymet R package
-  \item Europe daily climate data (gridded: NetCDF-4) retrieved from E-OBS 
-  \end{itemize}
-  \end{itemize}
-  \item Estimate climate overlap using daily climate data using Overlap R package
-  \item Calculated GDDs, spherical distances
-  \item Corrected for DOY (taking the difference), will put results in supplement
-  \item Mixed effects modelling: We used hierarchical Bayesian models where we partially pooled by species.
-We estimated effects of continent and species type (angiosperm VS gymnosperm) from posterior estimates, in relation to single predictors (including provenance latitude, MAT) and two predictors (climate overlap percentage and standard deviation).
-\end{itemize}
-
-\section{Results}
-%AZZNov04: all subsection titles are placeholders
-
-\subsection{Provenance latitude and MAT do not affect spring event timing}
-
-Overall, our models show that spring events are not related to provenance latitude or mean annual temperature in North America, and only weakly related in Europe (Figure \#). 
-\\
-\verb@Alina Says:@ Need to elaborate here. Note that all figures are subject to update for better resolution. I might add climate overlap plots here too.
-
-\begin{figure}[!h] 
-    \centering
- \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/springfall_latmat.png}
-    \caption{Spring event Day of Year (DOY) in relation to provenance latitude and MAT, coded by symbol for species and color for garden with linear fits from hierarchical Bayesian models.
-Spring events shown on top and fall event at the bottom.}
-    \label{figure:springfall_latmat}
-\end{figure}
-
-\newpage
-
-\subsubsection {Similar results across provenance latitude, latitude difference, and distance}
-\verb@Alina says: need to find a name for each metric and stick to it.@
-Along with (a) provenance latitude, we also looked at how spring event timing is related to (b) the absolute difference between provenance latitude and garden latitude , as well as (c) the spherical distance between the garden and each provenance.
-All three metrics depict little to no relationship between spring event timing and the geographical location of the provenance \verb@(Supplement Figure #)@.
-% can show that figure in supplement
-
-\subsubsection {Climate overlap does not predict event dates much better than provenance latitude or MAT}
-
-While comparing how similarity in climate relates to event dates, we observed very weak effects of climate overlap on spring events, nearly identical across angiosperms and gymnosperms.
-Fall events diverge as climate overlap declines for both angiosperms and gymnosperms, but slightly more strongly for gymnosperms.
-% add figure here
-
-\subsubsection {Strong relationship between provenance latitude, MAT, and GDDs}
-
-I am unsure about how to commment on what we are seeing in this plot.
-There is a strong relationship between GDDs of each event day recorded and the latitude and MAT, but isn't that self-apparent?
-also I don't know how to link (Fig.
-\#) in text with the corresponding figure being displayed.
-
-% all figures will be updated (color, font size, resolution)
-\begin{figure}[!h] 
-    \centering
- \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/gdd_ms.png}
-    \caption{Growing Degree Days (GDD)on each day of spring event in relation to provenance latitude and MAT, coded by symbol for species and color for garden with linear fits from hierarchical Bayesian models.}
-    \label{figure:gdd}
-\end{figure}
-
-\newpage
-
-\subsection{Stronger clines in fall events observed in North America than Europe}
-
-% old graphs can be found in C:\Users\alina\Documents\git\localadaptclim\Output\plotMay15_two predictors_experiment
-% refer to analyses\script_model_continent_spp_type_effect
-% https://github.com/lizzieinvancouver/localadaptclim/issues/16
-
-% need to replot ofc
-
-%  extract the mean estimate of the slope and report that instead of ~4 or such
-% hmmm i forgot the code :)))
-
-We find that fall events (budset, leaf senescence, leaf abcission) advance strongly with provenance latitude and mean annual temperature, meaning fall events are earlier where provenance mean annual temperature is lower (higher, more northern latitudes).
-This relationship, however, is observed mostly in North America where fall events advance 4.2 days per degree we move north, or 6.4 days when the MAT decreases by one $^{\circ}$C.
-In Europe, such relationship is weak: advance 0.5 days per degree we move north, or 0.6 days when the MAT decreases by one $^{\circ}$C.
-% add figure here
-
-\subsection{Effects of MAT on spring events diverge across angiosperms and gymnosperms}
-
-Effects of provenance latitude on both spring and fall events are similar across angiosperms and gymnosperms.
-However, effects of MAT on spring events weakly diverge: spring events get earlier as MAT increases in angiosperms and delay as MAT increases in gymnosperms, except for \emph{Pseudotsuga menziesii}.
-Fall events delay in warmer locations for both species types, but slightly more so for gymnosperms (3.7 days VS.
-6.2 days).
-
-\section{Discussion}
-
-The weak relationship between spring event dates and provenance latitude and MAT that we find in European studies might be explained by the higher extent of climate overlap in those studies.
-The more similar the climate is between provenances and gardens, the less difference between spring event dates.
-\\
-
-The inconsistent and weak clines in spring events that we found suggest high plasticity in spring phenology across continents and species.
-Fall events, on the other hand, exhibit stronger clines which suggest more local adaptation, especially in North America.
-Overall, our results predict that warming springs will continue to be tracked more closely phenologically by trees than warming fall temperatures.
-\\
-
-In contrast to spring events, we found strong latitudinal clines in fall events across both continents, with local adaptation appearing much stronger in North America than in Europe.
-Our results show that spring events are highly plastic, and thus may shift with warming, but data on more species and greater information on important factors, such as their geographic location in relation to their origins and elevation, are needed for forecasting.
-
-\bibliography{..//..//refs/ranges}
-\section{Figures}
+\subsection*{Corrected DOY results}
+\subsection*{Fitting each species \& common garden instead of just species}
+\subsection*{MAT difference}
+\subsection*{Might want to include how the model looked when we included studies with disagreeing provenance \& gardens}
+% if we indeed include this.. need to update results section in manuscript
 
 \end{document}
-
-% formatting code resources
-% https://github.com/lizzieinvancouver/ospree/blob/master/docs/ranges/ranges_outline.tex
-% https://github.com/lizzieinvancouver/ospree/blob/master/docs/traits/Traitors_Manuscript_supp.Rnw

### 2022-11-04 Analyses/script_model_continent_spp_type_effect.R

@@ -13,6 +13,8 @@ library(rstanarm)
 library(bayesplot)
 library(dplyr)
 library(ggplot2)
+library(viridis)
+install.packages("bayesplot")
 
 setwd("C:/Users/alina/Documents/git/localadaptclim")
 d <- read.csv("input/percentage_overlap_doy_difference_earth_calculated_garden_identifier_adjusted_fall_diffo_included.csv", header = TRUE)  #680
@@ -22,7 +24,8 @@ d$spring_event <- as.numeric(d$spring_event)
 #fit5 <- stan_glmer(spring_event~(lat_prov|species), data = d) 
 
 # fitA_spring_lat
-fitA_spring_lat <- stan_glmer(spring_event~(lat_prov|species), data = d)
+fitA_spring_lat <- readRDS("output/model_fit/fitA_spring_lat.RDS")
+fitA_spring_mat <- readRDS("output/model_fit/fitA_spring_mat.RDS")
 
 draws <- as.matrix(fitA_spring_lat)
 # the full posterior for Alnus rubra slope:
@@ -32,8 +35,6 @@ hist(alnrubslope)
 # To get the estimate for continent or evergreen/deciduous:
 # Here I show an incomplete example for evergreen/deciduous,
 # but we especially want to do for continent (group species by continent)
-d <- dplyr::group_by(d, prov_continent)
-d <- dplyr::group_by(d, Provenance_continent)
 # Add the posteriors of all the species in one group ...
 # Example ... many ways to do this, I just created different vectors (I made very incomplete lists!)
 colnames(draws)
@@ -87,8 +88,6 @@ mcmc_areas(leaftypeplot)+
   scale_y_discrete(labels = rev(new_labels_leaftype))
 dev.off()
 
-
-
 # for continental effects
 northamerican <- c("Alnus_rubra","Picea_engelmannii","Picea_sitchensis","Pinus_albicaulis",     
                    "Populus_trichocarpa","Tsuga_heterophylla", "Populus_balsamifera","Pseudotsuga_menziesii",
@@ -96,6 +95,7 @@ northamerican <- c("Alnus_rubra","Picea_engelmannii","Picea_sitchensis","Pinus_a
 
 european <- c("Fraxinus_excelsior", "Fagus_sylvatica" ,"Picea_abies", "Quercus_petraea")
 
+coef()
 
 # Get all the species of one type now
 northamericandraws <- matrix(data=NA,
@@ -119,6 +119,8 @@ europeanpost <- rowMeans(europeandraws)
 
[…]

### 2022-11-08 Docs/manuscript.Rnw
updated ms

@@
 \usepackage{graphicx}
+\usepackage{lipsum}                     % Dummytext % https://tex.stackexchange.com/questions/9796/how-to-add-todo-notes
+\usepackage{xargs}                      % Use more than one optional parameter in a new commands
+\usepackage[pdftex,dvipsnames]{xcolor}  % Coloured text etc.
+
+\usepackage[colorinlistoftodos,prependcaption,textsize=tiny]{todonotes}
+\newcommandx{\unsure}[2][1=]{\todo[linecolor=red,backgroundcolor=red!25,bordercolor=red,#1]{#2}}
+\newcommandx{\change}[2][1=]{\todo[linecolor=blue,backgroundcolor=blue!25,bordercolor=blue,#1]{#2}}
+\newcommandx{\info}[2][1=]{\todo[linecolor=OliveGreen,backgroundcolor=OliveGreen!25,bordercolor=OliveGreen,#1]{#2}}
+\newcommandx{\improvement}[2][1=]{\todo[linecolor=Plum,backgroundcolor=Plum!25,bordercolor=Plum,#1]{#2}}
+\newcommandx{\thiswillnotshow}[2][1=]{\todo[disable,#1]{#2}}
 
@@
 \section*{Abstract}
+
 \section{Introduction}
-
 \begin{itemize}
 \item Define spring event, fall event
-\item Background
+% \item Background \unsure{What else?}
 \item Research question (links to results)
@@
 \section{Methods}
-\begin{itemize}
-\item Literature search
-\item Provenance location and spring/fall DOY data gathering using ImageJ
-% note that we primarily focused on papers with spring data
+Our methods are divided into four main steps.
+
+% \item Literature search
+\subsection{Literature Search}
+We searched Web of Science, Google Scholar, and Connected Papers for common garden experiments that documented the timing of spring events of woody plant species.
+To be included, studies needed to (a) focus on woody plants originating from either Europe or North America, (b) report spring event in units of calendar days instead of a quantitative scale on a particular day, (c) report locations of provenances and gardens.
+\change{will reword things later}
   \begin{itemize}
@@
 \end{itemize}
-\item Present how many gardens, species, provenance we have using a table, which also documents the type of spring event they looked at, year, etc. 
-\item Insert mapped locations of studies
-\item Climate data gathering
+
+\subsection{Dataset Assembly}
+We reviewed all returned results and assembled a dataset of spring event dates, during which we used ImageJ to scrape data from figures whenever possible and necessary.
+We primarily focused on papers with spring data. \improvement{will elaborate later}
+
+\subsection{Climate Data Gathering \& Analyses}
   \begin{itemize}
@@
   \end{itemize}
-  \item Estimate climate overlap using daily climate data using Overlap R package
-  \item Calculated GDDs, spherical distances
-  \item Corrected for DOY (taking the difference), will put results in supplement
-  \item Mixed effects modelling: We used hierarchical Bayesian models where we partially pooled by species.
+
+\subsubsection{Estimate climate overlap using daily climate data using Overlap R package}
+\subsubsection{Calculated GDDs, spherical distances}
+\subsubsection{Corrected for DOY (taking the difference)}\todo{will put results in supplement}
+\subsection{Mixed effects modelling}
+
+We used hierarchical Bayesian models where we partially pooled by species.
 We estimated effects of continent and species type (angiosperm VS gymnosperm) from posterior estimates, in relation to single predictors (including provenance latitude, MAT) and two predictors (climate overlap percentage and standard deviation).
-\end{itemize}
+\improvement{will elaborate later}
 
 \section{Results}
-%AZZNov04: all subsection titles are placeholders
+\todo[inline]{all subsection titles are placeholders}
+
+Out of the 59 papers on common garden experiments and spring phenology that we reviewed, only 22 supplied data met our criteria to be included: (a) focus on woody plants originating from either Europe or North America, (b) report spring event in units of calendar days instead of a quantitative scale on a particular day, (c) report locations of provenances and gardens.
+Out of the 22, four studies had provenances and common gardens on different continents.
+These studies were excluded from our main analysis, but we document them in our supplements.\change{need to reword}
+Our final dataset captures seven angiosperm and eight gymnosperm species from 19 common gardens, encompassing data from 384 North American provenances and 101 European provenances with observations from 1962 to 2019.
+Seven species had fall event information available.
+It is worth noting that most common gardens included in this study only had observations of one species.
+Therefore, it is difficult for a model to interpret if variations across gardens are due to the species or the garden’s location and climate.
+\improvement{can refer to the fit we did where we fitted each species \& common garden instead of just species} We thus focused on estimating species effects because we expect a larger effect from species.
+
+\todo[inline]{add map here}
+\todo[inline]{add a table documenting studies attributes}
 
@@
 
-Overall, our models show that spring events are not related to provenance latitude or mean annual temperature in North America, and only weakly related in Europe (Figure \#). 
-\\
-\verb@Alina Says:@ Need to elaborate here. Note that all figures are subject to update for better resolution. I might add climate overlap plots here too.
+Overall, our models show that spring events are not related to provenance latitude or mean annual temperature in North America, and only weakly related in Europe (Fig.
+\ref{figure:springfall_latmat}).
+\unsure{Should I add climate overlap plots here too?}
 
@@
 \subsubsection {Similar results across provenance latitude, latitude difference, and distance}
-\verb@Alina says: need to find a name for each metric and stick to it.@
+\todo[inline]{need to find a name for each metric and stick to it.}
 Along with (a) provenance latitude, we also looked at how spring event timing is related to (b) the absolute difference between provenance latitude and garden latitude , as well as (c) the spherical distance between the garden and each provenance.
@@
 
-I am unsure about how to commment on what we are seeing in this plot.
-There is a strong relationship between GDDs of each event day recorded and the latitude and MAT, but isn't that self-apparent?
-also I don't know how to link (Fig.
-\#) in text with the corresponding figure being displayed.
+\todo[inline]{Flagged for Lizzie: I am unsure about how to commment on what we are seeing in this plot.
+\newline There is a strong relationship between GDDs of each event day recorded and the latitude and MAT, but isn't that self-apparent?
+}
+
+Placeholder text (Fig. \ref{figure:gdd}).
 

### 2022-11-08 Docs/manuscript.tex
updated ms

@@
 \section*{Abstract}
+
 \section{Introduction}
-
 \begin{itemize}
 \item Define spring event, fall event
-\item Background
+\item Background \unsure{What else?}
 \item Research question (links to results)
@@
 \section{Methods}
+Our methods are divided into four main steps.
+
+% \item Literature search
+\subsection{Literature Search}
+We searched Web of Science, Google Scholar, and Connected Papers for common garden experiments that documented the timing of spring events of woody plant species.
+To be included, studies needed to (a) focus on woody plants originating from either Europe or North America, (b) report spring event in units of calendar days instead of a quantitative scale on a particular day, (c) report locations of provenances and gardens.
+\change{will reword things later}
   \begin{itemize}
-\item Literature search
-\item Provenance location and spring/fall DOY data gathering using ImageJ
-% note that we primarily focused on papers with spring data
-  \begin{itemize}
-\item Read SB 2015 and A 2013, took paper with spring doy and provenance coordinates from them
+\item Read AB 2015 and A 2013, took paper with spring doy and provenance coordinates from them
 \item Searched online for more papers (explain criteria here)
@@
 
-Overall, our models show that spring events are not related to provenance latitude or mean annual temperature in North America, and only weakly related in Europe (Figure \#). 
-\\
-\verb@Alina Says: Need to elaborate here. Note that all figures are subject to update for better resolution. I might add climate overlap plots here too.@
+Overall, our models show that spring events are not related to provenance latitude or mean annual temperature in North America, and only weakly related in Europe (Fig.
+\ref{figure:springfall_latmat}).
+\unsure{Should I add climate overlap plots here too?}
 

### 2022-11-13 Docs/supplement.Rnw
cross ref ms and supp, model outputs

@@
 % required 
+\usepackage{xr-hyper}
+\usepackage{hyperref}
+\externaldocument[manuscript-]{manuscript}
+\usepackage{booktabs,siunitx}
 \usepackage{Sweave}
 \usepackage{graphicx}
+\usepackage{lipsum}                     % Dummytext % https://tex.stackexchange.com/questions/9796/how-to-add-todo-notes
+\usepackage{xargs}                      % Use more than one optional parameter in a new commands
+\usepackage[pdftex,dvipsnames]{xcolor}  % Coloured text etc.
+
+\usepackage[colorinlistoftodos,prependcaption,textsize=tiny]{todonotes}
+\newcommandx{\unsure}[2][1=]{\todo[linecolor=red,backgroundcolor=red!25,bordercolor=red,#1]{#2}}
+\newcommandx{\change}[2][1=]{\todo[linecolor=blue,backgroundcolor=blue!25,bordercolor=blue,#1]{#2}}
+\newcommandx{\info}[2][1=]{\todo[linecolor=OliveGreen,backgroundcolor=OliveGreen!25,bordercolor=OliveGreen,#1]{#2}}
+\newcommandx{\improvement}[2][1=]{\todo[linecolor=Plum,backgroundcolor=Plum!25,bordercolor=Plum,#1]{#2}}
+\newcommandx{\thiswillnotshow}[2][1=]{\todo[disable,#1]{#2}}
 
@@
 {0pt}{3.25ex plus 1ex minus .2ex}{1.5ex plus .2ex}
+
 % recommended! Uncomment the below line and change the path for your computer!
-\SweaveOpts{prefix.string=/Users/alina/Documents/git/localadaptclim/Docs/figure_supp/demoFig, eps=FALSE} 
+\SweaveOpts{prefix.string=/Users/alina/Documents/git/localadaptclim/Docs/figure_ms/demoFig, eps=FALSE} 
 %put your figures in one place! Also, note that here 'figures' is the folder and 'demoFig' is what each 
@@
 \fancyhead[LO]{November 2022}
-\fancyhead[RO]{Supplement}
+\fancyhead[RO]{Manuscript}
 % more optionals! %
-\usepackage[hyphens]{url} % this wraps my URL versus letting it spill across the page, a bad habit LaTeX has
+%\usepackage[hyphens]{url} % this wraps my URL versus letting it spill across the page, a bad habit LaTeX has
 
@@
 
+\subsection*{Spring Lat VS. Fall Lat}
+\begin{table}
+\centering
+\caption{Model summary for the relationship between event day of year (DOY)and provenance latitude, fitted by species, in spring (left) and fall (right).}
+\todo[inline]{i'm not sure how to word this}
+\begin{tabular}[t]{lcc}
+\toprule
+  & $DOY_{Spring}~(Latitude|Species)$ & $DOY_{Fall}~(Latitude|Species)$\\
+\midrule
+Intercept & \num{114.219} & \num{316.736}\\
+ & {}[\num{100.511}, \num{127.680}] & {}[\num{272.545}, \num{415.373}]\\
+Sigma[Species × Intercept,Intercept] & \num{1148.104} & \num{12381.073}\\
+ & {}[\num{435.227}, \num{2825.532}] & {}[\num{6176.075}, \num{24554.731}]\\
+Sigma[Species × Latitude,Intercept] & \num{-13.454} & \num{-307.256}\\
+ & {}[\num{-41.733}, \num{-1.069}] & {}[\num{-621.049}, \num{-122.796}]\\
+Sigma[Species × Latitude,Latitude] & \num{0.374} & \num{10.159}\\
+ & {}[\num{0.111}, \num{1.030}] & {}[\num{5.266}, \num{33.131}]\\
+\midrule
+Num.Obs. & \num{671} & \num{349}\\
+R2 & \num{0.903} & \num{0.961}\\
+R2 Adj. & \num{0.902} & \num{0.960}\\
+R2 Marg. & \num{0.000} & \num{0.000}\\
+Log.Lik. & \num{-2340.362} & \num{-1217.949}\\
+ELPD & \num{-2358.2} & \num{-1231.9}\\
+ELPD s.e. & \num{27.8} & \num{22.1}\\
+LOOIC & \num{4716.4} & \num{2463.8}\\
+LOOIC s.e. & \num{55.6} & \num{44.2}\\
+WAIC & \num{4716.2} & \num{2463.3}\\
+RMSE & \num{7.87} & \num{8.44}\\
+r2.adjusted.marginal & \num{0.902} & \num{0.960}\\
+\bottomrule
+\end{tabular}
+\label{table:model_sf_lat}
+\end{table}
+
 \subsection*{Corrected DOY results}

### 2022-11-13 Docs/supplement.tex
start sweaving
cross ref ms and supp, model outputs

--- 2022-11-04 Docs/supplement.Rnw
@@
 {0pt}{3.25ex plus 1ex minus .2ex}{1.5ex plus .2ex}
+
 % recommended! Uncomment the below line and change the path for your computer!
-\SweaveOpts{prefix.string=/Users/alina/Documents/git/localadaptclim/Docs/figure_supp/demoFig, eps=FALSE} 
+
 %put your figures in one place! Also, note that here 'figures' is the folder and 'demoFig' is what each 
@@
 \begin{document}
-\SweaveOpts{concordance=TRUE} % For RStudio hiccups
+\input{/Users/alina/Documents/git/localadaptclim/Docs/figure_ms/demoFig-concordance} % For RStudio hiccups
 \title{{\huge Supplements for: Changes and trends in budburst and leaf flush across Europe and North America} \\A meta-analysis of local adaptation in spring phenology studies}

### 2022-11-13 Analyses/script_stan_glmer_NEW_fit_plots_are_in_here_SEPT16_MODELLING_SCRIPT.R

@@ -3,6 +3,22 @@
 # 4/May/2022
 # alinazengziyun@yahoo.com
 
+
+# updated Nov-13, 2022 to try out model outputs
+
+install.packages('modelsummary')
+library(modelsummary)
+
+# test 
+modelsummary(fitA_spring_lat, statistic = "conf.int")
+
+
+# two models side by side
+models <- list(fitA_spring_lat, fitA_fall_lat)
+modelsummary(models,statistic = "conf.int")
+library(flextable)
+modelsummary(models,statistic = "conf.int",output = "tabletest.tex")
+
 # Housekeeping ----
 rm(list=ls()) 
 options(stringsAsFactors = FALSE)

### 2022-11-15 Docs/manuscript.Rnw
cross ref ms and supp, model outputs
update manuscript
started bibtex

@@
 % required 
+\usepackage{natbib}
+\bibliographystyle{plainnat} %{unsrtnat}
+\usepackage{xr-hyper}
+\usepackage{hyperref}
+\externaldocument[supplement-]{supplement}
+\usepackage{booktabs,siunitx}
 \usepackage{Sweave}
@@
 % more optionals! %
-\usepackage[hyphens]{url} % this wraps my URL versus letting it spill across the page, a bad habit LaTeX has
-
+% \usepackage[hyphens]{url} % this wraps my URL versus letting it spill across the page, a bad habit LaTeX has
 %%% end preambling. %%%
@@
 \item Define spring event, fall event
-% \item Background \unsure{What else?}
+\item Background 
 \item Research question (links to results)
 \end{itemize}
+
+\todo[inline]{Good background information.}
+\citep{AitkenBemmels16}\\
+"Geographic variation in trees has been investigated since the mid-18th century.
+Similar patterns of clinal variation have been observed along latitudinal and elevational gradients in common garden experiments for many temperate and boreal species."\\
+"Scientists have studied relationships between tree populations and environmental characteristics of their provenances in common garden experiments for over 250 years."\\
+\\
+\citep{Alberto13}\\
+"Evolutionary responses are required for tree populations to be able to track climate change.
+Results of 250 years of common garden experiments show that most forest trees have evolved local adaptation, as evidenced by the adaptive differentiation of populations in quantitative traits, reflecting environmental conditions of population origins."\\
+"Thus, responding to climate change will likely require that the quantitative traits of populations again match their environments.
+We examine what kind of information is needed for evaluating the potential to respond, and what information is already available.
+We review the genetic models related to selection responses, and what is known currently about the genetic basis of the traits.
+We address special problems to be found at the range margins, and highlight the need for more modeling to understand specific issues at southern and northern margins.
+We need new common garden experiments for less known species.
+"\\
+"Modeling work on the potential of populations and species to respond genetically to recent climate change is advancing (see Hoffmann \& Sgrò, 2011; Franks \& Hoffmann, 2012; Shaw \& Etterson, 2012 for recent reviews).
+"\\
+
+(see e.g., Valladares et al., 2007; Caffarra et al., 2011; Hänninen \& Tanino, 2011)
+"The plastic response of different traits (e.g., phenology in trees) to variation in climate is, however, often much more complex than in heuristic models of adaptation."\\
+
+\citep{Loarie09}\\ 
+"It is also now understood that the rate of adaptation required by climate change varies among geographic regions." \improvement{need to read again}\\
+
+\citep{Nicotra10}\\ 
+"The immediate responses via phenotypic plasticity have also been considered in the context of climate change." \improvement{need to read again}\\
+
+\citep{Menzel99} \textbf{Growing season extended in Europe.}\improvement{good Europe background paper}\\
+\citep{Menzel06} \textbf{European phenological response to climate change matches the warming pattern.}\improvement{another good Europe background paper}\\
+\citep{Parmesan06}\\
+"Trees exhibit a high degree of phenotypic plasticity with respect to climatic variation.
+Phenological shifts of bud flush in response to recent increases in temperatures have been widely documented ." \improvement{can cite more papers on this matter}
+
+\citep{Gill15} \textbf{Changes in autumn senescence in northern hemisphere deciduous trees: a meta-analysis of autumn phenology studies.} \improvement{GREAT paper for us to base our writing on: similar meta-analysis but focuses on fall events}\\
+"the effects of climate change on the phenology of vegetation have received increased attention over the past several decades as any factor that alters the timing of early growing season leaf-out and senescence has the potential to affect a variety of ecosystem properties.
+For example, the timing of leaf-out and senescence of deciduous plants has been shown to affect plant competition (Fridley, 2012), plant growth (Myneni et al., 1997) and ecosystem carbon uptake (Barichivich et al., 2012).\\
+
+"While leaf-out has been shown to advance over the past century as a result of increasing air temperatures (Linderholm, 2006; Polgar and Primack, 2011), the relationship between temperature and leaf senescence remains less well understood (García-Plazaola et al., 2003; Richardson et al., 2013)."
+
+"Thus, an extension of the growing season can contribute to reduced atmospheric CO2 concentrations due to enhanced carbon sequestration in terrestrial plants (Penuelas et al., 2009; Richardson et al., 2013).
+However, the increase in carbon uptake may be partially offset by increased rates of ecosystem respiration (Piao et al., 2008)"
+
+"The lack of consistent relationships between air temperature and timing of leaf senescence suggests that autumn senescence may be influenced by a variety of factors that obscure its relationship with temperature."
+
+\todo[inline]{Cite some papers about the documented trends}
+
+tbc...
+
+\todo[inline]{Alina says: I found AitkenBemmels2016's research question to be wonderfully phrased and think we can do something similar}
+\citep{AitkenBemmels16}\\
+"To what extent are patterns of local adaptation along climatic gradients similar among species within a given geographic region?
+If clines in phenotypic traits are similar, can average patterns of sympatric variation guide AGF or identify critical climatic variables as a first approximation for untested species?
+We reanalysed data from the literature on provenance trials in temperate tree species from western North America to address these questions.
+This is an ideal region of focus because of the many provenance trials conducted on native tree species that are important for forestry, and because its high topographic complexity means that turnover in climate occurs rapidly and heterogeneously over short geographic spaces.
+Clines along climate gradients are thus espcially likely to reflect local adaptation to climate, rather than non-adaptive phenomena resulting from population demographic history."
 
@@
 Overall, our models show that spring events are not related to provenance latitude or mean annual temperature in North America, and only weakly related in Europe (Fig.
-\ref{figure:springfall_latmat}).
+\ref{figure:springfall_latmat} \& Supplement Table.\ref{supplement-table:model_sf_lat}).
 \unsure{Should I add climate overlap plots here too?}
@@
 
-\bibliography{..//..//refs/ranges}
+\citet{Keir11}
+
 \section{Figures}
 
+\bibliography{bibliography_local_adaptation.bib}
+
 \end{document}
+
+% helping with abstract language
+% Alberto16
+% "On the basis of the patterns of quantitative variation for 19 adaptation-related traits studied in 59 tree species (mostly temperate and boreal species from the Northern hemisphere), we found that genetic differentiation between populations and clinal variation along environmental gradients were very common (respectively, 90% and 78% of cases)."
 

### 2022-11-15 Docs/manuscript.tex
cross ref ms and supp, model outputs
update manuscript
started bibtex

@@
 \item Define spring event, fall event
-\item Background \unsure{What else?}
+\item Background 
 \item Research question (links to results)
 \end{itemize}
+
+\todo[inline]{Good background information.}
+\citep{AitkenBemmels16}\\
+"Geographic variation in trees has been investigated since the mid-18th century.
+Similar patterns of clinal variation have been observed along latitudinal and elevational gradients in common garden experiments for many temperate and boreal species."\\
+"Scientists have studied relationships between tree populations and environmental characteristics of their provenances in common garden experiments for over 250 years."\\
+\\
+\citep{Alberto13}\\
+"Evolutionary responses are required for tree populations to be able to track climate change.
+Results of 250 years of common garden experiments show that most forest trees have evolved local adaptation, as evidenced by the adaptive differentiation of populations in quantitative traits, reflecting environmental conditions of population origins."\\
+"Thus, responding to climate change will likely require that the quantitative traits of populations again match their environments.
+We examine what kind of information is needed for evaluating the potential to respond, and what information is already available.
+We review the genetic models related to selection responses, and what is known currently about the genetic basis of the traits.
+We address special problems to be found at the range margins, and highlight the need for more modeling to understand specific issues at southern and northern margins.
+We need new common garden experiments for less known species.
+"\\
+"Modeling work on the potential of populations and species to respond genetically to recent climate change is advancing (see Hoffmann \& Sgrò, 2011; Franks \& Hoffmann, 2012; Shaw \& Etterson, 2012 for recent reviews).
+"\\
+
+(see e.g., Valladares et al., 2007; Caffarra et al., 2011; Hänninen \& Tanino, 2011)
+"The plastic response of different traits (e.g., phenology in trees) to variation in climate is, however, often much more complex than in heuristic models of adaptation."\\
+
+\citep{Loarie09}\\ 
+"It is also now understood that the rate of adaptation required by climate change varies among geographic regions." \improvement{need to read again}\\
+
+\citep{Nicotra10}\\ 
+"The immediate responses via phenotypic plasticity have also been considered in the context of climate change." \improvement{need to read again}\\
+
+\citep{Menzel99} \textbf{Growing season extended in Europe.}\improvement{good Europe background paper}\\
+\citep{Menzel06} \textbf{European phenological response to climate change matches the warming pattern.}\improvement{another good Europe background paper}\\
+\citep{Parmesan06}\\
+"Trees exhibit a high degree of phenotypic plasticity with respect to climatic variation.
+Phenological shifts of bud flush in response to recent increases in temperatures have been widely documented ." \improvement{can cite more papers on this matter}
+
+\citep{Gill15} \textbf{Changes in autumn senescence in northern hemisphere deciduous trees: a meta-analysis of autumn phenology studies.} \improvement{GREAT paper for us to base our writing on: similar meta-analysis but focuses on fall events}\\
+"the effects of climate change on the phenology of vegetation have received increased attention over the past several decades as any factor that alters the timing of early growing season leaf-out and senescence has the potential to affect a variety of ecosystem properties.
+For example, the timing of leaf-out and senescence of deciduous plants has been shown to affect plant competition (Fridley, 2012), plant growth (Myneni et al., 1997) and ecosystem carbon uptake (Barichivich et al., 2012).\\
+
+"While leaf-out has been shown to advance over the past century as a result of increasing air temperatures (Linderholm, 2006; Polgar and Primack, 2011), the relationship between temperature and leaf senescence remains less well understood (García-Plazaola et al., 2003; Richardson et al., 2013)."
+
+"Thus, an extension of the growing season can contribute to reduced atmospheric CO2 concentrations due to enhanced carbon sequestration in terrestrial plants (Penuelas et al., 2009; Richardson et al., 2013).
+However, the increase in carbon uptake may be partially offset by increased rates of ecosystem respiration (Piao et al., 2008)"
+
+"The lack of consistent relationships between air temperature and timing of leaf senescence suggests that autumn senescence may be influenced by a variety of factors that obscure its relationship with temperature."
+
+\todo[inline]{Cite some papers about the documented trends}
+
+tbc...
+
+\todo[inline]{Alina says: I found AitkenBemmels2016's research question to be wonderfully phrased and think we can do something similar}
+\citep{AitkenBemmels16}\\
+"To what extent are patterns of local adaptation along climatic gradients similar among species within a given geographic region?
+If clines in phenotypic traits are similar, can average patterns of sympatric variation guide AGF or identify critical climatic variables as a first approximation for untested species?
+We reanalysed data from the literature on provenance trials in temperate tree species from western North America to address these questions.
+This is an ideal region of focus because of the many provenance trials conducted on native tree species that are important for forestry, and because its high topographic complexity means that turnover in climate occurs rapidly and heterogeneously over short geographic spaces.
+Clines along climate gradients are thus espcially likely to reflect local adaptation to climate, rather than non-adaptive phenomena resulting from population demographic history."
 

### 2022-11-15 Analyses/script_stan_glmer_NEW_fit_plots_are_in_here_SEPT16_MODELLING_SCRIPT.R

@@ -10,12 +10,55 @@ install.packages('modelsummary')
 library(modelsummary)
 
 # test 
+fitA_spring_lat <- readRDS("output/model_fit/fitA_spring_lat.RDS")
+fitA_fall_lat <- readRDS("output/model_fit/fitA_fall_lat.RDS")
 modelsummary(fitA_spring_lat, statistic = "conf.int")
 
 
 # two models side by side
 models <- list(fitA_spring_lat, fitA_fall_lat)
 modelsummary(models,statistic = "conf.int")
+
+# this looks very good
+modelsummary::datasummary(fitA_spring_lat)
+datasummary_skim(fitA_spring_lat)
+
+#experiment
+datasummary_skim( (b[(Intercept) species:Alnus_rubra])~Mean + SD + P5 + P50 + P95, data = fitA_spring_lat)
+
+
+
+# another method https://cran.r-project.org/web/packages/tidybayes/vignettes/tidy-rstanarm.html
+
+install.packages("ggdist","tidybayes")
+install.packages("knitr")
+library(ggdist)
+library(tidybayes)
+library(tidyr)
+library(knitr)
+m = fitA_spring_lat
+get_variables(m) #  get a list of raw model variables names so that we know what variables we can extract from the model:
+
+m %>%   # tidy up the names for each row
+  spread_draws(b[term,group]) %>%
+  head(10)
+
+m %>%
+  spread_draws(b[,group]) %>%  #no need to have the term column
+  head(10)
+
+# m %>% 
+# spread_draws(b[,group]) %>%   # separate out species
+# separate(group, c("group", "species"), ":") %>%
+# head(10)
+
+#perfect!!!
+m %>%
+  spread_draws(b[,group]) %>% 
+  summarise_draws()     # https://cran.r-project.org/web/packages/tidybayes/vignettes/tidy-rstanarm.html
+                        # https://mc-stan.org/posterior/reference/draws_summary.html
+
+
 library(flextable)
 modelsummary(models,statistic = "conf.int",output = "tabletest.tex")
 

### 2022-11-25 Docs/manuscript.Rnw
added plots, revised methods

@@
 
+We conducted the first cross-continental meta-analysis of published studies from the peer-reviewed literature that reported spring event dates for a mix of angiosperm and gymnosperm tree species in the northern hemisphere, capturing data from 384 North American provenances and 101 European provenances with observations from 1962 to 2019.
+
 \section{Introduction}
@@
 \item Background
+\item Importance: An accurate understanding of how species distribution might be influenced by environmental conditions at a continental scale is critical for future range shift prediction. 
+\item Knowledge Gap: Anecdotally, the spring phenology of trees from European provenances appeared to exhibit stronger clines and less plasticity than in North American studies.
+Despite a growing interest in predicting local adaptation across locations, no study has comprehensively examined clines for spring and fall events and what factors may underlie differences observed across studies.
 \item Research question (links to results)
@@
 
+We ask: Across common garden studies, how strong are clines in spring and fall phenology?
+
+To better understand these clines, we also aim to examine:
+\begin{itemize}
+\item If differences in clines exist across Europe and North America.
+\item If differences in clines exist across angiosperm and gymnosperm species.
+\item If climate overlap (see Figure 1) between the provenance and garden explains the similarities and differences in clines.
+\end{itemize}
+To our knowledge, this marks the first study to synthesize spring phenology data retrieved from studies across North America and Europe.
+
+\subsection*{Ref notes}
 \todo[inline]{Good background information.}
@@
 \subsection{Literature Search}
-We searched Web of Science, Google Scholar, and Connected Papers for common garden experiments that documented the timing of spring events of woody plant species.
-To be included, studies needed to (a) focus on woody plants originating from either Europe or North America, (b) report spring event in units of calendar days instead of a quantitative scale on a particular day, (c) report locations of provenances and gardens.
-\change{will reword things later}
+
+We searched the peer-reviewed literature for common garden experiments that documented the timing of spring events of woody plant species.
+We searched on the Web of Science (Thompson Reuters, New York, NY) using all combinations of the following terms as ‘titles’ including ‘common garden*’ or “provenance*’ and ‘leafout*’ or ‘leaf out*’ or ‘budburst’ or ‘spring phenolog*’.
+Our search was limited to journal articles published before November 3, 2021.
+The search returned 94 publications.
+\\
+\todo[inline]{add these citations to bibtex?? do we cite them?}
+We read the methods and results of all publications returned and limited our subsequent analysis to the ones that (a) focused on woody plants originating from either Europe or North America, (b) reported spring events in units of calendar days (day of year or DOY) instead of a quantitative scale on a particular day, and (c) reported latitude and longitude of provenances and gardens.
+We had to exclude studies that reported spring events on a quantitative scale because (1) such studies usually only assessed where on the scale the spring event of a tree fell onto on the same days across different years (e.g.
+Robson et al., 2013; Vander et al., 2015; Santini et al., 2014; Schueler \& Liesebach, 2004), and (2) scales are not always consistent across different studies.
+Such factors made it impossible to back convert the quantitative scale to DOY.
+\\
+\\
+Some studies did not provide the exact latitude and longitude of the common garden (Bongarten, 1978) or the provenances (Hall et al., 2007; Soolanayakanahally et al., 2013), or they did not link the latitude and longitude of each provenance to the DOY of spring events (Deans and Harvey, 1996).
+We had to exclude these papers.
+Finally, studies that focused on examining how provenance altitude influenced local adaptation in spring phenology were excluded since we are interested in testing latitudinal effects (Vitasse et al., 2009; Vitasse et al., 2010; Li et al., 1997; Alberto et al., 2011; Acevedo-Rodríguez et al., 2006).
+
   \begin{itemize}
-\item Read AB 2015 and A 2013, took paper with spring doy and provenance coordinates from them
-\item Searched online for more papers (explain criteria here)
-\item Out of 59 papers related to common garden/spring phenology that I read, only 22 can supply data that we need; Out of the 22, 4 studies have provenance and gardens on different continents, which were subsequently archived (can document in supplements)).
+\item Read AB 2015 and A 2013 for anything that we missed, took paper with spring doy and provenance coordinates from them
+
+\item Out of XX papers related to common garden/spring phenology that we read, only 22 ultimately supplied the data that we needed (DOY of spring events, latitude and longitude of provenances and common gardens); Out of the 22, four studies had provenance and gardens on different continents, which were subsequently archived (can document in supplements)).
 \end{itemize}
@@
 \subsection{Dataset Assembly}
-We reviewed all returned results and assembled a dataset of spring event dates, during which we used ImageJ to scrape data from figures whenever possible and necessary.
-We primarily focused on papers with spring data. \improvement{will elaborate later}
+We assembled a dataset that documents the DOY of spring events and coordinates of provenances, during which we used ImageJ (version XXXX, citation) to scrape data from figures whenever possible and necessary.
+\\
+Everything was converted to day of year (some experiment counted not from january 1 but from march 1, etc.
+Need to talk about how we included data from field and indoor-experiment studies.\improvement{expand on this and give examples}
+\\
+\\
+It is worth noting that most common garden studies that supplied data only had observations of one species.
+Therefore, it is difficult for a model to interpret if variations across gardens are due to the species or the garden’s location and climate.
+We thus focused on estimating species effects because we expect a larger effect from species.
+Therefore, we considered unique gardens and tree species to represent independent data within publications, and fitted each species \& common gardens instead of just species.
+\improvement{need to phrase this better, and maybe put this somewhere in the modelling methods}
+\\
+\\
+Studies were distributed throughout North America and Europe, with the majority of data concentrated in XXXXX (Figure X).
+The distribution was skewed toward North America (n = X), with only X located in Europe.
+We primarily focused our analysis on spring events, and made use of the fall event DOYs present in X of the studies we included rather than searching for publications entirely on fall events online.
 
@@
 \section{Results}
+
+\verb@Research questions recap@
+\\
+Despite a growing interest in predicting local adaptation across locations, no study has comprehensively examined clines for spring and fall events and what factors may underlie differences observed across studies.
+
+We ask: Across common garden studies, how strong are clines in spring and fall phenology?
+
+To better understand these clines, we also aim to examine:
+\begin{itemize}
+\item If differences in clines exist across Europe and North America.
+\item If differences in clines exist across angiosperm and gymnosperm species.
+\item If climate overlap between the provenance and garden explains the similarities and differences in clines.
+\end{itemize}
+
 \todo[inline]{all subsection titles are placeholders}
-
-Out of the 59 papers on common garden experiments and spring phenology that we reviewed, only 22 supplied data met our criteria to be included: (a) focus on woody plants originating from either Europe or North America, (b) report spring event in units of calendar days instead of a quantitative scale on a particular day, (c) report locations of provenances and gardens.
+Out of the X papers on common garden experiments and spring phenology that we reviewed, only 22 supplied data met our criteria to be included: (a) focus on woody plants originating from either Europe or North America, (b) report spring event in units of calendar days instead of a quantitative scale on a particular day, (c) report locations of provenances and gardens.
 Out of the 22, four studies had provenances and common gardens on different continents.
 These studies were excluded from our main analysis, but we document them in our supplements.\change{need to reword}
+\\
+\\
 Our final dataset captures seven angiosperm and eight gymnosperm species from 19 common gardens, encompassing data from 384 North American provenances and 101 European provenances with observations from 1962 to 2019.
 Seven species had fall event information available.
+\\
+\\
 It is worth noting that most common gardens included in this study only had observations of one species.
@@
 Along with (a) provenance latitude, we also looked at how spring event timing is related to (b) the absolute difference between provenance latitude and garden latitude , as well as (c) the spherical distance between the garden and each provenance.
-All three metrics depict little to no relationship between spring event timing and the geographical location of the provenance \verb@(Supplement Figure #)@.
-% can show that figure in supplement
+All three metrics depict little to no relationship between spring event timing and the geographical location of the provenance (Supplement Fig.\ref{supplement-figure:lat_distance}).
 
+\newpage
 \subsubsection {Climate overlap does not predict event dates much better than provenance latitude or MAT}
@@
 While comparing how similarity in climate relates to event dates, we observed very weak effects of climate overlap on spring events, nearly identical across angiosperms and gymnosperms.
-Fall events diverge as climate overlap declines for both angiosperms and gymnosperms, but slightly more strongly for gymnosperms.
-% add figure here
+Fall events diverge as climate overlap declines for both angiosperms and gymnosperms, but slightly more strongly for gymnosperms (Fig.\ref{figure:overlap}).
 
-\subsubsection {Strong relationship between provenance latitude, MAT, and GDDs}
-
-\todo[inline]{Flagged for Lizzie: I am unsure about how to commment on what we are seeing in this plot.
-\newline There is a strong relationship between GDDs of each event day recorded and the latitude and MAT, but isn't that self-apparent?
-}
-
-Placeholder text (Fig. \ref{figure:gdd}).
-
-% all figures will be updated (color, font size, resolution)
 \begin{figure}[!h] 
     \centering
- \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/gdd_ms.png}
-    \caption{Growing Degree Days (GDD)on each day of spring event in relation to provenance latitude and MAT, coded by symbol for species and color for garden with linear fits from hierarchical Bayesian models.}
-    \label{figure:gdd}
+ \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/climate_overlap.png}
+    \caption{Caption placeholder}
+    \label{figure:overlap}
 \end{figure}
-
-\newpage
 
@@
 We find that fall events (budset, leaf senescence, leaf abcission) advance strongly with provenance latitude and mean annual temperature, meaning fall events are earlier where provenance mean annual temperature is lower (higher, more northern latitudes).
-This relationship, however, is observed mostly in North America where fall events advance 4.2 days per degree we move north, or 6.4 days when the MAT decreases by one $^{\circ}$C.
-In Europe, such relationship is weak: advance 0.5 days per degree we move north, or 0.6 days when the MAT decreases by one $^{\circ}$C.
-% add figure here
+This relationship, however, is observed mostly in North America where fall events advance 4.2 days per degree we move north, or 6.4 days when the MAT decreases by 1 $^{\circ}$C (Fig.
+\ref{figure:continent}).
+In Europe, such relationship is weak: advance 0.5 days per degree we move north, or 0.6 days when the MAT decreases by 1 $^{\circ}$C(Fig.
+\ref{figure:continent}).
+
+\begin{figure}[!h] 
+    \centering
+ \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/continent_histogram.png}
+    \caption{Caption placeholder}
+    \label{figure:continent}
+\end{figure}
 
@@
 Fall events delay in warmer locations for both species types, but slightly more so for gymnosperms (3.7 days VS.
-6.2 days).
+6.2 days)(Fig.
+\ref{figure:spp_type}).
+
+\begin{figure}[!h] 
+    \centering
+ \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/spp_type_histogram.png}
+    \caption{Caption placeholder}
+    \label{figure:spp_type}
+\end{figure}
 

### 2022-11-25 Docs/supplement.Rnw
started bibtex
added plots, revised methods

@@
 % required 
+\usepackage{natbib}
 \usepackage{xr-hyper}
@@
 
+\subsection*{Corrected DOY results}
+\subsection*{Fitting each species \& common garden instead of just species}
+\subsection*{MAT difference}
+\subsection*{Might want to include how the model looked when we included studies with disagreeing provenance \& gardens}
+% if we indeed include this.. need to update results section in manuscript
+
+\subsection*{Similar results across provenance latitude, latitude difference, and distance}
+Placeholder text (Fig. \ref{figure:lat_distance}).
+\begin{figure}[!h] 
+    \centering
+ \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_supp/lat_latdiff_earth_distance.png}
+    \caption{Similar results across provenance latitude, latitude difference, and distance}
+    \label{figure:lat_distance}
+\end{figure}
+\todo[inline]{change caption later}
+
+\subsection* {Strong relationship between provenance latitude, MAT, and GDDs}
+\todo[inline]{Flagged for Lizzie: I am unsure about how to commment on what we are seeing in this plot.
+\newline There is a strong relationship between GDDs of each event day recorded and the latitude and MAT, but isn't that self-apparent?
+}
+
+Placeholder text (Fig. \ref{figure:gdd}).
+
+% all figures will be updated (color, font size, resolution)
+\begin{figure}[!h] 
+    \centering
+ \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/gdd_ms.png}
+    \caption{Growing Degree Days (GDD)on each day of spring event in relation to provenance latitude and MAT, coded by symbol for species and color for garden with linear fits from hierarchical Bayesian models.}
+    \label{figure:gdd}
+\end{figure}
+
+\section {Don't have a specific sections for these ones yet}
+
+\begin{figure}[!h] 
+    \centering
+ \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_supp/climate_overlap_continent_comparison.png}
+    \caption{The closer a garden is to a provenance, the more overlap in temperature. 
+Higher extend of climate overlap in European studies.}
+    \label{figure:overlap_continent}
+\end{figure}
+
+\begin{figure}[!h] 
+    \centering
+ \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_supp/prov_lat_mat_continent.png}
+    \caption{Placeholder.}
+    \label{figure:lat_mat_continent}
+\end{figure}
+
+\begin{figure}[!h] 
+    \centering
+ \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_supp/spp_type_continent.png}
+    \caption{Placeholder}
+    \label{figure:spp_continent}
+\end{figure}
+
 \subsection*{Spring Lat VS. Fall Lat}
@@
 
-\subsection*{Corrected DOY results}
-\subsection*{Fitting each species \& common garden instead of just species}
-\subsection*{MAT difference}
-\subsection*{Might want to include how the model looked when we included studies with disagreeing provenance \& gardens}
-% if we indeed include this.. need to update results section in manuscript
-
 \end{document}

### 2022-12-15 Docs/manuscript.Rnw
Worked on methods

@@
 % required 
+\usepackage{amsmath}
 \usepackage{natbib}
@@
 \todo[inline]{Good background information.}
+
 \citep{AitkenBemmels16}\\
@@
 \\
+
+\todo[inline]{question for lizzie: I later checked 2022 publications too to make sure we didn't miss anything...
+Do I just say at the very beginning that our search is limited to publications before Dec 14, 2022, or do I say 'we later checked the literature again on XX-XX-2022 to ensure we did not miss any new publications?}
+\todo[inline]{I also think the reviewers might question why we had so few search results....
+maybe I should mention that I also checked on Google Scholar and Connected Papers, and Aitken Bammals 2016/Alberto 2013, and emailed our list of papers to established researchers such as Aitken and Guy, etc.
+to make sure we didn't miss anything?}
+
+We read the methods and results of all publications returned and limited our subsequent analysis to the ones that (a) focused on woody plants originating from either Europe or North America, (b) had provenance trails and common gardens on the same continent, (c) reported spring events in units of calendar days (day of year or DOY) instead of a quantitative scale on a particular day, and (d) reported latitude and longitude of provenances and gardens.
+\\
+We had to exclude studies that reported spring events on a quantitative scale because (1) such studies usually only assessed where on the scale the spring event of a tree fell onto on the same days across different years (e.g.
+Robson et al., 2013; Vander et al., 2015; Santini et al., 2014; Schueler \& Liesebach, 2004), and (2) scales are not always consistent across different studies (Chmura \& Rozkowski 2002; Dhont et al., 2010; Wang et al., 2022).
+Such factors made it impossible to back convert the quantitative scale to DOY.
 \todo[inline]{add these citations to bibtex?? do we cite them?}
-We read the methods and results of all publications returned and limited our subsequent analysis to the ones that (a) focused on woody plants originating from either Europe or North America, (b) reported spring events in units of calendar days (day of year or DOY) instead of a quantitative scale on a particular day, and (c) reported latitude and longitude of provenances and gardens.
-We had to exclude studies that reported spring events on a quantitative scale because (1) such studies usually only assessed where on the scale the spring event of a tree fell onto on the same days across different years (e.g.
-Robson et al., 2013; Vander et al., 2015; Santini et al., 2014; Schueler \& Liesebach, 2004), and (2) scales are not always consistent across different studies.
-Such factors made it impossible to back convert the quantitative scale to DOY.
-\\
-\\
-Some studies did not provide the exact latitude and longitude of the common garden (Bongarten, 1978) or the provenances (Hall et al., 2007; Soolanayakanahally et al., 2013), or they did not link the latitude and longitude of each provenance to the DOY of spring events (Deans and Harvey, 1996).
+
+Some studies did not provide the exact latitude and longitude of the common garden (Bongarten, 1978) or the provenances (Hall et al., 2007; Soolanayakanahally et al., 2013), or they did not link the latitude and longitude of each provenance to the DOY of spring events (Deans \& Harvey, 1996).
 We had to exclude these papers.
-Finally, studies that focused on examining how provenance altitude influenced local adaptation in spring phenology were excluded since we are interested in testing latitudinal effects (Vitasse et al., 2009; Vitasse et al., 2010; Li et al., 1997; Alberto et al., 2011; Acevedo-Rodríguez et al., 2006).
+We also left out studies in which woody plants from North American provenances were planted in common gardens in Europe because we wanted to test continental variations (Cannell et al., 1987; Lavadinovic et al., 2013).
+Finally, we excluded studies that focused on examining how provenance altitude influenced local adaptation in spring phenology since we are interested in testing latitudinal effects (Vitasse et al., 2009; Vitasse et al., 2010; Li et al., 1997; Alberto et al., 2011; Acevedo-Rodríguez et al., 2006).
 
-  \begin{itemize}
-\item Read AB 2015 and A 2013 for anything that we missed, took paper with spring doy and provenance coordinates from them
-
-\item Out of XX papers related to common garden/spring phenology that we read, only 22 ultimately supplied the data that we needed (DOY of spring events, latitude and longitude of provenances and common gardens); Out of the 22, four studies had provenance and gardens on different continents, which were subsequently archived (can document in supplements)).
-\end{itemize}
+% Read AB 2015 and A 2013 for anything that we missed, took paper with spring doy and provenance coordinates from them
 
 \subsection{Dataset Assembly}
-We assembled a dataset that documents the DOY of spring events and coordinates of provenances, during which we used ImageJ (version XXXX, citation) to scrape data from figures whenever possible and necessary.
-\\
-Everything was converted to day of year (some experiment counted not from january 1 but from march 1, etc.
-Need to talk about how we included data from field and indoor-experiment studies.\improvement{expand on this and give examples}
-\\
-\\
-It is worth noting that most common garden studies that supplied data only had observations of one species.
-Therefore, it is difficult for a model to interpret if variations across gardens are due to the species or the garden’s location and climate.
-We thus focused on estimating species effects because we expect a larger effect from species.
-Therefore, we considered unique gardens and tree species to represent independent data within publications, and fitted each species \& common gardens instead of just species.
-\improvement{need to phrase this better, and maybe put this somewhere in the modelling methods}
-\\
-\\
-Studies were distributed throughout North America and Europe, with the majority of data concentrated in XXXXX (Figure X).
+
+Out of XX papers related to common garden/spring phenology that we read, only 18 (?) ultimately supplied the data that we needed (DOY of spring events, latitude and longitude of provenances and common gardens).
+\improvement{need to rewrite this; can display a table here} Studies were distributed throughout North America and Europe, with the majority of data concentrated in XXXXX (Figure X).
 The distribution was skewed toward North America (n = X), with only X located in Europe.
 We primarily focused our analysis on spring events, and made use of the fall event DOYs present in X of the studies we included rather than searching for publications entirely on fall events online.
+\\
+% We retained studies that supplied DOY of spring events and geographic coordinates of provenances and common gardens.
+
+We assembled a data set that documents event DOY and geographic coordinates of provenances and gardens, during which we used ImageJ (version 1.53k; Rasband, 1997-2022) to extract values from figures whenever possible and necessary.
+We pooled the timing of budburst and leaf flush into a single category of ‘spring events’ and the timing of bud set, leaf senescence, growth cessation, and leaf abscission into ‘fall events.’ Such pooling is justified because of the shared pressures from natural selection that govern these events (Gill et al., 2015).
+Some studies documented event dates relative to a reference date that is not January 1 (e.g.
+Cannel et al., 1987; Rehfeldt, 1994).
+In this case, we converted such dates to DOY using the Lubridate Package in R for standardization purposes.
+\\
+
+Some studies had more than one gardens (), two studies shared the same garden ().\improvement{can mention this in table caption}
+
+Most common garden studies that supplied data only had observations of one tree species, which makes it difficult for a model to interpret if variations across gardens are due to the species or the garden’s location and climate.
+We thus focused on estimating species effects because we expect a larger effect from species.
+Within publications, we considered unique gardens and species to represent independent data, and fitted each species \& common gardens instead of just species.
+\improvement{need to phrase this better, and maybe put this somewhere in the modelling methods}
 
@@
 \subsubsection{Estimate climate overlap using daily climate data using Overlap R package}
-\subsubsection{Calculated GDDs, spherical distances}
+\subsubsection{Calculate GDDs}
+
+Growing degree days (GDD) is a commonly used heat accumulation measure to forecast phenological development in plants (Miller et al., 2001).
+We calculated GDD based on the accumulation of mean daily temperatures ($T_{m}$)from 2011 to 2022 above a baseline of 0$^{\circ}$C, from January 1 until budburst and leaf flush with the following formula: GDD = $\sum$($T_{m}$ - $0^{\circ}$C for $T_{m}$ $\ge$ $0^{\circ}$C; 0 for $T_{m}$ $\le$ $0^{\circ}$C)
+
+\subsubsection{Spherical distances}
+
 \subsubsection{Corrected for DOY (taking the difference)}\todo{will put results in supplement}
@@
 \todo[inline]{all subsection titles are placeholders}
-Out of the X papers on common garden experiments and spring phenology that we reviewed, only 22 supplied data met our criteria to be included: (a) focus on woody plants originating from either Europe or North America, (b) report spring event in units of calendar days instead of a quantitative scale on a particular day, (c) report locations of provenances and gardens.
-Out of the 22, four studies had provenances and common gardens on different continents.
+Out of the X papers on common garden experiments and spring phenology that we reviewed, only 18(?) supplied data met our criteria to be included: (a) focus on woody plants originating from either Europe or North America, (b) report spring event in units of calendar days instead of a quantitative scale on a particular day, (c) report locations of provenances and gardens.
+
+% Out of the 22, four studies had provenances and common gardens on different continents.
 These studies were excluded from our main analysis, but we document them in our supplements.\change{need to reword}
-\\
-\\
 Our final dataset captures seven angiosperm and eight gymnosperm species from 19 common gardens, encompassing data from 384 North American provenances and 101 European provenances with observations from 1962 to 2019.
 Seven species had fall event information available.
-\\
 \\
@@
 
+Rasband, W.S. (1997-2016). Image J.U.S. National Institute of Health, Bethesda, Maryland, USA.
+http://imagej.nih.gov/ij
+
 \bibliography{bibliography_local_adaptation.bib}

### 2022-12-15 Analyses/script_UPDATED_fit_garden_spp_Dec13_2022.R

@@ -1,0 +1,332 @@
+# This script documents updated fits that we need to run
+# All results are for unique sp x garden (not just unique spp)
+# Dec 13, 2022
+
+
+# for previous fits, refer to script_stan_glmer_NEW_fit_plots_are_in_here
+
+
+
+#most updated version
+d<- read.csv("input/percentage_overlap_doy_difference_earth_calculated_garden_identifier_adjusted_fall_diffo_included_slope_intercept_gdd_Sept27.csv",header = TRUE)
+
+
+# write first so i don't lose progress
+name<-"input/percentage_overlap_doy_difference_earth_calculated_garden_identifier_adjusted_fall_diffo_included_slope_intercept_gdd_Sept27.csv"
+write.csv(d,name, row.names = FALSE)
+
+
+
+# spring or fall DOY ~ lat or mat
+fitC_spring_lat <- stan_glmer(spring_event~(lat_prov|species_garden), data = d)
+fitC_fall_lat <- stan_glmer(fall_event~(lat_prov|species_garden), data = d)
+fitC_spring_mat <- stan_glmer(spring_event~(MAT_prov|species_garden), data = d)
+fitC_fall_mat <- stan_glmer(fall_event~(MAT_prov|species_garden), data = d)
+
+# lat diffo
+fitC_spring_lat_diffo <- stan_glmer(spring_event~(distance_from_garden|species_garden), data = d)
+fitC_fall_lat_diffo <- stan_glmer(fall_event~(distance_from_garden|species_garden), data = d)
+fitC_spring_earth_distance <- stan_glmer(spring_event~(earth_distance_from_garden|species_garden), data = d)
+fitC_fall_earth_distance <- stan_glmer(fall_event~(earth_distance_from_garden|species_garden), data = d)
+
+# mat diffo
+fitC_spring_mat_diffo <- stan_glmer(spring_event~(MAT_diffo|species_garden), data = d)
+fitC_fall_mat_diffo <- stan_glmer(fall_event~(MAT_diffo|species_garden), data = d)
+
+# climate overlap
+fitC_spring_overlap <- stan_glmer(spring_event~(percentage|species_garden), data = d)
+fitC_fall_overlap <- stan_glmer(fall_event~(percentage|species_garden), data = d)
+
+# gdd
+fitD_spring_lat <- stan_glmer(gdd_10yr_mean~(lat_prov|species_garden), data = d)
[…]

### 2022-12-15 Analyses/script_stan_glmer_NEW_latmat_difference_SEPT16_PLOTTING_SCRIPT.R

@@ -6,6 +6,7 @@ library(Rcpp)
 library(rstanarm)
 library(bayesplot)
 library(ggplot2)
+library(viridis)
 
 
 

### 2022-12-23 Docs/manuscript.Rnw
updated methods

@@
 \subsection{Climate Data Gathering \& Analyses}
-  \begin{itemize}
-  \item Simple climate matrix (MAT \& MSP/MAP) gathering
-  \begin{itemize}
-  \item ClimateNA by Dr. Tongli Wang
-  \item Climate Information Tool by FAO
-  \end{itemize}
-  \item Daily climate data gathering: Gridded daily climate data downloaded for all provenances and gardens: We hope that coarse metrics such as latitude and MAT ultimately represent how similar the climates are between the provenances and gardens in times that matter for the events.
+With information about provenance latitude, longitude, and elevation from original publications, we estimated the mean annual temperature (MAT) for each provenance using the Climate Information Tool by FAO (Food and Agriculture Organization of the United Nations, 2022) and ClimateWNA (Wang et al., 2012).
+\\
+To assess climate overlap percentage (define this better), we extracted gridded daily climate data for 2011 to 2020 for all European and North American provenances and gardens from E-OBS and the built-in Daymet package in R respectively (Cornes et al., 2018; Thornton et al., 2020).
+
+\subsubsection{Estimate climate overlap using daily climate data using Overlap R package}
+We hope that coarse metrics such as latitude and MAT ultimately represent how similar the climates are between the provenances and gardens in times that matter for the events.
 If climates are very similar, then we would expect similar timings [add more here].
 To this end, we estimated climate overlap in relevant months: For spring events, we considered overlap across March to May.
-  \begin{itemize}
-  \item North American daily climate data from Daymet R package
-  \item Europe daily climate data (gridded: NetCDF-4) retrieved from E-OBS 
-  \end{itemize}
-  \end{itemize}
+\\
+Using the Overlap package in R, we calculated the percentage overlap of the daily temperature of each provenance in spring months (March to May) from 2011 to 2020 in relation to the daily temperature of their corresponding gardens.
 
-\subsubsection{Estimate climate overlap using daily climate data using Overlap R package}
 \subsubsection{Calculate GDDs}
@@
 
+hmmm need to better define this... 
+On top of taking the difference between provenance latitude and garden latitude (maybe I could write this in subscript), we also use the dism() function in the Geosphere package in R to estimate the geospherical distance between locations (accounting for Earth curvature).\improvement{maybe there are different/better terms}
+
 \subsubsection{Corrected for DOY (taking the difference)}\todo{will put results in supplement}
+(need to better define this as well)
+
 \subsection{Mixed effects modelling}

### 2023-01-24 Docs/manuscript.Rnw
Edits for methods and results

@@
 \subsection{Literature Search}
+%emw24Jan -- this section looks good; once you're happy with it, I can slim it down a little and suggest a few things to cut or move to supp.
 
-We searched the peer-reviewed literature for common garden experiments that documented the timing of spring events of woody plant species.
-We searched on the Web of Science (Thompson Reuters, New York, NY) using all combinations of the following terms as ‘titles’ including ‘common garden*’ or “provenance*’ and ‘leafout*’ or ‘leaf out*’ or ‘budburst’ or ‘spring phenolog*’.
-Our search was limited to journal articles published before November 3, 2021.
+We searched the peer-reviewed literature for common garden experiments that documented the timing of spring events of woody plant species on 14 December 2022 using Web of Science (Thompson Reuters, New York, NY).
+Our search terms using all combinations of the following terms as ‘titles’ including ‘common garden*’ or “provenance*’ and ‘leafout*’ or ‘leaf out*’ or ‘budburst’ or ‘spring phenolog*’.
 The search returned 94 publications.
+%emw24Jan -- see edits that address your question.
+We can just give the last date we searched.
+Also, can you check your terms?
+It's not clear what the search was.
+...
+I would just write it out as we have in other papers.
+Here's an example:
+% TOPIC = (budburst OR leafout) AND (photoperiod or daylength) AND temperature*, which yielded 85 publications
 \\
 
-\todo[inline]{question for lizzie: I later checked 2022 publications too to make sure we didn't miss anything...
-Do I just say at the very beginning that our search is limited to publications before Dec 14, 2022, or do I say 'we later checked the literature again on XX-XX-2022 to ensure we did not miss any new publications?}
 \todo[inline]{I also think the reviewers might question why we had so few search results....
@@
 to make sure we didn't miss anything?}
+%emw24Jan -- I agree!
+You can say you also searched google scholar (but then you have to say how and when), and contacted authors of previous review papers to help further search the literature.
+The other thing you could do is your search with TOPIC instead of TITLE, as reviewers will find TITLE suspicious -- even though I suspect you have found all the main papers.
 
@@
 Some studies had more than one gardens (), two studies shared the same garden ().\improvement{can mention this in table caption}
-
+%emw24Jan -- yes, we can move the below to modeling section
 Most common garden studies that supplied data only had observations of one tree species, which makes it difficult for a model to interpret if variations across gardens are due to the species or the garden’s location and climate.
@@
 
+%emw24Jan -- I think we can delete this subsection (or move to supp for now). Just define clearly in caption what you mean. 
 \subsubsection{Spherical distances}
@@
 
+%emw24Jan -- I can add to this. 
 \subsection{Mixed effects modelling}
@@
 \todo[inline]{all subsection titles are placeholders}
-Out of the X papers on common garden experiments and spring phenology that we reviewed, only 18(?) supplied data met our criteria to be included: (a) focus on woody plants originating from either Europe or North America, (b) report spring event in units of calendar days instead of a quantitative scale on a particular day, (c) report locations of provenances and gardens.
+%emw24Jan -- the below is in the methods so not needed here. 
+% Out of the X papers on common garden experiments and spring phenology that we reviewed, only 18(?) supplied data met our criteria to be included: (a) focus on woody plants originating from either Europe or North America, (b) report spring event in units of calendar days instead of a quantitative scale on a particular day, (c) report locations of provenances and gardens.
 
@@
 These studies were excluded from our main analysis, but we document them in our supplements.\change{need to reword}
+%emw24Jan -- below is great opening!
 Our final dataset captures seven angiosperm and eight gymnosperm species from 19 common gardens, encompassing data from 384 North American provenances and 101 European provenances with observations from 1962 to 2019.
@@
 \\
-It is worth noting that most common gardens included in this study only had observations of one species.
+%emw24Jan -- you can basically always delete 'we note' or anything similar. Also, I think this is in the methods or should be, so remove from here. 
+Most common gardens included in this study only had observations of one species.
 Therefore, it is difficult for a model to interpret if variations across gardens are due to the species or the garden’s location and climate.
@@
 \subsection{Provenance latitude and MAT do not affect spring event timing}
-
+%emw24Jan -- below is good, can you also explain the direction of relationship in europe? Do events get later/earlier with higher/lower latitude and MAT?
 Overall, our models show that spring events are not related to provenance latitude or mean annual temperature in North America, and only weakly related in Europe (Fig.
-\ref{figure:springfall_latmat} \& Supplement Table.\ref{supplement-table:model_sf_lat}).
-\unsure{Should I add climate overlap plots here too?}
+\ref{figure:springfall_latmat} \& Supplement Table.
+\ref{supplement-table:model_sf_lat}).
+\unsure{Should I add climate overlap plots here too?} %emw24Jan -- do you mean add a table of climate overlap to the supp?
 
@@
 \subsection{Stronger clines in fall events observed in North America than Europe}
+%emw24Jan -- Nice! Also, results seem good, I can help trim text and collapse some sections.
 

### 2023-01-28 Docs/manuscript.Rnw
updated methods

--- 2022-12-15 Docs/manuscript.Rnw
@@
 \author{Ziyun Zeng \& E. M. Wolkovich}
-\date{November 2022}
+\date{2023}
 \maketitle
@@
 Clines along climate gradients are thus espcially likely to reflect local adaptation to climate, rather than non-adaptive phenomena resulting from population demographic history."
+\\
+
+% alina_Jan27 -- I think we should have a section to define "spring events, fall events, distance between garden locations, difference of latitude, etc.
+Where do you think this section should go?
+Methods?
+
+% I am pasting a paragraph of the method section of Ettinger below to make reference to later on when I write the definition section. 
+"Defining budburst. Most studies defined budburst as initial ‘green tips’ (33 out of
+49 papers). Select studies defined budburst as a specific increment of growth (for
+example, ‘0.5 cm of new growth’) or as bud swell, leaf emergence, leaf unfolded,
+open bud scales or petiole emerged. The remaining papers (4 of 49) did not
+include a definition of budburst. Most papers using the above definitions (34 of 49)
+required only one bud to have met the defined criteria of budburst; however, the
+remaining studies implemented specific thresholds to be met (10–100\% of all buds
+on an individual needed to have bursted bud). For studies that quantified multiple
+measurements of percentage budburst over time (days), we extracted one value of
+‘days to budburst’ of these multiple measurements to make them comparable to
+other studies. To extract this summary value, we selected the days to budburst when
+percentage budburst was closest to 90\%, including estimates as low as 49.5\% budburst"
 
@@
 \subsection{Literature Search}
+%emw24Jan -- this section looks good; once you're happy with it, I can slim it down a little and suggest a few things to cut or move to supp.
 
-We searched the peer-reviewed literature for common garden experiments that documented the timing of spring events of woody plant species.
-We searched on the Web of Science (Thompson Reuters, New York, NY) using all combinations of the following terms as ‘titles’ including ‘common garden*’ or “provenance*’ and ‘leafout*’ or ‘leaf out*’ or ‘budburst’ or ‘spring phenolog*’.
-Our search was limited to journal articles published before November 3, 2021.
-The search returned 94 publications.
+We searched the peer-reviewed literature for common garden experiments that documented the timing of spring events of woody plant species on 14 December 2022 using Web of Science (Thompson Reuters, New York, NY)with the following terms:
 \\
-
-\todo[inline]{question for lizzie: I later checked 2022 publications too to make sure we didn't miss anything...
-Do I just say at the very beginning that our search is limited to publications before Dec 14, 2022, or do I say 'we later checked the literature again on XX-XX-2022 to ensure we did not miss any new publications?}
-\todo[inline]{I also think the reviewers might question why we had so few search results....
-maybe I should mention that I also checked on Google Scholar and Connected Papers, and Aitken Bammals 2016/Alberto 2013, and emailed our list of papers to established researchers such as Aitken and Guy, etc.
-to make sure we didn't miss anything?}
-
-We read the methods and results of all publications returned and limited our subsequent analysis to the ones that (a) focused on woody plants originating from either Europe or North America, (b) had provenance trails and common gardens on the same continent, (c) reported spring events in units of calendar days (day of year or DOY) instead of a quantitative scale on a particular day, and (d) reported latitude and longitude of provenances and gardens.
+\\
+TOPIC = (common garden* OR provenance*) AND (leafout* OR leaf out* OR budburst OR spring phenolog*).
+The search returned 122 publications.
+We also contacted authors of previous review papers, including S.N.
+Aitken and R.D.
+Guy to help further search the literature.
+\\
+\\
+We read the methods and results of all publications and limited our subsequent analysis to the ones that (a) focused on woody plants originating from either Europe or North America, (b) had provenance trails and common gardens on the same continent, (c) reported spring events in units of calendar days (day of year or DOY) instead of a quantitative scale on a particular day, and (d) reported latitude and longitude of provenances and gardens.
+\\
 \\
@@
 Such factors made it impossible to back convert the quantitative scale to DOY.
-\todo[inline]{add these citations to bibtex?? do we cite them?}
+\\
+\\
+@ alina_jan27: do we need to cite those paper indicated above that we didnt end up using through bibtex??
 
@@
 
-% Read AB 2015 and A 2013 for anything that we missed, took paper with spring doy and provenance coordinates from them
-
 \subsection{Dataset Assembly}
 
-Out of XX papers related to common garden/spring phenology that we read, only 18 (?) ultimately supplied the data that we needed (DOY of spring events, latitude and longitude of provenances and common gardens).
-\improvement{need to rewrite this; can display a table here} Studies were distributed throughout North America and Europe, with the majority of data concentrated in XXXXX (Figure X).
+Out of the 122 common garden studies on spring phenology that we read, only 18 (?) ultimately supplied the data that we needed (DOY of spring events, latitude and longitude of provenances and common gardens).
+Studies were distributed throughout North America and Europe, with the majority of data concentrated in XXXXX (Figure X).
 The distribution was skewed toward North America (n = X), with only X located in Europe.
 We primarily focused our analysis on spring events, and made use of the fall event DOYs present in X of the studies we included rather than searching for publications entirely on fall events online.
+\improvement{display a table and a map here, and in the caption, clarify that some studies had more than one gardens and two studies shared the same garden}
 \\
-% We retained studies that supplied DOY of spring events and geographic coordinates of provenances and common gardens.
-
+\\
 We assembled a data set that documents event DOY and geographic coordinates of provenances and gardens, during which we used ImageJ (version 1.53k; Rasband, 1997-2022) to extract values from figures whenever possible and necessary.
@@
 In this case, we converted such dates to DOY using the Lubridate Package in R for standardization purposes.
-\\
-
-Some studies had more than one gardens (), two studies shared the same garden ().\improvement{can mention this in table caption}
-
-Most common garden studies that supplied data only had observations of one tree species, which makes it difficult for a model to interpret if variations across gardens are due to the species or the garden’s location and climate.
-We thus focused on estimating species effects because we expect a larger effect from species.
-Within publications, we considered unique gardens and species to represent independent data, and fitted each species \& common gardens instead of just species.
-\improvement{need to phrase this better, and maybe put this somewhere in the modelling methods}
 
 \subsection{Climate Data Gathering \& Analyses}
-  \begin{itemize}
-  \item Simple climate matrix (MAT \& MSP/MAP) gathering
-  \begin{itemize}
-  \item ClimateNA by Dr. Tongli Wang
-  \item Climate Information Tool by FAO
-  \end{itemize}
-  \item Daily climate data gathering: Gridded daily climate data downloaded for all provenances and gardens: We hope that coarse metrics such as latitude and MAT ultimately represent how similar the climates are between the provenances and gardens in times that matter for the events.
+\subsubsection {placeholder}
+With information about provenance latitude, longitude, and elevation from original publications, we estimated the mean annual temperature (MAT) for each provenance using the Climate Information Tool by FAO (Food and Agriculture Organization of the United Nations, 2022) and ClimateWNA (Wang et al., 2012).
+\\
+\\
+We hope that coarse metrics such as latitude and MAT ultimately represent how similar the climates are between the provenances and gardens in times that matter for the events.
 If climates are very similar, then we would expect similar timings [add more here].
 To this end, we estimated climate overlap in relevant months: For spring events, we considered overlap across March to May.
-  \begin{itemize}
-  \item North American daily climate data from Daymet R package
-  \item Europe daily climate data (gridded: NetCDF-4) retrieved from E-OBS 
-  \end{itemize}
-  \end{itemize}
+To assess climate overlap percentage (define this better), we extracted gridded daily temperature data for 2011 to 2020 for all European and North American provenances and gardens from E-OBS and the built-in Daymet package in R respectively (Cornes et al., 2018; Thornton et al., 2020).
+Using the Overlap package in R, we calculated the percentage overlap of the daily temperature of each provenance in spring months (March to May) from 2011 to 2020 in relation to the daily temperature of their corresponding gardens.
 
-\subsubsection{Estimate climate overlap using daily climate data using Overlap R package}
 \subsubsection{Calculate GDDs}
@@
 We calculated GDD based on the accumulation of mean daily temperatures ($T_{m}$)from 2011 to 2022 above a baseline of 0$^{\circ}$C, from January 1 until budburst and leaf flush with the following formula: GDD = $\sum$($T_{m}$ - $0^{\circ}$C for $T_{m}$ $\ge$ $0^{\circ}$C; 0 for $T_{m}$ $\le$ $0^{\circ}$C)
+\\
 
-\subsubsection{Spherical distances}
+define this in caption, no need to have is as a section: Spherical distances 
+hmmm need to better define this... 
+On top of taking the difference between provenance latitude and garden latitude (maybe I could write this in subscript), we also use the dism() function in the Geosphere package in R to estimate the geospherical distance between locations (accounting for Earth curvature).\improvement{maybe there are different/better terms}
 
 \subsubsection{Corrected for DOY (taking the difference)}\todo{will put results in supplement}
+(need to better define this as well)
+
+%emw24Jan -- I can add to this. 
 \subsection{Mixed effects modelling}
+
+Most tree species are only represented in one common garden study in our dataset.
+In such instances, it is difficult to statistically interpret if variations across gardens are due to the species or the garden’s location and climate.
+To address this, we focused on estimating species effects as we expect a larger effect from species.
+Within publications, we considered unique gardens and species to represent independent data, and fitted each species \& common gardens instead of just species.
 
@@
 \todo[inline]{all subsection titles are placeholders}
-Out of the X papers on common garden experiments and spring phenology that we reviewed, only 18(?) supplied data met our criteria to be included: (a) focus on woody plants originating from either Europe or North America, (b) report spring event in units of calendar days instead of a quantitative scale on a particular day, (c) report locations of provenances and gardens.
 
-% Out of the 22, four studies had provenances and common gardens on different continents.
-These studies were excluded from our main analysis, but we document them in our supplements.\change{need to reword}
+%emw24Jan -- below is great opening!
 Our final dataset captures seven angiosperm and eight gymnosperm species from 19 common gardens, encompassing data from 384 North American provenances and 101 European provenances with observations from 1962 to 2019.
@@
 \\
-It is worth noting that most common gardens included in this study only had observations of one species.
-Therefore, it is difficult for a model to interpret if variations across gardens are due to the species or the garden’s location and climate.
-\improvement{can refer to the fit we did where we fitted each species \& common garden instead of just species} We thus focused on estimating species effects because we expect a larger effect from species.
 
@@
 \subsection{Provenance latitude and MAT do not affect spring event timing}
-
+%emw24Jan -- below is good, can you also explain the direction of relationship in europe?
+Do events get later/earlier with higher/lower latitude and MAT?
+% alina_jan27 -- I believe I described that in later paragraphs.
+Would you like me to move it up?
 Overall, our models show that spring events are not related to provenance latitude or mean annual temperature in North America, and only weakly related in Europe (Fig.
-\ref{figure:springfall_latmat} \& Supplement Table.\ref{supplement-table:model_sf_lat}).
-\unsure{Should I add climate overlap plots here too?}
+\ref{figure:springfall_latmat} \& Supplement Table.
+\ref{supplement-table:model_sf_lat}).
+\unsure{Should I add climate overlap plots here too?} %emw24Jan -- do you mean add a table of climate overlap to the supp?
+% alina_jan27 -- I meant if the plots where I plotted climate overlap percentage against DOY should be included here too, so easier for people to compare perhaps?
 

### 2023-01-28 Docs/manuscript.tex
updated methods

--- 2022-12-15 Docs/manuscript.tex
@@
 Such factors made it impossible to back convert the quantitative scale to DOY.
-\todo[inline]{add these citations to bibtex?? do we cite them?}
+\\
+\\
 

### 2023-02-05 Docs/manuscript.Rnw
Replies to questions
updated ms

@@
 Methods?
+EMW replies -- I think defining the events would be good, distance between garden locations and latitude etc should be SHORT (for example we use XX R package to calculate yy).
+I would put all this in the methods.
+The definition of events can go in the dataset assembly and I would start a section on 'analyses' where you include most of 2.3 *and* 2.4.
+I would move up where you got climate data from to above -- think of a 'Data' section with some subheaders and an 'analyses' section with some subheaders.
 
@@
 \section{Methods}
-Our methods are divided into four main steps.
 
 % \item Literature search
-\subsection{Literature Search}
+\subsection{Data Collection}
+\subsubsection{Literature Search}
 %emw24Jan -- this section looks good; once you're happy with it, I can slim it down a little and suggest a few things to cut or move to supp. 
-
 We searched the peer-reviewed literature for common garden experiments that documented the timing of spring events of woody plant species on 14 December 2022 using Web of Science (Thompson Reuters, New York, NY)with the following terms:
@@
 @ alina_jan27: do we need to cite those paper indicated above that we didnt end up using through bibtex??
-
+% EMW replies -- no, some journals will require a spreadsheet (what you discarded and why) though -- but until we pick a journal and know what's required we don't need to do any of that.
 Some studies did not provide the exact latitude and longitude of the common garden (Bongarten, 1978) or the provenances (Hall et al., 2007; Soolanayakanahally et al., 2013), or they did not link the latitude and longitude of each provenance to the DOY of spring events (Deans \& Harvey, 1996).
@@
 
-\subsection{Dataset Assembly}
-
+\subsubsection{Dataset Assembly}
 Out of the 122 common garden studies on spring phenology that we read, only 18 (?) ultimately supplied the data that we needed (DOY of spring events, latitude and longitude of provenances and common gardens).
@@
 
-\subsection{Climate Data Gathering \& Analyses}
-\subsubsection {placeholder}
+\subsubsection {Climate Data Collection}
 With information about provenance latitude, longitude, and elevation from original publications, we estimated the mean annual temperature (MAT) for each provenance using the Climate Information Tool by FAO (Food and Agriculture Organization of the United Nations, 2022) and ClimateWNA (Wang et al., 2012).
@@
 To assess climate overlap percentage (define this better), we extracted gridded daily temperature data for 2011 to 2020 for all European and North American provenances and gardens from E-OBS and the built-in Daymet package in R respectively (Cornes et al., 2018; Thornton et al., 2020).
+
+\subsection{Analyses}
+\subsubsection{Variable preparation}
 Using the Overlap package in R, we calculated the percentage overlap of the daily temperature of each provenance in spring months (March to May) from 2011 to 2020 in relation to the daily temperature of their corresponding gardens.
-
-\subsubsection{Calculate GDDs}
-
-Growing degree days (GDD) is a commonly used heat accumulation measure to forecast phenological development in plants (Miller et al., 2001).
-We calculated GDD based on the accumulation of mean daily temperatures ($T_{m}$)from 2011 to 2022 above a baseline of 0$^{\circ}$C, from January 1 until budburst and leaf flush with the following formula: GDD = $\sum$($T_{m}$ - $0^{\circ}$C for $T_{m}$ $\ge$ $0^{\circ}$C; 0 for $T_{m}$ $\le$ $0^{\circ}$C)
-\\
-
-define this in caption, no need to have is as a section: Spherical distances 
-hmmm need to better define this... 
-On top of taking the difference between provenance latitude and garden latitude (maybe I could write this in subscript), we also use the dism() function in the Geosphere package in R to estimate the geospherical distance between locations (accounting for Earth curvature).\improvement{maybe there are different/better terms}
-
-\subsubsection{Corrected for DOY (taking the difference)}\todo{will put results in supplement}
-(need to better define this as well)
-
+We calculated Growing degree days (GDD), a commonly used heat accumulation measure to forecast phenological development in plants (Miller et al., 2001), based on the accumulation of mean daily temperatures ($T_{m}$)from 2011 to 2022 above a baseline of 0$^{\circ}$C, from January 1 until budburst and leaf flush with the following formula: GDD = $\sum$($T_{m}$ - $0^{\circ}$C for $T_{m}$ $\ge$ $0^{\circ}$C; 0 for $T_{m}$ $\le$ $0^{\circ}$C)
 %emw24Jan -- I can add to this. 
-\subsection{Mixed effects modelling}
-
+\subsubsection{Mixed effects modelling}
 Most tree species are only represented in one common garden study in our dataset.
@@
 \subsection{Provenance latitude and MAT do not affect spring event timing}
-%emw24Jan -- below is good, can you also explain the direction of relationship in europe?
-Do events get later/earlier with higher/lower latitude and MAT?
-% alina_jan27 -- I believe I described that in later paragraphs.
-Would you like me to move it up?
-Overall, our models show that spring events are not related to provenance latitude or mean annual temperature in North America, and only weakly related in Europe (Fig.
+Overall, our models show that spring events are not related to provenance latitude or mean annual temperature in North America, and only weakly related in Europe, where springs events are slightly earlier where provenance mean annual temperature is lower (higher, more northern latitudes)(Fig.
 \ref{figure:springfall_latmat} \& Supplement Table.
 \ref{supplement-table:model_sf_lat}).
-\unsure{Should I add climate overlap plots here too?} %emw24Jan -- do you mean add a table of climate overlap to the supp?
-% alina_jan27 -- I meant if the plots where I plotted climate overlap percentage against DOY should be included here too, so easier for people to compare perhaps?
 
@@
 
-\subsubsection {Similar results across provenance latitude, latitude difference, and distance}
-\todo[inline]{need to find a name for each metric and stick to it.}
-Along with (a) provenance latitude, we also looked at how spring event timing is related to (b) the absolute difference between provenance latitude and garden latitude , as well as (c) the spherical distance between the garden and each provenance.
+\subsubsection {Similar results across provenance latitude, absolute value of difference between provenance and garden latitude, and spherical distance}
+Along with (a) provenance latitude, we also looked at how spring event timing is related to (b) the absolute value of difference between provenance and garden latitude , as well as (c) the spherical distance between the garden and each provenance.
 All three metrics depict little to no relationship between spring event timing and the geographical location of the provenance (Supplement Fig.\ref{supplement-figure:lat_distance}).

### 2023-02-05 Docs/manuscript.tex
updated ms

@@
 \\
-
 Some studies did not provide the exact latitude and longitude of the common garden (Bongarten, 1978) or the provenances (Hall et al., 2007; Soolanayakanahally et al., 2013), or they did not link the latitude and longitude of each provenance to the DOY of spring events (Deans \& Harvey, 1996).

### 2023-02-05 Docs/supplement.Rnw
Worked on methods
updated ms

@@
 
-\subsection*{Similar results across provenance latitude, latitude difference, and distance}
+\subsection*{Similar results across provenance latitude, absolute value of difference between provenance and garden latitude, and spherical distance}
 Placeholder text (Fig. \ref{figure:lat_distance}).
@@
  \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_supp/lat_latdiff_earth_distance.png}
-    \caption{Similar results across provenance latitude, latitude difference, and distance}
+    \caption{Similar results across provenance latitude, absolute value of difference between provenance and garden latitude, and spherical distance}
     \label{figure:lat_distance}

### 2023-03-03 Docs/manuscript.Rnw
Lots of work on methods ....
See %emw2Mar for comments. I still need to do results!

@@
 \subsection{Data Collection}
-\subsubsection{Literature Search}
-%emw24Jan -- this section looks good; once you're happy with it, I can slim it down a little and suggest a few things to cut or move to supp. 
-We searched the peer-reviewed literature for common garden experiments that documented the timing of spring events of woody plant species on 14 December 2022 using Web of Science (Thompson Reuters, New York, NY)with the following terms:
+%emw2Mar: Try to open with WHY (goal/aim) you did something more than what you did, not a rule, but try to do when possible. 
+To locate common garden studies that documented the timing of spring events of woody plant species we searched and reviewed the peer-reviewed literature.
+On 14 December 2022 we searched Web of Science (Thompson Reuters, New York, NY) using the following terms:
+\begin{quote}
+TOPIC = (common garden* OR provenance*) AND (leafout* OR leaf out* OR budburst OR spring phenolog*)
+\end{quote}
+which returned 122 publications.
+We also contacted authors of previous review papers (CITES), to help further search the literature.
+We then reviewed the methods and results of all publications to refine to only studies that met the following criteria: (a) focused on woody plants originating from either Europe or North America, (b) had provenance trials/common gardens on the same continent, (c) reported spring events in units of calendar days (day of year or DOY) or could be converted into DOY (see `Additional Methods' section in the Supplement), and (d) reported latitude and longitude of provenances and gardens.
+%emw2Mar -- From above ...
+We don't usually name the people, we would cite the papers and acknowledge the people in the acknowledgements section.
+So add the citations above in CITES and add the below to the acknowledgements
+% We thank S. Aitken,  I. Chuine, R. Guy, C Korner and Y. Vitasse for reviewing our list of papers for possible additional common garden studies. 
+\\
+%emw2Mar: Below and throughout ...
+You should switch to citing references using bibtex throughout!
+Making that change now will be much easier than later.
+Ask Deirdre if you get stuck (doing it soon would also be good as she will be away starting in May and so busier in April).
+%emw2Mar -- mv the below paragraphs to an `Additional Methods' section of supp (also note how I added topic sentences here to guide the reader). 
+% START: Additional methods section:
+We had to exclude several studies that reported spring events on a quantitative scale.
+This was because (1) such studies usually only assessed where on the scale the spring event of a tree fell onto on the same days across different years (e.g.
+Robson et al., 2013; Vander et al., 2015; Santini et al., 2014; Schueler \& Liesebach, 2004), and (2) scales are not always consistent across different studies (Chmura \& Rozkowski 2002; Dhont et al., 2010; Wang et al., 2022).
+Such factors made it impossible to convert the quantitative scale to DOY.
 \\
 \\
-TOPIC = (common garden* OR provenance*) AND (leafout* OR leaf out* OR budburst OR spring phenolog*).
-The search returned 122 publications.
-We also contacted authors of previous review papers, including S.N.
-Aitken and R.D.
-Guy to help further search the literature.
+% @ alina_jan27: do we need to cite those paper indicated above that we didnt end up using through bibtex??
+% EMW replies -- no, some journals will require a spreadsheet (what you discarded and why) though -- but until we pick a journal and know what's required we don't need to do any of that.
+We additionally excluded several studies because we could not pinpoint their location, or because they focused on non-native species or elevational trends.
+We excluded studies that did not provide the exact latitude and longitude of the common garden (Bongarten, 1978) or the provenances (Hall et al., 2007; Soolanayakanahally et al., 2013), or they did not link the latitude and longitude of each provenance to the DOY of spring events (Deans \& Harvey, 1996).
+We also left out studies in which woody plants from North American provenances were planted in common gardens in Europe (Cannell et al., 1987; Lavadinovic et al., 2013) because we wanted to test continental variations.
+Finally, as we are focused on latitudinal trends we excluded studies that examined only provenance altitude (Vitasse et al., 2009; Vitasse et al., 2010; Li et al., 1997; Alberto et al., 2011; Acevedo-Rodríguez et al., 2006).
+% END: Additional methods section:
+
+Based on these criteria we found 18 (?) provenance/common garden studies, distributed throughout North America and Europe, with the majority of data concentrated in XXXXX (Figure X).
+From each paper we extracted phenological data on spring events (budburst and leaf flush in DOY) and, when present in the same paper, fall events (bud set, leaf senescence, growth cessation, and leaf abscission) by species, provenance and garden location (geographic coordinates), using ImageJ (version 1.53k; Rasband, 1997-2022) to extract values from figures whenever necessary.
+For studies that reported event dates relative to a reference date other than 1 January (e.g.
+Cannel et al., 1987; Rehfeldt, 1994), we converted such dates to DOY using the Lubridate Package in R.
+\improvement{display a table and a map here, and in the caption, clarify that some studies had more than one gardens and two studies shared the same garden}
+%emw2Mar -- mv the following sentence to the caption of the map figure: The distribution was skewed toward North America (n = X), with only X located in Europe. 
 \\
 \\
-We read the methods and results of all publications and limited our subsequent analysis to the ones that (a) focused on woody plants originating from either Europe or North America, (b) had provenance trails and common gardens on the same continent, (c) reported spring events in units of calendar days (day of year or DOY) instead of a quantitative scale on a particular day, and (d) reported latitude and longitude of provenances and gardens.
-\\
-\\
-We had to exclude studies that reported spring events on a quantitative scale because (1) such studies usually only assessed where on the scale the spring event of a tree fell onto on the same days across different years (e.g.
-Robson et al., 2013; Vander et al., 2015; Santini et al., 2014; Schueler \& Liesebach, 2004), and (2) scales are not always consistent across different studies (Chmura \& Rozkowski 2002; Dhont et al., 2010; Wang et al., 2022).
-Such factors made it impossible to back convert the quantitative scale to DOY.
-\\
-\\
-@ alina_jan27: do we need to cite those paper indicated above that we didnt end up using through bibtex??
-% EMW replies -- no, some journals will require a spreadsheet (what you discarded and why) though -- but until we pick a journal and know what's required we don't need to do any of that.
-Some studies did not provide the exact latitude and longitude of the common garden (Bongarten, 1978) or the provenances (Hall et al., 2007; Soolanayakanahally et al., 2013), or they did not link the latitude and longitude of each provenance to the DOY of spring events (Deans \& Harvey, 1996).
-We had to exclude these papers.
-We also left out studies in which woody plants from North American provenances were planted in common gardens in Europe because we wanted to test continental variations (Cannell et al., 1987; Lavadinovic et al., 2013).
-Finally, we excluded studies that focused on examining how provenance altitude influenced local adaptation in spring phenology since we are interested in testing latitudinal effects (Vitasse et al., 2009; Vitasse et al., 2010; Li et al., 1997; Alberto et al., 2011; Acevedo-Rodríguez et al., 2006).
+%emw2Mar -- save the below for now; may fit in the discussion or could add if reveiwers complain.
+For now, I added the events above -- it's quicker and IMHO we don't currently need to justify why we pooled these.
+% We pooled the timing of budburst and leaf flush into a single category of ‘spring events’ and the timing of bud set, leaf senescence, growth cessation, and leaf abscission into ‘fall events.’ Such pooling is justified because of the shared pressures from natural selection that govern these events (Gill et al., 2015).
 
-\subsubsection{Dataset Assembly}
-Out of the 122 common garden studies on spring phenology that we read, only 18 (?) ultimately supplied the data that we needed (DOY of spring events, latitude and longitude of provenances and common gardens).
-Studies were distributed throughout North America and Europe, with the majority of data concentrated in XXXXX (Figure X).
-The distribution was skewed toward North America (n = X), with only X located in Europe.
-We primarily focused our analysis on spring events, and made use of the fall event DOYs present in X of the studies we included rather than searching for publications entirely on fall events online.
-\improvement{display a table and a map here, and in the caption, clarify that some studies had more than one gardens and two studies shared the same garden}
-\\
-\\
-We assembled a data set that documents event DOY and geographic coordinates of provenances and gardens, during which we used ImageJ (version 1.53k; Rasband, 1997-2022) to extract values from figures whenever possible and necessary.
-We pooled the timing of budburst and leaf flush into a single category of ‘spring events’ and the timing of bud set, leaf senescence, growth cessation, and leaf abscission into ‘fall events.’ Such pooling is justified because of the shared pressures from natural selection that govern these events (Gill et al., 2015).
-Some studies documented event dates relative to a reference date that is not January 1 (e.g.
-Cannel et al., 1987; Rehfeldt, 1994).
-In this case, we converted such dates to DOY using the Lubridate Package in R for standardization purposes.
+To understand how climatic differences, in addition to geographical differences, shape local adaptation in spring events we extracted several types of climate data using information about provenance latitude, longitude, and elevation from original publications.
+We estimated the mean annual temperature (MAT) for each provenance using the Climate Information Tool by FAO (Food and Agriculture Organization of the United Nations, 2022) and ClimateWNA (Wang et al., 2012).
+To examine more explicitly climate near spring events, we used gridded daily temperature data for 2011 to 2020 for all European and North American provenances and gardens from E-OBS and the built-in Daymet package in R respectively (Cornes et al., 2018; Thornton et al., 2020).
+Using these data we estimated how much the daily temperatures overlapped between garden and provenance locations, which we call `climate overlap.' For this we used the Overlap package in R to calculate the percentage overlap of the daily temperature of each provenance in spring months (March to May) from 2011 to 2020 in relation to the daily temperature of their corresponding gardens.
+Finally, using the daily data, we also calculated Growing degree days (GDD), a commonly used heat accumulation measure to forecast phenological development in plants (Miller et al., 2001), based on the accumulation of mean daily temperatures ($T_{m}$)from 2011 to 2022 above a baseline of 0$^{\circ}$C, from January 1 until budburst and leaf flush with the following formula: GDD = $\sum$($T_{m}$ - $0^{\circ}$C for $T_{m}$ $\ge$ $0^{\circ}$C; 0 for $T_{m}$ $\le$ $0^{\circ}$C).
 
-\subsubsection {Climate Data Collection}
-With information about provenance latitude, longitude, and elevation from original publications, we estimated the mean annual temperature (MAT) for each provenance using the Climate Information Tool by FAO (Food and Agriculture Organization of the United Nations, 2022) and ClimateWNA (Wang et al., 2012).
-\\
-\\
-We hope that coarse metrics such as latitude and MAT ultimately represent how similar the climates are between the provenances and gardens in times that matter for the events.
+%emw2Mar -- there's a delicate balance of how much rationale you can give for your methods in the methods (and similar for the results). I would move the below to the discussion section. 
+% While coarse metrics such as latitude and MAT may represent how similar the climates are between the provenances and gardens in times that matter for the events.
 If climates are very similar, then we would expect similar timings [add more here].
-To this end, we estimated climate overlap in relevant months: For spring events, we considered overlap across March to May.
-To assess climate overlap percentage (define this better), we extracted gridded daily temperature data for 2011 to 2020 for all European and North American provenances and gardens from E-OBS and the built-in Daymet package in R respectively (Cornes et al., 2018; Thornton et al., 2020).
 
 \subsection{Analyses}
-\subsubsection{Variable preparation}
-Using the Overlap package in R, we calculated the percentage overlap of the daily temperature of each provenance in spring months (March to May) from 2011 to 2020 in relation to the daily temperature of their corresponding gardens.
-We calculated Growing degree days (GDD), a commonly used heat accumulation measure to forecast phenological development in plants (Miller et al., 2001), based on the accumulation of mean daily temperatures ($T_{m}$)from 2011 to 2022 above a baseline of 0$^{\circ}$C, from January 1 until budburst and leaf flush with the following formula: GDD = $\sum$($T_{m}$ - $0^{\circ}$C for $T_{m}$ $\ge$ $0^{\circ}$C; 0 for $T_{m}$ $\le$ $0^{\circ}$C)
-%emw24Jan -- I can add to this. 
-\subsubsection{Mixed effects modelling}
-Most tree species are only represented in one common garden study in our dataset.
-In such instances, it is difficult to statistically interpret if variations across gardens are due to the species or the garden’s location and climate.
-To address this, we focused on estimating species effects as we expect a larger effect from species.
-Within publications, we considered unique gardens and species to represent independent data, and fitted each species \& common gardens instead of just species.
-
-We used hierarchical Bayesian models where we partially pooled by species.
-We estimated effects of continent and species type (angiosperm VS gymnosperm) from posterior estimates, in relation to single predictors (including provenance latitude, MAT) and two predictors (climate overlap percentage and standard deviation).
-\improvement{will elaborate later}
+To estimate clines in spring and fall phenological events across species we used hierarchical Bayesian models.
+We regressed day of year (DOY) of events against geographical and climate predictors with partial pooling (sometimes called `random effects') on the intercept and slope for each species within each garden.
+Because most tree species were present in only one common garden in our dataset, it was impossible to fit garden and species separately, thus we treat each species within a garden as a unique group.
+Using posterior estimates for each each species within a garden, we estimated effects of continent and species type (angiosperm vs.
+gymnosperm).
+\improvement{will elaborate later ...
+ADD: All models were fit in XX package (give version number), using default priors, with 4 chains and XX sampling iterations per chain.
+We checked for model fit by ...
+}
+%emw2Mar -- Check: Did you use 4 chains? And add info. It would be good if you have tidy code you can post with the data so folks can fit the models themselves. 
+%emw2Mar -- Did you use default priors?
+If so, can you figure out how to increase them threefold and see if answers change (they change run to run, but should not change dramatically)?
+Ideally they don't and then we can add 'increasing priors three-fold did not change results.'
 
 \section{Results}
-
+% START HERE Lizzie ... 
 \verb@Research questions recap@

### 2023-03-06 Docs/manuscript.Rnw
edited ms and scripts

@@
 \documentclass{article}
-
 % required 
 \usepackage{amsmath}
-\usepackage{natbib}
+\usepackage[semicolon]{natbib}
 \bibliographystyle{plainnat} %{unsrtnat}
@@
 \usepackage[pdftex,dvipsnames]{xcolor}  % Coloured text etc.
-
 \usepackage[colorinlistoftodos,prependcaption,textsize=tiny]{todonotes}
@@
 \newcommandx{\thiswillnotshow}[2][1=]{\todo[disable,#1]{#2}}
-
 % https://tex.stackexchange.com/questions/60209/how-to-add-an-extra-level-of-sections-with-headings-below-subsubsection
@@
 % figure produced will be titled plus its number or label (e.g., demoFig-nqpbetter.pdf')
-
 % make your captioning look better
@@
 \setlength{\belowcaptionskip}{10pt}
-
 % optional: muck with spacing
@@
 \section{Methods}
-
-% \item Literature search
 \subsection{Data Collection}
@@
 which returned 122 publications.
-We also contacted authors of previous review papers (CITES), to help further search the literature.
+We also contacted authors of previous review papers \citep{AitkenBemmels16, Alberto13}, to help further search the literature.
 We then reviewed the methods and results of all publications to refine to only studies that met the following criteria: (a) focused on woody plants originating from either Europe or North America, (b) had provenance trials/common gardens on the same continent, (c) reported spring events in units of calendar days (day of year or DOY) or could be converted into DOY (see `Additional Methods' section in the Supplement), and (d) reported latitude and longitude of provenances and gardens.
-%emw2Mar -- From above ...
-We don't usually name the people, we would cite the papers and acknowledge the people in the acknowledgements section.
-So add the citations above in CITES and add the below to the acknowledgements
-% We thank S. Aitken,  I. Chuine, R. Guy, C Korner and Y. Vitasse for reviewing our list of papers for possible additional common garden studies. 
-\\
-%emw2Mar: Below and throughout ...
-You should switch to citing references using bibtex throughout!
-Making that change now will be much easier than later.
-Ask Deirdre if you get stuck (doing it soon would also be good as she will be away starting in May and so busier in April).
-%emw2Mar -- mv the below paragraphs to an `Additional Methods' section of supp (also note how I added topic sentences here to guide the reader). 
-% START: Additional methods section:
-We had to exclude several studies that reported spring events on a quantitative scale.
-This was because (1) such studies usually only assessed where on the scale the spring event of a tree fell onto on the same days across different years (e.g.
-Robson et al., 2013; Vander et al., 2015; Santini et al., 2014; Schueler \& Liesebach, 2004), and (2) scales are not always consistent across different studies (Chmura \& Rozkowski 2002; Dhont et al., 2010; Wang et al., 2022).
-Such factors made it impossible to convert the quantitative scale to DOY.
 \\
 \\
-% @ alina_jan27: do we need to cite those paper indicated above that we didnt end up using through bibtex??
-% EMW replies -- no, some journals will require a spreadsheet (what you discarded and why) though -- but until we pick a journal and know what's required we don't need to do any of that.
-We additionally excluded several studies because we could not pinpoint their location, or because they focused on non-native species or elevational trends.
-We excluded studies that did not provide the exact latitude and longitude of the common garden (Bongarten, 1978) or the provenances (Hall et al., 2007; Soolanayakanahally et al., 2013), or they did not link the latitude and longitude of each provenance to the DOY of spring events (Deans \& Harvey, 1996).
-We also left out studies in which woody plants from North American provenances were planted in common gardens in Europe (Cannell et al., 1987; Lavadinovic et al., 2013) because we wanted to test continental variations.
-Finally, as we are focused on latitudinal trends we excluded studies that examined only provenance altitude (Vitasse et al., 2009; Vitasse et al., 2010; Li et al., 1997; Alberto et al., 2011; Acevedo-Rodríguez et al., 2006).
-% END: Additional methods section:
-
-Based on these criteria we found 18 (?) provenance/common garden studies, distributed throughout North America and Europe, with the majority of data concentrated in XXXXX (Figure X).
-From each paper we extracted phenological data on spring events (budburst and leaf flush in DOY) and, when present in the same paper, fall events (bud set, leaf senescence, growth cessation, and leaf abscission) by species, provenance and garden location (geographic coordinates), using ImageJ (version 1.53k; Rasband, 1997-2022) to extract values from figures whenever necessary.
+Based on these criteria we found 18 provenance/common garden studies, distributed throughout North America and Europe, with the majority of data concentrated in North America (Figure X).
+From each paper we extracted phenological data on spring events (budburst and leaf flush in DOY) and, when present in the same paper, fall events (bud set, leaf senescence, growth cessation, and leaf abscission) by species, provenance and garden location (geographic coordinates).
+We used ImageJ (version 1.53k; \citealp{schneider_rasband_eliceiri_2012}) to extract values from figures whenever necessary.
 For studies that reported event dates relative to a reference date other than 1 January (e.g.
-Cannel et al., 1987; Rehfeldt, 1994), we converted such dates to DOY using the Lubridate Package in R.
+\citealp{Rehfeldt1994}), we converted such dates to DOY using the Lubridate Package in R.
 \improvement{display a table and a map here, and in the caption, clarify that some studies had more than one gardens and two studies shared the same garden}
 %emw2Mar -- mv the following sentence to the caption of the map figure: The distribution was skewed toward North America (n = X), with only X located in Europe.
+%alinaMar5:think abt displaying number of studies or provenances
 \\
 \\
-%emw2Mar -- save the below for now; may fit in the discussion or could add if reveiwers complain.
-For now, I added the events above -- it's quicker and IMHO we don't currently need to justify why we pooled these.
-% We pooled the timing of budburst and leaf flush into a single category of ‘spring events’ and the timing of bud set, leaf senescence, growth cessation, and leaf abscission into ‘fall events.’ Such pooling is justified because of the shared pressures from natural selection that govern these events (Gill et al., 2015).
-
 To understand how climatic differences, in addition to geographical differences, shape local adaptation in spring events we extracted several types of climate data using information about provenance latitude, longitude, and elevation from original publications.
-We estimated the mean annual temperature (MAT) for each provenance using the Climate Information Tool by FAO (Food and Agriculture Organization of the United Nations, 2022) and ClimateWNA (Wang et al., 2012).
-To examine more explicitly climate near spring events, we used gridded daily temperature data for 2011 to 2020 for all European and North American provenances and gardens from E-OBS and the built-in Daymet package in R respectively (Cornes et al., 2018; Thornton et al., 2020).
+We estimated the mean annual temperature (MAT) for each provenance using the Climate Information Tool by Food and Agriculture Organization of the United Nations \citep{FAO2022}and ClimateWNA \citep{wang2016}.
+To examine more explicitly climate near spring events, we used gridded daily temperature data for 2011 to 2020 for all European and North American provenances and gardens from E-OBS and the daymetr R package respectively \citep{cornes2018,hufkens2018}.
 Using these data we estimated how much the daily temperatures overlapped between garden and provenance locations, which we call `climate overlap.' For this we used the Overlap package in R to calculate the percentage overlap of the daily temperature of each provenance in spring months (March to May) from 2011 to 2020 in relation to the daily temperature of their corresponding gardens.
 Finally, using the daily data, we also calculated Growing degree days (GDD), a commonly used heat accumulation measure to forecast phenological development in plants (Miller et al., 2001), based on the accumulation of mean daily temperatures ($T_{m}$)from 2011 to 2022 above a baseline of 0$^{\circ}$C, from January 1 until budburst and leaf flush with the following formula: GDD = $\sum$($T_{m}$ - $0^{\circ}$C for $T_{m}$ $\ge$ $0^{\circ}$C; 0 for $T_{m}$ $\le$ $0^{\circ}$C).
-
-%emw2Mar -- there's a delicate balance of how much rationale you can give for your methods in the methods (and similar for the results). I would move the below to the discussion section. 
-% While coarse metrics such as latitude and MAT may represent how similar the climates are between the provenances and gardens in times that matter for the events.
-If climates are very similar, then we would expect similar timings [add more here].
 
@@
 Because most tree species were present in only one common garden in our dataset, it was impossible to fit garden and species separately, thus we treat each species within a garden as a unique group.
-Using posterior estimates for each each species within a garden, we estimated effects of continent and species type (angiosperm vs.
+Using posterior estimates for each species within a garden, we estimated effects of continent (North America vs.
+Europe) and species type (angiosperm vs.
 gymnosperm).
-\improvement{will elaborate later ...
-ADD: All models were fit in XX package (give version number), using default priors, with 4 chains and XX sampling iterations per chain.
+All models were fit in rstanarm package (version 2.21.3; \citealp{brilleman2018}), using default priors, with 4 chains and 1000 sampling iterations per chain.
 We checked for model fit by ...
-}
 %emw2Mar -- Check: Did you use 4 chains? And add info. It would be good if you have tidy code you can post with the data so folks can fit the models themselves. 
@@
 \section{Discussion}
+%emw2Mar -- there's a delicate balance of how much rationale you can give for your methods in the methods (and similar for the results). I would move the below to the discussion section. 
+While coarse metrics such as latitude and MAT may represent how similar the climates are between the provenances and gardens in times that matter for the events.
+If climates are very similar, then we would expect similar timings [add more here].
+
+%emw2Mar -- save the below for now; may fit in the discussion or could add if reveiwers complain.
+For now, I added the events above -- it's quicker and IMHO we don't currently need to justify why we pooled these.
+We pooled the timing of budburst and leaf flush into a single category of ‘spring events’ and the timing of bud set, leaf senescence, growth cessation, and leaf abscission into ‘fall events.’ Such pooling is justified because of the shared pressures from natural selection that govern these events \citep{Gill15}.
 
@@
 
+\section{Acknowledgement}
+We thank S. Aitken,  I. Chuine, R. Guy, C Korner and Y. Vitasse for reviewing our list of papers for possible additional common garden studies.
+
 \end{document}

### 2023-03-06 Docs/manuscript.tex
edited ms and scripts

@@
 \section{Methods}
-
-% \item Literature search
 \subsection{Data Collection}
-\subsubsection{Literature Search}
-%emw24Jan -- this section looks good; once you're happy with it, I can slim it down a little and suggest a few things to cut or move to supp. 
-We searched the peer-reviewed literature for common garden experiments that documented the timing of spring events of woody plant species on 14 December 2022 using Web of Science (Thompson Reuters, New York, NY)with the following terms:
+%emw2Mar: Try to open with WHY (goal/aim) you did something more than what you did, not a rule, but try to do when possible. 
+To locate common garden studies that documented the timing of spring events of woody plant species we searched and reviewed the peer-reviewed literature.
+On 14 December 2022 we searched Web of Science (Thompson Reuters, New York, NY) using the following terms:
+\begin{quote}
+TOPIC = (common garden* OR provenance*) AND (leafout* OR leaf out* OR budburst OR spring phenolog*)
+\end{quote}
+which returned 122 publications.
+We also contacted authors of previous review papers \citep{AitkenBemmels16, Alberto13}, to help further search the literature.
+We then reviewed the methods and results of all publications to refine to only studies that met the following criteria: (a) focused on woody plants originating from either Europe or North America, (b) had provenance trials/common gardens on the same continent, (c) reported spring events in units of calendar days (day of year or DOY) or could be converted into DOY (see `Additional Methods' section in the Supplement), and (d) reported latitude and longitude of provenances and gardens.
 \\
 \\
-TOPIC = (common garden* OR provenance*) AND (leafout* OR leaf out* OR budburst OR spring phenolog*).
-The search returned 122 publications.
-We also contacted authors of previous review papers, including S.N.
-Aitken and R.D.
-Guy to help further search the literature.
+Based on these criteria we found 18 provenance/common garden studies, distributed throughout North America and Europe, with the majority of data concentrated in North America (Figure X).
+From each paper we extracted phenological data on spring events (budburst and leaf flush in DOY) and, when present in the same paper, fall events (bud set, leaf senescence, growth cessation, and leaf abscission) by species, provenance and garden location (geographic coordinates).
+We used ImageJ (version 1.53k; \citealp{schneider_rasband_eliceiri_2012}) to extract values from figures whenever necessary.
+For studies that reported event dates relative to a reference date other than 1 January (e.g.
+\citealp{Rehfeldt1994}), we converted such dates to DOY using the Lubridate Package in R.
+\improvement{display a table and a map here, and in the caption, clarify that some studies had more than one gardens and two studies shared the same garden}
+%emw2Mar -- mv the following sentence to the caption of the map figure: The distribution was skewed toward North America (n = X), with only X located in Europe.
+%alinaMar5:think abt displaying number of studies or provenances
 \\
 \\
-We read the methods and results of all publications and limited our subsequent analysis to the ones that (a) focused on woody plants originating from either Europe or North America, (b) had provenance trails and common gardens on the same continent, (c) reported spring events in units of calendar days (day of year or DOY) instead of a quantitative scale on a particular day, and (d) reported latitude and longitude of provenances and gardens.
-\\
-\\
-We had to exclude studies that reported spring events on a quantitative scale because (1) such studies usually only assessed where on the scale the spring event of a tree fell onto on the same days across different years (e.g.
-Robson et al., 2013; Vander et al., 2015; Santini et al., 2014; Schueler \& Liesebach, 2004), and (2) scales are not always consistent across different studies (Chmura \& Rozkowski 2002; Dhont et al., 2010; Wang et al., 2022).
-Such factors made it impossible to back convert the quantitative scale to DOY.
-\\
-\\
-Some studies did not provide the exact latitude and longitude of the common garden (Bongarten, 1978) or the provenances (Hall et al., 2007; Soolanayakanahally et al., 2013), or they did not link the latitude and longitude of each provenance to the DOY of spring events (Deans \& Harvey, 1996).
-We had to exclude these papers.
-We also left out studies in which woody plants from North American provenances were planted in common gardens in Europe because we wanted to test continental variations (Cannell et al., 1987; Lavadinovic et al., 2013).
-Finally, we excluded studies that focused on examining how provenance altitude influenced local adaptation in spring phenology since we are interested in testing latitudinal effects (Vitasse et al., 2009; Vitasse et al., 2010; Li et al., 1997; Alberto et al., 2011; Acevedo-Rodríguez et al., 2006).
-
-\subsubsection{Dataset Assembly}
-Out of the 122 common garden studies on spring phenology that we read, only 18 (?) ultimately supplied the data that we needed (DOY of spring events, latitude and longitude of provenances and common gardens).
-Studies were distributed throughout North America and Europe, with the majority of data concentrated in XXXXX (Figure X).
-The distribution was skewed toward North America (n = X), with only X located in Europe.
-We primarily focused our analysis on spring events, and made use of the fall event DOYs present in X of the studies we included rather than searching for publications entirely on fall events online.
-\improvement{display a table and a map here, and in the caption, clarify that some studies had more than one gardens and two studies shared the same garden}
-\\
-\\
-We assembled a data set that documents event DOY and geographic coordinates of provenances and gardens, during which we used ImageJ (version 1.53k; Rasband, 1997-2022) to extract values from figures whenever possible and necessary.
-We pooled the timing of budburst and leaf flush into a single category of ‘spring events’ and the timing of bud set, leaf senescence, growth cessation, and leaf abscission into ‘fall events.’ Such pooling is justified because of the shared pressures from natural selection that govern these events (Gill et al., 2015).
-Some studies documented event dates relative to a reference date that is not January 1 (e.g.
-Cannel et al., 1987; Rehfeldt, 1994).
-In this case, we converted such dates to DOY using the Lubridate Package in R for standardization purposes.
-
-\subsubsection {Climate Data Collection}
-With information about provenance latitude, longitude, and elevation from original publications, we estimated the mean annual temperature (MAT) for each provenance using the Climate Information Tool by FAO (Food and Agriculture Organization of the United Nations, 2022) and ClimateWNA (Wang et al., 2012).
-\\
-\\
-We hope that coarse metrics such as latitude and MAT ultimately represent how similar the climates are between the provenances and gardens in times that matter for the events.
-If climates are very similar, then we would expect similar timings [add more here].
-To this end, we estimated climate overlap in relevant months: For spring events, we considered overlap across March to May.
-To assess climate overlap percentage (define this better), we extracted gridded daily temperature data for 2011 to 2020 for all European and North American provenances and gardens from E-OBS and the built-in Daymet package in R respectively (Cornes et al., 2018; Thornton et al., 2020).
+To understand how climatic differences, in addition to geographical differences, shape local adaptation in spring events we extracted several types of climate data using information about provenance latitude, longitude, and elevation from original publications.
+We estimated the mean annual temperature (MAT) for each provenance using the Climate Information Tool by Food and Agriculture Organization of the United Nations \citep{FAO2022}and ClimateWNA \citep{wang2016}.
+To examine more explicitly climate near spring events, we used gridded daily temperature data for 2011 to 2020 for all European and North American provenances and gardens from E-OBS and the daymetr R package respectively \citep{cornes2018,hufkens2018}.
+Using these data we estimated how much the daily temperatures overlapped between garden and provenance locations, which we call `climate overlap.' For this we used the Overlap package in R to calculate the percentage overlap of the daily temperature of each provenance in spring months (March to May) from 2011 to 2020 in relation to the daily temperature of their corresponding gardens.
+Finally, using the daily data, we also calculated Growing degree days (GDD), a commonly used heat accumulation measure to forecast phenological development in plants (Miller et al., 2001), based on the accumulation of mean daily temperatures ($T_{m}$)from 2011 to 2022 above a baseline of 0$^{\circ}$C, from January 1 until budburst and leaf flush with the following formula: GDD = $\sum$($T_{m}$ - $0^{\circ}$C for $T_{m}$ $\ge$ $0^{\circ}$C; 0 for $T_{m}$ $\le$ $0^{\circ}$C).
 
 \subsection{Analyses}
-\subsubsection{Variable preparation}
-Using the Overlap package in R, we calculated the percentage overlap of the daily temperature of each provenance in spring months (March to May) from 2011 to 2020 in relation to the daily temperature of their corresponding gardens.
-We calculated Growing degree days (GDD), a commonly used heat accumulation measure to forecast phenological development in plants (Miller et al., 2001), based on the accumulation of mean daily temperatures ($T_{m}$)from 2011 to 2022 above a baseline of 0$^{\circ}$C, from January 1 until budburst and leaf flush with the following formula: GDD = $\sum$($T_{m}$ - $0^{\circ}$C for $T_{m}$ $\ge$ $0^{\circ}$C; 0 for $T_{m}$ $\le$ $0^{\circ}$C)
-%emw24Jan -- I can add to this. 
-\subsubsection{Mixed effects modelling}
-Most tree species are only represented in one common garden study in our dataset.
-In such instances, it is difficult to statistically interpret if variations across gardens are due to the species or the garden’s location and climate.
-To address this, we focused on estimating species effects as we expect a larger effect from species.
-Within publications, we considered unique gardens and species to represent independent data, and fitted each species \& common gardens instead of just species.
-
-We used hierarchical Bayesian models where we partially pooled by species.
-We estimated effects of continent and species type (angiosperm VS gymnosperm) from posterior estimates, in relation to single predictors (including provenance latitude, MAT) and two predictors (climate overlap percentage and standard deviation).
-\improvement{will elaborate later}
+To estimate clines in spring and fall phenological events across species we used hierarchical Bayesian models.
+We regressed day of year (DOY) of events against geographical and climate predictors with partial pooling (sometimes called `random effects') on the intercept and slope for each species within each garden.
+Because most tree species were present in only one common garden in our dataset, it was impossible to fit garden and species separately, thus we treat each species within a garden as a unique group.
+Using posterior estimates for each species within a garden, we estimated effects of continent (North America vs.
+Europe) and species type (angiosperm vs.
+gymnosperm).
+All models were fit in rstanarm package (version 2.21.3; \citealp{brilleman2018}), using default priors, with 4 chains and 1000 sampling iterations per chain.
+We checked for model fit by ...
+%emw2Mar -- Check: Did you use 4 chains? And add info. It would be good if you have tidy code you can post with the data so folks can fit the models themselves. 
+%emw2Mar -- Did you use default priors?
+If so, can you figure out how to increase them threefold and see if answers change (they change run to run, but should not change dramatically)?
+Ideally they don't and then we can add 'increasing priors three-fold did not change results.'
 
 \section{Results}
-
+% START HERE Lizzie ... 
 \verb@Research questions recap@

### 2023-03-06 Docs/supplement.Rnw
edited ms and scripts

@@
 \section*{Methods}
+
+\subsection*{Additional Methods}
+We had to exclude several studies that reported spring events on a quantitative scale.
+This was because (1) such studies usually only assessed where on the scale the spring event of a tree fell onto on the same days across different years (e.g.
+Robson et al., 2013; Vander et al., 2015; Santini et al., 2014; Schueler \& Liesebach, 2004), and (2) scales are not always consistent across different studies (Chmura \& Rozkowski 2002; Dhont et al., 2010; Wang et al., 2022).
+Such factors made it impossible to convert the quantitative scale to DOY.
+\\
+\\
+% @ alina_jan27: do we need to cite those paper indicated above that we didnt end up using through bibtex??
+% EMW replies -- no, some journals will require a spreadsheet (what you discarded and why) though -- but until we pick a journal and know what's required we don't need to do any of that.
+We additionally excluded several studies because we could not pinpoint their location, or because they focused on non-native species or elevational trends.
+We excluded studies that did not provide the exact latitude and longitude of the common garden (Bongarten, 1978) or the provenances (Hall et al., 2007; Soolanayakanahally et al., 2013), or they did not link the latitude and longitude of each provenance to the DOY of spring events (Deans \& Harvey, 1996).
+We also left out studies in which woody plants from North American provenances were planted in common gardens in Europe (Cannell et al., 1987; Lavadinovic et al., 2013) because we wanted to test continental variations.
+Finally, as we are focused on latitudinal trends we excluded studies that examined only provenance altitude (Vitasse et al., 2009; Vitasse et al., 2010; Li et al., 1997; Alberto et al., 2011; Acevedo-Rodríguez et al., 2006).
 \subsection*{Table for all studies}

### 2023-03-06 Analyses/script_UPDATED_fit_garden_spp_Dec13_2022.R

@@ -3,7 +3,7 @@
 # Dec 13, 2022
 
 
-# for previous fits, refer to script_stan_glmer_NEW_fit_plots_are_in_here
+# for previous fits, refer to script_rstanarm::stan_glmer_NEW_fit_plots_are_in_here
 
 
 
@@ -16,30 +16,40 @@ name<-"input/percentage_overlap_doy_difference_earth_calculated_garden_identifie
 write.csv(d,name, row.names = FALSE)
 
 
+library(dplyr)
+library(tidyr)
+library(Rcpp)
+library(rstanarm)
+library(bayesplot)
+library(ggplot2)
+library(viridis)
+
 
 # spring or fall DOY ~ lat or mat
-fitC_spring_lat <- stan_glmer(spring_event~(lat_prov|species_garden), data = d)
-fitC_fall_lat <- stan_glmer(fall_event~(lat_prov|species_garden), data = d)
-fitC_spring_mat <- stan_glmer(spring_event~(MAT_prov|species_garden), data = d)
-fitC_fall_mat <- stan_glmer(fall_event~(MAT_prov|species_garden), data = d)
-
-# lat diffo
-fitC_spring_lat_diffo <- stan_glmer(spring_event~(distance_from_garden|species_garden), data = d)
-fitC_fall_lat_diffo <- stan_glmer(fall_event~(distance_from_garden|species_garden), data = d)
-fitC_spring_earth_distance <- stan_glmer(spring_event~(earth_distance_from_garden|species_garden), data = d)
-fitC_fall_earth_distance <- stan_glmer(fall_event~(earth_distance_from_garden|species_garden), data = d)
-
-# mat diffo
-fitC_spring_mat_diffo <- stan_glmer(spring_event~(MAT_diffo|species_garden), data = d)
-fitC_fall_mat_diffo <- stan_glmer(fall_event~(MAT_diffo|species_garden), data = d)
+fitC_spring_lat <- rstanarm::stan_glmer(spring_event~(lat_prov|species_garden), data = d)
+fitC_fall_lat <- rstanarm::stan_glmer(fall_event~(lat_prov|species_garden), data = d)
+# already did lat ones and saved to d
+fitC_spring_mat <- rstanarm::stan_glmer(spring_event~(MAT_prov|species_garden), data = d)
+fitC_fall_mat <- rstanarm::stan_glmer(fall_event~(MAT_prov|species_garden), data = d)
 
[…]

### 2023-03-06 Analyses/script_stan_glmer_NEW_fit_plots_are_in_here_SEPT16_MODELLING_SCRIPT.R

@@ -82,7 +82,7 @@ d$spring_event <- as.numeric(d$spring_event)
 
 #export d cuz i have information abt intercept and slope
 
-name<-paste("input/percentage_overlap_doy_difference_earth_calculated_garden_identifier_adjusted_fall_diffo_included_slope_intercept_May5.csv",sep="")
+name<-paste("input/percentage_overlap_doy_difference_earth_calculated_garden_identifier_adjusted_fall_diffo_included_slope_intercept_gdd_Sept27",sep="")
 write.csv(d,name, row.names = FALSE)
 
 ### New bits by Lizzie on 1 May 2022
@@ -159,9 +159,6 @@ fitA_fall_diffo_sd <- stan_glmer(fall_event_difference~(sd|species), data = d)
 # plotting for two predictors
 fitB_spring_percentage_sd <- stan_glmer(spring_event~((percentage*sd)|species), data = d)
 fitB_fall_percentage_sd <- stan_glmer(fall_event~((percentage*sd)|species), data = d)
-
-
-
 
 # new addition Sept 16, 2022
 fitA_spring_lat_diffo <- stan_glmer(spring_event~(distance_from_garden|Species), data = d)

### 2023-03-06 Docs/manuscript.Rnw
Results are edited, but we need to add model fits inline in the text and update tables

@@
 \pagestyle{fancy}
-\fancyhead[LO]{November 2022}
+\fancyhead[LO]{March 2023}
 \fancyhead[RO]{Manuscript}
@@
 gymnosperm).
-All models were fit in rstanarm package (version 2.21.3; \citealp{brilleman2018}), using default priors, with 4 chains and 1000 sampling iterations per chain.
+All models were fit in rstanarm package (version 2.21.3; \citealp{brilleman2018}), using default priors, with 4 chains and 1000 sampling iterations per chain for a total of 4000 samples.
 We checked for model fit by ...
-%emw2Mar -- Check: Did you use 4 chains? And add info. It would be good if you have tidy code you can post with the data so folks can fit the models themselves. 
+[ADD here].
+We present estimates as mean $\pm$ 90\% uncertainty intervals, unless otherwise stated.
+%emw2Mar -- It would be good if you have tidy code you can post with the data so folks can fit the models themselves. 
 %emw2Mar -- Did you use default priors?
@@
 \section{Results}
-% START HERE Lizzie ... 
+%emw6Mar -- below opening and questions are good, but not appropriate for results. Move up to be at end of introduction. 
 \verb@Research questions recap@
@@
 %emw24Jan -- below is great opening!
-Our final dataset captures seven angiosperm and eight gymnosperm species from 19 common gardens, encompassing data from 384 North American provenances and 101 European provenances with observations from 1962 to 2019.
-Seven species had fall event information available.
+%emw6Mar -- results are in past tense generally
+Our final dataset captured seven angiosperm and eight gymnosperm species from 19 common gardens, encompassing 384 North American provenances and 101 European provenances, with observations from 1962 to 2019.
+Seven species also had fall event information available.
 \\
@@
 \todo[inline]{add map here}
-\todo[inline]{add a table documenting studies attributes}
-
-\subsection{Provenance latitude and MAT do not affect spring event timing}
-Overall, our models show that spring events are not related to provenance latitude or mean annual temperature in North America, and only weakly related in Europe, where springs events are slightly earlier where provenance mean annual temperature is lower (higher, more northern latitudes)(Fig.
+\todo[inline]{add a table documenting studies attributes} %emw6Mar -- Tables should go in supplement.
+But we should add estimates to statements we make throughout.
+These should be added as mean +/- 90% unvertainty (posterior) intervals (UI) parenthetically using /Sexpr so they update automatically.
+Overall, spring events were not related to provenance latitude or mean annual temperature in North America (insert the mean slope here, followed by the 90\% intervals ), and only weakly related in Europe (insert slope and UI), where springs events are slightly earlier where provenance mean annual temperature is lower (i.e., higher, more northern latitudes)(Fig.
 \ref{figure:springfall_latmat} \& Supplement Table.
 \ref{supplement-table:model_sf_lat}).
+Results were similar using other distance metrics in lieu of latitude (see Supplement Fig.\ref{supplement-figure:lat_distance} for results using the absolute value of the distance and the spherical distance difference between provenance and garden latitude).
+%emw6Mar -- this text from below -- "advanced strongly with provenance latitude and mean annual temperature, meaning fall events are earlier where provenance mean annual temperature is lower" is good, can you copy that phrasing here instead "where springs events are slightly earlier where provenance mean annual temperature is lower"?
+I think it would be something like 'delay slightly ...'
 
@@
 
-\newpage
+In contrast, fall events (budset, leaf senescence, leaf abcission) advanced strongly with provenance latitude and mean annual temperature, meaning fall events are earlier where provenance mean annual temperature is lower (i.e., higher, more northern latitudes).
+This relationship, however, was observed mostly in North America where fall events advance 4.2 days per degree (UI), or 6.4 days (UI) per degree decline in MAT$^{\circ}$C (Fig.
+\ref{figure:continent}).
+In Europe, such relationship is weak: advance 0.5 days (UI) per degree we move north, or 0.6 days (UI) per degree decline in MAT $^{\circ}$C(Fig.
+\ref{figure:continent}).
+%emw6Mar -- above numbers (e.g., 4.2 should be done with /Sexpr and include UI in parentheses)
 
-\subsubsection {Similar results across provenance latitude, absolute value of difference between provenance and garden latitude, and spherical distance}
-Along with (a) provenance latitude, we also looked at how spring event timing is related to (b) the absolute value of difference between provenance and garden latitude , as well as (c) the spherical distance between the garden and each provenance.
-All three metrics depict little to no relationship between spring event timing and the geographical location of the provenance (Supplement Fig.\ref{supplement-figure:lat_distance}).
+\begin{figure}[!h] 
+    \centering
+ \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/continent_histogram.png}
+    \caption{Caption placeholder}
+    \label{figure:continent}
+\end{figure}
 
-\newpage
-\subsubsection {Climate overlap does not predict event dates much better than provenance latitude or MAT}
+Effects of provenance latitude on both spring and fall events were similar across angiosperms and gymnosperms.
+However, effects of MAT on spring events weakly diverge: spring events get earlier as MAT increases in angiosperms and delay as MAT increases in gymnosperms, except for \emph{Pseudotsuga menziesii}.
+Fall events delay in warmer locations for both species types, but slightly more so for gymnosperms (3.7 days VS.
+6.2 days)(Fig.
+\ref{figure:spp_type}).
+%emw6Mar -- Again, need numbers using /Sexpr here.
 
-While comparing how similarity in climate relates to event dates, we observed very weak effects of climate overlap on spring events, nearly identical across angiosperms and gymnosperms.
+We also tested whether climate effects on events may be stronger using metrics from seasonal windows closer to event dates, but results were not qualitatively different than using MAT.
+We observed very weak effects of climate overlap on spring events, nearly identical across angiosperms and gymnosperms.
 Fall events diverge as climate overlap declines for both angiosperms and gymnosperms, but slightly more strongly for gymnosperms (Fig.\ref{figure:overlap}).
+
+%emw6Mar -- I think we can skip subheaders in the results (yay!) but I need to read it again once you had added all the numbers and intervals. 
+\begin{figure}[!h] 
+    \centering
+ \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/spp_type_histogram.png}
+    \caption{Caption placeholder}
+    \label{figure:spp_type}
+\end{figure}
 
@@
 \end{figure}
+%emw6Mar -- Climate overlap figures can go in supp.
 
-\subsection{Stronger clines in fall events observed in North America than Europe}
+%emw6Mar -- I could combine the current fig 3-4. Just be sure to label each panel.
+
 %emw24Jan -- Nice! Also, results seem good, I can help trim text and collapse some sections.
@@
 % hmmm i forgot the code :)))
-
-We find that fall events (budset, leaf senescence, leaf abcission) advance strongly with provenance latitude and mean annual temperature, meaning fall events are earlier where provenance mean annual temperature is lower (higher, more northern latitudes).
-This relationship, however, is observed mostly in North America where fall events advance 4.2 days per degree we move north, or 6.4 days when the MAT decreases by 1 $^{\circ}$C (Fig.
-\ref{figure:continent}).
-In Europe, such relationship is weak: advance 0.5 days per degree we move north, or 0.6 days when the MAT decreases by 1 $^{\circ}$C(Fig.
-\ref{figure:continent}).
-
-\begin{figure}[!h] 
-    \centering
- \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/continent_histogram.png}
-    \caption{Caption placeholder}
-    \label{figure:continent}
-\end{figure}
-
-\subsection{Effects of MAT on spring events diverge across angiosperms and gymnosperms}
-
-Effects of provenance latitude on both spring and fall events are similar across angiosperms and gymnosperms.
-However, effects of MAT on spring events weakly diverge: spring events get earlier as MAT increases in angiosperms and delay as MAT increases in gymnosperms, except for \emph{Pseudotsuga menziesii}.
-Fall events delay in warmer locations for both species types, but slightly more so for gymnosperms (3.7 days VS.
-6.2 days)(Fig.
-\ref{figure:spp_type}).
-
-\begin{figure}[!h] 
-    \centering
- \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/spp_type_histogram.png}
-    \caption{Caption placeholder}
-    \label{figure:spp_type}
-\end{figure}
 

### 2023-03-13 Docs/manuscript.Rnw
added /sexpr{}

@@
 
+%emw6Mar -- below opening and questions are good, but not appropriate for results. Move up to be at end of introduction. 
+\verb@Research questions recap@
+\\
+Despite a growing interest in predicting local adaptation across locations, no study has comprehensively examined clines for spring and fall events and what factors may underlie differences observed across studies.
+
+We ask: Across common garden studies, how strong are clines in spring and fall phenology?
+
+To better understand these clines, we also aim to examine:
+\begin{itemize}
+\item If differences in clines exist across Europe and North America.
+\item If differences in clines exist across angiosperm and gymnosperm species.
+\item If climate overlap between the provenance and garden explains the similarities and differences in clines.
+\end{itemize}
+
 \section{Methods}
@@
 \section{Results}
-%emw6Mar -- below opening and questions are good, but not appropriate for results. Move up to be at end of introduction. 
-\verb@Research questions recap@
-\\
-Despite a growing interest in predicting local adaptation across locations, no study has comprehensively examined clines for spring and fall events and what factors may underlie differences observed across studies.
-
-We ask: Across common garden studies, how strong are clines in spring and fall phenology?
-
-To better understand these clines, we also aim to examine:
-\begin{itemize}
-\item If differences in clines exist across Europe and North America.
-\item If differences in clines exist across angiosperm and gymnosperm species.
-\item If climate overlap between the provenance and garden explains the similarities and differences in clines.
-\end{itemize}
-
-\todo[inline]{all subsection titles are placeholders}
-
 %emw24Jan -- below is great opening!
@@
 But we should add estimates to statements we make throughout.
-These should be added as mean +/- 90% unvertainty (posterior) intervals (UI) parenthetically using /Sexpr so they update automatically.
-Overall, spring events were not related to provenance latitude or mean annual temperature in North America (insert the mean slope here, followed by the 90\% intervals ), and only weakly related in Europe (insert slope and UI), where springs events are slightly earlier where provenance mean annual temperature is lower (i.e., higher, more northern latitudes)(Fig.
+These should be added as mean +/- 90% uncertainty (posterior) intervals (UI) parenthetically using /Sexpr so they update automatically.
+Overall, spring events were not related to provenance latitude (insert the mean slope here, followed by the 90\% intervals) or mean annual temperature in North America (insert the mean slope here, followed by the 90\% intervals).
+Spring events advanced slightly with provenance latitude and mean annual temperature in Europe (insert slope and UI), meaning spring events are slightly earlier where provenance mean annual temperature is lower (i.e., higher, more northern latitudes)(Fig.
 \ref{figure:springfall_latmat} \& Supplement Table.
@@
 Results were similar using other distance metrics in lieu of latitude (see Supplement Fig.\ref{supplement-figure:lat_distance} for results using the absolute value of the distance and the spherical distance difference between provenance and garden latitude).
-%emw6Mar -- this text from below -- "advanced strongly with provenance latitude and mean annual temperature, meaning fall events are earlier where provenance mean annual temperature is lower" is good, can you copy that phrasing here instead "where springs events are slightly earlier where provenance mean annual temperature is lower"?
-I think it would be something like 'delay slightly ...'
 
@@
 
-In contrast, fall events (budset, leaf senescence, leaf abcission) advanced strongly with provenance latitude and mean annual temperature, meaning fall events are earlier where provenance mean annual temperature is lower (i.e., higher, more northern latitudes).
-This relationship, however, was observed mostly in North America where fall events advance 4.2 days per degree (UI), or 6.4 days (UI) per degree decline in MAT$^{\circ}$C (Fig.
+<<label=source_fall_models, echo=FALSE>>=
+#load models and data to report numbers of things
+#getwd()
+source("../Analyses/source_fall_models_for_sweave.R")
+@
+
+In contrast, fall events (budset, leaf senescence, leaf abcission) advanced strongly with provenance latitude and mean annual temperature, meaning fall events are earlier where provenance mean annual temperature is lower.
+This relationship, however, was observed mostly in North America where fall events advance \Sexpr{format(abs(median(northamericanpost_fitC_fall_lat)),digits = 2)} (UI = \Sexpr{format(abs(quantile(northamericanpost_fitC_fall_lat, 0.9)), digits = 2)}) days per degree we move north, or \Sexpr{format(median(northamericanpost_fitC_fall_mat),digits = 2)} days (UI) per degree decline in MAT$^{\circ}$C (Fig.
 \ref{figure:continent}).
-In Europe, such relationship is weak: advance 0.5 days (UI) per degree we move north, or 0.6 days (UI) per degree decline in MAT $^{\circ}$C(Fig.
+In Europe, such relationship is weak: advance \Sexpr{format(abs(median(europeanpost_fitC_fall_lat)),digits = 1)} (UI) days per degree we move north, or \Sexpr{format(median(europeanpost_fitC_fall_mat),digits = 1)} days (UI) per degree decline in MAT $^{\circ}$C(Fig.
 \ref{figure:continent}).
-%emw6Mar -- above numbers (e.g., 4.2 should be done with /Sexpr and include UI in parentheses)
 
@@
 However, effects of MAT on spring events weakly diverge: spring events get earlier as MAT increases in angiosperms and delay as MAT increases in gymnosperms, except for \emph{Pseudotsuga menziesii}.
-Fall events delay in warmer locations for both species types, but slightly more so for gymnosperms (3.7 days VS.
-6.2 days)(Fig.
+Fall events delay in warmer locations for both species types, but slightly more so for gymnosperms (\Sexpr{format(median(deciduouspost_fitC_fall_mat),digits = 2)} days VS.
+\Sexpr{format(median(evergreenspost_fitC_fall_mat),digits = 2)} days)(Fig.
 \ref{figure:spp_type}).
-%emw6Mar -- Again, need numbers using /Sexpr here.
+
+%(3.7 days VS. 6.2 days)
 
@@
 
-%emw24Jan -- Nice! Also, results seem good, I can help trim text and collapse some sections.
-
 % old graphs can be found in C:\Users\alina\Documents\git\localadaptclim\Output\plotMay15_two predictors_experiment
@@
 % need to replot ofc
-
-%  extract the mean estimate of the slope and report that instead of ~4 or such
-% hmmm i forgot the code :)))
 
@@
 
+% Must cover in discussion:
+
+% Counter vs. co gradient literature
+% Maybe also discuss? ice sheet dynamics in Europe/North America, which could impact rates of local adaptation.
+
 \citet{Keir11}

### 2023-03-13 Docs/manuscript.tex
added /sexpr{}

@@
 \section{Results}
-% START HERE Lizzie ... 
-\verb@Research questions recap@
-\\
-Despite a growing interest in predicting local adaptation across locations, no study has comprehensively examined clines for spring and fall events and what factors may underlie differences observed across studies.
-
-We ask: Across common garden studies, how strong are clines in spring and fall phenology?
-
-To better understand these clines, we also aim to examine:
-\begin{itemize}
-\item If differences in clines exist across Europe and North America.
-\item If differences in clines exist across angiosperm and gymnosperm species.
-\item If climate overlap between the provenance and garden explains the similarities and differences in clines.
-\end{itemize}
-
-\todo[inline]{all subsection titles are placeholders}
-
 %emw24Jan -- below is great opening!
-Our final dataset captures seven angiosperm and eight gymnosperm species from 19 common gardens, encompassing data from 384 North American provenances and 101 European provenances with observations from 1962 to 2019.
-Seven species had fall event information available.
+%emw6Mar -- results are in past tense generally
+Our final dataset captured seven angiosperm and eight gymnosperm species from 19 common gardens, encompassing 384 North American provenances and 101 European provenances, with observations from 1962 to 2019.
+Seven species also had fall event information available.
 \\
@@
 \todo[inline]{add map here}
-\todo[inline]{add a table documenting studies attributes}
-
-\subsection{Provenance latitude and MAT do not affect spring event timing}
-Overall, our models show that spring events are not related to provenance latitude or mean annual temperature in North America, and only weakly related in Europe, where springs events are slightly earlier where provenance mean annual temperature is lower (higher, more northern latitudes)(Fig.
+\todo[inline]{add a table documenting studies attributes} %emw6Mar -- Tables should go in supplement.
+But we should add estimates to statements we make throughout.
+These should be added as mean +/- 90% uncertainty (posterior) intervals (UI) parenthetically using /Sexpr so they update automatically.
+Overall, spring events were not related to provenance latitude (insert the mean slope here, followed by the 90\% intervals) or mean annual temperature in North America (insert the mean slope here, followed by the 90\% intervals).
+Spring events advanced slightly with provenance latitude and mean annual temperature in Europe (insert slope and UI), meaning spring events are slightly earlier where provenance mean annual temperature is lower (i.e., higher, more northern latitudes)(Fig.
 \ref{figure:springfall_latmat} \& Supplement Table.
 \ref{supplement-table:model_sf_lat}).
+Results were similar using other distance metrics in lieu of latitude (see Supplement Fig.\ref{supplement-figure:lat_distance} for results using the absolute value of the distance and the spherical distance difference between provenance and garden latitude).
 
@@
 
-\newpage
-
-\subsubsection {Similar results across provenance latitude, absolute value of difference between provenance and garden latitude, and spherical distance}
-Along with (a) provenance latitude, we also looked at how spring event timing is related to (b) the absolute value of difference between provenance and garden latitude , as well as (c) the spherical distance between the garden and each provenance.
-All three metrics depict little to no relationship between spring event timing and the geographical location of the provenance (Supplement Fig.\ref{supplement-figure:lat_distance}).
-
-\newpage
-\subsubsection {Climate overlap does not predict event dates much better than provenance latitude or MAT}
-
-While comparing how similarity in climate relates to event dates, we observed very weak effects of climate overlap on spring events, nearly identical across angiosperms and gymnosperms.
-Fall events diverge as climate overlap declines for both angiosperms and gymnosperms, but slightly more strongly for gymnosperms (Fig.\ref{figure:overlap}).
-
-\begin{figure}[!h] 
-    \centering
- \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/climate_overlap.png}
-    \caption{Caption placeholder}
-    \label{figure:overlap}
-\end{figure}
-
-\subsection{Stronger clines in fall events observed in North America than Europe}
-%emw24Jan -- Nice! Also, results seem good, I can help trim text and collapse some sections.
-
-% old graphs can be found in C:\Users\alina\Documents\git\localadaptclim\Output\plotMay15_two predictors_experiment
-% refer to analyses\script_model_continent_spp_type_effect
-% https://github.com/lizzieinvancouver/localadaptclim/issues/16
-
-% need to replot ofc
-
-%  extract the mean estimate of the slope and report that instead of ~4 or such
-% hmmm i forgot the code :)))
-
-We find that fall events (budset, leaf senescence, leaf abcission) advance strongly with provenance latitude and mean annual temperature, meaning fall events are earlier where provenance mean annual temperature is lower (higher, more northern latitudes).
-This relationship, however, is observed mostly in North America where fall events advance 4.2 days per degree we move north, or 6.4 days when the MAT decreases by 1 $^{\circ}$C (Fig.
+In contrast, fall events (budset, leaf senescence, leaf abcission) advanced strongly with provenance latitude and mean annual temperature, meaning fall events are earlier where provenance mean annual temperature is lower.
+This relationship, however, was observed mostly in North America where fall events advance 4.2 (UI = 4) days per degree we move north, or 6.4 days (UI) per degree decline in MAT$^{\circ}$C (Fig.
 \ref{figure:continent}).
-In Europe, such relationship is weak: advance 0.5 days per degree we move north, or 0.6 days when the MAT decreases by 1 $^{\circ}$C(Fig.
+In Europe, such relationship is weak: advance 0.5 (UI) days per degree we move north, or 0.7 days (UI) per degree decline in MAT $^{\circ}$C(Fig.
 \ref{figure:continent}).
@@
 
-\subsection{Effects of MAT on spring events diverge across angiosperms and gymnosperms}
-
-Effects of provenance latitude on both spring and fall events are similar across angiosperms and gymnosperms.
+Effects of provenance latitude on both spring and fall events were similar across angiosperms and gymnosperms.
 However, effects of MAT on spring events weakly diverge: spring events get earlier as MAT increases in angiosperms and delay as MAT increases in gymnosperms, except for \emph{Pseudotsuga menziesii}.
@@
 
+%(3.7 days VS. 6.2 days)
+
+We also tested whether climate effects on events may be stronger using metrics from seasonal windows closer to event dates, but results were not qualitatively different than using MAT.
+We observed very weak effects of climate overlap on spring events, nearly identical across angiosperms and gymnosperms.
+Fall events diverge as climate overlap declines for both angiosperms and gymnosperms, but slightly more strongly for gymnosperms (Fig.\ref{figure:overlap}).
+
+%emw6Mar -- I think we can skip subheaders in the results (yay!) but I need to read it again once you had added all the numbers and intervals. 
 \begin{figure}[!h] 
@@
 \end{figure}
+
+\begin{figure}[!h] 
+    \centering
+ \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/climate_overlap.png}
+    \caption{Caption placeholder}
+    \label{figure:overlap}
+\end{figure}
+%emw6Mar -- Climate overlap figures can go in supp.
+
+%emw6Mar -- I could combine the current fig 3-4. Just be sure to label each panel.
+
+% old graphs can be found in C:\Users\alina\Documents\git\localadaptclim\Output\plotMay15_two predictors_experiment
+% refer to analyses\script_model_continent_spp_type_effect
+% https://github.com/lizzieinvancouver/localadaptclim/issues/16
+
+% need to replot ofc
 

### 2023-03-13 Analyses/script_UPDATED_fit_garden_spp_Dec13_2022.R

@@ -23,14 +23,53 @@ library(rstanarm)
 library(bayesplot)
 library(ggplot2)
 library(viridis)
+library(rstantools)
 
+
+# attempting to make tables March 7
+summary(fitC_spring_mat)
+coef(fitC_spring_mat)$species
+posterior_interval(fitC_spring_mat, prob=0.80) # not like this...
+# following Issue # 16 to sum and average posteriors of one model
+
+draws <- as.matrix(fitC_spring_mat)
+all_spp_garden <- c("Alnus rubra A",
+                "Betula papyrifera K",
+                "Betula papyrifera L",
+                "Betula papyrifera M","Fagus sylvatica R*","Fagus sylvatica T*",
+                "Fraxinus excelsior Q*","Picea abies S*","Picea engelmannii B",
+                "Picea mariana I","Picea sitchensis D","Pinus albicaulis C","Pinus ponderosa J",
+                "Populus balsamifera F","Populus trichocarpa D","Pseudotsuga menziesii H",
+                "Quercus petraea U*","Quercus petraea V*","Tsuga heterophylla E","Tsuga heterophylla G") # put the rest of the species names in ... 
+
+library(stringr)
+all_spp_garden <- str_replace_all(all_spp_garden," ","_")
+
+# Get all the species of one type now
+all_spp_garden_draws <- matrix(data=NA,
+                          nrow=nrow(draws),
+                          ncol=length(all_spp_garden))
+
+for (i in c(1:length(all_spp_garden))){
+  all_spp_garden_draws[,i] <- draws[,paste("b[MAT_prov species_garden:", all_spp_garden[i], "]", sep="")]
+}
+
+# Now you can manipulate these posteriors with basic math ...  
+pe -- you need to take the rowMeans beacause the samples are related and so you keep the rows together
+all_spp_garden_post <- rowMeans(all_spp_garden_draws)
+
+quantile(all_spp_garden_post, probs = c(.1, .5, .9))
+
+10%         50%         90% 
+-0.28666551 -0.10987028  0.07123643 
 
 # spring or fall DOY ~ lat or mat
 fitC_spring_lat <- rstanarm::stan_glmer(spring_event~(lat_prov|species_garden), data = d)
[…]

### 2023-03-13 Analyses/script_model_continent_spp_type_effect_UPDATED.R

@@ -1,5 +1,5 @@
 # script model for continent and leaf type effects
-# May-7, 2022
+# updated on March 12, 2023
 
 
 ## Extracting model fits sidebar ...
@@ -21,15 +21,242 @@ d<- read.csv("input/percentage_overlap_doy_difference_earth_calculated_garden_id
 d$fall_event <- as.numeric(d$fall_event)
 d$spring_event <- as.numeric(d$spring_event)
 
-#fit5 <- stan_glmer(spring_event~(lat_prov|species), data = d) 
+# fall first since we are more interested in those... well I need to tidy up this entire script
 
-# fitA_spring_lat
-fitA_spring_lat <- readRDS("output/model_fit/fitA_spring_lat.RDS")
-fitA_spring_mat <- readRDS("output/model_fit/fitA_spring_mat.RDS")
 
-draws <- as.matrix(fitA_spring_lat)
+# fitC_fall_lat
+fitC_fall_lat <- readRDS("output/model_fit/fitC_fall_lat.RDS")
+draws_fitC_fall_lat <- as.matrix(fitC_fall_lat)
+
+evergreens <- c("Picea engelmannii B","Picea sitchensis D","Tsuga heterophylla E")
+library(stringr)
+evergreens <- str_replace_all(evergreens," ","_")
+deciduous <- c("Alnus rubra A", "Fagus sylvatica R*","Fraxinus excelsior Q*",
+               "Populus trichocarpa D")
+deciduous <- str_replace_all(deciduous," ","_")
+
+# Get all the species of one type now
+evergreensdraws_fitC_fall_lat <- matrix(data=NA,
+                                        nrow=nrow(draws_fitC_fall_lat),
+                                        ncol=length(evergreens))
+for (i in c(1:length(evergreens))){
+  evergreensdraws_fitC_fall_lat[,i] <- draws_fitC_fall_lat[,paste("b[lat_prov species_garden:", evergreens[i], "]", sep="")]
+}
+
+deciduousdraws_fitC_fall_lat <- matrix(data=NA,
+                                       nrow=nrow(draws_fitC_fall_lat),
+                                       ncol=length(deciduous))
+for (i in c(1:length(deciduous))){
+  deciduousdraws_fitC_fall_lat[,i] <- draws_fitC_fall_lat[,paste("b[lat_prov species_garden:", deciduous[i], "]", sep="")]
+}
+
+# Now you can manipulate these posteriors with basic math ...  
[…]

### 2023-03-13 Analyses/script_model_continent_spp_type_effect_two_predictors.R

@@ -14,7 +14,7 @@ library(bayesplot)
 library(dplyr)
 library(ggplot2)
 
-d <- read.csv("input/percentage_overlap_doy_difference_earth_calculated_garden_identifier_adjusted_fall_diffo_included.csv", header = TRUE)  #680
+d<- read.csv("input/percentage_overlap_doy_difference_earth_calculated_garden_identifier_adjusted_fall_diffo_included_slope_intercept_gdd_Sept27.csv",header = TRUE)
 d$fall_event <- as.numeric(d$fall_event)
 d$spring_event <- as.numeric(d$spring_event)
 
@@ -29,7 +29,7 @@ alnrubslope <- draws[,"b[percentage:sd species:Alnus_rubra]"]
 hist(alnrubslope)
 
 
-
+# draws <- as.matrix()
 # To get the estimate for continent or evergreen/deciduous:
 # Here I show an incomplete example for evergreen/deciduous,
 # but we especially want to do for continent (group species by continent)

### 2023-03-13 Analyses/source_fall_models_for_sweave.R

@@ -1,0 +1,226 @@
+
+# this document acts as a source file which I will read usinng ms.rnw
+# it contains
+
+# fitC_fall_lat
+# fitC_fall_mat
+
+library(dplyr)
+library(tidyr)
+library(Rcpp)
+library(rstanarm)
+library(bayesplot)
+library(ggplot2)
+library(viridis)
+library(rstantools)
+
+
+# fall evergreen vs deciduous
+evergreens <- c("Picea engelmannii B","Picea sitchensis D","Tsuga heterophylla E")
+library(stringr)
+evergreens <- str_replace_all(evergreens," ","_")
+deciduous <- c("Alnus rubra A", "Fagus sylvatica R*","Fraxinus excelsior Q*",
+               "Populus trichocarpa D")
+deciduous <- str_replace_all(deciduous," ","_")
+
+# fall europeans vs north americans
+northamerican <- c("Alnus rubra A", "Populus trichocarpa D",
+                   "Picea engelmannii B","Picea sitchensis D", "Tsuga heterophylla E")
+library(stringr)
+northamerican <- str_replace_all(northamerican," ","_")
+european <- c("Fagus sylvatica R*","Fraxinus excelsior Q*")
+european<- str_replace_all(european," ","_")
+
+
+
+
+# fitC_fall_mat
+fitC_fall_mat <- readRDS("../output/model_fit/fitC_fall_mat.RDS")
+draws_fitC_fall_mat <- as.matrix(fitC_fall_mat)
+# Get all the species of one type now
+evergreensdraws_fitC_fall_mat <- matrix(data=NA,
+                                        nrow=nrow(draws_fitC_fall_mat),
+                                        ncol=length(evergreens))
+for (i in c(1:length(evergreens))){
+  evergreensdraws_fitC_fall_mat[,i] <- draws_fitC_fall_mat[,paste("b[MAT_prov species_garden:", evergreens[i], "]", sep="")]
+}
+
+deciduousdraws_fitC_fall_mat <- matrix(data=NA,
+                                       nrow=nrow(draws_fitC_fall_mat),
+                                       ncol=length(deciduous))
+for (i in c(1:length(deciduous))){
+  deciduousdraws_fitC_fall_mat[,i] <- draws_fitC_fall_mat[,paste("b[MAT_prov species_garden:", deciduous[i], "]", sep="")]
+}
+
+# Now you can manipulate these posteriors with basic math ...  
[…]

### 2023-03-25 Docs/manuscript.Rnw
added Sexpr in ms

@@
 [ADD here].
-We present estimates as mean $\pm$ 90\% uncertainty intervals, unless otherwise stated.
+We present estimates as mean $\pm$ 90\% uncertainty intervals given parenthetically, unless otherwise stated.
 %emw2Mar -- It would be good if you have tidy code you can post with the data so folks can fit the models themselves. 
@@
 These should be added as mean +/- 90% uncertainty (posterior) intervals (UI) parenthetically using /Sexpr so they update automatically.
-Overall, spring events were not related to provenance latitude (insert the mean slope here, followed by the 90\% intervals) or mean annual temperature in North America (insert the mean slope here, followed by the 90\% intervals).
-Spring events advanced slightly with provenance latitude and mean annual temperature in Europe (insert slope and UI), meaning spring events are slightly earlier where provenance mean annual temperature is lower (i.e., higher, more northern latitudes)(Fig.
+
+<<label=source_spring_models, echo=FALSE>>=
+#load models and data to report numbers of things
+# "C:/Users/alina/Documents/git/localadaptclim"
+# setwd("C:/Users/alina/Documents/git/localadaptclim/Docs")
+#getwd()
+source("../Analyses/source_spring_models_for_sweave.R")
+@
+
+Overall, spring events were not related to provenance latitude (mean slope = \Sexpr{format(mean(northamerican_springpost_fitC_spring_lat),digits = 1)} $\pm$
+\Sexpr{format(abs((quantile(northamerican_springpost_fitC_spring_lat, 0.95)-(quantile(northamerican_springpost_fitC_spring_lat, 0.05)))*0.5), digits = 2)}) or MAT (mean slope = \Sexpr{format(mean(northamerican_springpost_fitC_spring_mat),digits = 1)} $\pm$
+  \Sexpr{format(abs((quantile(northamerican_springpost_fitC_spring_mat, 0.95)-(quantile(northamerican_springpost_fitC_spring_mat, 0.05)))*0.5), digits = 2)}) in North America.
+\todo{add figure here}  Spring events advanced slightly with provenance latitude (mean slope = \Sexpr{format(mean(european_springpost_fitC_spring_lat),digits = 1)} $\pm$
+  \Sexpr{format(abs((quantile(european_springpost_fitC_spring_lat, 0.95)-(quantile(european_springpost_fitC_spring_lat, 0.05)))*0.5), digits = 2)}) and MAT (mean slope = \Sexpr{format(mean(european_springpost_fitC_spring_mat),digits = 1)} $\pm$
+  \Sexpr{format(abs((quantile(european_springpost_fitC_spring_mat, 0.95)-(quantile(european_springpost_fitC_spring_mat, 0.05)))*0.5), digits = 2)}) in Europe, meaning spring events are slightly earlier where provenance MAT is lower (i.e., higher, more northern latitudes)(Fig.
 \ref{figure:springfall_latmat} \& Supplement Table.
@@
 #load models and data to report numbers of things
+# "C:/Users/alina/Documents/git/localadaptclim"
+# setwd("C:/Users/alina/Documents/git/localadaptclim/Docs")
 #getwd()
@@
 
-In contrast, fall events (budset, leaf senescence, leaf abcission) advanced strongly with provenance latitude and mean annual temperature, meaning fall events are earlier where provenance mean annual temperature is lower.
-This relationship, however, was observed mostly in North America where fall events advance \Sexpr{format(abs(median(northamericanpost_fitC_fall_lat)),digits = 2)} (UI = \Sexpr{format(abs(quantile(northamericanpost_fitC_fall_lat, 0.9)), digits = 2)}) days per degree we move north, or \Sexpr{format(median(northamericanpost_fitC_fall_mat),digits = 2)} days (UI) per degree decline in MAT$^{\circ}$C (Fig.
+In contrast, fall events (budset, leaf senescence, leaf abcission) advanced strongly with provenance latitude and MAT, meaning fall events are earlier where provenance MAT is lower.
+This relationship, however, was observed mostly in North America where fall events advance \Sexpr{format(abs(mean(northamericanpost_fitC_fall_lat)),digits = 2)} (\Sexpr{format(abs(quantile(northamericanpost_fitC_fall_lat, 0.95)), digits = 2)}-\Sexpr{format(abs(quantile(northamericanpost_fitC_fall_lat, 0.05)), digits = 2)}) days per degree northward, or \Sexpr{format(mean(northamericanpost_fitC_fall_mat),digits = 2)} days (\Sexpr{format(abs(quantile(northamericanpost_fitC_fall_mat, 0.05)), digits = 2)}-\Sexpr{format(abs(quantile(northamericanpost_fitC_fall_mat, 0.95)), digits = 2)}) per degree decline in MAT$^{\circ}$C (Fig.
 \ref{figure:continent}).
-In Europe, such relationship is weak: advance \Sexpr{format(abs(median(europeanpost_fitC_fall_lat)),digits = 1)} (UI) days per degree we move north, or \Sexpr{format(median(europeanpost_fitC_fall_mat),digits = 1)} days (UI) per degree decline in MAT $^{\circ}$C(Fig.
+In Europe, such relationship is weak: advance \Sexpr{format(abs(mean(europeanpost_fitC_fall_lat)),digits = 1)} (\Sexpr{format(abs(quantile(europeanpost_fitC_fall_lat, 0.95)), digits = 1)}-\Sexpr{format(abs(quantile(europeanpost_fitC_fall_lat, 0.05)), digits = 2)}) days per degree northward, or \Sexpr{format(mean(europeanpost_fitC_fall_mat),digits = 1)} days (\Sexpr{format(abs(quantile(europeanpost_fitC_fall_mat, 0.05)), digits = 2)}-\Sexpr{format(abs(quantile(europeanpost_fitC_fall_mat, 0.95)), digits = 2)}) per degree decline in MAT $^{\circ}$C(Fig.
 \ref{figure:continent}).
@@
 However, effects of MAT on spring events weakly diverge: spring events get earlier as MAT increases in angiosperms and delay as MAT increases in gymnosperms, except for \emph{Pseudotsuga menziesii}.
-Fall events delay in warmer locations for both species types, but slightly more so for gymnosperms (\Sexpr{format(median(deciduouspost_fitC_fall_mat),digits = 2)} days VS.
-\Sexpr{format(median(evergreenspost_fitC_fall_mat),digits = 2)} days)(Fig.
+Fall events delay in warmer locations for both species types, but slightly more so for gymnosperms (\Sexpr{format(mean(deciduouspost_fitC_fall_mat),digits = 2)} days VS.
+\Sexpr{format(mean(evergreenspost_fitC_fall_mat),digits = 2)} days)(Fig.
 \ref{figure:spp_type}).

### 2023-03-25 Docs/manuscript.tex
added Sexpr in ms

@@
 These should be added as mean +/- 90% uncertainty (posterior) intervals (UI) parenthetically using /Sexpr so they update automatically.
-Overall, spring events were not related to provenance latitude (insert the mean slope here, followed by the 90\% intervals) or mean annual temperature in North America (insert the mean slope here, followed by the 90\% intervals).
-Spring events advanced slightly with provenance latitude and mean annual temperature in Europe (insert slope and UI), meaning spring events are slightly earlier where provenance mean annual temperature is lower (i.e., higher, more northern latitudes)(Fig.
+
+Overall, spring events were not related to provenance latitude (mean slope = 0.1 $\pm$
+0.16) or MAT (mean slope = -0.09 $\pm$
+  0.27) in North America. \todo{add figure here}  Spring events advanced slightly with provenance latitude (mean slope = 0.1 $\pm$
+  0.33) and MAT (mean slope = -0.2 $\pm$
+  0.39) in Europe, meaning spring events are slightly earlier where provenance MAT is lower (i.e., higher, more northern latitudes)(Fig.
 \ref{figure:springfall_latmat} \& Supplement Table.
@@
 
-In contrast, fall events (budset, leaf senescence, leaf abcission) advanced strongly with provenance latitude and mean annual temperature, meaning fall events are earlier where provenance mean annual temperature is lower.
-This relationship, however, was observed mostly in North America where fall events advance 4.2 (UI = 4) days per degree we move north, or 6.4 days (UI) per degree decline in MAT$^{\circ}$C (Fig.
+In contrast, fall events (budset, leaf senescence, leaf abcission) advanced strongly with provenance latitude and MAT, meaning fall events are earlier where provenance MAT is lower.
+This relationship, however, was observed mostly in North America where fall events advance 4.2 (3.9-4.5) days per degree northward, or 6.4 days (5.8-7) per degree decline in MAT$^{\circ}$C (Fig.
 \ref{figure:continent}).
-In Europe, such relationship is weak: advance 0.5 (UI) days per degree we move north, or 0.7 days (UI) per degree decline in MAT $^{\circ}$C(Fig.
+In Europe, such relationship is weak: advance 0.5 (0.2-1.2) days per degree northward, or 0.7 days (1-2.4) per degree decline in MAT $^{\circ}$C(Fig.
 \ref{figure:continent}).

### 2023-03-25 Analyses/source_spring_models_for_sweave.R

@@ -1,0 +1,244 @@
+
+# this document acts as a source file for spring models which I will read usinng ms.rnw
+# it contains
+
+# fitC_spring_lat
+# fitC_spring_mat
+
+library(dplyr)
+library(tidyr)
+library(Rcpp)
+library(rstanarm)
+library(bayesplot)
+library(ggplot2)
+library(viridis)
+library(rstantools)
+library(stringr)
+
+# spring evergreen vs deciduous_spring
+evergreens_spring <- c("Picea abies S*","Picea engelmannii B",
+                       "Picea mariana I","Picea sitchensis D","Pinus albicaulis C","Pinus ponderosa J",
+                       "Pseudotsuga menziesii H", "Tsuga heterophylla E","Tsuga heterophylla G")
+evergreens_spring <- stringr::str_replace_all(evergreens_spring," ","_")
+deciduous_spring <-  c("Alnus rubra A",
+                       "Betula papyrifera K",
+                       "Betula papyrifera L",
+                       "Betula papyrifera M","Fagus sylvatica R*","Fagus sylvatica T*",
+                       "Fraxinus excelsior Q*",
+                       "Populus balsamifera F","Populus trichocarpa D",
+                       "Quercus petraea U*","Quercus petraea V*")
+deciduous_spring <- stringr::str_replace_all(deciduous_spring," ","_")
+
+# spring europeans vs north americans
+northamerican_spring <- c("Alnus rubra A","Betula papyrifera K",
+                          "Betula papyrifera L", "Picea engelmannii B","Picea sitchensis D", "Pinus albicaulis C",
+                          "Populus trichocarpa D", "Tsuga heterophylla E","Tsuga heterophylla G", "Populus balsamifera F",
+                          "Picea mariana I","Pinus ponderosa J", "Pseudotsuga menziesii H", "Betula papyrifera M")
+library(stringr)
+northamerican_spring <- stringr::str_replace_all(northamerican_spring," ","_")
+european_spring <- c("Fagus sylvatica R*","Fagus sylvatica T*",
+                     "Fraxinus excelsior Q*","Picea abies S*", "Quercus petraea U*","Quercus petraea V*")
+european_spring<- stringr::str_replace_all(european_spring," ","_")
+
+
+
+
+# fitC_spring_mat
+fitC_spring_mat <- readRDS("../output/model_fit/fitC_spring_mat.RDS")
[…]

### 2023-04-04 Docs/manuscript.Rnw
updated results

@@
 
-Overall, spring events were not related to provenance latitude (mean slope = \Sexpr{format(mean(northamerican_springpost_fitC_spring_lat),digits = 1)} $\pm$
-\Sexpr{format(abs((quantile(northamerican_springpost_fitC_spring_lat, 0.95)-(quantile(northamerican_springpost_fitC_spring_lat, 0.05)))*0.5), digits = 2)}) or MAT (mean slope = \Sexpr{format(mean(northamerican_springpost_fitC_spring_mat),digits = 1)} $\pm$
-  \Sexpr{format(abs((quantile(northamerican_springpost_fitC_spring_mat, 0.95)-(quantile(northamerican_springpost_fitC_spring_mat, 0.05)))*0.5), digits = 2)}) in North America.
-\todo{add figure here}  Spring events advanced slightly with provenance latitude (mean slope = \Sexpr{format(mean(european_springpost_fitC_spring_lat),digits = 1)} $\pm$
-  \Sexpr{format(abs((quantile(european_springpost_fitC_spring_lat, 0.95)-(quantile(european_springpost_fitC_spring_lat, 0.05)))*0.5), digits = 2)}) and MAT (mean slope = \Sexpr{format(mean(european_springpost_fitC_spring_mat),digits = 1)} $\pm$
-  \Sexpr{format(abs((quantile(european_springpost_fitC_spring_mat, 0.95)-(quantile(european_springpost_fitC_spring_mat, 0.05)))*0.5), digits = 2)}) in Europe, meaning spring events are slightly earlier where provenance MAT is lower (i.e., higher, more northern latitudes)(Fig.
+Overall, spring events were not related to provenance latitude (\Sexpr{format(mean(northamerican_springpost_fitC_spring_lat),digits = 1)} [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.05), digits = 1)}-\Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.95), digits = 2)}]) or MAT (\Sexpr{format(mean(northamerican_springpost_fitC_spring_mat),digits = 1)} [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.05), digits = 2)}-\Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.95), digits = 2)}]) in North America.
+\todo{add figure here} Spring events advanced slightly with provenance latitude (\Sexpr{format(mean(european_springpost_fitC_spring_lat),digits = 1)} [\Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.05), digits = 2)}-\Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.95), digits = 2)}]) and MAT (\Sexpr{format(mean(european_springpost_fitC_spring_mat),digits = 1)} [\Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.05), digits = 2)}-\Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.95), digits = 2)}]) in Europe, meaning spring events are slightly earlier where provenance MAT is lower (i.e., higher, more northern latitudes)(Fig.
 \ref{figure:springfall_latmat} \& Supplement Table.
-\ref{supplement-table:model_sf_lat}).
+\ref{supplement-table:model_sf_lat}), though uncertainty intervals strongly overlapped zero.
 Results were similar using other distance metrics in lieu of latitude (see Supplement Fig.\ref{supplement-figure:lat_distance} for results using the absolute value of the distance and the spherical distance difference between provenance and garden latitude).

### 2023-04-04 Docs/manuscript.tex
updated results

@@
 
-Overall, spring events were not related to provenance latitude (mean slope = 0.1 $\pm$
-0.16) or MAT (mean slope = -0.09 $\pm$
-  0.27) in North America. \todo{add figure here}  Spring events advanced slightly with provenance latitude (mean slope = 0.1 $\pm$
-  0.33) and MAT (mean slope = -0.2 $\pm$
-  0.39) in Europe, meaning spring events are slightly earlier where provenance MAT is lower (i.e., higher, more northern latitudes)(Fig.
+Overall, spring events were not related to provenance latitude (0.1 [-0.06-0.26]) or MAT (-0.09 [-0.36-0.18]) in North America.
+\todo{add figure here} Spring events advanced slightly with provenance latitude (0.1 [-0.23-0.42]) and MAT (-0.2 [-0.55-0.23]) in Europe, meaning spring events are slightly earlier where provenance MAT is lower (i.e., higher, more northern latitudes)(Fig.
 \ref{figure:springfall_latmat} \& Supplement Table.
-\ref{supplement-table:model_sf_lat}).
+\ref{supplement-table:model_sf_lat}), though uncertainty intervals strongly overlapped zero.
 Results were similar using other distance metrics in lieu of latitude (see Supplement Fig.\ref{supplement-figure:lat_distance} for results using the absolute value of the distance and the spherical distance difference between provenance and garden latitude).

### 2023-04-04 Docs/manuscript.Rnw
edited results

@@
 
-Overall, spring events were not related to provenance latitude (\Sexpr{format(mean(northamerican_springpost_fitC_spring_lat),digits = 1)} [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.05), digits = 1)}-\Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.95), digits = 2)}]) or MAT (\Sexpr{format(mean(northamerican_springpost_fitC_spring_mat),digits = 1)} [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.05), digits = 2)}-\Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.95), digits = 2)}]) in North America.
-\todo{add figure here} Spring events advanced slightly with provenance latitude (\Sexpr{format(mean(european_springpost_fitC_spring_lat),digits = 1)} [\Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.05), digits = 2)}-\Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.95), digits = 2)}]) and MAT (\Sexpr{format(mean(european_springpost_fitC_spring_mat),digits = 1)} [\Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.05), digits = 2)}-\Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.95), digits = 2)}]) in Europe, meaning spring events are slightly earlier where provenance MAT is lower (i.e., higher, more northern latitudes)(Fig.
+Overall, spring events were not related to provenance latitude (\Sexpr{format(round(mean(northamerican_springpost_fitC_spring_lat),digits = 2), nsmall=2)} days/degree [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.05), digits = 1)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.95), digits = 2)}]) or MAT (\Sexpr{format(mean(northamerican_springpost_fitC_spring_mat),digits = 1)} days/degree [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.95), digits = 2)}]) in North America.
+\todo{add figure here} Spring events advanced slightly with provenance latitude (\Sexpr{format(round(mean(european_springpost_fitC_spring_lat),digits = 2),nsmall=2)} days/degree [\Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.95), digits = 2)}]) and MAT (\Sexpr{format(round(mean(european_springpost_fitC_spring_mat),digits = 2),nsmall=2)} days/degree [\Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.95), digits = 2)}]) in Europe, meaning spring events are slightly earlier where provenance MAT is lower (i.e., higher, more northern latitudes)(Fig.
 \ref{figure:springfall_latmat} \& Supplement Table.
@@
 In contrast, fall events (budset, leaf senescence, leaf abcission) advanced strongly with provenance latitude and MAT, meaning fall events are earlier where provenance MAT is lower.
-This relationship, however, was observed mostly in North America where fall events advance \Sexpr{format(abs(mean(northamericanpost_fitC_fall_lat)),digits = 2)} (\Sexpr{format(abs(quantile(northamericanpost_fitC_fall_lat, 0.95)), digits = 2)}-\Sexpr{format(abs(quantile(northamericanpost_fitC_fall_lat, 0.05)), digits = 2)}) days per degree northward, or \Sexpr{format(mean(northamericanpost_fitC_fall_mat),digits = 2)} days (\Sexpr{format(abs(quantile(northamericanpost_fitC_fall_mat, 0.05)), digits = 2)}-\Sexpr{format(abs(quantile(northamericanpost_fitC_fall_mat, 0.95)), digits = 2)}) per degree decline in MAT$^{\circ}$C (Fig.
+This relationship, however, was observed mostly in North America where fall events advance \Sexpr{format(round(abs(mean(northamericanpost_fitC_fall_lat)),digits = 2),nsmall=2)} (\Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree northward, or \Sexpr{format(round(mean(northamericanpost_fitC_fall_mat),digits = 2),nsmall=2)} days (\Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_mat, 0.95)), digits = 2),nsmall=2)}) per degree decline in MAT$^{\circ}$C (Fig.
 \ref{figure:continent}).
-In Europe, such relationship is weak: advance \Sexpr{format(abs(mean(europeanpost_fitC_fall_lat)),digits = 1)} (\Sexpr{format(abs(quantile(europeanpost_fitC_fall_lat, 0.95)), digits = 1)}-\Sexpr{format(abs(quantile(europeanpost_fitC_fall_lat, 0.05)), digits = 2)}) days per degree northward, or \Sexpr{format(mean(europeanpost_fitC_fall_mat),digits = 1)} days (\Sexpr{format(abs(quantile(europeanpost_fitC_fall_mat, 0.05)), digits = 2)}-\Sexpr{format(abs(quantile(europeanpost_fitC_fall_mat, 0.95)), digits = 2)}) per degree decline in MAT $^{\circ}$C(Fig.
+In Europe, such relationship is weak: advance \Sexpr{format(round(abs(mean(europeanpost_fitC_fall_lat)),digits = 2),nsmall=2)} (\Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree northward, or \Sexpr{format(round(mean(europeanpost_fitC_fall_mat),digits = 2),nsmall=2)} days (\Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_mat, 0.95)), digits = 2),nsmall=2)}) per degree decline in MAT $^{\circ}$C(Fig.
 \ref{figure:continent}).
@@
 However, effects of MAT on spring events weakly diverge: spring events get earlier as MAT increases in angiosperms and delay as MAT increases in gymnosperms, except for \emph{Pseudotsuga menziesii}.
-Fall events delay in warmer locations for both species types, but slightly more so for gymnosperms (\Sexpr{format(mean(deciduouspost_fitC_fall_mat),digits = 2)} days VS.
-\Sexpr{format(mean(evergreenspost_fitC_fall_mat),digits = 2)} days)(Fig.
+Fall events delay in warmer locations for both species types, but slightly more so for gymnosperms (\Sexpr{format(round(mean(deciduouspost_fitC_fall_mat),digits = 2),nsmall=2)} days VS.
+\Sexpr{format(round(mean(evergreenspost_fitC_fall_mat),digits = 2),nsmall=2)} days)(Fig.
 \ref{figure:spp_type}).

### 2023-04-04 Docs/manuscript.tex
edited results

@@
 
-Overall, spring events were not related to provenance latitude (0.1 [-0.06-0.26]) or MAT (-0.09 [-0.36-0.18]) in North America.
-\todo{add figure here} Spring events advanced slightly with provenance latitude (0.1 [-0.23-0.42]) and MAT (-0.2 [-0.55-0.23]) in Europe, meaning spring events are slightly earlier where provenance MAT is lower (i.e., higher, more northern latitudes)(Fig.
+Overall, spring events were not related to provenance latitude (0.10 days/degree [-0.06 - 0.26]) or MAT (-0.09 days/degree [-0.36 - 0.18]) in North America.
+\todo{add figure here} Spring events advanced slightly with provenance latitude (0.10 days/degree [-0.23 - 0.42]) and MAT (-0.16 days/degree [-0.55 - 0.23]) in Europe, meaning spring events are slightly earlier where provenance MAT is lower (i.e., higher, more northern latitudes)(Fig.
 \ref{figure:springfall_latmat} \& Supplement Table.
@@
 In contrast, fall events (budset, leaf senescence, leaf abcission) advanced strongly with provenance latitude and MAT, meaning fall events are earlier where provenance MAT is lower.
-This relationship, however, was observed mostly in North America where fall events advance 4.2 (3.9-4.5) days per degree northward, or 6.4 days (5.8-7) per degree decline in MAT$^{\circ}$C (Fig.
+This relationship, however, was observed mostly in North America where fall events advance 4.24 (3.95 - 4.53) days per degree northward, or 6.41 days (5.78 - 7.04) per degree decline in MAT$^{\circ}$C (Fig.
 \ref{figure:continent}).
-In Europe, such relationship is weak: advance 0.5 (0.2-1.2) days per degree northward, or 0.7 days (1-2.4) per degree decline in MAT $^{\circ}$C(Fig.
+In Europe, such relationship is weak: advance 0.47 (0.21 - 1.17) days per degree northward, or 0.70 days (1.04 - 2.42) per degree decline in MAT $^{\circ}$C(Fig.
 \ref{figure:continent}).
@@
 However, effects of MAT on spring events weakly diverge: spring events get earlier as MAT increases in angiosperms and delay as MAT increases in gymnosperms, except for \emph{Pseudotsuga menziesii}.
-Fall events delay in warmer locations for both species types, but slightly more so for gymnosperms (3.7 days VS.
-6.2 days)(Fig.
+Fall events delay in warmer locations for both species types, but slightly more so for gymnosperms (3.69 days VS.
+6.23 days)(Fig.
 \ref{figure:spp_type}).

### 2023-04-20 Docs/manuscript.Rnw
Some notes on results

@@
 
+%EMWApr20 -- please organize your first outline using \enumerate, which is much easier for me to follow and refer to #s then
 \section{Introduction}
@@
 %emw2Mar: Try to open with WHY (goal/aim) you did something more than what you did, not a rule, but try to do when possible. 
-To locate common garden studies that documented the timing of spring events of woody plant species we searched and reviewed the peer-reviewed literature.
+To locate common garden studies that reported the timing of spring events of woody plant species we searched and reviewed the peer-reviewed literature.
 On 14 December 2022 we searched Web of Science (Thompson Reuters, New York, NY) using the following terms:
@@
 %alinaMar5:think abt displaying number of studies or provenances
+%EMWApr20 -- table in supp and map in main text (figure caption should refer to table in supp)
 \\
@@
 \subsection{Analyses}
-To estimate clines in spring and fall phenological events across species we used hierarchical Bayesian models.
+To estimate clines in spring and fall phenological events across species we used Bayesian hierarchical models.
 We regressed day of year (DOY) of events against geographical and climate predictors with partial pooling (sometimes called `random effects') on the intercept and slope for each species within each garden.
@@
 %emw6Mar -- results are in past tense generally
-Our final dataset captured seven angiosperm and eight gymnosperm species from 19 common gardens, encompassing 384 North American provenances and 101 European provenances, with observations from 1962 to 2019.
+Our final dataset included seven angiosperm and eight gymnosperm species from 19 common gardens, encompassing 384 North American provenances and 101 European provenances, with observations from 1962 to 2019.
 Seven species also had fall event information available.
@@
 
-Overall, spring events were not related to provenance latitude (\Sexpr{format(round(mean(northamerican_springpost_fitC_spring_lat),digits = 2), nsmall=2)} days/degree [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.05), digits = 1)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.95), digits = 2)}]) or MAT (\Sexpr{format(mean(northamerican_springpost_fitC_spring_mat),digits = 1)} days/degree [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.95), digits = 2)}]) in North America.
-\todo{add figure here} Spring events advanced slightly with provenance latitude (\Sexpr{format(round(mean(european_springpost_fitC_spring_lat),digits = 2),nsmall=2)} days/degree [\Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.95), digits = 2)}]) and MAT (\Sexpr{format(round(mean(european_springpost_fitC_spring_mat),digits = 2),nsmall=2)} days/degree [\Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.95), digits = 2)}]) in Europe, meaning spring events are slightly earlier where provenance MAT is lower (i.e., higher, more northern latitudes)(Fig.
+%EMWApr20 -- if you look at the intervals the results in NA and Europe seem similar so I changed the text. Also -- you should report the OVERALL model estimate across both europe and NA ...
+Overall, spring events were not related to provenance latitude or MAT, neither across continents (Fig.
 \ref{figure:springfall_latmat} \& Supplement Table.
-\ref{supplement-table:model_sf_lat}), though uncertainty intervals strongly overlapped zero.
+\ref{supplement-table:model_sf_lat}, ADD latitude: stats, MAT: stats), nor in North America (latitude: \Sexpr{format(round(mean(northamerican_springpost_fitC_spring_lat),digits = 2), nsmall=2)} days/degree [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.05), digits = 1)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT \Sexpr{format(mean(northamerican_springpost_fitC_spring_mat),digits = 1)} days/degree [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.95), digits = 2)}]) or Europe (latitude: \Sexpr{format(round(mean(european_springpost_fitC_spring_lat),digits = 2),nsmall=2)} days/degree [\Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT:\Sexpr{format(round(mean(european_springpost_fitC_spring_mat),digits = 2),nsmall=2)} days/degree [\Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.95), digits = 2)}]).
 Results were similar using other distance metrics in lieu of latitude (see Supplement Fig.\ref{supplement-figure:lat_distance} for results using the absolute value of the distance and the spherical distance difference between provenance and garden latitude).
+\\
+\todo{add figure here} 
+% Spring events advanced slightly with provenance latitude in Europe spring events are slightly earlier where provenance MAT is lower (i.e., higher, more northern latitudes), though uncertainty intervals strongly overlapped zero.
 
+%EMWApr20 -- reminder that Fig 1 needs a way to tell apart Europe and NA garden -- you can say in caption for now how the GardenID relates and eventually show on figure (add some text and a bracket to gardenID legend I think)
 \begin{figure}[!h] 
@@
 
+%EMWApr20 -- add stats for model (across Europe and NA) in first sentence. 
 In contrast, fall events (budset, leaf senescence, leaf abcission) advanced strongly with provenance latitude and MAT, meaning fall events are earlier where provenance MAT is lower.
-This relationship, however, was observed mostly in North America where fall events advance \Sexpr{format(round(abs(mean(northamericanpost_fitC_fall_lat)),digits = 2),nsmall=2)} (\Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree northward, or \Sexpr{format(round(mean(northamericanpost_fitC_fall_mat),digits = 2),nsmall=2)} days (\Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_mat, 0.95)), digits = 2),nsmall=2)}) per degree decline in MAT$^{\circ}$C (Fig.
+This relationship, however, was observed mostly in North America where fall events advance \Sexpr{format(round(abs(mean(northamericanpost_fitC_fall_lat)),digits = 2),nsmall=2)} (\Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree northward, or \Sexpr{format(round(mean(northamericanpost_fitC_fall_mat),digits = 2),nsmall=2)} days (\Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_mat, 0.95)), digits = 2),nsmall=2)}) per degree decline in MAT ($^{\circ}$C) (Fig.
+\ref{figure:continent}), whereas in Europe these relationships were weaker: advance of \Sexpr{format(round(abs(mean(europeanpost_fitC_fall_lat)),digits = 2),nsmall=2)} (\Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree northward, or \Sexpr{format(round(mean(europeanpost_fitC_fall_mat),digits = 2),nsmall=2)} days (\Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_mat, 0.95)), digits = 2),nsmall=2)}) per degree decline in MAT $^{\circ}$C(Fig.
 \ref{figure:continent}).
-In Europe, such relationship is weak: advance \Sexpr{format(round(abs(mean(europeanpost_fitC_fall_lat)),digits = 2),nsmall=2)} (\Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree northward, or \Sexpr{format(round(mean(europeanpost_fitC_fall_mat),digits = 2),nsmall=2)} days (\Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_mat, 0.95)), digits = 2),nsmall=2)}) per degree decline in MAT $^{\circ}$C(Fig.
-\ref{figure:continent}).
+\\
 
@@
 \end{figure}
+%EMWApr20 -- please fix all figure edits and add captions for next draft so we can finalize all the tweaks needed for submitting to journal
 
+%EMWApr20 -- add the stats in below paragraph also. Annoying I know, but we're almost done!
 Effects of provenance latitude on both spring and fall events were similar across angiosperms and gymnosperms.
-However, effects of MAT on spring events weakly diverge: spring events get earlier as MAT increases in angiosperms and delay as MAT increases in gymnosperms, except for \emph{Pseudotsuga menziesii}.
+However, effects of MAT on spring events weakly diverged: spring events shifted earlier as MAT increases in angiosperms and delay as MAT increases in gymnosperms, except for \emph{Pseudotsuga menziesii}.
 Fall events delay in warmer locations for both species types, but slightly more so for gymnosperms (\Sexpr{format(round(mean(deciduouspost_fitC_fall_mat),digits = 2),nsmall=2)} days VS.
 \Sexpr{format(round(mean(evergreenspost_fitC_fall_mat),digits = 2),nsmall=2)} days)(Fig.
-\ref{figure:spp_type}).
+\ref{figure:spp_type}).\\
 
@@
 
-We also tested whether climate effects on events may be stronger using metrics from seasonal windows closer to event dates, but results were not qualitatively different than using MAT.
+%EMWApr20 -- need to refer to some supp figures or tables or such below.
+Once you have the supp so I can see better what we're referring to here in first sentence we should likely edit this paragraph a little.
+We also tested whether climate effects on events were stronger when using metrics from seasonal windows closer to event dates, but results were not qualitatively different than using MAT (See supp?
+Not sure what results we mean here -- overlap?).
 We observed very weak effects of climate overlap on spring events, nearly identical across angiosperms and gymnosperms.
@@
 
+%EMWApr20 -- update figures with a) Spring events and continent ...
+b) Fall etc.
+so they look like ms figures.
+You also need to add the correct x axis -- they are likely days (spring or fall) per degree latitude or per degree MAT (C) ...
 \begin{figure}[!h] 

### 2023-05-06 Docs/manuscript.Rnw
updated ms and supp

--- 2023-04-04 Docs/manuscript.Rnw
@@
 
-We conducted the first cross-continental meta-analysis of published studies from the peer-reviewed literature that reported spring event dates for a mix of angiosperm and gymnosperm tree species in the northern hemisphere, capturing data from 384 North American provenances and 101 European provenances with observations from 1962 to 2019.
+% We conducted the first cross-continental meta-analysis of published studies from the peer-reviewed literature that reported spring event dates for a mix of angiosperm and gymnosperm tree species in the northern hemisphere, capturing data from 384 North American provenances and 101 European provenances with observations from 1962 to 2019.
 
 \section{Introduction}
-\begin{itemize}
-\item Define spring event, fall event
-\item Background
-\item Importance: An accurate understanding of how species distribution might be influenced by environmental conditions at a continental scale is critical for future range shift prediction. 
-\item Knowledge Gap: Anecdotally, the spring phenology of trees from European provenances appeared to exhibit stronger clines and less plasticity than in North American studies.
-Despite a growing interest in predicting local adaptation across locations, no study has comprehensively examined clines for spring and fall events and what factors may underlie differences observed across studies.
-\item Research question (links to results)
-\end{itemize}
-
-We ask: Across common garden studies, how strong are clines in spring and fall phenology?
-
-To better understand these clines, we also aim to examine:
-\begin{itemize}
-\item If differences in clines exist across Europe and North America.
-\item If differences in clines exist across angiosperm and gymnosperm species.
-\item If climate overlap (see Figure 1) between the provenance and garden explains the similarities and differences in clines.
-\end{itemize}
-To our knowledge, this marks the first study to synthesize spring phenology data retrieved from studies across North America and Europe.
-
-\subsection*{Ref notes}
-\todo[inline]{Good background information.}
-
-\citep{AitkenBemmels16}\\
-"Geographic variation in trees has been investigated since the mid-18th century.
-Similar patterns of clinal variation have been observed along latitudinal and elevational gradients in common garden experiments for many temperate and boreal species."\\
-"Scientists have studied relationships between tree populations and environmental characteristics of their provenances in common garden experiments for over 250 years."\\
+To optimize ecosystem services for habitat conservation and understand how plants cope with environmental variations, predicting the biological impacts of climate change is of critical importance \citep{botero15}.
+One of the most important parts of accurate prediction is understanding the relative importance of local adaptation versus plasticity \citep{chevin10}.
+Local adaptation - how well the growth and adaptive characteristics of plants align with the local climate conditions - influences their ability to thrive and reproduce  \citep{casmey22, kawecki04, anderson12, kim13}.
+When the selection intensity exceeds gene flow in species that spread over diverse areas, clinal variations among environmental factors can result in populations with unique adaptations to the local climate and, therefore, greater fitness \citep{kremer12, rossi15, salmela13}.
+On the other hand, phenotypic plasticity describes an individual genotype’s ability to exhibit different phenotypes in response to environmental variations, enabling plants to adjust to new environmental conditions and survive in this rapidly changing world (also known as ‘plastic rescue’)  \citep{chevin10, chevin102, snell18}.
 \\
-\citep{Alberto13}\\
-"Evolutionary responses are required for tree populations to be able to track climate change.
-Results of 250 years of common garden experiments show that most forest trees have evolved local adaptation, as evidenced by the adaptive differentiation of populations in quantitative traits, reflecting environmental conditions of population origins."\\
-"Thus, responding to climate change will likely require that the quantitative traits of populations again match their environments.
-We examine what kind of information is needed for evaluating the potential to respond, and what information is already available.
-We review the genetic models related to selection responses, and what is known currently about the genetic basis of the traits.
-We address special problems to be found at the range margins, and highlight the need for more modeling to understand specific issues at southern and northern margins.
-We need new common garden experiments for less known species.
-"\\
-"Modeling work on the potential of populations and species to respond genetically to recent climate change is advancing (see Hoffmann \& Sgrò, 2011; Franks \& Hoffmann, 2012; Shaw \& Etterson, 2012 for recent reviews).
-"\\
-
-(see e.g., Valladares et al., 2007; Caffarra et al., 2011; Hänninen \& Tanino, 2011)
-"The plastic response of different traits (e.g., phenology in trees) to variation in climate is, however, often much more complex than in heuristic models of adaptation."\\
-
-\citep{Loarie09}\\ 
-"It is also now understood that the rate of adaptation required by climate change varies among geographic regions." \improvement{need to read again}\\
-
-\citep{Nicotra10}\\ 
-"The immediate responses via phenotypic plasticity have also been considered in the context of climate change." \improvement{need to read again}\\
-
-\citep{Menzel99} \textbf{Growing season extended in Europe.}\improvement{good Europe background paper}\\
-\citep{Menzel06} \textbf{European phenological response to climate change matches the warming pattern.}\improvement{another good Europe background paper}\\
-\citep{Parmesan06}\\
-"Trees exhibit a high degree of phenotypic plasticity with respect to climatic variation.
-Phenological shifts of bud flush in response to recent increases in temperatures have been widely documented ." \improvement{can cite more papers on this matter}
-
-\citep{Gill15} \textbf{Changes in autumn senescence in northern hemisphere deciduous trees: a meta-analysis of autumn phenology studies.} \improvement{GREAT paper for us to base our writing on: similar meta-analysis but focuses on fall events}\\
-"the effects of climate change on the phenology of vegetation have received increased attention over the past several decades as any factor that alters the timing of early growing season leaf-out and senescence has the potential to affect a variety of ecosystem properties.
-For example, the timing of leaf-out and senescence of deciduous plants has been shown to affect plant competition (Fridley, 2012), plant growth (Myneni et al., 1997) and ecosystem carbon uptake (Barichivich et al., 2012).\\
-
-"While leaf-out has been shown to advance over the past century as a result of increasing air temperatures (Linderholm, 2006; Polgar and Primack, 2011), the relationship between temperature and leaf senescence remains less well understood (García-Plazaola et al., 2003; Richardson et al., 2013)."
-
-"Thus, an extension of the growing season can contribute to reduced atmospheric CO2 concentrations due to enhanced carbon sequestration in terrestrial plants (Penuelas et al., 2009; Richardson et al., 2013).
-However, the increase in carbon uptake may be partially offset by increased rates of ecosystem respiration (Piao et al., 2008)"
-
-"The lack of consistent relationships between air temperature and timing of leaf senescence suggests that autumn senescence may be influenced by a variety of factors that obscure its relationship with temperature."
-
-\todo[inline]{Cite some papers about the documented trends}
-
-tbc...
-
-\todo[inline]{Alina says: I found AitkenBemmels2016's research question to be wonderfully phrased and think we can do something similar}
-\citep{AitkenBemmels16}\\
-"To what extent are patterns of local adaptation along climatic gradients similar among species within a given geographic region?
-If clines in phenotypic traits are similar, can average patterns of sympatric variation guide AGF or identify critical climatic variables as a first approximation for untested species?
-We reanalysed data from the literature on provenance trials in temperate tree species from western North America to address these questions.
-This is an ideal region of focus because of the many provenance trials conducted on native tree species that are important for forestry, and because its high topographic complexity means that turnover in climate occurs rapidly and heterogeneously over short geographic spaces.
-Clines along climate gradients are thus espcially likely to reflect local adaptation to climate, rather than non-adaptive phenomena resulting from population demographic history."
 \\
-
-% alina_Jan27 -- I think we should have a section to define "spring events, fall events, distance between garden locations, difference of latitude, etc.
-Where do you think this section should go?
-Methods?
-EMW replies -- I think defining the events would be good, distance between garden locations and latitude etc should be SHORT (for example we use XX R package to calculate yy).
-I would put all this in the methods.
-The definition of events can go in the dataset assembly and I would start a section on 'analyses' where you include most of 2.3 *and* 2.4.
-I would move up where you got climate data from to above -- think of a 'Data' section with some subheaders and an 'analyses' section with some subheaders.
-
-% I am pasting a paragraph of the method section of Ettinger below to make reference to later on when I write the definition section. 
-"Defining budburst. Most studies defined budburst as initial ‘green tips’ (33 out of
-49 papers). Select studies defined budburst as a specific increment of growth (for
-example, ‘0.5 cm of new growth’) or as bud swell, leaf emergence, leaf unfolded,
-open bud scales or petiole emerged. The remaining papers (4 of 49) did not
-include a definition of budburst. Most papers using the above definitions (34 of 49)
-required only one bud to have met the defined criteria of budburst; however, the
-remaining studies implemented specific thresholds to be met (10–100\% of all buds
-on an individual needed to have bursted bud). For studies that quantified multiple
-measurements of percentage budburst over time (days), we extracted one value of
-‘days to budburst’ of these multiple measurements to make them comparable to
-other studies. To extract this summary value, we selected the days to budburst when
-percentage budburst was closest to 90\%, including estimates as low as 49.5\% budburst"
-
-%emw6Mar -- below opening and questions are good, but not appropriate for results. Move up to be at end of introduction. 
-\verb@Research questions recap@
+Among the various population-level characteristics critical in synchronizing plant growth with the available growing season, phenology traits hold particular significance \citep{primack80, oneil97, chuine01}.
+Phenology governs the timing of plant transitions between dormancy and active growing, allowing them to time reproduction and utilize each growing season as best as possible \citep{chuine10, hanninen11}.
+Additionally, phenology provides insight into how climate change affects plant life cycles, regeneration capacity, and ecosystem management.\\
+\\
+Spring phenology, specifically the timing of budburst and leaf flush, plays a significant role in determining plant fitness and has been widely adopted as a useful tool to investigate plant adaptation mechanisms \citep{guo22, chuine01}.
+In temperate and boreal environments, the optimal timing for spring growth initiation is determined by a balance of avoiding late spring frost and having a long growing season \citep{alberto11, lenz16, allevato19}.
+Utilizing the early portion of the growing season can be especially critical for species in colder regions \citep{morin07, dantec15} and the ones with suboptimal shade tolerance \citep{richardson09}.
+Plants that budburst earlier can take advantage of a longer growing season and avoid competition  \citep{guo22}.
+\\
+\\
+In recent decades, there has been growing interest in how climate change influences plant phenology \citep{Gill15}.
+Trees exhibit immediate responses via phenotypic plasticity with the warming climate \citep{Nicotra10}, while the degree of adaptation dictated by climate change differs across different geographical areas \citep{Loarie09}.
+Phenological shifts in spring phenology due to recent temperature increases have been extensively observed \citep{Parmesan06}.
+For example, leaf unfolding in Europe advanced up to six days since the early 1960s due to the rise in spring temperature, and spring events have generally shifted more than fall events \citep{menzel99}.
+Studies have found that species that shift more with warming also appear to perform better \citep{Gill15}.
+For instance, research has demonstrated that the timing of budburst and senescence in deciduous plants can impact plant competition \citep{fridley12}, plant growth \citep{myneni97}, and the uptake of carbon by the ecosystem \citep{Barichivich12}.
+As leaf phenology shifts in response to climate, increasing evidence suggests how much a species shifts its phenology with warming may predict fitness consequences: Individuals, populations, and phenological events (spring versus fall) with more plasticity might fare better with climate change.
+Plasticity, however, likely varies across species and phenophases.
+An accurate understanding of how environmental conditions might influence species distribution at a continental scale is critical for future range shift prediction.
+\\
+\\
+Much spring phenology is driven by plasticity.
+For more than 250 years, researchers have been comparing populations of trees from different geographical origins under the same environmental conditions through common garden experiments to disentangle the effects of environmental and genetic variation on trees’ phenology and phenotype \citep{AitkenBemmels16, Alberto13}.
+Over time, comparable clinal variations have been observed in these experiments along latitudinal and elevational gradients for many temperate and boreal species.
+Fall bud set phenology and cold hardiness exhibit strong population differentiation and strong clines along different latitudes, altitudes, and continentality gradients \citep{Alberto13}.
+For example, low-latitude populations have shown significantly stronger delays in leaf senescence over time than high-latitude populations \citep{Gill15}.
+Spring events (budburst, leaf flush, etc.) show more substantial phenotypic plasticity and higher variability \citep{AitkenBemmels16}.\\
 \\
 Despite a growing interest in predicting local adaptation across locations, no study has comprehensively examined clines for spring and fall events and what factors may underlie differences observed across studies.
-
-We ask: Across common garden studies, how strong are clines in spring and fall phenology?
-
-To better understand these clines, we also aim to examine:
+We ask: To what extent are patterns of phenotypic plasticity and local adaptation along climate gradients similar across common garden studies?
+To better understand these clines, we also aim to examine the following:
 \begin{itemize}
-\item If differences in clines exist across Europe and North America.
-\item If differences in clines exist across angiosperm and gymnosperm species.
-\item If climate overlap between the provenance and garden explains the similarities and differences in clines.
+\item How different are clines between spring and fall phenology? 
+\item How different are clines across Europe and North America?
+\item If differences in clines exist across angiosperm and gymnosperm species?
+\item If climate overlap between the provenance and garden explains the similarities and differences in clines?
 \end{itemize}
+To address these questions, we reanalyzed data from the literature on common garden experiments in temperate tree species across Europe and North America.
 
@@
 \subsection{Data Collection}
-%emw2Mar: Try to open with WHY (goal/aim) you did something more than what you did, not a rule, but try to do when possible. 
-To locate common garden studies that documented the timing of spring events of woody plant species we searched and reviewed the peer-reviewed literature.
+To locate common garden studies that reported the timing of spring events of woody plant species we searched and reviewed the peer-reviewed literature.
 On 14 December 2022 we searched Web of Science (Thompson Reuters, New York, NY) using the following terms:
@@
 We also contacted authors of previous review papers \citep{AitkenBemmels16, Alberto13}, to help further search the literature.
-We then reviewed the methods and results of all publications to refine to only studies that met the following criteria: (a) focused on woody plants originating from either Europe or North America, (b) had provenance trials/common gardens on the same continent, (c) reported spring events in units of calendar days (day of year or DOY) or could be converted into DOY (see `Additional Methods' section in the Supplement), and (d) reported latitude and longitude of provenances and gardens.
+We then reviewed the methods and results of all publications to refine to only studies that met the following criteria: (a) focused on woody plants originating from either Europe or North America, (b) had provenance trials/common gardens on the same continent, (c) reported latitude and longitude of provenances and gardens, and (d) reported spring events in units of calendar days (day of year or DOY) or could be converted into DOY (see \nameref{supplement-section:addmethods} in the Supplements).
 \\
 \\
-Based on these criteria we found 18 provenance/common garden studies, distributed throughout North America and Europe, with the majority of data concentrated in North America (Figure X).
-From each paper we extracted phenological data on spring events (budburst and leaf flush in DOY) and, when present in the same paper, fall events (bud set, leaf senescence, growth cessation, and leaf abscission) by species, provenance and garden location (geographic coordinates).
+Based on these criteria we found 19 common gardens distributed throughout North America and Europe, with the majority of data concentrated in North America (Fig.
+\ref{figure:map_gardens} \& Supplement Table.\ref{supplement-table:all_studies}).
+From each common garden study we extracted phenological data on spring events (budburst and leaf flush) in DOY and, when present in the same paper, fall events (bud set, leaf senescence, growth cessation, and leaf abscission) by species and the geographic information of provenances and gardens.
 We used ImageJ (version 1.53k; \citealp{schneider_rasband_eliceiri_2012}) to extract values from figures whenever necessary.
 For studies that reported event dates relative to a reference date other than 1 January (e.g.
-\citealp{Rehfeldt1994}), we converted such dates to DOY using the Lubridate Package in R.
-\improvement{display a table and a map here, and in the caption, clarify that some studies had more than one gardens and two studies shared the same garden}
-%emw2Mar -- mv the following sentence to the caption of the map figure: The distribution was skewed toward North America (n = X), with only X located in Europe.
-%alinaMar5:think abt displaying number of studies or provenances
+\citealp{Rehfeldt1994}), we converted such dates to DOY using the 'lubridate' package in R \citep{Grolemund11}.
 \\
-\\
+\begin{figure}[!h] 
+    \centering
+ \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/map_gardens.jpg}
+    \caption{Distribution of common gardens (denoted as stars) and provenances (denoted as circles) included in this meta-analysis.
+The distribution was skewed toward North America (number of studies = 12).
+See Table.\ref{supplement-table:all_studies} in Supplements for information on selected studies.}
+    \label{figure:map_gardens}
+\end{figure}
+\newline
 To understand how climatic differences, in addition to geographical differences, shape local adaptation in spring events we extracted several types of climate data using information about provenance latitude, longitude, and elevation from original publications.
 We estimated the mean annual temperature (MAT) for each provenance using the Climate Information Tool by Food and Agriculture Organization of the United Nations \citep{FAO2022}and ClimateWNA \citep{wang2016}.
-To examine more explicitly climate near spring events, we used gridded daily temperature data for 2011 to 2020 for all European and North American provenances and gardens from E-OBS and the daymetr R package respectively \citep{cornes2018,hufkens2018}.
-Using these data we estimated how much the daily temperatures overlapped between garden and provenance locations, which we call `climate overlap.' For this we used the Overlap package in R to calculate the percentage overlap of the daily temperature of each provenance in spring months (March to May) from 2011 to 2020 in relation to the daily temperature of their corresponding gardens.
-Finally, using the daily data, we also calculated Growing degree days (GDD), a commonly used heat accumulation measure to forecast phenological development in plants (Miller et al., 2001), based on the accumulation of mean daily temperatures ($T_{m}$)from 2011 to 2022 above a baseline of 0$^{\circ}$C, from January 1 until budburst and leaf flush with the following formula: GDD = $\sum$($T_{m}$ - $0^{\circ}$C for $T_{m}$ $\ge$ $0^{\circ}$C; 0 for $T_{m}$ $\le$ $0^{\circ}$C).
+To examine more explicitly climate near spring events, we used gridded daily temperature data for 2011 to 2020 for all European and North American provenances and gardens from E-OBS and the 'daymetr' package in R respectively \citep{cornes2018,hufkens2018}.
+Using these data we estimated how much the daily temperatures overlapped between garden and provenance locations, which we call `climate overlap.' For this we used the 'overlap' package in R to calculate the percentage overlap of the daily temperature of each provenance and their corresponding gardens in spring months (March to May) from 2011 to 2020.
+Finally, using the daily temperature data, we also calculated Growing degree days (GDD), a commonly used heat accumulation measure to forecast phenological development in plants (Miller et al., 2001), based on the accumulation of mean daily temperatures ($T_{m}$)from 2011 to 2020 above a baseline of 0$^{\circ}$C, from January 1 until budburst and leaf flush with the following formula: GDD = $\sum$($T_{m}$ - $0^{\circ}$C for $T_{m}$ $\ge$ $0^{\circ}$C; 0 for $T_{m}$ $\le$ $0^{\circ}$C).
 
 \subsection{Analyses}
-To estimate clines in spring and fall phenological events across species we used hierarchical Bayesian models.
-We regressed day of year (DOY) of events against geographical and climate predictors with partial pooling (sometimes called `random effects') on the intercept and slope for each species within each garden.
+To estimate clines in spring and fall phenological events across species we used Bayesian hierarchical models.
+We regressed DOY of events against geographical and climate predictors with partial pooling (sometimes called `random effects') on the intercept and slope for each species within each garden.
 Because most tree species were present in only one common garden in our dataset, it was impossible to fit garden and species separately, thus we treat each species within a garden as a unique group.
@@
 gymnosperm).
-All models were fit in rstanarm package (version 2.21.3; \citealp{brilleman2018}), using default priors, with 4 chains and 1000 sampling iterations per chain for a total of 4000 samples.
+All models were fit in rstanarm package (version 2.21.3; \citealp{brilleman2018}) using default priors, with 4 chains and 1000 sampling iterations per chain for a total of 4000 samples.
 We checked for model fit by ...
@@
 We present estimates as mean $\pm$ 90\% uncertainty intervals given parenthetically, unless otherwise stated.
-%emw2Mar -- It would be good if you have tidy code you can post with the data so folks can fit the models themselves. 
-%emw2Mar -- Did you use default priors?
-If so, can you figure out how to increase them threefold and see if answers change (they change run to run, but should not change dramatically)?
-Ideally they don't and then we can add 'increasing priors three-fold did not change results.'
+%emw2Mar -- It would be good if you have tidy code you can post with the data so folks can fit the models themselves
 
 \section{Results}
-%emw24Jan -- below is great opening!
-%emw6Mar -- results are in past tense generally
-Our final dataset captured seven angiosperm and eight gymnosperm species from 19 common gardens, encompassing 384 North American provenances and 101 European provenances, with observations from 1962 to 2019.
+Our final dataset included seven angiosperm and eight gymnosperm species from 19 common gardens, encompassing 384 North American provenances and 101 European provenances, with observations from 1962 to 2019.
 Seven species also had fall event information available.
 \\
-
-\todo[inline]{add map here}
-\todo[inline]{add a table documenting studies attributes} %emw6Mar -- Tables should go in supplement.
-But we should add estimates to statements we make throughout.
-These should be added as mean +/- 90% uncertainty (posterior) intervals (UI) parenthetically using /Sexpr so they update automatically.
-
 <<label=source_spring_models, echo=FALSE>>=
@@
 
-Overall, spring events were not related to provenance latitude (\Sexpr{format(mean(northamerican_springpost_fitC_spring_lat),digits = 1)} [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.05), digits = 1)}-\Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.95), digits = 2)}]) or MAT (\Sexpr{format(mean(northamerican_springpost_fitC_spring_mat),digits = 1)} [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.05), digits = 2)}-\Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.95), digits = 2)}]) in North America.
-\todo{add figure here} Spring events advanced slightly with provenance latitude (\Sexpr{format(mean(european_springpost_fitC_spring_lat),digits = 1)} [\Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.05), digits = 2)}-\Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.95), digits = 2)}]) and MAT (\Sexpr{format(mean(european_springpost_fitC_spring_mat),digits = 1)} [\Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.05), digits = 2)}-\Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.95), digits = 2)}]) in Europe, meaning spring events are slightly earlier where provenance MAT is lower (i.e., higher, more northern latitudes)(Fig.
-\ref{figure:springfall_latmat} \& Supplement Table.
-\ref{supplement-table:model_sf_lat}), though uncertainty intervals strongly overlapped zero.
-Results were similar using other distance metrics in lieu of latitude (see Supplement Fig.\ref{supplement-figure:lat_distance} for results using the absolute value of the distance and the spherical distance difference between provenance and garden latitude).
+Overall, spring events were not related to provenance latitude or MAT, neither across continents (latitude: \Sexpr{format(round(mean(all_species_springpost_fitC_spring_lat),digits = 2), nsmall=2)} days/degree [\Sexpr{format(quantile(all_species_springpost_fitC_spring_lat, 0.05), digits = 1)} - \Sexpr{format(quantile(all_species_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(round(mean(all_species_springpost_fitC_spring_mat),digits = 2), nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(all_species_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(all_species_springpost_fitC_spring_mat, 0.95), digits = 2)}]) (Fig.
+\ref{figure:springfall_latmat}, Table.
+\ref{supplement-table:model_spring_lat} \& \ref{supplement-table:model_spring_mat} in Supplements), nor in North America (latitude: \Sexpr{format(round(mean(northamerican_springpost_fitC_spring_lat),digits = 2), nsmall=2)} days/degree [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.05), digits = 1)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(mean(northamerican_springpost_fitC_spring_mat),digits = 1)} days/$^{\circ}$C [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.95), digits = 2)}]) or Europe (latitude: \Sexpr{format(round(mean(european_springpost_fitC_spring_lat),digits = 2),nsmall=2)} days/degree [\Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT:\Sexpr{format(round(mean(european_springpost_fitC_spring_mat),digits = 2),nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.95), digits = 2)}]).
+Results were similar using other distance metrics in lieu of latitude (see Supplement Fig.\ref{supplement-figure:lat_distance} for results using the difference between provenance and garden latitude, and the spherical distance between provenance and garden).
+\\
+% Spring events advanced slightly with provenance latitude in Europe spring events are slightly earlier where provenance MAT is lower (i.e., higher, more northern latitudes), though uncertainty intervals strongly overlapped zero.
+--- no longer the case
 
@@
     \centering
- \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/springfall_latmat.png}
-    \caption{Spring event Day of Year (DOY) in relation to provenance latitude and MAT, coded by symbol for species and color for garden with linear fits from hierarchical Bayesian models.
-Spring events shown on top and fall event at the bottom.}
+ \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/springfall_latmat.jpg}
+    \caption{Event day of year (DOY) in relation to provenance latitude and MAT, coded by symbol for species and color for garden with linear fits from hierarchical Bayesian models.
+Spring events shown on top and fall events at the bottom.
+European Gardens and species are bolded and denoted by an asterisk (*).
+}
     \label{figure:springfall_latmat}
@@
 
-In contrast, fall events (budset, leaf senescence, leaf abcission) advanced strongly with provenance latitude and MAT, meaning fall events are earlier where provenance MAT is lower.
-This relationship, however, was observed mostly in North America where fall events advance \Sexpr{format(abs(mean(northamericanpost_fitC_fall_lat)),digits = 2)} (\Sexpr{format(abs(quantile(northamericanpost_fitC_fall_lat, 0.95)), digits = 2)}-\Sexpr{format(abs(quantile(northamericanpost_fitC_fall_lat, 0.05)), digits = 2)}) days per degree northward, or \Sexpr{format(mean(northamericanpost_fitC_fall_mat),digits = 2)} days (\Sexpr{format(abs(quantile(northamericanpost_fitC_fall_mat, 0.05)), digits = 2)}-\Sexpr{format(abs(quantile(northamericanpost_fitC_fall_mat, 0.95)), digits = 2)}) per degree decline in MAT$^{\circ}$C (Fig.
-\ref{figure:continent}).
-In Europe, such relationship is weak: advance \Sexpr{format(abs(mean(europeanpost_fitC_fall_lat)),digits = 1)} (\Sexpr{format(abs(quantile(europeanpost_fitC_fall_lat, 0.95)), digits = 1)}-\Sexpr{format(abs(quantile(europeanpost_fitC_fall_lat, 0.05)), digits = 2)}) days per degree northward, or \Sexpr{format(mean(europeanpost_fitC_fall_mat),digits = 1)} days (\Sexpr{format(abs(quantile(europeanpost_fitC_fall_mat, 0.05)), digits = 2)}-\Sexpr{format(abs(quantile(europeanpost_fitC_fall_mat, 0.95)), digits = 2)}) per degree decline in MAT $^{\circ}$C(Fig.
-\ref{figure:continent}).
+In contrast, fall events (budset, leaf senescence, leaf abscission) advanced strongly with provenance latitude (\Sexpr{format(round(abs(mean(all_species_fallpost_fitC_fall_lat)),digits = 2), nsmall=2)} days/degree [\Sexpr{format(round(abs(quantile(all_species_fallpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)}-\Sexpr{format(round(abs(quantile(all_species_fallpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}]) and the decrease in MAT (\Sexpr{format(round(abs(mean(all_species_fallpost_fitC_fall_mat)),digits = 2), nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(all_species_fallpost_fitC_fall_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(all_species_fallpost_fitC_fall_mat, 0.95), digits = 2)}]), meaning fall events WEre earlier where provenance MAT was lower (Fig.
+\ref{figure:springfall_latmat}, Table.\ref{supplement-table:model_fall_lat} \& \ref{supplement-table:model_fall_mat} in Supplements ).
+This relationship, however, was observed mostly in North America where fall events advanced \Sexpr{format(round(abs(mean(northamericanpost_fitC_fall_lat)),digits = 2),nsmall=2)} (\Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree northward, or \Sexpr{format(round(mean(northamericanpost_fitC_fall_mat),digits = 2),nsmall=2)} days (\Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_mat, 0.95)), digits = 2),nsmall=2)}) per degree decline in MAT ($^{\circ}$C), whereas in Europe these relationships were weaker: advance of \Sexpr{format(round(abs(mean(europeanpost_fitC_fall_lat)),digits = 2),nsmall=2)} (\Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree northward, or \Sexpr{format(round(mean(europeanpost_fitC_fall_mat),digits = 2),nsmall=2)} days (\Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_mat, 0.95)), digits = 2),nsmall=2)}) per degree decline in MAT $^{\circ}$C(Fig.
+\ref{figure:continental_spptype_effect}A).
+\\
 
@@
     \centering
- \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/continent_histogram.png}
-    \caption{Caption placeholder}
-    \label{figure:continent}
+ \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/continental_spptype_effect.jpg}
+    \caption{Posterior distributions for the effect of MAT across different continents and species types.
+Zero -- no effect -- is shown with a dashed line.
+Solid line and shading in posteriors represent mean and 50 percent interval.
+(A) Effect of MAT and continent on spring and fall event date (DOY).
+Fall events advanced strongly with decreasing MAT, particularly notably in North America.
+Effect of MAT and leaf type on spring and fall event date (DOY) weakly diverged.
+(B) Effect of MAT and leaf type on spring and fall event date (DOY).
+As MAT increased, spring events slightly advanced in angiosperms and delayed in gymnosperms.
+Fall events delayed in warmer locations for both species types.}
+    \label{figure:continental_spptype_effect}
 \end{figure}
@@
 Effects of provenance latitude on both spring and fall events were similar across angiosperms and gymnosperms.
-However, effects of MAT on spring events weakly diverge: spring events get earlier as MAT increases in angiosperms and delay as MAT increases in gymnosperms, except for \emph{Pseudotsuga menziesii}.
-Fall events delay in warmer locations for both species types, but slightly more so for gymnosperms (\Sexpr{format(mean(deciduouspost_fitC_fall_mat),digits = 2)} days VS.
-\Sexpr{format(mean(evergreenspost_fitC_fall_mat),digits = 2)} days)(Fig.
-\ref{figure:spp_type}).
+Spring events slightly delayed \Sexpr{format(round(abs(mean(deciduous_springpost_fitC_spring_lat)),digits = 2), nsmall=2)} (\Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_lat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_lat, 0.95)), digits = 2),nsmall=2)}) days per degree north in angiosperms and \Sexpr{format(round(abs(mean(evergreens_springpost_fitC_spring_lat)),digits = 2), nsmall=2)} (\Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_lat, 0.05)), digits = 2),nsmall=2)}) days per degree north in gymnosperms.
+Fall events advanced \Sexpr{format(round(abs(mean(deciduouspost_fitC_fall_lat)),digits = 2), nsmall=2)} (\Sexpr{format(round(abs(quantile(deciduouspost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(deciduouspost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree north in angiosperms and \Sexpr{format(round(abs(mean(evergreenspost_fitC_fall_lat)),digits = 2), nsmall=2)} (\Sexpr{format(round(abs(quantile(evergreenspost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)}-\Sexpr{format(round(abs(quantile(evergreenspost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree north in gymnosperms.
+However, effects of MAT on spring events weakly diverged (Fig.\ref{figure:continental_spptype_effect}B).
+Spring events advanced \Sexpr{format(round(abs(mean(deciduous_springpost_fitC_spring_mat)),digits = 2), nsmall=2)} (\Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_mat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_mat, 0.05)), digits = 2),nsmall=2)}) days/$^{\circ}$C as MAT increased in angiosperms and delayed \Sexpr{format(round(abs(mean(evergreens_springpost_fitC_spring_mat)),digits = 2), nsmall=2)} (\Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_mat, 0.95)), digits = 2),nsmall=2)}) days/$^{\circ}$C as MAT increased in gymnosperms.
+Fall events delayed in warmer locations for both species types, but slightly more so for gymnosperms (\Sexpr{format(round(mean(evergreenspost_fitC_fall_mat),digits = 2),nsmall=2)} days) than angiosperms (\Sexpr{format(round(mean(deciduouspost_fitC_fall_mat),digits = 2),nsmall=2)} days) (Fig.
+\ref{figure:continental_spptype_effect}B).\\
 
-%(3.7 days VS. 6.2 days)
-
-We also tested whether climate effects on events may be stronger using metrics from seasonal windows closer to event dates, but results were not qualitatively different than using MAT.
-We observed very weak effects of climate overlap on spring events, nearly identical across angiosperms and gymnosperms.
-Fall events diverge as climate overlap declines for both angiosperms and gymnosperms, but slightly more strongly for gymnosperms (Fig.\ref{figure:overlap}).
-
-%emw6Mar -- I think we can skip subheaders in the results (yay!) but I need to read it again once you had added all the numbers and intervals. 
-\begin{figure}[!h] 
-    \centering
- \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/spp_type_histogram.png}
-    \caption{Caption placeholder}
-    \label{figure:spp_type}
-\end{figure}
-
-\begin{figure}[!h] 
-    \centering
- \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/climate_overlap.png}
-    \caption{Caption placeholder}
-    \label{figure:overlap}
-\end{figure}
-%emw6Mar -- Climate overlap figures can go in supp.
-
-%emw6Mar -- I could combine the current fig 3-4. Just be sure to label each panel.
+%EMWApr20 -- need to refer to some supp figures or tables or such below.
+Once you have the supp so I can see better what we're referring to here in first sentence we should likely edit this paragraph a little.
+While we believe that coarse metrics such as latitude and MAT would ultimately represent how similar the climates are between the provenances and gardens, we also estimated climate overlap in months that matter for the events to further test how much climate similarity between provenances and gardens predicts local adaptation.
+For spring events, we considered overlap across March to May.
+However, results were not qualitatively different than using MAT (See Fig.
+\ref{supplement-figure:overlap_scatterplot} in Supplements).
+We observed very weak effects of climate overlap on spring events (\Sexpr{format(round(abs(mean(all_species_springpost_fitC_spring_overlap)),digits = 2), nsmall=2)} [\Sexpr{format(round(abs(quantile(all_species_springpost_fitC_spring_overlap, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(all_species_springpost_fitC_spring_overlap, 0.95)), digits = 2),nsmall=2)}] days per one per cent increase in climate overlap), nearly identical across angiosperms (\Sexpr{format(round(abs(mean(deciduous_springpost_fitC_spring_overlap)),digits = 2), nsmall=2)} [\Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_overlap, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_overlap, 0.05)), digits = 2),nsmall=2)}]) and gymnosperms (\Sexpr{format(round(abs(mean(evergreens_springpost_fitC_spring_overlap)),digits = 2), nsmall=2)} [\Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_overlap, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_overlap, 0.95)), digits = 2),nsmall=2)}]).
+Fall events advanced as climate overlap declined, but slightly more strongly for gymnosperms (advancing \Sexpr{format(round(abs(mean(evergreenspost_fitC_fall_overlap)),digits = 2), nsmall=2)} [\Sexpr{format(round(abs(quantile(evergreenspost_fitC_fall_overlap, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(evergreenspost_fitC_fall_overlap, 0.95)), digits = 2),nsmall=2)}] days per one per cent decline in climate overlap) (Fig.\ref{supplement-figure:overlap_posterior} in Supplements).
 
@@
 % https://github.com/lizzieinvancouver/localadaptclim/issues/16
-
-% need to replot ofc
 
@@
 
-\citet{Keir11}
-
 \section{Figures}
-
-Rasband, W.S. (1997-2016). Image J.U.S. National Institute of Health, Bethesda, Maryland, USA.
-http://imagej.nih.gov/ij
 

### 2023-05-06 Docs/manuscript.tex
updated ms and supp

--- 2023-03-25 Docs/manuscript.tex
@@
 We present estimates as mean $\pm$ 90\% uncertainty intervals given parenthetically, unless otherwise stated.
-%emw2Mar -- It would be good if you have tidy code you can post with the data so folks can fit the models themselves. 
-%emw2Mar -- Did you use default priors?
-If so, can you figure out how to increase them threefold and see if answers change (they change run to run, but should not change dramatically)?
-Ideally they don't and then we can add 'increasing priors three-fold did not change results.'
+%emw2Mar -- It would be good if you have tidy code you can post with the data so folks can fit the models themselves
 
 \section{Results}
-%emw24Jan -- below is great opening!
-%emw6Mar -- results are in past tense generally
-Our final dataset captured seven angiosperm and eight gymnosperm species from 19 common gardens, encompassing 384 North American provenances and 101 European provenances, with observations from 1962 to 2019.
+Our final dataset included seven angiosperm and eight gymnosperm species from 19 common gardens, encompassing 384 North American provenances and 101 European provenances, with observations from 1962 to 2019.
 Seven species also had fall event information available.
@@
 
-\todo[inline]{add map here}
-\todo[inline]{add a table documenting studies attributes} %emw6Mar -- Tables should go in supplement.
-But we should add estimates to statements we make throughout.
-These should be added as mean +/- 90% uncertainty (posterior) intervals (UI) parenthetically using /Sexpr so they update automatically.
-
-Overall, spring events were not related to provenance latitude (mean slope = 0.1 $\pm$
-0.16) or MAT (mean slope = -0.09 $\pm$
-  0.27) in North America. \todo{add figure here}  Spring events advanced slightly with provenance latitude (mean slope = 0.1 $\pm$
-  0.33) and MAT (mean slope = -0.2 $\pm$
-  0.39) in Europe, meaning spring events are slightly earlier where provenance MAT is lower (i.e., higher, more northern latitudes)(Fig.
-\ref{figure:springfall_latmat} \& Supplement Table.
-\ref{supplement-table:model_sf_lat}).
-Results were similar using other distance metrics in lieu of latitude (see Supplement Fig.\ref{supplement-figure:lat_distance} for results using the absolute value of the distance and the spherical distance difference between provenance and garden latitude).
+Overall, spring events were not related to provenance latitude or MAT, neither across continents (latitude: 0.10 days/degree [-0.05 - 0.25]; MAT: -0.11 days/$^{\circ}$C [-0.34 - 0.12]) (Fig.
+\ref{figure:springfall_latmat}, Table.
+\ref{supplement-table:model_spring_lat} \& \ref{supplement-table:model_spring_mat} in Supplements), nor in North America (latitude: 0.10 days/degree [-0.06 - 0.26]; MAT: -0.09 days/$^{\circ}$C [-0.36 - 0.18]) or Europe (latitude: 0.10 days/degree [-0.23 - 0.42]; MAT:-0.16 days/$^{\circ}$C [-0.55 - 0.23]).
+Results were similar using other distance metrics in lieu of latitude (see Supplement Fig.\ref{supplement-figure:lat_distance} for results using the difference between provenance and garden latitude, and the spherical distance between provenance and garden).
+\\
+% Spring events advanced slightly with provenance latitude in Europe spring events are slightly earlier where provenance MAT is lower (i.e., higher, more northern latitudes), though uncertainty intervals strongly overlapped zero.
+--- no longer the case
 
@@
 
-In contrast, fall events (budset, leaf senescence, leaf abcission) advanced strongly with provenance latitude and MAT, meaning fall events are earlier where provenance MAT is lower.
-This relationship, however, was observed mostly in North America where fall events advance 4.2 (3.9-4.5) days per degree northward, or 6.4 days (5.8-7) per degree decline in MAT$^{\circ}$C (Fig.
-\ref{figure:continent}).
-In Europe, such relationship is weak: advance 0.5 (0.2-1.2) days per degree northward, or 0.7 days (1-2.4) per degree decline in MAT $^{\circ}$C(Fig.
-\ref{figure:continent}).
+In contrast, fall events (budset, leaf senescence, leaf abscission) advanced strongly with provenance latitude (3.16 days/degree [2.87-3.45]) and the decrease in MAT (4.78 days/$^{\circ}$C [4.1 - 5.4]), meaning fall events WEre earlier where provenance MAT was lower (Fig.
+\ref{figure:springfall_latmat}, Table.\ref{supplement-table:model_fall_lat} \& \ref{supplement-table:model_fall_mat} in Supplements ).
+This relationship, however, was observed mostly in North America where fall events advanced 4.24 (3.95 - 4.53) days per degree northward, or 6.41 days (5.78 - 7.04) per degree decline in MAT ($^{\circ}$C), whereas in Europe these relationships were weaker: advance of 0.47 (0.21 - 1.17) days per degree northward, or 0.70 days (1.04 - 2.42) per degree decline in MAT $^{\circ}$C(Fig.
+\ref{figure:continental_spptype_effect}A).
+\\
 
@@
 Effects of provenance latitude on both spring and fall events were similar across angiosperms and gymnosperms.
-However, effects of MAT on spring events weakly diverge: spring events get earlier as MAT increases in angiosperms and delay as MAT increases in gymnosperms, except for \emph{Pseudotsuga menziesii}.
-Fall events delay in warmer locations for both species types, but slightly more so for gymnosperms (3.7 days VS.
-6.2 days)(Fig.
-\ref{figure:spp_type}).
+Spring events slightly delayed 0.37 (0.15 - 0.59) days per degree north in angiosperms and 0.23 (0.00 - 0.46) days per degree north in gymnosperms.
+Fall events advanced 3.18 (2.76 - 3.62) days per degree north in angiosperms and 3.14 (2.81-3.47) days per degree north in gymnosperms.
+However, effects of MAT on spring events weakly diverged (Fig.\ref{figure:continental_spptype_effect}B).
+Spring events advanced 0.82 (0.54 - 1.11) days/$^{\circ}$C as MAT increased in angiosperms and delayed 0.76 (0.37 - 1.14) days/$^{\circ}$C as MAT increased in gymnosperms.
+Fall events delayed in warmer locations for both species types, but slightly more so for gymnosperms (6.23 days) than angiosperms (3.69 days) (Fig.
+\ref{figure:continental_spptype_effect}B).\\
 
-%(3.7 days VS. 6.2 days)
-
-We also tested whether climate effects on events may be stronger using metrics from seasonal windows closer to event dates, but results were not qualitatively different than using MAT.
-We observed very weak effects of climate overlap on spring events, nearly identical across angiosperms and gymnosperms.
-Fall events diverge as climate overlap declines for both angiosperms and gymnosperms, but slightly more strongly for gymnosperms (Fig.\ref{figure:overlap}).
-
-%emw6Mar -- I think we can skip subheaders in the results (yay!) but I need to read it again once you had added all the numbers and intervals. 
-\begin{figure}[!h] 
-    \centering
- \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/spp_type_histogram.png}
-    \caption{Caption placeholder}
-    \label{figure:spp_type}
-\end{figure}
-
-\begin{figure}[!h] 
-    \centering
- \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/climate_overlap.png}
-    \caption{Caption placeholder}
-    \label{figure:overlap}
-\end{figure}
-%emw6Mar -- Climate overlap figures can go in supp.
-
-%emw6Mar -- I could combine the current fig 3-4. Just be sure to label each panel.
+%EMWApr20 -- need to refer to some supp figures or tables or such below.
+Once you have the supp so I can see better what we're referring to here in first sentence we should likely edit this paragraph a little.
+While we believe that coarse metrics such as latitude and MAT would ultimately represent how similar the climates are between the provenances and gardens, we also estimated climate overlap in months that matter for the events to further test how much climate similarity between provenances and gardens predicts local adaptation.
+For spring events, we considered overlap across March to May.
+However, results were not qualitatively different than using MAT (See Fig.
+\ref{supplement-figure:overlap_scatterplot} in Supplements).
+We observed very weak effects of climate overlap on spring events (0.01 [0.02 - 0.03] days per one per cent increase in climate overlap), nearly identical across angiosperms (0.02 [0.00 - 0.05]) and gymnosperms (0.04 [0.00 - 0.09]).
+Fall events advanced as climate overlap declined, but slightly more strongly for gymnosperms (advancing 0.72 [0.51 - 0.92] days per one per cent decline in climate overlap) (Fig.\ref{supplement-figure:overlap_posterior} in Supplements).
 

### 2023-05-06 Docs/supplement.Rnw
updated ms and supp

@@
 % required 
+\usepackage{rotating}
+\usepackage{color}
+\usepackage{tabularray}
 \usepackage{amsmath}
+\usepackage{array}
 \usepackage{natbib}
@@
 
+\usepackage{colortbl}
+
 % make your captioning look better
@@
 \pagestyle{fancy}
-\fancyhead[LO]{November 2022}
+\fancyhead[LO]{2023}
 \fancyhead[RO]{Manuscript}
@@
 \author{Ziyun Zeng \& E. M. Wolkovich}
-\date{November 2022}
+\date{2023}
 \maketitle
 
-\section*{Methods}
-
-\subsection*{Additional Methods}
+\section{Additional Methods}
+\label{section:addmethods}
+% @ alina_jan27: do we need to cite those paper indicated above that we didnt end up using through bibtex??
+% EMW replies -- no, some journals will require a spreadsheet (what you discarded and why) though -- but until we pick a journal and know what's required we don't need to do any of that.
 We had to exclude several studies that reported spring events on a quantitative scale.
@@
 \\
-% @ alina_jan27: do we need to cite those paper indicated above that we didnt end up using through bibtex??
-% EMW replies -- no, some journals will require a spreadsheet (what you discarded and why) though -- but until we pick a journal and know what's required we don't need to do any of that.
 We additionally excluded several studies because we could not pinpoint their location, or because they focused on non-native species or elevational trends.
@@
 Finally, as we are focused on latitudinal trends we excluded studies that examined only provenance altitude (Vitasse et al., 2009; Vitasse et al., 2010; Li et al., 1997; Alberto et al., 2011; Acevedo-Rodríguez et al., 2006).
-\subsection*{Table for all studies}
-\subsection*{Mapped locations for all studies}
 
-\section*{Results}
+% \section*{Supplemental Figures and Tables}
 
-\subsection*{Corrected DOY results}
-\subsection*{Fitting each species \& common garden instead of just species}
-\subsection*{MAT difference}
-\subsection*{Might want to include how the model looked when we included studies with disagreeing provenance \& gardens}
-% if we indeed include this.. need to update results section in manuscript
+\begin{sidewaystable}
+\centering
+\caption{This table includes all publications that met search criteria for this meta-analysis. Some studies had more than one gardens and two studies shared the same garden (D)}
+\arrayrulecolor{black}
+\resizebox{\linewidth}{!}{%
+\begin{tabular}{!{\color{black}\vrule}>{\hspace{0pt}}m{0.046\linewidth}!{\color{black}\vrule}>{\hspace{0pt}}m{0.183\linewidth}!{\color{black}\vrule}>{\hspace{0pt}}m{0.09\linewidth}!{\color{black}\vrule}>{\hspace{0pt}}m{0.075\linewidth}!{\color{black}\vrule}>{\hspace{0pt}}m{0.108\linewidth}!{\color{black}\vrule}>{\hspace{0pt}}m{0.11\linewidth}!{\color{black}\vrule}>{\hspace{0pt}}m{0.131\linewidth}!{\color{black}\vrule}>{\hspace{0pt}}m{0.06\linewidth}!{\color{black}\vrule}>{\hspace{0pt}}m{0.133\linewidth}!{\color{black}\vrule}}
+\hline
+\textbf{No.} & \textbf{Publication}   & \textbf{Continent}                                      & \begin{tabular}[c]{@{}l@{}}\textbf{Garden } \\\textbf{ID}\end{tabular} & \textbf{Species}                                                                   & \begin{tabular}[c]{@{}l@{}}\textbf{Species } \\\textbf{Type}\end{tabular} & \begin{tabular}[c]{@{}l@{}}\textbf{Spring} \\\textbf{Event Definition}\end{tabular} & \begin{tabular}[c]{@{}l@{}}\textbf{Fall } \\\textbf{Event}\end{tabular} & \begin{tabular}[c]{@{}l@{}}\textbf{Fall Event } \\\textbf{Definition}\end{tabular}  \\
+\hline
+1            & Hamann et al., 1998    & \begin{tabular}[c]{@{}l@{}}North \\America\end{tabular} & A                                                                      & \begin{tabular}[c]{@{}l@{}}\textit{Alnus }\\\textit{rubra}\end{tabular}            & Angiosperm                                                                & Bud burst                                                                           & Yes                                                                     & \begin{tabular}[c]{@{}l@{}}Leaf \\abscission\end{tabular}                           \\
+\hline
+2            & Rehfeldt, 1994         & \begin{tabular}[c]{@{}l@{}}North \\America\end{tabular} & B                                                                      & \begin{tabular}[c]{@{}l@{}}\textit{Picea } \\\textit{engelmannii}\end{tabular}     & Gymnosperm                                                                & Bud burst                                                                           & Yes                                                                     & \begin{tabular}[c]{@{}l@{}}Leaf \\cessation\end{tabular}                            \\
+\hline
+3            & Bower  Aitken, 2008    & \begin{tabular}[c]{@{}l@{}}North \\America\end{tabular} & C                                                                      & \begin{tabular}[c]{@{}l@{}}\textit{Pinus } \\\textit{albicaulis}\end{tabular}      & Gymnosperm                                                                & Leaf flush                                                                          & No                                                                      & n.a.
+\\
+\hline
+4            & McKown et al., 2013    & \begin{tabular}[c]{@{}l@{}}North \\America\end{tabular} & D                                                                      & \begin{tabular}[c]{@{}l@{}}\textit{Populus } \\\textit{trichocarpa}\end{tabular}   & Angiosperm                                                                & \begin{tabular}[c]{@{}l@{}}Bud burst \\Leaf flush\end{tabular}                      & Yes                                                                     & Bud set                                                                             \\
+\hline
+5            & Mimura  Aitken 2007    & \begin{tabular}[c]{@{}l@{}}North \\America\end{tabular} & D                                                                      & \begin{tabular}[c]{@{}l@{}}\textit{Picea} \\\textit{sitchensis}\end{tabular}       & Gymnosperm                                                                & Bud burst                                                                           & Yes                                                                     & Bud set                                                                             \\
+\hline
+6            & Kuser, 1980            & \begin{tabular}[c]{@{}l@{}}North \\America\end{tabular} & E                                                                      & \begin{tabular}[c]{@{}l@{}}\textit{Tsuga } \\\textit{heterophylla}\end{tabular}    & Gymnosperm                                                                & Bud burst                                                                           & Yes                                                                     & Bud set                                                                             \\
+\hline
+7            & Farmer, 1993           & \begin{tabular}[c]{@{}l@{}}North \\America\end{tabular} & F                                                                      & \begin{tabular}[c]{@{}l@{}}\textit{Populus } \\\textit{balsamifera}\end{tabular}   & Angiosperm                                                                & Bud burst                                                                           & No                                                                      & n.a.
+\\
+\hline
+8            & Hannerz et al., 1999   & \begin{tabular}[c]{@{}l@{}}North \\America\end{tabular} & G                                                                      & \begin{tabular}[c]{@{}l@{}}\textit{Tsuga} \\\textit{heterophylla}\end{tabular}     & Gymnosperm                                                                & Bud burst                                                                           & No                                                                      & n.a.
+\\
+\hline
+9            & White et al., 1979     & \begin{tabular}[c]{@{}l@{}}North \\America\end{tabular} & H                                                                      & \begin{tabular}[c]{@{}l@{}}\textit{Pseudotsuga } \\\textit{menziesii}\end{tabular} & Gymnosperm                                                                & Bud burst                                                                           & No                                                                      & n.a.
+\\
+\hline
+10           & Guo et al., 2021       & \begin{tabular}[c]{@{}l@{}}North \\America\end{tabular} & I                                                                      & \begin{tabular}[c]{@{}l@{}}\textit{Picea } \\\textit{mariana}\end{tabular}         & Gymnosperm                                                                & Bud burst                                                                           & No                                                                      & n.a.
+\\
+\hline
+11           & Dixit et al., 2020     & \begin{tabular}[c]{@{}l@{}}North \\America\end{tabular} & J                                                                      & \begin{tabular}[c]{@{}l@{}}\textit{Pinus } \\\textit{ponderosa}\end{tabular}       & Gymnosperm                                                                & Bud burst                                                                           & No                                                                      & n.a.
+\\
+\hline
+12           & Hawkins  Dhar 2012     & \begin{tabular}[c]{@{}l@{}}North \\America\end{tabular} & K/L/M                                                                  & \begin{tabular}[c]{@{}l@{}}\textit{Betula } \\\textit{papyrifera}\end{tabular}     & Angiosperm                                                                & Bud burst                                                                           & No                                                                      & n.a.
+\\
+\hline
+13           & Rosique-Esplugas, 2021 & Europe                                                  & Q*                                                                     & \begin{tabular}[c]{@{}l@{}}\textit{Fraxinus } \\\textit{excelsior}\end{tabular}    & Angiosperm                                                                & Leaf flush                                                                          & Yes                                                                     & Leaf senescence                                                                     \\
+\hline
+14           & Petkova et al., 2017   & Europe                                                  & R*                                                                     & \begin{tabular}[c]{@{}l@{}}\textit{Fagus } \\\textit{sylvatica}\end{tabular}       & Angiosperm                                                                & Bud burst                                                                           & Yes                                                                     & Leaf senescence                                                                     \\
+\hline
+15           & Sogaard et al., 2008   & Europe                                                  & S*                                                                     & \begin{tabular}[c]{@{}l@{}}\textit{Picea } \\\textit{abies}\end{tabular}           & Gymnosperm                                                                & Bud burst                                                                           & No                                                                      & n.a.
+\\
+\hline
+16           & Gömöry  Paule 2011     & Europe                                                  & T*                                                                     & \begin{tabular}[c]{@{}l@{}}\textit{Fagus } \\\textit{sylvatica}\end{tabular}       & Angiosperm                                                                & Bud burst                                                                           & No                                                                      & n.a.
+\\
+\hline
+17           & Alberto et al., 2011   & Europe                                                  & U*/V*                                                                  & \begin{tabular}[c]{@{}l@{}}\textit{Quercus } \\\textit{petraea}\end{tabular}       & Angiosperm                                                                & Bud burst                                                                           & No                                                                      & n.a.
+\\
+\hline
+\end{tabular}
+}
+\arrayrulecolor{black}
+\label{table:all_studies}
+\end{sidewaystable}
 
-\subsection*{Similar results across provenance latitude, absolute value of difference between provenance and garden latitude, and spherical distance}
-Placeholder text (Fig. \ref{figure:lat_distance}).
+\definecolor{Gallery}{rgb}{0.929,0.929,0.929}
+\begin{table}
+\centering
+\caption{Model summary of the relationship between spring event day of year (DOY)and provenance latitude (lat\_prov), fitted by different species within a garden (species\_garden).
+European gardens and species are denoted by an asterisk(*).}
+\begin{tblr}{
+  row{even} = {Gallery},
+  column{4} = {r},
+  column{5} = {r},
+  column{6} = {r},
+  cell{2}{2} = {r},
+  cell{2}{3} = {r},
+  cell{3}{2} = {r},
+  cell{3}{3} = {r},
+  cell{4}{2} = {r},
+  cell{4}{3} = {r},
+  cell{5}{2} = {r},
+  cell{5}{3} = {r},
+  cell{6}{2} = {r},
+  cell{6}{3} = {r},
+  cell{7}{2} = {r},
+  cell{7}{3} = {r},
+  cell{8}{2} = {r},
+  cell{8}{3} = {r},
+  cell{9}{2} = {r},
+  cell{9}{3} = {r},
+  cell{10}{2} = {r},
+  cell{10}{3} = {r},
+  cell{11}{2} = {r},
+  cell{11}{3} = {r},
+  cell{12}{2} = {r},
+  cell{12}{3} = {r},
+  cell{13}{2} = {r},
+  cell{13}{3} = {r},
+  cell{14}{2} = {r},
+  cell{14}{3} = {r},
+  cell{15}{2} = {r},
+  cell{15}{3} = {r},
+  cell{16}{2} = {r},
+  cell{16}{3} = {r},
+  cell{17}{2} = {r},
+  cell{17}{3} = {r},
+  cell{18}{2} = {r},
+  cell{18}{3} = {r},
+  cell{19}{2} = {r},
+  cell{19}{3} = {r},
+  cell{20}{2} = {r},
+  cell{20}{3} = {r},
+  cell{21}{2} = {r},
+  cell{21}{3} = {r},
+  cell{22}{2} = {r},
+  cell{22}{3} = {r},
+  cell{23}{2} = {r},
+  cell{23}{3} = {r},
+  cell{24}{2} = {r},
+  cell{24}{3} = {r},
+  cell{25}{2} = {r},
+  cell{25}{3} = {r},
+  cell{26}{2} = {r},
+  cell{26}{3} = {r},
+}
+\textbf{Parameter}                                     & \textbf{mean} & \textbf{sd} & \textbf{10\%} & \textbf{50\%} & \textbf{90\%} \\
+(Intercept)                                            & 112.6         & 5.4         & 105.8         & 112.6         & 119.4         \\
+b[lat\_prov species\_garden:\textit{Alnus\_rubra\_A]}           & -0.7   & 0.3   & -1.1  & -0.7   & -0.4   \\
+b[lat\_prov species\_garden:\textit{Betula\_papyrifera\_K]}     & 1.1    & 0.4   & 0.5   & 1.1    & 1.6    \\
+b[lat\_prov species\_garden:\textit{Betula\_papyrifera\_L]}     & 0.9    & 0.4   & 0.4   & 0.9    & 1.4    \\
+b[lat\_prov species\_garden:\textit{Betula\_papyrifera\_M]}     & 0.8    & 0.4   & 0.3   & 0.8    & 1.3    \\
+b[lat\_prov species\_garden:\textit{Fagus\_sylvatica\_R*]}      & 0.2    & 0.3   & -0.2  & 0.2    & 0.7    \\
+b[lat\_prov species\_garden:\textit{Fagus\_sylvatica\_T*]}      & 0.2    & 0.2   & -0.1  & 0.2    & 0.4    \\
+b[lat\_prov species\_garden:\textit{Fraxinus\_excelsior\_Q*]}   & 0.9    & 0.3   & 0.4   & 0.9    & 1.3    \\
+b[lat\_prov species\_garden:\textit{Picea\_abies\_S*]}          & -0.4   & 0.5   & -1    & -0.3   & 0.2    \\
+b[lat\_prov species\_garden:\textit{Picea\_engelmannii\_B]}     & 0      & 0.2   & -0.3  & 0      & 0.2    \\
+b[lat\_prov species\_garden:\textit{Picea\_mariana\_I]}         & -0.3   & 0.5   & -0.9  & -0.3   & 0.2    \\
+b[lat\_prov species\_garden:\textit{Picea\_sitchensis\_D]}      & -0.1   & 0.2   & -0.4  & -0.1   & 0.1    \\
+b[lat\_prov species\_garden:\textit{Pinus\_albicaulis\_C]}      & -1.2   & 0.3   & -1.6  & -1.2   & -0.7   \\
+b[lat\_prov species\_garden:\textit{Pinus\_ponderosa\_J]}       & -0.5   & 0.7   & -1.4  & -0.5   & 0.4    \\
+b[lat\_prov species\_garden:\textit{Populus\_balsamifera\_F]}   & 0.2    & 0.4   & -0.3  & 0.2    & 0.8    \\
+b[lat\_prov species\_garden:\textit{Populus\_trichocarpa\_D]}   & 0.7    & 0.2   & 0.5   & 0.7    & 1      \\
+b[lat\_prov species\_garden:\textit{Pseudotsuga\_menziesii\_H]} & 0.6    & 0.3   & 0.2   & 0.6    & 1      \\
+b[lat\_prov species\_garden:\textit{Quercus\_petraea\_U*]}      & -0.1   & 0.7   & -1    & -0.1   & 0.8    \\
+b[lat\_prov species\_garden:\textit{Quercus\_petraea\_V*]}      & -0.1   & 0.7   & -1    & -0.1   & 0.7    \\
+b[lat\_prov species\_garden:\textit{Tsuga\_heterophylla\_E]}    & -0.8   & 0.3   & -1.2  & -0.8   & -0.4   \\
+b[lat\_prov species\_garden:\textit{Tsuga\_heterophylla\_G]}    & 0.6    & 0.5   & 0     & 0.6    & 1.3    \\
+sigma                                                           & 5.1    & 0.1   & 4.9   & 5.1    & 5.3    \\
+Sigma[species\_garden:(Intercept),(Intercept)]                  & 1604.9 & 600.8 & 957.6 & 1491.2 & 2411.2 \\
+Sigma[species\_garden:lat\_prov,(Intercept)]                    & -23    & 10.3  & -36.6 & -21.2  & -11.7  \\
+Sigma[species\_garden:lat\_prov,lat\_prov]                      & 0.5    & 0.2   & 0.3   & 0.5    & 0.8    
+\end{tblr}
+\label{table:model_spring_lat}
+\end{table}
+
+\definecolor{Gallery}{rgb}{0.929,0.929,0.929}
+\begin{table}
+\centering
+\caption{Model summary of the relationship between spring event day of year (DOY)and provenance MAT (MAT\_prov), fitted by different species within a garden (species\_garden).
+European gardens and species are denoted by an asterisk(*).}
+\begin{tblr}{
+  row{even} = {Gallery},
+  column{4} = {r},
+  column{5} = {r},
+  column{6} = {r},
+  cell{2}{2} = {r},
+  cell{2}{3} = {r},
+  cell{3}{2} = {r},
+  cell{3}{3} = {r},
+  cell{4}{2} = {r},
+  cell{4}{3} = {r},
+  cell{5}{2} = {r},
+  cell{5}{3} = {r},
+  cell{6}{2} = {r},
+  cell{6}{3} = {r},
+  cell{7}{2} = {r},
+  cell{7}{3} = {r},
+  cell{8}{2} = {r},
+  cell{8}{3} = {r},
+  cell{9}{2} = {r},
+  cell{9}{3} = {r},
+  cell{10}{2} = {r},
+  cell{10}{3} = {r},
+  cell{11}{2} = {r},
+  cell{11}{3} = {r},
+  cell{12}{2} = {r},
+  cell{12}{3} = {r},
+  cell{13}{2} = {r},
+  cell{13}{3} = {r},
+  cell{14}{2} = {r},
+  cell{14}{3} = {r},
+  cell{15}{2} = {r},
+  cell{15}{3} = {r},
+  cell{16}{2} = {r},
+  cell{16}{3} = {r},
+  cell{17}{2} = {r},
+  cell{17}{3} = {r},
+  cell{18}{2} = {r},
+  cell{18}{3} = {r},
+  cell{19}{2} = {r},
+  cell{19}{3} = {r},
+  cell{20}{2} = {r},
+  cell{20}{3} = {r},
+  cell{21}{2} = {r},
+  cell{21}{3} = {r},
+  cell{22}{2} = {r},
+  cell{22}{3} = {r},
+  cell{23}{2} = {r},
+  cell{23}{3} = {r},
+  cell{24}{2} = {r},
+  cell{24}{3} = {r},
+  cell{25}{2} = {r},
+  cell{25}{3} = {r},
+  cell{26}{2} = {r},
+  cell{26}{3} = {r},
+}
+\textbf{Parameter}                                     & \textbf{mean} & \textbf{sd} & \textbf{10\%} & \textbf{50\%} & \textbf{90\%} \\
+(Intercept)                                            & 115.1 & 5.1   & 108.7 & 115   & 121.7 \\
+b[MAT\_prov species\_garden:\textit{Alnus\_rubra\_A]}           & 0.9   & 0.5   & 0.2   & 0.9   & 1.5   \\
+b[MAT\_prov species\_garden:\textit{Betula\_papyrifera\_K]}     & -3.4  & 0.7   & -4.4  & -3.4  & -2.5  \\
+b[MAT\_prov species\_garden:\textit{Betula\_papyrifera\_L]}     & -2.1  & 0.7   & -3    & -2.1  & -1.2  \\
+b[MAT\_prov species\_garden:\textit{Betula\_papyrifera\_M]}     & -1.4  & 0.7   & -2.3  & -1.4  & -0.5  \\
+b[MAT\_prov species\_garden:\textit{Fagus\_sylvatica\_R*]}      & -0.5  & 0.7   & -1.4  & -0.5  & 0.4   \\
+b[MAT\_prov species\_garden:\textit{Fagus\_sylvatica\_T*]}      & 1     & 0.3   & 0.6   & 1     & 1.5   \\
+b[MAT\_prov species\_garden:\textit{Fraxinus\_excelsior\_Q*]}   & -1.2  & 0.6   & -1.9  & -1.2  & -0.4  \\
+b[MAT\_prov species\_garden:\textit{Picea\_abies\_S*]}          & 1.1   & 0.9   & 0     & 1.1   & 2.2   \\
+b[MAT\_prov species\_garden:\textit{Picea\_engelmannii\_B]}     & 0.1   & 0.3   & -0.4  & 0.1   & 0.5   \\
+b[MAT\_prov species\_garden:\textit{Picea\_mariana\_I]}         & 0.6   & 0.6   & -0.2  & 0.6   & 1.3   \\
+b[MAT\_prov species\_garden:\textit{Picea\_sitchensis\_D]}      & 0.3   & 0.4   & -0.2  & 0.3   & 0.8   \\
+b[MAT\_prov species\_garden:\textit{Pinus\_albicaulis\_C]}      & 2.3   & 0.7   & 1.4   & 2.3   & 3.2   \\
+b[MAT\_prov species\_garden:\textit{Pinus\_ponderosa\_J]}       & -0.1  & 0.5   & -0.7  & -0.1  & 0.5   \\
+b[MAT\_prov species\_garden:\textit{Populus\_balsamifera\_F]}   & -0.3  & 0.6   & -1    & -0.3  & 0.4   \\
+b[MAT\_prov species\_garden:\textit{Populus\_trichocarpa\_D]}   & -0.6  & 0.2   & -0.9  & -0.6  & -0.3  \\
+b[MAT\_prov species\_garden:\textit{Pseudotsuga\_menziesii\_H]} & -0.4  & 0.3   & -0.8  & -0.4  & 0     \\
+b[MAT\_prov species\_garden:\textit{Quercus\_petraea\_U*]}      & -0.5  & 0.4   & -1    & -0.5  & -0.1  \\
+b[MAT\_prov species\_garden:\textit{Quercus\_petraea\_V*]}      & -0.8  & 0.4   & -1.3  & -0.8  & -0.4  \\
+b[MAT\_prov species\_garden:\textit{Tsuga\_heterophylla\_E]}    & 3.9   & 0.7   & 3.1   & 3.9   & 4.8   \\
+b[MAT\_prov species\_garden:\textit{Tsuga\_heterophylla\_G]}    & -0.8  & 1.3   & -2.5  & -0.8  & 0.8   \\
+sigma                                                           & 4.9   & 0.1   & 4.8   & 4.9   & 5.1   \\
+Sigma[species\_garden:(Intercept),(Intercept)]                  & 537.4 & 170.4 & 349.2 & 508.8 & 760.8 \\
+Sigma[species\_garden:MAT\_prov,(Intercept)]                    & -9.7  & 10    & -22.2 & -8.7  & 1.6   \\
+Sigma[species\_garden:MAT\_prov,MAT\_prov]                      & 3.4   & 1.5   & 1.9   & 3.2   & 5.3   \\
+\end{tblr}
+\label{table:model_spring_mat}
+\end{table}
+
+% \usepackage{color}
+% \usepackage{tabularray}
+\definecolor{Gallery}{rgb}{0.929,0.929,0.929}
+\begin{table}
+\centering
+\caption{Model summary of the relationship between fall event day of year (DOY)and provenance latitude (lat\_prov), fitted by different species within a garden (species\_garden).
+European gardens and species are denoted by an asterisk(*).}
+\begin{tblr}{
+  row{even} = {Gallery},
+  column{4} = {r},
+  column{5} = {r},
+  column{6} = {r},
+  cell{2}{2} = {r},
+  cell{2}{3} = {r},
+  cell{3}{2} = {r},
+  cell{3}{3} = {r},
+  cell{4}{2} = {r},
+  cell{4}{3} = {r},
+  cell{5}{2} = {r},
+  cell{5}{3} = {r},
+  cell{6}{2} = {r},
+  cell{6}{3} = {r},
+  cell{7}{2} = {r},
+  cell{7}{3} = {r},
+  cell{8}{2} = {r},
+  cell{8}{3} = {r},
+  cell{9}{2} = {r},
+  cell{9}{3} = {r},
+  cell{10}{2} = {r},
+  cell{10}{3} = {r},
+  cell{11}{2} = {r},
+  cell{11}{3} = {r},
+  cell{12}{2} = {r},
+  cell{12}{3} = {r},
+  cell{13}{2} = {r},
+  cell{13}{3} = {r},
+}
+\textbf{Parameter}                                   & \textbf{mean} & \textbf{sd} & \textbf{10\%} & \textbf{50\%} & \textbf{90\%} \\
+(Intercept)                                          & 325.5         & 35          & 287.1         & 318.9         & 373.9         \\
+b[lat\_prov species\_garden:\textit{Alnus\_rubra\_A]}         & -2.3          & 0.5         & -3            & -2.3          & -1.6          \\
+b[lat\_prov species\_garden:\textit{Fagus\_sylvatica\_R*]}    & 0.2           & 0.6         & -0.5          & 0.3           & 1             \\
+b[lat\_prov species\_garden:\textit{Fraxinus\_excelsior\_Q*]} & -1.2          & 0.6         & -1.9          & -1.2          & -0.5          \\
+b[lat\_prov species\_garden:\textit{Picea\_engelmannii\_B]}   & -0.1          & 0.3         & -0.5          & -0.1          & 0.3           \\
+b[lat\_prov species\_garden:\textit{Picea\_sitchensis\_D]}    & -5.5          & 0.3         & -5.9          & -5.5          & -5.2          \\
+b[lat\_prov species\_garden:\textit{Populus\_trichocarpa\_D]} & -9.5          & 0.3         & -9.9          & -9.5          & -9.1          \\
+b[lat\_prov species\_garden:\textit{Tsuga\_heterophylla\_E]}  & -3.8          & 0.4         & -4.3          & -3.8          & -3.3          \\
+sigma                                                         & 8.2           & 0.3         & 7.8           & 8.2           & 8.6           \\
+Sigma[species\_garden:(Intercept),(Intercept)]                & 13147.6       & 4738.9      & 7953.6        & 12418.3       & 19373.1       \\
+Sigma[species\_garden:lat\_prov,(Intercept)]                  & -326.3        & 126.5       & -481.1        & -309.2        & -191.2        \\
+Sigma[species\_garden:lat\_prov,lat\_prov]                    & 12.2          & 8.4         & 6.5           & 10.2          & 19.2          
+\end{tblr}
+\label{table:model_fall_lat}
+\end{table}
+
+\definecolor{Gallery}{rgb}{0.929,0.929,0.929}
+\begin{table}
+\centering
+\caption{Model summary of the relationship between fall event day of year (DOY)and provenance MAT (MAT\_prov), fitted by different species within a garden (species\_garden).
+European gardens and species are denoted by an asterisk(*).}
+\begin{tblr}{
+  row{even} = {Gallery},
+  column{4} = {r},
+  column{5} = {r},
+  column{6} = {r},
+  cell{2}{2} = {r},
+  cell{2}{3} = {r},
+  cell{3}{2} = {r},
+  cell{3}{3} = {r},
+  cell{4}{2} = {r},
+  cell{4}{3} = {r},
+  cell{5}{2} = {r},
+  cell{5}{3} = {r},
+  cell{6}{2} = {r},
+  cell{6}{3} = {r},
+  cell{7}{2} = {r},
+  cell{7}{3} = {r},
+  cell{8}{2} = {r},
+  cell{8}{3} = {r},
+  cell{9}{2} = {r},
+  cell{9}{3} = {r},
+  cell{10}{2} = {r},
+  cell{10}{3} = {r},
+  cell{11}{2} = {r},
+  cell{11}{3} = {r},
+  cell{12}{2} = {r},
+  cell{12}{3} = {r},
+  cell{13}{2} = {r},
+  cell{13}{3} = {r},
+}
+\textbf{Parameter}                                   & \textbf{mean} & \textbf{sd} & \textbf{10\%} & \textbf{50\%} & \textbf{90\%} \\
+(Intercept)                                          & 270.3         & 23.1        & 238.7         & 273.6         & 296.9         \\
+b[MAT\_prov species\_garden:\textit{Alnus\_rubra\_A]}         & 2.8    & 1.1    & 1.4    & 2.8    & 4.3    \\
+b[MAT\_prov species\_garden:\textit{Fagus\_sylvatica\_R*]}    & -0.2   & 1.6    & -2.2   & -0.2   & 1.9    \\
+b[MAT\_prov species\_garden:\textit{Fraxinus\_excelsior\_Q*]} & 1.6    & 1.3    & -0.1   & 1.6    & 3.4    \\
+b[MAT\_prov species\_garden:\textit{Picea\_engelmannii\_B]}   & 0.3    & 0.8    & -0.7   & 0.2    & 1.2    \\
+b[MAT\_prov species\_garden:\textit{Picea\_sitchensis\_D]}    & 11.2   & 0.9    & 10.1   & 11.1   & 12.3   \\
+b[MAT\_prov species\_garden:\textit{Populus\_trichocarpa\_D]} & 10.5   & 0.5    & 9.9    & 10.5   & 11.2   \\
+b[MAT\_prov species\_garden:\textit{Tsuga\_heterophylla\_E]}  & 7.3    & 0.9    & 6.2    & 7.3    & 8.5    \\
+sigma                                                         & 10.7   & 0.4    & 10.2   & 10.7   & 11.3   \\
+Sigma[species\_garden:(Intercept),(Intercept)]                & 4746.1 & 2270.5 & 2455.2 & 4247.7 & 7676.8 \\
+Sigma[species\_garden:MAT\_prov,(Intercept)]                  & -328   & 177.8  & -545.8 & -301.2 & -150.8 \\
+Sigma[species\_garden:MAT\_prov,MAT\_prov]                    & 40.8   & 28.3   & 19.4   & 33.4   & 68.8   
+\end{tblr}
+\label{table:model_fall_mat}
+\end{table}
+
 \begin{figure}[!h] 
     \centering
- \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_supp/lat_latdiff_earth_distance.png}
-    \caption{Similar results across provenance latitude, absolute value of difference between provenance and garden latitude, and spherical distance}
+ \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_supp/lat_latdiff_earth_distance.jpg}
+    \caption{Results were similar across different distance metrics: provenance latitude, difference between provenance and garden latitude, and spherical distance between provenance and garden.}
     \label{figure:lat_distance}
 \end{figure}
-\todo[inline]{change caption later}
 
-\subsection* {Strong relationship between provenance latitude, MAT, and GDDs}
-\todo[inline]{Flagged for Lizzie: I am unsure about how to commment on what we are seeing in this plot.
-\newline There is a strong relationship between GDDs of each event day recorded and the latitude and MAT, but isn't that self-apparent?
-}
+\begin{figure}[!h] 
+    \centering
+ \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_supp/overlap_scatterplot.jpg}
+    \caption{Results from using climate overlap were not qualitatively different than using MAT.}
+    \label{figure:overlap_scatterplot}
+\end{figure}
 
-Placeholder text (Fig. \ref{figure:gdd}).
+\begin{figure}[!h] 
+    \centering
+ \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_supp/overlap_posterior.jpg}
+    \caption{We observed very weak effects of climate overlap on spring events (0.01 [0.02 - 0.03] days per one per cent increase in climate overlap), nearly identical across angiosperms (0.02 [0.00 - 0.05]) and gymnosperms (0.04 [0.00 - 0.09]).
+Fall events advanced as climate overlap declined, but slightly more strongly for gymnosperms
+(advancing 0.72 [0.51 - 0.92] days per one per cent decline in climate overlap).}
+    \label{figure:overlap_posterior}
+\end{figure}
 
-% all figures will be updated (color, font size, resolution)
 \begin{figure}[!h] 
@@
 \end{figure}
-
-\section {Don't have a specific sections for these ones yet}
 
@@
 
-\subsection*{Spring Lat VS. Fall Lat}
-\begin{table}
-\centering
-\caption{Model summary for the relationship between event day of year (DOY)and provenance latitude, fitted by species, in spring (left) and fall (right).}
-\todo[inline]{i'm not sure how to word this}
-\begin{tabular}[t]{lcc}
-\toprule
-  & $DOY_{Spring}~(Latitude|Species)$ & $DOY_{Fall}~(Latitude|Species)$\\
-\midrule
-Intercept & \num{114.219} & \num{316.736}\\
- & {}[\num{100.511}, \num{127.680}] & {}[\num{272.545}, \num{415.373}]\\
-Sigma[Species × Intercept,Intercept] & \num{1148.104} & \num{12381.073}\\
- & {}[\num{435.227}, \num{2825.532}] & {}[\num{6176.075}, \num{24554.731}]\\
-Sigma[Species × Latitude,Intercept] & \num{-13.454} & \num{-307.256}\\
- & {}[\num{-41.733}, \num{-1.069}] & {}[\num{-621.049}, \num{-122.796}]\\
-Sigma[Species × Latitude,Latitude] & \num{0.374} & \num{10.159}\\
- & {}[\num{0.111}, \num{1.030}] & {}[\num{5.266}, \num{33.131}]\\
-\midrule
-Num.Obs. & \num{671} & \num{349}\\
-R2 & \num{0.903} & \num{0.961}\\
-R2 Adj. & \num{0.902} & \num{0.960}\\
-R2 Marg. & \num{0.000} & \num{0.000}\\
-Log.Lik. & \num{-2340.362} & \num{-1217.949}\\
-ELPD & \num{-2358.2} & \num{-1231.9}\\
-ELPD s.e. & \num{27.8} & \num{22.1}\\
-LOOIC & \num{4716.4} & \num{2463.8}\\
-LOOIC s.e. & \num{55.6} & \num{44.2}\\
-WAIC & \num{4716.2} & \num{2463.3}\\
-RMSE & \num{7.87} & \num{8.44}\\
-r2.adjusted.marginal & \num{0.902} & \num{0.960}\\
-\bottomrule
-\end{tabular}
-\label{table:model_sf_lat}
-\end{table}
-
 \end{document}

### 2023-05-06 Analyses/source_fall_models_for_sweave.R

@@ -32,11 +32,32 @@ european <- c("Fagus sylvatica R*","Fraxinus excelsior Q*")
 european<- str_replace_all(european," ","_")
 
 
+# may 4, 2023 -> overall model results
+all_species_fall <- c("Picea engelmannii B","Picea sitchensis D","Tsuga heterophylla E","Alnus rubra A", "Fagus sylvatica R*","Fraxinus excelsior Q*",
+                      "Populus trichocarpa D")
+library(stringr)
+all_species_fall <- str_replace_all(all_species_fall," ","_")
+
 
 
 # fitC_fall_mat
 fitC_fall_mat <- readRDS("../output/model_fit/fitC_fall_mat.RDS")
+# fitC_fall_mat <- readRDS("C:/Users/alina/Documents/git/localadaptclim/output/model_fit/fitC_fall_mat.RDS")
 draws_fitC_fall_mat <- as.matrix(fitC_fall_mat)
+
+
+# Get all the species of one type now
+all_species_falldraws_fitC_fall_mat <- matrix(data=NA,
+                                              nrow=nrow(draws_fitC_fall_mat),
+                                              ncol=length(all_species_fall))
+for (i in c(1:length(all_species_fall))){
+  all_species_falldraws_fitC_fall_mat[,i] <- draws_fitC_fall_mat[,paste("b[MAT_prov species_garden:", all_species_fall[i], "]", sep="")]
+}
+
+# Now you can manipulate these posteriors with basic math ...  
+all_species_fallpost_fitC_fall_mat <- rowMeans(all_species_falldraws_fitC_fall_mat)
+
+
 # Get all the species of one type now
 evergreensdraws_fitC_fall_mat <- matrix(data=NA,
                                         nrow=nrow(draws_fitC_fall_mat),
@@ -76,12 +97,12 @@ png(filename="leaf_effect_mat_fall.png",
     res=300)
 leaftypeplot_fitC_fall_mat <- as.matrix(cbind(evergreenspost_fitC_fall_mat, deciduouspost_fitC_fall_mat))
 plot_title <- ggtitle("Leaf Type Effects on Fall DOY in Relation to Provenance MAT")
-color_scheme_set("viridis") 
+color_scheme_set("yellow")
 new_labels_leaftype <- c("Angiosperm","Gymnosperm")
-mcmc_areas(leaftypeplot_fitC_fall_mat)+ 
+mcmc_areas(leaftypeplot_fitC_fall_mat)+
[…]

### 2023-05-06 Analyses/source_spring_models_for_sweave.R

@@ -4,6 +4,7 @@
 
 # fitC_spring_lat
 # fitC_spring_mat
+# fitC_spring_overlap
 
 library(dplyr)
 library(tidyr)
@@ -41,11 +42,44 @@ european_spring <- c("Fagus sylvatica R*","Fagus sylvatica T*",
 european_spring<- stringr::str_replace_all(european_spring," ","_")
 
 
+# May 4 across continents All species
+all_species_spring <- c("Alnus rubra A","Betula papyrifera K",
+                        "Betula papyrifera L", "Picea engelmannii B","Picea sitchensis D", "Pinus albicaulis C",
+                        "Populus trichocarpa D", "Tsuga heterophylla E","Tsuga heterophylla G", "Populus balsamifera F",
+                        "Picea mariana I","Pinus ponderosa J", "Pseudotsuga menziesii H", "Betula papyrifera M",
+                        "Fagus sylvatica R*","Fagus sylvatica T*",
+                        "Fraxinus excelsior Q*","Picea abies S*", "Quercus petraea U*","Quercus petraea V*")
+library(stringr)
+all_species_spring<- stringr::str_replace_all(all_species_spring," ","_")
+# all_species_spring
+
+
 
 
 # fitC_spring_mat
 fitC_spring_mat <- readRDS("../output/model_fit/fitC_spring_mat.RDS")
+
+# fitC_spring_mat <- readRDS("C:/Users/alina/Documents/git/localadaptclim/Output/model_fit/fitC_spring_mat.RDS")
+# summary(fitC_spring_mat)
 draws_fitC_spring_mat <- as.matrix(fitC_spring_mat)
+
+
+# Get all the species of one type now
+all_species_springdraws_fitC_spring_mat <- matrix(data=NA,
+                                                  nrow=nrow(draws_fitC_spring_mat),
+                                                  ncol=length(all_species_spring))
+for (i in c(1:length(all_species_spring))){
+  all_species_springdraws_fitC_spring_mat[,i] <- draws_fitC_spring_mat[,paste("b[MAT_prov species_garden:", all_species_spring[i], "]", sep="")]
+}
+
+# Now you can manipulate these posteriors with basic math ...  
+all_species_springpost_fitC_spring_mat <- rowMeans(all_species_springdraws_fitC_spring_mat)
+
+
+
+
+
+
 # Get all the species of one type now
 evergreens_springdraws_fitC_spring_mat <- matrix(data=NA,
[…]

### 2023-05-25 Docs/manuscript.Rnw
Big edit of intro and some refs

@@
 \newpage
+% Journal ideas: Global Change Biology; Functional Ecology or Journal of Ecology next seem good ...
+maybe a biogeography journal (though I am not sure if we can afford one) ...
+maybe New Phytologist but I think they could be picky.
 
-\section*{Abstract}
+% EMW recommends parskip now instead of \\ too much: https://www.overleaf.com/learn/latex/Articles/How_to_change_paragraph_spacing_in_LaTeX#The_parskip_package ...
+add \usepackage{parskip} to preamble, delete \\ and see how it looks
 
-% We conducted the first cross-continental meta-analysis of published studies from the peer-reviewed literature that reported spring event dates for a mix of angiosperm and gymnosperm tree species in the northern hemisphere, capturing data from 384 North American provenances and 101 European provenances with observations from 1962 to 2019.
+% Just jotting down and putting some bits you have for abstract for now. 
+\begin{abstract}
+Climate change forecasting ....
+plasticity....
+adaptation.
+Spring tree phenology is critical to forecasting ...
+Despite a long history of study, however, the balance of local adaptation versus plasticity in determining environmental responses is not well understood, especially for spring events.
+We conducted the first cross-continental meta-analysis of published studies from the peer-reviewed literature that reported spring event dates for a mix of angiosperm and gymnosperm tree species in the northern hemisphere, capturing data from 384 North American provenances and 101 European provenances with observations from 1962 to 2019.
+We found ...
+This is important because...
+\end{abstract}
 
 \section{Introduction}
-To optimize ecosystem services for habitat conservation and understand how plants cope with environmental variations, predicting the biological impacts of climate change is of critical importance \citep{botero15}.
+%emw: FYI this is a first draft -- meaning it still has lots of words, sentences and other bits to fix; but best to do that as we go. 
+%emw: I noticed a lot of papers on local adaptation don't always use the term 'local adaptation' so we may want to avoid using it too much?
+I suspect it may be contested.
+I also wonder if my advice about definitions was wrong ...
+but let's definitely save that text for now.
+%emw: We need to 'funnel' in our intros -- until we submit to a plant journal focusing on plants too early is getting too narrow too fast.
+
+% To optimize ecosystem services for habitat conservation and understand how plants cope with environmental variations, predicting the biological impacts of climate change is of critical importance \citep{botero15}.
 One of the most important parts of accurate prediction is understanding the relative importance of local adaptation versus plasticity \citep{chevin10}.
-Local adaptation - how well the growth and adaptive characteristics of plants align with the local climate conditions - influences their ability to thrive and reproduce  \citep{casmey22, kawecki04, anderson12, kim13}.
+The increasing need to predict the biological impacts of climate change has made understanding how organisms cope with environmental variation more urgent \citep{botero15}.
+In particular, the relative importance of genetic adaptation versus plasticity is of paramount importance for prediction \citep{chevin10}, with plasticity expected to allow species to shift more rapidly with climate change than environmental responses based on local adaptations (CITES).
+
+% Local adaptation---how well the growth and adaptive characteristics align with the local climate conditions---influences their ability to thrive and reproduce  \citep{casmey22, kawecki04, anderson12, kim13}.
 When the selection intensity exceeds gene flow in species that spread over diverse areas, clinal variations among environmental factors can result in populations with unique adaptations to the local climate and, therefore, greater fitness \citep{kremer12, rossi15, salmela13}.
-On the other hand, phenotypic plasticity describes an individual genotype’s ability to exhibit different phenotypes in response to environmental variations, enabling plants to adjust to new environmental conditions and survive in this rapidly changing world (also known as ‘plastic rescue’)  \citep{chevin10, chevin102, snell18}.
-\\
-\\
-Among the various population-level characteristics critical in synchronizing plant growth with the available growing season, phenology traits hold particular significance \citep{primack80, oneil97, chuine01}.
-Phenology governs the timing of plant transitions between dormancy and active growing, allowing them to time reproduction and utilize each growing season as best as possible \citep{chuine10, hanninen11}.
-Additionally, phenology provides insight into how climate change affects plant life cycles, regeneration capacity, and ecosystem management.\\
-\\
-Spring phenology, specifically the timing of budburst and leaf flush, plays a significant role in determining plant fitness and has been widely adopted as a useful tool to investigate plant adaptation mechanisms \citep{guo22, chuine01}.
-In temperate and boreal environments, the optimal timing for spring growth initiation is determined by a balance of avoiding late spring frost and having a long growing season \citep{alberto11, lenz16, allevato19}.
-Utilizing the early portion of the growing season can be especially critical for species in colder regions \citep{morin07, dantec15} and the ones with suboptimal shade tolerance \citep{richardson09}.
-Plants that budburst earlier can take advantage of a longer growing season and avoid competition  \citep{guo22}.
-\\
-\\
-In recent decades, there has been growing interest in how climate change influences plant phenology \citep{Gill15}.
-Trees exhibit immediate responses via phenotypic plasticity with the warming climate \citep{Nicotra10}, while the degree of adaptation dictated by climate change differs across different geographical areas \citep{Loarie09}.
-Phenological shifts in spring phenology due to recent temperature increases have been extensively observed \citep{Parmesan06}.
-For example, leaf unfolding in Europe advanced up to six days since the early 1960s due to the rise in spring temperature, and spring events have generally shifted more than fall events \citep{menzel99}.
-Studies have found that species that shift more with warming also appear to perform better \citep{Gill15}.
-For instance, research has demonstrated that the timing of budburst and senescence in deciduous plants can impact plant competition \citep{fridley12}, plant growth \citep{myneni97}, and the uptake of carbon by the ecosystem \citep{Barichivich12}.
-As leaf phenology shifts in response to climate, increasing evidence suggests how much a species shifts its phenology with warming may predict fitness consequences: Individuals, populations, and phenological events (spring versus fall) with more plasticity might fare better with climate change.
-Plasticity, however, likely varies across species and phenophases.
-An accurate understanding of how environmental conditions might influence species distribution at a continental scale is critical for future range shift prediction.
-\\
-\\
-Much spring phenology is driven by plasticity.
-For more than 250 years, researchers have been comparing populations of trees from different geographical origins under the same environmental conditions through common garden experiments to disentangle the effects of environmental and genetic variation on trees’ phenology and phenotype \citep{AitkenBemmels16, Alberto13}.
-Over time, comparable clinal variations have been observed in these experiments along latitudinal and elevational gradients for many temperate and boreal species.
-Fall bud set phenology and cold hardiness exhibit strong population differentiation and strong clines along different latitudes, altitudes, and continentality gradients \citep{Alberto13}.
-For example, low-latitude populations have shown significantly stronger delays in leaf senescence over time than high-latitude populations \citep{Gill15}.
-Spring events (budburst, leaf flush, etc.) show more substantial phenotypic plasticity and higher variability \citep{AitkenBemmels16}.\\
-\\
-Despite a growing interest in predicting local adaptation across locations, no study has comprehensively examined clines for spring and fall events and what factors may underlie differences observed across studies.
-We ask: To what extent are patterns of phenotypic plasticity and local adaptation along climate gradients similar across common garden studies?
-To better understand these clines, we also aim to examine the following:
+On the other hand, phenotypic plasticity describes an individual genotype’s ability to exhibit different phenotypes in response to environmental variations, enabling plants to adjust to new environmental conditions and survive in this rapidly changing world (also known as `plastic rescue')  \citep{chevin10, chevin102, snell18}.
+
+Many of the currently observed responses to climate change appear to be mostly plastic (CITES), including the most reported biological response to climate change---shifting phenology.
+Phenology (timing of recurring seasonal events) governs the timing of transitions between dormancy and active growth for many organisms, allowing them to time reproduction and fully exploit the resources of each growing season \citep{chuine10,hanninen11}.
+% add some animal refs -- Chuine Annual Review would be good: Process-Based Models of Phenology for Plants and Animals
+As such, phenology plays a significant role in determining fitness (CITES).
+Shifted phenology in recent decades---with many events moving several days per decade (CITES)---has thus led to concerns about fitness consequences, and the limits of possible future shifts.
+Future phenological shifts will depend on how much phenology is determined by adaptation versus plasticity will determine, yet our understanding of the balance of these two approaches to variation is limited, even for species groups that are critical to both forecasting and have been well studied, such as trees.
+
+Tree phenology is critically important to climate change forecasting and has been an area of long-term study of adaptation and plasticity.
+The timing of budburst and senescence can impact plant competition \citep{fridley12}, growth \citep{myneni97}, and carbon sequestration \citep{Barichivich12}---thus effecting climate change itself.
+Hundreds of years of study have established common clines in fall phenology, as locations with shorter growing season (poleward, higher elevation) exhibit earlier stopping of growth (e.g., via budset), and connected these to an underlying proximate mechanism of changing photoperiod cues, driven by local adaptation to the local growing season and relevant photoperiod to stop growth (CITES).
+In contrast, spring phenology appears far more plastic (CITES) and determined more strongly by temperature (CITES).
+Many studies, however, have argued that spring phenology shows important levels of adaptation.
+
+Evidence for local adaptation in spring phenology comes mostly from common garden studies.
+For more than 250 years, researchers have been comparing populations of trees from different geographical origins under the same environmental conditions through these experiments to disentangle the effects of environmental and genetic variation on trees’ phenology and phenotype \citep{AitkenBemmels16, Alberto13}.
+Studies have documented provenance differences of XX-XX days per (degree C or latitude??) for some species (\emph{Populus} and \emph{Fagus}?) where??
+while others have not found any trends in budburst (give species and where).
+This has led to some debate over the prevalence and importance of adaptation in spring tree phenology.
+While clines of spring phenology have been found in studies in both Europe (CITES) and North America (CITES) they appear potentially more common in European studies.
+
+Continental differences in patterns of adaptation versus plasticity could be driven by climatic differences, as North American springs are more variable across years than European ones (CITES).
+Such high temporal variability means that distant sites can effectively experience the same spring climate, but in different years.
+Studies of spring phenology in arboreta suggest cues for budburst may vary depending on continental climate (CITES), but are poorly controlled compared to traditional provenance trials, making them difficult to use for tests of plasticity versus adaptation.
+Even for more carefully designed common garden differences in study design---especially studied species---may complicate understanding what underlies potential trends across continents, with European studies focusing more on deciduous angiosperms and North American studies focusing often on conifers.
+% Alberto shows elevation cline 
+% Soolananayakanahally et al 2013 finds weak cline ... I think there should also be Fagus studies. Vitasse has some elevational studies... 
+To test for evidence of adaptation in spring phenology and what factors may underlie differences observed across studies, we comprehensively examined clines for spring and fall events.
+We test for evidence of adaptation via provenance trends with latitude and climate.
+Further, to better understand these clines, we also examine the following:
 \begin{itemize}
 \item How different are clines between spring and fall phenology? 
-\item How different are clines across Europe and North America?
-\item If differences in clines exist across angiosperm and gymnosperm species?
-\item If climate overlap between the provenance and garden explains the similarities and differences in clines?
+\item How different are clines between Europe and North America?
+\item How different are clines between angiosperm and gymnosperm species?
 \end{itemize}
-To address these questions, we reanalyzed data from the literature on common garden experiments in temperate tree species across Europe and North America.
+To address these questions, we combined a new meta-analysis of all common garden experiments in temperate tree species across Europe and North America reporting spring phenology with Bayesian hierarchical models.
+
+ %REFS
+ % https://link.springer.com/article/10.1007/s00484-014-0900-y
+ % emw could not open bibliography_local_adaptation.bib so just adding my refs as separate file for now. 
+ % kharouba2018 for estimates of change across species per decade (could go with your menzel ref well)
+ % \cite{Cleland:2007or} classic on phenology and why it matters with climate change, carbon etc. -- easy read so I recommend it!
+ % \cite{chuinearees} covers both insect and plant phenology and discusses plasticity and adaptation
+ % bonamour2019 -- good paper on plasticity (whole issue of this journal was good) 
+ % savolainenAREES -- have not read recently but sounds relevant
+ % gauzere2020 -- cite and read
+ % \cite{ospreebbms,ospreephoto} -- relevant papers from the lab
+ % \cite{zohner2017} cover variation in spring climate across continents
+ % \cite{chang2021} and \cite{chuine2016} -- relevant for discussion of chilling
+ % See also Salk 2020 paper and some papers mentioned above
 
@@
 
+\begin{enumerate}
+\item Opening: 1-2 paragraph overview of our findings (spring phenology show no real clines, some variation by continent and gymnosperm/angiosperm but hard to fully disentangle).
+Suggests some small clines of spring phenology in some studies, but no consistent effects
+ \item Subheader: Adaptation in spring vs fall phenology: Implications for climate change responses
+ \begin{enumerate}
+ \item Mostly unorganized list of things we may want to say in this section
+ \begin{enumerate}
+ \item Current advances in spring phenology may continue, but delays in fall phenology will depend on speed of adaptation
+ \item If current advances in spring phenology continues it could mean shifts in spring frost etc.
+  \item Spring phenology, specifically the timing of budburst and leaf flush, plays a significant role in determining plant fitness and has been widely adopted as a useful tool to investigate plant adaptation mechanisms \citep{guo22, chuine01}.
+In temperate and boreal environments, the optimal timing for spring growth initiation is determined by a balance of avoiding late spring frost and having a long growing season \citep{alberto11, lenz16, allevato19}.
+Utilizing the early portion of the growing season can be especially critical for species in colder regions \citep{morin07, dantec15} and the ones with suboptimal shade tolerance \citep{richardson09}.
+Plants that budburst earlier can take advantage of a longer growing season and avoid competition  \citep{guo22}.
+ \item Clines in fall phenology are stronger and more consistent while clines in spring phenology are weaker and can go either way  (highlight that some studies and we should include WHICH ones do show clines by XX days).
+ \item Mention some of of the co and counter-gradient studies here
+  \item  An accurate understanding of how environmental conditions might influence species distribution at a continental scale is critical for future range shift prediction.
+ \item Implications of small differences in spring phenology could mean different gene flow etc. 
+  \item the degree of adaptation dictated by climate change differs across different geographical areas \citep{Loarie09}. 
+  \item Need to cite elevation studies and discuss them somewhere here.
+  \item Somewhere mention results from chamber studies and local adaptation (which are hard to interpret given issues with chilling) 
+ \item What else?
+ \end{enumerate}
+  \end{enumerate}
+ \item Subheader: Variation across continents and species groupings
+ \begin{enumerate}
+ \item  Unorganized list of things we may want to say in this section
+ \begin{enumerate}
+ \item No strong evidence of continental differences, could make sense based on long-term climate history perhaps ... 
+\item discuss xylem (ring versus diffuse porous spp) and embolism but might be too in the weeds (see Salk article)
+ \end{enumerate}
+ \end{enumerate}
+\end{enumerate}
+
 % Must cover in discussion:

### 2023-05-29 Docs/manuscript.Rnw
updated intro

@@
 % required
+
+% EMW recommends parskip now instead of \\ too much: https://www.overleaf.com/learn/latex/Articles/How_to_change_paragraph_spacing_in_LaTeX#The_parskip_package ...
+add \usepackage{parskip} to preamble, delete \\ and see how it looks
+\usepackage{parskip}
 \usepackage{amsmath}
@@
 
-% EMW recommends parskip now instead of \\ too much: https://www.overleaf.com/learn/latex/Articles/How_to_change_paragraph_spacing_in_LaTeX#The_parskip_package ...
-add \usepackage{parskip} to preamble, delete \\ and see how it looks
-
 % Just jotting down and putting some bits you have for abstract for now. 
@@
 The increasing need to predict the biological impacts of climate change has made understanding how organisms cope with environmental variation more urgent \citep{botero15}.
-In particular, the relative importance of genetic adaptation versus plasticity is of paramount importance for prediction \citep{chevin10}, with plasticity expected to allow species to shift more rapidly with climate change than environmental responses based on local adaptations (CITES).
+In particular, the relative importance of genetic adaptation versus plasticity is of paramount importance for prediction \citep{chevin10}, with plasticity expected to allow species to shift more rapidly with climate change than environmental responses based on local adaptations \citep{chevin102, snell18}.
 
@@
 
-Many of the currently observed responses to climate change appear to be mostly plastic (CITES), including the most reported biological response to climate change---shifting phenology.
+Many of the currently observed responses to climate change appear to be mostly plastic \citep{burton22,zettle21,bonamour19, king17}, including the most reported biological response to climate change---shifting phenology.
 Phenology (timing of recurring seasonal events) governs the timing of transitions between dormancy and active growth for many organisms, allowing them to time reproduction and fully exploit the resources of each growing season \citep{chuine10,hanninen11}.
-% add some animal refs -- Chuine Annual Review would be good: Process-Based Models of Phenology for Plants and Animals
-As such, phenology plays a significant role in determining fitness (CITES).
-Shifted phenology in recent decades---with many events moving several days per decade (CITES)---has thus led to concerns about fitness consequences, and the limits of possible future shifts.
+Many terrestrial animals time their breeding season to coincide with the availability of food resources.
+For example, migratory birds arrive at breeding grounds amidst the peaking of insects \citep{hallfors20}.
+Many butterfly species ensure their larvae have an ample food supply by synchronizing their egg-laying with the phenology of their host plants \citep{rytteri21,posle18}.
+Similarly, the migration and spawning of many marine organisms are closely tied to water temperature and the availability of planktonic food sources \citep{genner10}.
+Such synchronization enables them to provide optimal food for their young during critical growth stages.
+Therefore, climate warming and the associated changes in spring phenology timing, such as earlier leafout or insect emergence due to climate change, can disrupt this synchronization and negatively impact animal populations \citep{wann19,renner18, chu17}.
+As such, phenology plays a significant role in determining fitness \citep{guo22, chuine01}.
+Shifted phenology in recent decades---with many events moving several days per decade \citep{vita21,khar18,Menzel06} ---has thus led to concerns about fitness consequences, and the limits of possible future shifts.
 Future phenological shifts will depend on how much phenology is determined by adaptation versus plasticity will determine, yet our understanding of the balance of these two approaches to variation is limited, even for species groups that are critical to both forecasting and have been well studied, such as trees.
 
-Tree phenology is critically important to climate change forecasting and has been an area of long-term study of adaptation and plasticity.
-The timing of budburst and senescence can impact plant competition \citep{fridley12}, growth \citep{myneni97}, and carbon sequestration \citep{Barichivich12}---thus effecting climate change itself.
-Hundreds of years of study have established common clines in fall phenology, as locations with shorter growing season (poleward, higher elevation) exhibit earlier stopping of growth (e.g., via budset), and connected these to an underlying proximate mechanism of changing photoperiod cues, driven by local adaptation to the local growing season and relevant photoperiod to stop growth (CITES).
-In contrast, spring phenology appears far more plastic (CITES) and determined more strongly by temperature (CITES).
+Tree phenology is critically important to climate change forecasting and has been an area of long-term study of adaptation and plasticity \citep{Cleland:2007or}.
+The timing of budburst and senescence can impact plant competition \citep{fridley12}, growth \citep{myneni97}, and carbon sequestration \citep{Barichivich12}---thus affecting climate change itself.
+Hundreds of years of study have established common clines in fall phenology, as locations with shorter growing season (poleward, higher elevation) exhibit earlier stopping of growth (e.g., via budset), and connected these to an underlying proximate mechanism of changing photoperiod cues, driven by local adaptation to the local growing season and relevant photoperiod to stop growth \citep{Alberto13,Savolainen07}.
+In contrast, spring phenology appears far more plastic \citep{AitkenBemmels16} and determined more strongly by temperature \citep{flynn18}.
 Many studies, however, have argued that spring phenology shows important levels of adaptation.
@@
 For more than 250 years, researchers have been comparing populations of trees from different geographical origins under the same environmental conditions through these experiments to disentangle the effects of environmental and genetic variation on trees’ phenology and phenotype \citep{AitkenBemmels16, Alberto13}.
-Studies have documented provenance differences of XX-XX days per (degree C or latitude??) for some species (\emph{Populus} and \emph{Fagus}?) where??
-while others have not found any trends in budburst (give species and where).
+Studies have documented provenance differences of 2.08-4.17 days per degree latitude for some species (\emph{Picea abies} in \citealp{sog08} and \emph{Quercus petraea} in \citealp{deans96}) while others have not found any trends in budburst along latitudinal gradients (For example,\emph{Picea sitchensis} in \citealp{mimura07}, \emph{Picea glauca} in \citealp{Li97}, and \emph{Populus balsamifera} in \citealp{farmer93}).
 This has led to some debate over the prevalence and importance of adaptation in spring tree phenology.
-While clines of spring phenology have been found in studies in both Europe (CITES) and North America (CITES) they appear potentially more common in European studies.
+While clines of spring phenology have been found in studies in both Europe \citep{sog08,deans96,von95} and North America \citep{rossi15, soo13, hannerz99}, they appear potentially more common in European studies.
 
-Continental differences in patterns of adaptation versus plasticity could be driven by climatic differences, as North American springs are more variable across years than European ones (CITES).
+Continental differences in patterns of adaptation versus plasticity could be driven by climatic differences, as North American springs are more variable across years than European ones \citep{tward21,zohner2017,schwartz00}.
 Such high temporal variability means that distant sites can effectively experience the same spring climate, but in different years.
-Studies of spring phenology in arboreta suggest cues for budburst may vary depending on continental climate (CITES), but are poorly controlled compared to traditional provenance trials, making them difficult to use for tests of plasticity versus adaptation.
-Even for more carefully designed common garden differences in study design---especially studied species---may complicate understanding what underlies potential trends across continents, with European studies focusing more on deciduous angiosperms and North American studies focusing often on conifers.
-% Alberto shows elevation cline 
-% Soolananayakanahally et al 2013 finds weak cline ... I think there should also be Fagus studies. Vitasse has some elevational studies... 
+Studies of spring phenology in arboreta suggest cues for budburst may vary depending on continental climate \citep{cat23, ospreebbms,ospreephoto}, but are poorly controlled compared to traditional provenance trials, making them difficult to use for tests of plasticity versus adaptation \citep{gauzere2020}.
+Even for more carefully designed common garden differences in study design---especially studied species---may complicate understanding what underlies potential trends across continents, with European studies focusing more on deciduous angiosperms and North American studies focusing often on evergreen gymnosperms.
 To test for evidence of adaptation in spring phenology and what factors may underlie differences observed across studies, we comprehensively examined clines for spring and fall events.
@@
  %REFS
- % https://link.springer.com/article/10.1007/s00484-014-0900-y
- % emw could not open bibliography_local_adaptation.bib so just adding my refs as separate file for now. 
- % kharouba2018 for estimates of change across species per decade (could go with your menzel ref well)
- % \cite{Cleland:2007or} classic on phenology and why it matters with climate change, carbon etc. -- easy read so I recommend it!
- % \cite{chuinearees} covers both insect and plant phenology and discusses plasticity and adaptation
- % bonamour2019 -- good paper on plasticity (whole issue of this journal was good) 
  % savolainenAREES -- have not read recently but sounds relevant
  % gauzere2020 -- cite and read
- % \cite{ospreebbms,ospreephoto} -- relevant papers from the lab
- % \cite{zohner2017} cover variation in spring climate across continents
  % \cite{chang2021} and \cite{chuine2016} -- relevant for discussion of chilling
@@
 We then reviewed the methods and results of all publications to refine to only studies that met the following criteria: (a) focused on woody plants originating from either Europe or North America, (b) had provenance trials/common gardens on the same continent, (c) reported latitude and longitude of provenances and gardens, and (d) reported spring events in units of calendar days (day of year or DOY) or could be converted into DOY (see \nameref{supplement-section:addmethods} in the Supplements).
-\\
-\\
+
 Based on these criteria we found 19 common gardens distributed throughout North America and Europe, with the majority of data concentrated in North America (Fig.
@@
 \citealp{Rehfeldt1994}), we converted such dates to DOY using the 'lubridate' package in R \citep{Grolemund11}.
-\\
 \begin{figure}[!h] 
@@
 Seven species also had fall event information available.
-\\
 <<label=source_spring_models, echo=FALSE>>=
@@
 Results were similar using other distance metrics in lieu of latitude (see Supplement Fig.\ref{supplement-figure:lat_distance} for results using the difference between provenance and garden latitude, and the spherical distance between provenance and garden).
-\\
 % Spring events advanced slightly with provenance latitude in Europe spring events are slightly earlier where provenance MAT is lower (i.e., higher, more northern latitudes), though uncertainty intervals strongly overlapped zero.
@@
 \ref{figure:continental_spptype_effect}A).
-\\
 
@@
 Fall events delayed in warmer locations for both species types, but slightly more so for gymnosperms (\Sexpr{format(round(mean(evergreenspost_fitC_fall_mat),digits = 2),nsmall=2)} days) than angiosperms (\Sexpr{format(round(mean(deciduouspost_fitC_fall_mat),digits = 2),nsmall=2)} days) (Fig.
-\ref{figure:continental_spptype_effect}B).\\
+\ref{figure:continental_spptype_effect}B).
 
@@
 The more similar the climate is between provenances and gardens, the less difference between spring event dates.
-\\
 
@@
 Overall, our results predict that warming springs will continue to be tracked more closely phenologically by trees than warming fall temperatures.
-\\
 

### 2023-05-29 Docs/manuscript.tex
updated intro

@@
 \newpage
+% Journal ideas: Global Change Biology; Functional Ecology or Journal of Ecology next seem good ...
+maybe a biogeography journal (though I am not sure if we can afford one) ...
+maybe New Phytologist but I think they could be picky.
 
-\section*{Abstract}
-
-% We conducted the first cross-continental meta-analysis of published studies from the peer-reviewed literature that reported spring event dates for a mix of angiosperm and gymnosperm tree species in the northern hemisphere, capturing data from 384 North American provenances and 101 European provenances with observations from 1962 to 2019.
+% Just jotting down and putting some bits you have for abstract for now. 
+\begin{abstract}
+Climate change forecasting ....
+plasticity....
+adaptation.
+Spring tree phenology is critical to forecasting ...
+Despite a long history of study, however, the balance of local adaptation versus plasticity in determining environmental responses is not well understood, especially for spring events.
+We conducted the first cross-continental meta-analysis of published studies from the peer-reviewed literature that reported spring event dates for a mix of angiosperm and gymnosperm tree species in the northern hemisphere, capturing data from 384 North American provenances and 101 European provenances with observations from 1962 to 2019.
+We found ...
+This is important because...
+\end{abstract}
 
 \section{Introduction}
-To optimize ecosystem services for habitat conservation and understand how plants cope with environmental variations, predicting the biological impacts of climate change is of critical importance \citep{botero15}.
+%emw: FYI this is a first draft -- meaning it still has lots of words, sentences and other bits to fix; but best to do that as we go. 
+%emw: I noticed a lot of papers on local adaptation don't always use the term 'local adaptation' so we may want to avoid using it too much?
+I suspect it may be contested.
+I also wonder if my advice about definitions was wrong ...
+but let's definitely save that text for now.
+%emw: We need to 'funnel' in our intros -- until we submit to a plant journal focusing on plants too early is getting too narrow too fast.
+
+% To optimize ecosystem services for habitat conservation and understand how plants cope with environmental variations, predicting the biological impacts of climate change is of critical importance \citep{botero15}.
 One of the most important parts of accurate prediction is understanding the relative importance of local adaptation versus plasticity \citep{chevin10}.
-Local adaptation - how well the growth and adaptive characteristics of plants align with the local climate conditions - influences their ability to thrive and reproduce  \citep{casmey22, kawecki04, anderson12, kim13}.
+The increasing need to predict the biological impacts of climate change has made understanding how organisms cope with environmental variation more urgent \citep{botero15}.
+In particular, the relative importance of genetic adaptation versus plasticity is of paramount importance for prediction \citep{chevin10}, with plasticity expected to allow species to shift more rapidly with climate change than environmental responses based on local adaptations \citep{chevin102, snell18}.
+
+% Local adaptation---how well the growth and adaptive characteristics align with the local climate conditions---influences their ability to thrive and reproduce  \citep{casmey22, kawecki04, anderson12, kim13}.
 When the selection intensity exceeds gene flow in species that spread over diverse areas, clinal variations among environmental factors can result in populations with unique adaptations to the local climate and, therefore, greater fitness \citep{kremer12, rossi15, salmela13}.
-On the other hand, phenotypic plasticity describes an individual genotype’s ability to exhibit different phenotypes in response to environmental variations, enabling plants to adjust to new environmental conditions and survive in this rapidly changing world (also known as ‘plastic rescue’)  \citep{chevin10, chevin102, snell18}.
-\\
-\\
-Among the various population-level characteristics critical in synchronizing plant growth with the available growing season, phenology traits hold particular significance \citep{primack80, oneil97, chuine01}.
-Phenology governs the timing of plant transitions between dormancy and active growing, allowing them to time reproduction and utilize each growing season as best as possible \citep{chuine10, hanninen11}.
-Additionally, phenology provides insight into how climate change affects plant life cycles, regeneration capacity, and ecosystem management.\\
-\\
-Spring phenology, specifically the timing of budburst and leaf flush, plays a significant role in determining plant fitness and has been widely adopted as a useful tool to investigate plant adaptation mechanisms \citep{guo22, chuine01}.
-In temperate and boreal environments, the optimal timing for spring growth initiation is determined by a balance of avoiding late spring frost and having a long growing season \citep{alberto11, lenz16, allevato19}.
-Utilizing the early portion of the growing season can be especially critical for species in colder regions \citep{morin07, dantec15} and the ones with suboptimal shade tolerance \citep{richardson09}.
-Plants that budburst earlier can take advantage of a longer growing season and avoid competition  \citep{guo22}.
-\\
-\\
-In recent decades, there has been growing interest in how climate change influences plant phenology \citep{Gill15}.
-Trees exhibit immediate responses via phenotypic plasticity with the warming climate \citep{Nicotra10}, while the degree of adaptation dictated by climate change differs across different geographical areas \citep{Loarie09}.
-Phenological shifts in spring phenology due to recent temperature increases have been extensively observed \citep{Parmesan06}.
-For example, leaf unfolding in Europe advanced up to six days since the early 1960s due to the rise in spring temperature, and spring events have generally shifted more than fall events \citep{menzel99}.
-Studies have found that species that shift more with warming also appear to perform better \citep{Gill15}.
-For instance, research has demonstrated that the timing of budburst and senescence in deciduous plants can impact plant competition \citep{fridley12}, plant growth \citep{myneni97}, and the uptake of carbon by the ecosystem \citep{Barichivich12}.
-As leaf phenology shifts in response to climate, increasing evidence suggests how much a species shifts its phenology with warming may predict fitness consequences: Individuals, populations, and phenological events (spring versus fall) with more plasticity might fare better with climate change.
-Plasticity, however, likely varies across species and phenophases.
-An accurate understanding of how environmental conditions might influence species distribution at a continental scale is critical for future range shift prediction.
-\\
-\\
-Much spring phenology is driven by plasticity.
-For more than 250 years, researchers have been comparing populations of trees from different geographical origins under the same environmental conditions through common garden experiments to disentangle the effects of environmental and genetic variation on trees’ phenology and phenotype \citep{AitkenBemmels16, Alberto13}.
-Over time, comparable clinal variations have been observed in these experiments along latitudinal and elevational gradients for many temperate and boreal species.
-Fall bud set phenology and cold hardiness exhibit strong population differentiation and strong clines along different latitudes, altitudes, and continentality gradients \citep{Alberto13}.
-For example, low-latitude populations have shown significantly stronger delays in leaf senescence over time than high-latitude populations \citep{Gill15}.
-Spring events (budburst, leaf flush, etc.) show more substantial phenotypic plasticity and higher variability \citep{AitkenBemmels16}.\\
-\\
-Despite a growing interest in predicting local adaptation across locations, no study has comprehensively examined clines for spring and fall events and what factors may underlie differences observed across studies.
-We ask: To what extent are patterns of phenotypic plasticity and local adaptation along climate gradients similar across common garden studies?
-To better understand these clines, we also aim to examine the following:
+On the other hand, phenotypic plasticity describes an individual genotype’s ability to exhibit different phenotypes in response to environmental variations, enabling plants to adjust to new environmental conditions and survive in this rapidly changing world (also known as `plastic rescue')  \citep{chevin10, chevin102, snell18}.
+
+Many of the currently observed responses to climate change appear to be mostly plastic \citep{burton22,zettle21,bonamour19, king17}, including the most reported biological response to climate change---shifting phenology.
+Phenology (timing of recurring seasonal events) governs the timing of transitions between dormancy and active growth for many organisms, allowing them to time reproduction and fully exploit the resources of each growing season \citep{chuine10,hanninen11}.
+Many terrestrial animals time their breeding season to coincide with the availability of food resources.
+For example, migratory birds arrive at breeding grounds amidst the peaking of insects \citep{hallfors20}.
+Many butterfly species ensure their larvae have an ample food supply by synchronizing their egg-laying with the phenology of their host plants \citep{rytteri21,posle18}.
+Similarly, the migration and spawning of many marine organisms are closely tied to water temperature and the availability of planktonic food sources \citep{genner10}.
+Such synchronization enables them to provide optimal food for their young during critical growth stages.
+Therefore, climate warming and the associated changes in spring phenology timing, such as earlier leafout or insect emergence due to climate change, can disrupt this synchronization and negatively impact animal populations \citep{wann19,renner18, chu17}.
+As such, phenology plays a significant role in determining fitness \citep{guo22, chuine01}.
+Shifted phenology in recent decades---with many events moving several days per decade \citep{vita21,khar18,Menzel06} ---has thus led to concerns about fitness consequences, and the limits of possible future shifts.
+Future phenological shifts will depend on how much phenology is determined by adaptation versus plasticity will determine, yet our understanding of the balance of these two approaches to variation is limited, even for species groups that are critical to both forecasting and have been well studied, such as trees.
+
+Tree phenology is critically important to climate change forecasting and has been an area of long-term study of adaptation and plasticity \citep{Cleland:2007or}.
+The timing of budburst and senescence can impact plant competition \citep{fridley12}, growth \citep{myneni97}, and carbon sequestration \citep{Barichivich12}---thus affecting climate change itself.
+Hundreds of years of study have established common clines in fall phenology, as locations with shorter growing season (poleward, higher elevation) exhibit earlier stopping of growth (e.g., via budset), and connected these to an underlying proximate mechanism of changing photoperiod cues, driven by local adaptation to the local growing season and relevant photoperiod to stop growth \citep{Alberto13,Savolainen07}.
+In contrast, spring phenology appears far more plastic \citep{AitkenBemmels16} and determined more strongly by temperature \citep{flynn18}.
+Many studies, however, have argued that spring phenology shows important levels of adaptation.
+
+Evidence for local adaptation in spring phenology comes mostly from common garden studies.
+For more than 250 years, researchers have been comparing populations of trees from different geographical origins under the same environmental conditions through these experiments to disentangle the effects of environmental and genetic variation on trees’ phenology and phenotype \citep{AitkenBemmels16, Alberto13}.
+Studies have documented provenance differences of 2.08-4.17 days per degree latitude for some species (\emph{Picea abies} in \citealp{sog08} and \emph{Quercus petraea} in \citealp{deans96}) while others have not found any trends in budburst along latitudinal gradients (For example,\emph{Picea sitchensis} in \citealp{mimura07}, \emph{Picea glauca} in \citealp{Li97}, and \emph{Populus balsamifera} in \citealp{farmer93}).
+This has led to some debate over the prevalence and importance of adaptation in spring tree phenology.
+While clines of spring phenology have been found in studies in both Europe \citep{sog08,deans96,von95} and North America \citep{rossi15, soo13, hannerz99}, they appear potentially more common in European studies.
+
+Continental differences in patterns of adaptation versus plasticity could be driven by climatic differences, as North American springs are more variable across years than European ones \citep{tward21,zohner2017,schwartz00}.
+Such high temporal variability means that distant sites can effectively experience the same spring climate, but in different years.
+Studies of spring phenology in arboreta suggest cues for budburst may vary depending on continental climate \citep{cat23, ospreebbms,ospreephoto}, but are poorly controlled compared to traditional provenance trials, making them difficult to use for tests of plasticity versus adaptation \citep{gauzere2020}.
+Even for more carefully designed common garden differences in study design---especially studied species---may complicate understanding what underlies potential trends across continents, with European studies focusing more on deciduous angiosperms and North American studies focusing often on evergreen gymnosperms.
+To test for evidence of adaptation in spring phenology and what factors may underlie differences observed across studies, we comprehensively examined clines for spring and fall events.
+We test for evidence of adaptation via provenance trends with latitude and climate.
+Further, to better understand these clines, we also examine the following:
 \begin{itemize}
 \item How different are clines between spring and fall phenology? 
-\item How different are clines across Europe and North America?
-\item If differences in clines exist across angiosperm and gymnosperm species?
-\item If climate overlap between the provenance and garden explains the similarities and differences in clines?
+\item How different are clines between Europe and North America?
+\item How different are clines between angiosperm and gymnosperm species?
 \end{itemize}
-To address these questions, we reanalyzed data from the literature on common garden experiments in temperate tree species across Europe and North America.
+To address these questions, we combined a new meta-analysis of all common garden experiments in temperate tree species across Europe and North America reporting spring phenology with Bayesian hierarchical models.
+
+ %REFS
+ % savolainenAREES -- have not read recently but sounds relevant
+ % gauzere2020 -- cite and read
+ % \cite{chang2021} and \cite{chuine2016} -- relevant for discussion of chilling
+ % See also Salk 2020 paper and some papers mentioned above
 
@@
 Seven species also had fall event information available.
-\\
 
@@
 Fall events delayed in warmer locations for both species types, but slightly more so for gymnosperms (6.23 days) than angiosperms (3.69 days) (Fig.
-\ref{figure:continental_spptype_effect}B).\\
+\ref{figure:continental_spptype_effect}B).
 
@@
 Our results show that spring events are highly plastic, and thus may shift with warming, but data on more species and greater information on important factors, such as their geographic location in relation to their origins and elevation, are needed for forecasting.
+
+\begin{enumerate}
+\item Opening: 1-2 paragraph overview of our findings (spring phenology show no real clines, some variation by continent and gymnosperm/angiosperm but hard to fully disentangle).
+Suggests some small clines of spring phenology in some studies, but no consistent effects
+ \item Subheader: Adaptation in spring vs fall phenology: Implications for climate change responses
+ \begin{enumerate}
+ \item Mostly unorganized list of things we may want to say in this section
+ \begin{enumerate}
+ \item Current advances in spring phenology may continue, but delays in fall phenology will depend on speed of adaptation
+ \item If current advances in spring phenology continues it could mean shifts in spring frost etc.
+  \item Spring phenology, specifically the timing of budburst and leaf flush, plays a significant role in determining plant fitness and has been widely adopted as a useful tool to investigate plant adaptation mechanisms \citep{guo22, chuine01}.
+In temperate and boreal environments, the optimal timing for spring growth initiation is determined by a balance of avoiding late spring frost and having a long growing season \citep{alberto11, lenz16, allevato19}.
+Utilizing the early portion of the growing season can be especially critical for species in colder regions \citep{morin07, dantec15} and the ones with suboptimal shade tolerance \citep{richardson09}.
+Plants that budburst earlier can take advantage of a longer growing season and avoid competition  \citep{guo22}.
+ \item Clines in fall phenology are stronger and more consistent while clines in spring phenology are weaker and can go either way  (highlight that some studies and we should include WHICH ones do show clines by XX days).
+ \item Mention some of of the co and counter-gradient studies here
+  \item  An accurate understanding of how environmental conditions might influence species distribution at a continental scale is critical for future range shift prediction.
+ \item Implications of small differences in spring phenology could mean different gene flow etc. 
+  \item the degree of adaptation dictated by climate change differs across different geographical areas \citep{Loarie09}. 
+  \item Need to cite elevation studies and discuss them somewhere here.
+  \item Somewhere mention results from chamber studies and local adaptation (which are hard to interpret given issues with chilling) 
+ \item What else?
+ \end{enumerate}
+  \end{enumerate}
+ \item Subheader: Variation across continents and species groupings
+ \begin{enumerate}
+ \item  Unorganized list of things we may want to say in this section
+ \begin{enumerate}
+ \item No strong evidence of continental differences, could make sense based on long-term climate history perhaps ... 
+\item discuss xylem (ring versus diffuse porous spp) and embolism but might be too in the weeds (see Salk article)
+ \end{enumerate}
+ \end{enumerate}
+\end{enumerate}
 

### 2023-06-04 Docs/manuscript.Rnw
Small edits to intro and some comments/tweaks throughout

@@
 One of the most important parts of accurate prediction is understanding the relative importance of local adaptation versus plasticity \citep{chevin10}.
-The increasing need to predict the biological impacts of climate change has made understanding how organisms cope with environmental variation more urgent \citep{botero15}.
+Predicting the biological impacts of climate change has made understanding how organisms cope with environmental variation more urgent \citep{botero15}.
 In particular, the relative importance of genetic adaptation versus plasticity is of paramount importance for prediction \citep{chevin10}, with plasticity expected to allow species to shift more rapidly with climate change than environmental responses based on local adaptations \citep{chevin102, snell18}.
@@
 
+% emwJun4: I really love the info you added here, but we have to be careful to not focus too deeply on synchrony as it is a very specific topic and our work is broader than that (sorry if I confused you by my request for 'refs' -- I mostly wanted just citations; I suspect we can use some of the text for the discussion so it will not go to waste).
+We also need to keep paragraphs themselves and our intro overall short, so I cut here.
+Thus I slimmed this way down and kept many references, but less of the examples.
 Many of the currently observed responses to climate change appear to be mostly plastic \citep{burton22,zettle21,bonamour19, king17}, including the most reported biological response to climate change---shifting phenology.
-Phenology (timing of recurring seasonal events) governs the timing of transitions between dormancy and active growth for many organisms, allowing them to time reproduction and fully exploit the resources of each growing season \citep{chuine10,hanninen11}.
-Many terrestrial animals time their breeding season to coincide with the availability of food resources.
-For example, migratory birds arrive at breeding grounds amidst the peaking of insects \citep{hallfors20}.
-Many butterfly species ensure their larvae have an ample food supply by synchronizing their egg-laying with the phenology of their host plants \citep{rytteri21,posle18}.
-Similarly, the migration and spawning of many marine organisms are closely tied to water temperature and the availability of planktonic food sources \citep{genner10}.
-Such synchronization enables them to provide optimal food for their young during critical growth stages.
-Therefore, climate warming and the associated changes in spring phenology timing, such as earlier leafout or insect emergence due to climate change, can disrupt this synchronization and negatively impact animal populations \citep{wann19,renner18, chu17}.
-As such, phenology plays a significant role in determining fitness \citep{guo22, chuine01}.
-Shifted phenology in recent decades---with many events moving several days per decade \citep{vita21,khar18,Menzel06} ---has thus led to concerns about fitness consequences, and the limits of possible future shifts.
-Future phenological shifts will depend on how much phenology is determined by adaptation versus plasticity will determine, yet our understanding of the balance of these two approaches to variation is limited, even for species groups that are critical to both forecasting and have been well studied, such as trees.
+Phenology (the timing of recurring seasonal events) governs the timing of transitions between dormancy and active growth for many organisms, allowing them to time reproduction and exploit the resources of each growing season \citep{chuine10,hanninen11,rytteri21,posle18}.
+As such, phenology plays a significant role in determining fitness for both plants \citep{guo22,chuine01} and animals \citep{wann19,renner18,chu17}.
+% emwJun4 -- check last three refs here are still relevant
+
+Shifted phenology in recent decades---with many events moving several days per decade \citep{vita21,khar18,Menzel06}---has led to concerns about fitness consequences, and the limits of possible future shifts.
+Future phenological shifts will depend on how much phenology is determined by adaptation versus plasticity; yet our understanding of the balance of these two approaches to variation is limited, even for species groups that are critical to both forecasting and have been well studied, such as trees.
 
@@
 The timing of budburst and senescence can impact plant competition \citep{fridley12}, growth \citep{myneni97}, and carbon sequestration \citep{Barichivich12}---thus affecting climate change itself.
-Hundreds of years of study have established common clines in fall phenology, as locations with shorter growing season (poleward, higher elevation) exhibit earlier stopping of growth (e.g., via budset), and connected these to an underlying proximate mechanism of changing photoperiod cues, driven by local adaptation to the local growing season and relevant photoperiod to stop growth \citep{Alberto13,Savolainen07}.
+Hundreds of years of study have established common clines in fall phenology, as locations with shorter growing season (poleward and higher elevations) exhibit earlier stopping of growth (e.g., budset).
+Research has connected these clines to an underlying proximate mechanism of changing photoperiod cues (i.e., shifts in the photoperiod threshold that trigger budset), driven by local adaptation to the local growing season \citep{Alberto13,Savolainen07}.
 In contrast, spring phenology appears far more plastic \citep{AitkenBemmels16} and determined more strongly by temperature \citep{flynn18}.
@@
 
+% emwJun4 -- check below where I added 'spring phenology' -- make sure that's accurate (I also removed the decimals)
 Evidence for local adaptation in spring phenology comes mostly from common garden studies.
 For more than 250 years, researchers have been comparing populations of trees from different geographical origins under the same environmental conditions through these experiments to disentangle the effects of environmental and genetic variation on trees’ phenology and phenotype \citep{AitkenBemmels16, Alberto13}.
-Studies have documented provenance differences of 2.08-4.17 days per degree latitude for some species (\emph{Picea abies} in \citealp{sog08} and \emph{Quercus petraea} in \citealp{deans96}) while others have not found any trends in budburst along latitudinal gradients (For example,\emph{Picea sitchensis} in \citealp{mimura07}, \emph{Picea glauca} in \citealp{Li97}, and \emph{Populus balsamifera} in \citealp{farmer93}).
+Studies have documented provenance differences of 2-4 days per degree latitude in spring phenology for some species (\emph{Picea abies} in \citealp{sog08} and \emph{Quercus petraea} in \citealp{deans96}) while others have not found any trends in budburst along latitudinal gradients (For example,\emph{Picea sitchensis} in \citealp{mimura07}, \emph{Picea glauca} in \citealp{Li97}, and \emph{Populus balsamifera} in \citealp{farmer93}).
 This has led to some debate over the prevalence and importance of adaptation in spring tree phenology.
 While clines of spring phenology have been found in studies in both Europe \citep{sog08,deans96,von95} and North America \citep{rossi15, soo13, hannerz99}, they appear potentially more common in European studies.
+% emwJun4 -- we need to justify why we say they are more common in Europe (I know I wrote this!
+But you should change anything I write that seems incorrect to you and let me know) ...
+if you can, please add an explanation and citations with the explanation (or we change 'they appear potentially more common in European studies,' with 'they could vary by continent').
 
@@
 Such high temporal variability means that distant sites can effectively experience the same spring climate, but in different years.
-Studies of spring phenology in arboreta suggest cues for budburst may vary depending on continental climate \citep{cat23, ospreebbms,ospreephoto}, but are poorly controlled compared to traditional provenance trials, making them difficult to use for tests of plasticity versus adaptation \citep{gauzere2020}.
+Studies of spring phenology in arboreta suggest cues for budburst may vary depending on continental climate \citep{zohner2017}, but are poorly controlled compared to traditional provenance trials, making them difficult to use for tests of plasticity versus adaptation \citep{gauzere2020}.
 Even for more carefully designed common garden differences in study design---especially studied species---may complicate understanding what underlies potential trends across continents, with European studies focusing more on deciduous angiosperms and North American studies focusing often on evergreen gymnosperms.
+
 To test for evidence of adaptation in spring phenology and what factors may underlie differences observed across studies, we comprehensively examined clines for spring and fall events.
@@
 \end{itemize}
-To address these questions, we combined a new meta-analysis of all common garden experiments in temperate tree species across Europe and North America reporting spring phenology with Bayesian hierarchical models.
+To address these questions, we combined Bayesian hierarchical models with a new meta-analysis of all common garden experiments in temperate tree species across Europe and North America reporting spring phenology.
 
@@
 For studies that reported event dates relative to a reference date other than 1 January (e.g.
-\citealp{Rehfeldt1994}), we converted such dates to DOY using the 'lubridate' package in R \citep{Grolemund11}.
+\citealp{Rehfeldt1994}), we converted such dates to DOY using the `lubridate' package in R \citep{Grolemund11}.
+% emwJun4 -- you need to use ` to open single quotes in Latex.
 \begin{figure}[!h] 
@@
 The distribution was skewed toward North America (number of studies = 12).
-See Table.\ref{supplement-table:all_studies} in Supplements for information on selected studies.}
+See Table.\ref{supplement-table:all_studies} in Supplements for information on selected studies.}  % emwJun4 -- add to caption how many studies in europe (7?)
     \label{figure:map_gardens}
@@
 We estimated the mean annual temperature (MAT) for each provenance using the Climate Information Tool by Food and Agriculture Organization of the United Nations \citep{FAO2022}and ClimateWNA \citep{wang2016}.
-To examine more explicitly climate near spring events, we used gridded daily temperature data for 2011 to 2020 for all European and North American provenances and gardens from E-OBS and the 'daymetr' package in R respectively \citep{cornes2018,hufkens2018}.
+
+To examine more explicitly climate near spring events, we used gridded daily temperature data for 2011 to 2020 for all European and North American provenances and gardens from E-OBS and the `daymetr' package in R respectively \citep{cornes2018,hufkens2018}.
 Using these data we estimated how much the daily temperatures overlapped between garden and provenance locations, which we call `climate overlap.' For this we used the 'overlap' package in R to calculate the percentage overlap of the daily temperature of each provenance and their corresponding gardens in spring months (March to May) from 2011 to 2020.
 Finally, using the daily temperature data, we also calculated Growing degree days (GDD), a commonly used heat accumulation measure to forecast phenological development in plants (Miller et al., 2001), based on the accumulation of mean daily temperatures ($T_{m}$)from 2011 to 2020 above a baseline of 0$^{\circ}$C, from January 1 until budburst and leaf flush with the following formula: GDD = $\sum$($T_{m}$ - $0^{\circ}$C for $T_{m}$ $\ge$ $0^{\circ}$C; 0 for $T_{m}$ $\le$ $0^{\circ}$C).
+% emwJun4 -- can you clarify in the text how you did GDD ...
+did you calculate it specific per tree or garden or ...
+?
+And what about data before 2011?
+(I know we discussed, but I have forgotten))
 
@@
 %emw2Mar -- It would be good if you have tidy code you can post with the data so folks can fit the models themselves
+% emwJun4 -- do you need help with ADD here ... if so, ping me in the old issue and we can work on it. It should be easy if you have cleared up all divergent transitions!
 
@@
 Spring events shown on top and fall events at the bottom.
-European Gardens and species are bolded and denoted by an asterisk (*).
+European gardens and species are bolded and denoted by an asterisk (*).
 }
@@
 
-In contrast, fall events (budset, leaf senescence, leaf abscission) advanced strongly with provenance latitude (\Sexpr{format(round(abs(mean(all_species_fallpost_fitC_fall_lat)),digits = 2), nsmall=2)} days/degree [\Sexpr{format(round(abs(quantile(all_species_fallpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)}-\Sexpr{format(round(abs(quantile(all_species_fallpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}]) and the decrease in MAT (\Sexpr{format(round(abs(mean(all_species_fallpost_fitC_fall_mat)),digits = 2), nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(all_species_fallpost_fitC_fall_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(all_species_fallpost_fitC_fall_mat, 0.95), digits = 2)}]), meaning fall events WEre earlier where provenance MAT was lower (Fig.
+In contrast, fall events (budset, leaf senescence, leaf abscission) advanced strongly with provenance latitude (\Sexpr{format(round(abs(mean(all_species_fallpost_fitC_fall_lat)),digits = 2), nsmall=2)} days/degree [\Sexpr{format(round(abs(quantile(all_species_fallpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)}-\Sexpr{format(round(abs(quantile(all_species_fallpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}]) and the decrease in MAT (\Sexpr{format(round(abs(mean(all_species_fallpost_fitC_fall_mat)),digits = 2), nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(all_species_fallpost_fitC_fall_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(all_species_fallpost_fitC_fall_mat, 0.95), digits = 2)}]), meaning fall events Were earlier where provenance MAT was lower (Fig.
 \ref{figure:springfall_latmat}, Table.\ref{supplement-table:model_fall_lat} \& \ref{supplement-table:model_fall_mat} in Supplements ).
-This relationship, however, was observed mostly in North America where fall events advanced \Sexpr{format(round(abs(mean(northamericanpost_fitC_fall_lat)),digits = 2),nsmall=2)} (\Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree northward, or \Sexpr{format(round(mean(northamericanpost_fitC_fall_mat),digits = 2),nsmall=2)} days (\Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_mat, 0.95)), digits = 2),nsmall=2)}) per degree decline in MAT ($^{\circ}$C), whereas in Europe these relationships were weaker: advance of \Sexpr{format(round(abs(mean(europeanpost_fitC_fall_lat)),digits = 2),nsmall=2)} (\Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree northward, or \Sexpr{format(round(mean(europeanpost_fitC_fall_mat),digits = 2),nsmall=2)} days (\Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_mat, 0.95)), digits = 2),nsmall=2)}) per degree decline in MAT $^{\circ}$C(Fig.
+This relationship, however, was observed mostly in North America where fall events advanced \Sexpr{format(round(abs(mean(northamericanpost_fitC_fall_lat)),digits = 2),nsmall=2)} (\Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree northward, or \Sexpr{format(round(mean(northamericanpost_fitC_fall_mat),digits = 2),nsmall=2)} days (\Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_mat, 0.95)), digits = 2),nsmall=2)}) per degree decline in MAT ($^{\circ}$C), whereas in Europe these relationships were weaker: advance of \Sexpr{format(round(abs(mean(europeanpost_fitC_fall_lat)),digits = 2),nsmall=2)} (\Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree northward, or \Sexpr{format(round(mean(europeanpost_fitC_fall_mat),digits = 2),nsmall=2)} days (\Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_mat, 0.95)), digits = 2),nsmall=2)}) per degree decline in MAT ($^{\circ}$C) (Fig.
 \ref{figure:continental_spptype_effect}A).
@@
 
+% emwJun4 -- can you reference the figures in the supp that have the latitude results?
 Effects of provenance latitude on both spring and fall events were similar across angiosperms and gymnosperms.
@@
 Once you have the supp so I can see better what we're referring to here in first sentence we should likely edit this paragraph a little.
+% emwJun4 -- give more explanation of what we mean by  'in months that matter for the events' ... is this in the methods? 
 While we believe that coarse metrics such as latitude and MAT would ultimately represent how similar the climates are between the provenances and gardens, we also estimated climate overlap in months that matter for the events to further test how much climate similarity between provenances and gardens predicts local adaptation.
@@
 Suggests some small clines of spring phenology in some studies, but no consistent effects
+% \item Implications of our results on animals: Many terrestrial animals time their breeding season to coincide with the availability of food resources.
+For example, migratory birds arrive at breeding grounds amidst the peaking of insects \citep{hallfors20}.
+Many butterfly species ensure their larvae have an ample food supply by synchronizing their egg-laying with the phenology of their host plants \citep{rytteri21,posle18}.
+Similarly, the migration and spawning of many marine organisms are closely tied to water temperature and the availability of planktonic food sources \citep{genner10}.
+Such synchronization enables them to provide optimal food for their young during critical growth stages.
+Therefore, climate warming and the associated changes in spring phenology timing, such as earlier leafout or insect emergence due to climate change, can disrupt this synchronization and negatively impact animal populations \citep{wann19,renner18, chu17}.
  \item Subheader: Adaptation in spring vs fall phenology: Implications for climate change responses
@@
 \item discuss xylem (ring versus diffuse porous spp) and embolism but might be too in the weeds (see Salk article)
+\item Angiosperm vs gymnosperm
  \end{enumerate}

### 2023-06-13 Docs/manuscript.Rnw
edited intro/discussion

@@
 As such, phenology plays a significant role in determining fitness for both plants \citep{guo22,chuine01} and animals \citep{wann19,renner18,chu17}.
-% emwJun4 -- check last three refs here are still relevant
 
@@
 
-% emwJun4 -- check below where I added 'spring phenology' -- make sure that's accurate (I also removed the decimals)
 Evidence for local adaptation in spring phenology comes mostly from common garden studies.
@@
 This has led to some debate over the prevalence and importance of adaptation in spring tree phenology.
-While clines of spring phenology have been found in studies in both Europe \citep{sog08,deans96,von95} and North America \citep{rossi15, soo13, hannerz99}, they appear potentially more common in European studies.
-% emwJun4 -- we need to justify why we say they are more common in Europe (I know I wrote this!
-But you should change anything I write that seems incorrect to you and let me know) ...
-if you can, please add an explanation and citations with the explanation (or we change 'they appear potentially more common in European studies,' with 'they could vary by continent').
+While clines of spring phenology have been found in studies in both Europe \citep{sog08,deans96,von95} and North America \citep{rossi15, soo13, hannerz99}, they could vary by continent.
 
@@
 To address these questions, we combined Bayesian hierarchical models with a new meta-analysis of all common garden experiments in temperate tree species across Europe and North America reporting spring phenology.
-
- %REFS
- % savolainenAREES -- have not read recently but sounds relevant
- % gauzere2020 -- cite and read
- % \cite{chang2021} and \cite{chuine2016} -- relevant for discussion of chilling
- % See also Salk 2020 paper and some papers mentioned above
 
@@
 \section{Discussion}
-%emw2Mar -- there's a delicate balance of how much rationale you can give for your methods in the methods (and similar for the results). I would move the below to the discussion section. 
-While coarse metrics such as latitude and MAT may represent how similar the climates are between the provenances and gardens in times that matter for the events.
-If climates are very similar, then we would expect similar timings [add more here].
-
-%emw2Mar -- save the below for now; may fit in the discussion or could add if reveiwers complain.
-For now, I added the events above -- it's quicker and IMHO we don't currently need to justify why we pooled these.
-We pooled the timing of budburst and leaf flush into a single category of ‘spring events’ and the timing of bud set, leaf senescence, growth cessation, and leaf abscission into ‘fall events.’ Such pooling is justified because of the shared pressures from natural selection that govern these events \citep{Gill15}.
-
-The weak relationship between spring event dates and provenance latitude and MAT that we find in European studies might be explained by the higher extent of climate overlap in those studies.
-The more similar the climate is between provenances and gardens, the less difference between spring event dates.
-
-The inconsistent and weak clines in spring events that we found suggest high plasticity in spring phenology across continents and species.
-Fall events, on the other hand, exhibit stronger clines which suggest more local adaptation, especially in North America.
-Overall, our results predict that warming springs will continue to be tracked more closely phenologically by trees than warming fall temperatures.
-
-In contrast to spring events, we found strong latitudinal clines in fall events across both continents, with local adaptation appearing much stronger in North America than in Europe.
-Our results show that spring events are highly plastic, and thus may shift with warming, but data on more species and greater information on important factors, such as their geographic location in relation to their origins and elevation, are needed for forecasting.
 
@@
 Suggests some small clines of spring phenology in some studies, but no consistent effects
+% The inconsistent and weak clines in spring events that we found suggest high plasticity in spring phenology across continents and species.
+Fall events, on the other hand, exhibit stronger clines which suggest more local adaptation, especially in North America.
+Overall, our results predict that warming springs will continue to be tracked more closely phenologically by trees than warming fall temperatures.
+The weak relationship between spring event dates and provenance latitude and MAT that we find in European studies might be explained by the higher extent of climate overlap in those studies.
+The more similar the climate is between provenances and gardens, the less difference between spring event dates.
+In contrast to spring events, we found strong latitudinal clines in fall events across both continents, with local adaptation appearing much stronger in North America than in Europe.
+Our results show that spring events are highly plastic, and thus may shift with warming, but data on more species and greater information on important factors, such as their geographic location in relation to their origins and elevation, are needed for forecasting.
+
 % \item Implications of our results on animals: Many terrestrial animals time their breeding season to coincide with the availability of food resources.
@@
  \item What else?
+ % AlinaJune10: Maybe also discuss ice sheet dynamics in Europe/North America, which could impact rates of local adaptation.
  \end{enumerate}
@@
 
-% Must cover in discussion:
+%emw2Mar -- there's a delicate balance of how much rationale you can give for your methods in the methods (and similar for the results). I would move the below to the discussion section. 
+While coarse metrics such as latitude and MAT may represent how similar the climates are between the provenances and gardens in times that matter for the events.
+If climates are very similar, then we would expect similar timings [add more here].
 
-% Counter vs. co gradient literature
-% Maybe also discuss? ice sheet dynamics in Europe/North America, which could impact rates of local adaptation.
+%emw2Mar -- save the below for now; may fit in the discussion or could add if reveiwers complain.
+For now, I added the events above -- it's quicker and IMHO we don't currently need to justify why we pooled these.
+We pooled the timing of budburst and leaf flush into a single category of ‘spring events’ and the timing of bud set, leaf senescence, growth cessation, and leaf abscission into ‘fall events.’ Such pooling is justified because of the shared pressures from natural selection that govern these events \citep{Gill15}.
+
+ %REFS
+ % savolainenAREES -- have not read recently but sounds relevant
+ % gauzere2020 -- cite and read
+ % \cite{chang2021} and \cite{chuine2016} -- relevant for discussion of chilling
+ % See also Salk 2020 paper
 

### 2023-06-13 Docs/manuscript.tex
edited intro/discussion

@@
 
+% emwJun4: I really love the info you added here, but we have to be careful to not focus too deeply on synchrony as it is a very specific topic and our work is broader than that (sorry if I confused you by my request for 'refs' -- I mostly wanted just citations; I suspect we can use some of the text for the discussion so it will not go to waste).
+We also need to keep paragraphs themselves and our intro overall short, so I cut here.
+Thus I slimmed this way down and kept many references, but less of the examples.
 Many of the currently observed responses to climate change appear to be mostly plastic \citep{burton22,zettle21,bonamour19, king17}, including the most reported biological response to climate change---shifting phenology.
-Phenology (timing of recurring seasonal events) governs the timing of transitions between dormancy and active growth for many organisms, allowing them to time reproduction and fully exploit the resources of each growing season \citep{chuine10,hanninen11}.
-Many terrestrial animals time their breeding season to coincide with the availability of food resources.
-For example, migratory birds arrive at breeding grounds amidst the peaking of insects \citep{hallfors20}.
-Many butterfly species ensure their larvae have an ample food supply by synchronizing their egg-laying with the phenology of their host plants \citep{rytteri21,posle18}.
-Similarly, the migration and spawning of many marine organisms are closely tied to water temperature and the availability of planktonic food sources \citep{genner10}.
-Such synchronization enables them to provide optimal food for their young during critical growth stages.
-Therefore, climate warming and the associated changes in spring phenology timing, such as earlier leafout or insect emergence due to climate change, can disrupt this synchronization and negatively impact animal populations \citep{wann19,renner18, chu17}.
-As such, phenology plays a significant role in determining fitness \citep{guo22, chuine01}.
-Shifted phenology in recent decades---with many events moving several days per decade \citep{vita21,khar18,Menzel06} ---has thus led to concerns about fitness consequences, and the limits of possible future shifts.
-Future phenological shifts will depend on how much phenology is determined by adaptation versus plasticity will determine, yet our understanding of the balance of these two approaches to variation is limited, even for species groups that are critical to both forecasting and have been well studied, such as trees.
+Phenology (the timing of recurring seasonal events) governs the timing of transitions between dormancy and active growth for many organisms, allowing them to time reproduction and exploit the resources of each growing season \citep{chuine10,hanninen11,rytteri21,posle18}.
+As such, phenology plays a significant role in determining fitness for both plants \citep{guo22,chuine01} and animals \citep{wann19,renner18,chu17}.
+
+Shifted phenology in recent decades---with many events moving several days per decade \citep{vita21,khar18,Menzel06}---has led to concerns about fitness consequences, and the limits of possible future shifts.
+Future phenological shifts will depend on how much phenology is determined by adaptation versus plasticity; yet our understanding of the balance of these two approaches to variation is limited, even for species groups that are critical to both forecasting and have been well studied, such as trees.
 
@@
 For more than 250 years, researchers have been comparing populations of trees from different geographical origins under the same environmental conditions through these experiments to disentangle the effects of environmental and genetic variation on trees’ phenology and phenotype \citep{AitkenBemmels16, Alberto13}.
-Studies have documented provenance differences of 2.08-4.17 days per degree latitude for some species (\emph{Picea abies} in \citealp{sog08} and \emph{Quercus petraea} in \citealp{deans96}) while others have not found any trends in budburst along latitudinal gradients (For example,\emph{Picea sitchensis} in \citealp{mimura07}, \emph{Picea glauca} in \citealp{Li97}, and \emph{Populus balsamifera} in \citealp{farmer93}).
+Studies have documented provenance differences of 2-4 days per degree latitude in spring phenology for some species (\emph{Picea abies} in \citealp{sog08} and \emph{Quercus petraea} in \citealp{deans96}) while others have not found any trends in budburst along latitudinal gradients (For example,\emph{Picea sitchensis} in \citealp{mimura07}, \emph{Picea glauca} in \citealp{Li97}, and \emph{Populus balsamifera} in \citealp{farmer93}).
 This has led to some debate over the prevalence and importance of adaptation in spring tree phenology.
-While clines of spring phenology have been found in studies in both Europe \citep{sog08,deans96,von95} and North America \citep{rossi15, soo13, hannerz99}, they appear potentially more common in European studies.
+While clines of spring phenology have been found in studies in both Europe \citep{sog08,deans96,von95} and North America \citep{rossi15, soo13, hannerz99}, they could vary by continent.
 
@@
 \end{itemize}
-To address these questions, we combined a new meta-analysis of all common garden experiments in temperate tree species across Europe and North America reporting spring phenology with Bayesian hierarchical models.
-
- %REFS
- % savolainenAREES -- have not read recently but sounds relevant
- % gauzere2020 -- cite and read
- % \cite{chang2021} and \cite{chuine2016} -- relevant for discussion of chilling
- % See also Salk 2020 paper and some papers mentioned above
+To address these questions, we combined Bayesian hierarchical models with a new meta-analysis of all common garden experiments in temperate tree species across Europe and North America reporting spring phenology.
 
@@
 
-In contrast, fall events (budset, leaf senescence, leaf abscission) advanced strongly with provenance latitude (3.16 days/degree [2.87-3.45]) and the decrease in MAT (4.78 days/$^{\circ}$C [4.1 - 5.4]), meaning fall events WEre earlier where provenance MAT was lower (Fig.
+In contrast, fall events (budset, leaf senescence, leaf abscission) advanced strongly with provenance latitude (3.16 days/degree [2.87-3.45]) and the decrease in MAT (4.78 days/$^{\circ}$C [4.1 - 5.4]), meaning fall events Were earlier where provenance MAT was lower (Fig.
 \ref{figure:springfall_latmat}, Table.\ref{supplement-table:model_fall_lat} \& \ref{supplement-table:model_fall_mat} in Supplements ).
-This relationship, however, was observed mostly in North America where fall events advanced 4.24 (3.95 - 4.53) days per degree northward, or 6.41 days (5.78 - 7.04) per degree decline in MAT ($^{\circ}$C), whereas in Europe these relationships were weaker: advance of 0.47 (0.21 - 1.17) days per degree northward, or 0.70 days (1.04 - 2.42) per degree decline in MAT $^{\circ}$C(Fig.
+This relationship, however, was observed mostly in North America where fall events advanced 4.24 (3.95 - 4.53) days per degree northward, or 6.41 days (5.78 - 7.04) per degree decline in MAT ($^{\circ}$C), whereas in Europe these relationships were weaker: advance of 0.47 (0.21 - 1.17) days per degree northward, or 0.70 days (1.04 - 2.42) per degree decline in MAT ($^{\circ}$C) (Fig.
 \ref{figure:continental_spptype_effect}A).
@@
 Suggests some small clines of spring phenology in some studies, but no consistent effects
+% The inconsistent and weak clines in spring events that we found suggest high plasticity in spring phenology across continents and species.
+Fall events, on the other hand, exhibit stronger clines which suggest more local adaptation, especially in North America.
+Overall, our results predict that warming springs will continue to be tracked more closely phenologically by trees than warming fall temperatures.
+The weak relationship between spring event dates and provenance latitude and MAT that we find in European studies might be explained by the higher extent of climate overlap in those studies.
+The more similar the climate is between provenances and gardens, the less difference between spring event dates.
+In contrast to spring events, we found strong latitudinal clines in fall events across both continents, with local adaptation appearing much stronger in North America than in Europe.
+Our results show that spring events are highly plastic, and thus may shift with warming, but data on more species and greater information on important factors, such as their geographic location in relation to their origins and elevation, are needed for forecasting.
+
+% \item Implications of our results on animals: Many terrestrial animals time their breeding season to coincide with the availability of food resources.
+For example, migratory birds arrive at breeding grounds amidst the peaking of insects \citep{hallfors20}.
+Many butterfly species ensure their larvae have an ample food supply by synchronizing their egg-laying with the phenology of their host plants \citep{rytteri21,posle18}.
+Similarly, the migration and spawning of many marine organisms are closely tied to water temperature and the availability of planktonic food sources \citep{genner10}.
+Such synchronization enables them to provide optimal food for their young during critical growth stages.
+Therefore, climate warming and the associated changes in spring phenology timing, such as earlier leafout or insect emergence due to climate change, can disrupt this synchronization and negatively impact animal populations \citep{wann19,renner18, chu17}.
  \item Subheader: Adaptation in spring vs fall phenology: Implications for climate change responses

### 2023-07-07 Docs/manuscript.Rnw
Lots of comments on discussion and some tweakd and edits throughout

@@
 
+%7July2023:
+% Zohner2023 throughout is: 
+% https://www.science.org/doi/10.1126/science.adf5098
+
 \section{Introduction}
@@
 
-Tree phenology is critically important to climate change forecasting and has been an area of long-term study of adaptation and plasticity \citep{Cleland:2007or}.
-The timing of budburst and senescence can impact plant competition \citep{fridley12}, growth \citep{myneni97}, and carbon sequestration \citep{Barichivich12}---thus affecting climate change itself.
-Hundreds of years of study have established common clines in fall phenology, as locations with shorter growing season (poleward and higher elevations) exhibit earlier stopping of growth (e.g., budset).
+Tree phenology is critically important to climate change forecasting at both the community and ecosystem levels.
+The timing of budburst and senescence can impact plant competition, plant invasions and community assembly \citep{fridley12}.
+These effects often occur as phenology effects growth \citep{myneni97}, scaling up to impact ecosystem-level carbon sequestration \citep{Barichivich12}, and thus is integrated into forecasts of climate change itself.
+Growing evidence, however, suggests links between growth and phenology are not as consistent as previously predicted---or currently modelled,
+% cite https://www.nature.com/articles/s41586-022-05092-3
+, with recent work suggesting how much spring versus fall events shift may determine impacts on tree growth. % cite Zohner2023
+Such connections between tree phenology and global forecasts make understanding how plastic---or not--it may be important.
+
+Studies of adaptation versus plasticity in tree phenology have been conducted for centuries \citep{Cleland:2007or}, through common garden studies \citep[i.e., studies---often for forestry purposes---where researchers grow trees of different geographical origins under the same environmental conditions to disentangle the effects of environmental and genetic variation on trees’ phenology and phenotype][]{AitkenBemmels16, Alberto13}.
+This work has established common clines in fall phenology, as source locations with shorter growing season (poleward and higher elevations) exhibit earlier stopping of growth (e.g., budset).
 Research has connected these clines to an underlying proximate mechanism of changing photoperiod cues (i.e., shifts in the photoperiod threshold that trigger budset), driven by local adaptation to the local growing season \citep{Alberto13,Savolainen07}.
@@
 
-Evidence for local adaptation in spring phenology comes mostly from common garden studies.
-For more than 250 years, researchers have been comparing populations of trees from different geographical origins under the same environmental conditions through these experiments to disentangle the effects of environmental and genetic variation on trees’ phenology and phenotype \citep{AitkenBemmels16, Alberto13}.
-Studies have documented provenance differences of 2-4 days per degree latitude in spring phenology for some species (\emph{Picea abies} in \citealp{sog08} and \emph{Quercus petraea} in \citealp{deans96}) while others have not found any trends in budburst along latitudinal gradients (For example,\emph{Picea sitchensis} in \citealp{mimura07}, \emph{Picea glauca} in \citealp{Li97}, and \emph{Populus balsamifera} in \citealp{farmer93}).
+Results, however, have not always been consistent.
+Studies have documented provenance differences of 2-4 days per degree latitude in spring phenology for some species (\emph{Picea abies} in \citealp{sog08} and \emph{Quercus petraea} in \citealp{deans96}) while others have not found any trends in budburst along latitudinal gradients (for example, \emph{Picea sitchensis} in \citealp{mimura07}, \emph{Picea glauca} in \citealp{Li97}, and \emph{Populus balsamifera} in \citealp{farmer93}).
 This has led to some debate over the prevalence and importance of adaptation in spring tree phenology.
@@
 To test for evidence of adaptation in spring phenology and what factors may underlie differences observed across studies, we comprehensively examined clines for spring and fall events.
-We test for evidence of adaptation via provenance trends with latitude and climate.
-Further, to better understand these clines, we also examine the following:
+We test for evidence of adaptation via provenance trends with latitude and climate and examine possible factors that underlie these clines, asking:
 \begin{itemize}
@@
 % emwJun4 -- do you need help with ADD here ... if so, ping me in the old issue and we can work on it. It should be easy if you have cleared up all divergent transitions!
+% We pooled the timing of budburst and leaf flush into a single category of ‘spring events’ and the timing of bud set, leaf senescence, growth cessation, and leaf abscission into ‘fall events.’ Such pooling is justified because of the shared pressures from natural selection that govern these events \citep{Gill15}.
 
@@
 \ref{figure:springfall_latmat}, Table.
-\ref{supplement-table:model_spring_lat} \& \ref{supplement-table:model_spring_mat} in Supplements), nor in North America (latitude: \Sexpr{format(round(mean(northamerican_springpost_fitC_spring_lat),digits = 2), nsmall=2)} days/degree [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.05), digits = 1)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(mean(northamerican_springpost_fitC_spring_mat),digits = 1)} days/$^{\circ}$C [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.95), digits = 2)}]) or Europe (latitude: \Sexpr{format(round(mean(european_springpost_fitC_spring_lat),digits = 2),nsmall=2)} days/degree [\Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT:\Sexpr{format(round(mean(european_springpost_fitC_spring_mat),digits = 2),nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.95), digits = 2)}]).
+\ref{supplement-table:model_spring_lat} \& \ref{supplement-table:model_spring_mat} in Supplements), nor within North America (latitude: \Sexpr{format(round(mean(northamerican_springpost_fitC_spring_lat),digits = 2), nsmall=2)} days/degree [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.05), digits = 1)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(mean(northamerican_springpost_fitC_spring_mat),digits = 1)} days/$^{\circ}$C [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.95), digits = 2)}]) or Europe (latitude: \Sexpr{format(round(mean(european_springpost_fitC_spring_lat),digits = 2),nsmall=2)} days/degree [\Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT:\Sexpr{format(round(mean(european_springpost_fitC_spring_mat),digits = 2),nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.95), digits = 2)}]).
 Results were similar using other distance metrics in lieu of latitude (see Supplement Fig.\ref{supplement-figure:lat_distance} for results using the difference between provenance and garden latitude, and the spherical distance between provenance and garden).
@@
 # "C:/Users/alina/Documents/git/localadaptclim"
-# setwd("C:/Users/alina/Documents/git/localadaptclim/Docs")
+# setwd("C:/Users/alina/Documents/git/localadaptclim/Docs")Aitken and Bammals (2015
 #getwd()
@@
 
-In contrast, fall events (budset, leaf senescence, leaf abscission) advanced strongly with provenance latitude (\Sexpr{format(round(abs(mean(all_species_fallpost_fitC_fall_lat)),digits = 2), nsmall=2)} days/degree [\Sexpr{format(round(abs(quantile(all_species_fallpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)}-\Sexpr{format(round(abs(quantile(all_species_fallpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}]) and the decrease in MAT (\Sexpr{format(round(abs(mean(all_species_fallpost_fitC_fall_mat)),digits = 2), nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(all_species_fallpost_fitC_fall_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(all_species_fallpost_fitC_fall_mat, 0.95), digits = 2)}]), meaning fall events Were earlier where provenance MAT was lower (Fig.
+In contrast, fall events (e.g., budset, leaf senescence, leaf abscission) advanced strongly with provenance latitude (\Sexpr{format(round(abs(mean(all_species_fallpost_fitC_fall_lat)),digits = 2), nsmall=2)} days/degree [\Sexpr{format(round(abs(quantile(all_species_fallpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)}-\Sexpr{format(round(abs(quantile(all_species_fallpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}]) and the decrease in MAT (\Sexpr{format(round(abs(mean(all_species_fallpost_fitC_fall_mat)),digits = 2), nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(all_species_fallpost_fitC_fall_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(all_species_fallpost_fitC_fall_mat, 0.95), digits = 2)}]), meaning fall events were earlier where provenance MAT was lower (Fig.
 \ref{figure:springfall_latmat}, Table.\ref{supplement-table:model_fall_lat} \& \ref{supplement-table:model_fall_mat} in Supplements ).
@@
 \ref{figure:continental_spptype_effect}A).
+
+% Note to us, see also https://github.com/lizzieinvancouver/localadaptclim/issues/25
+%7July2023: Add to this please! Add refs to figure and add the correct bibtex refs.
+ \item Clines in fall phenology were stronger and more consistent whereas clines in spring phenology are weaker and somewhat varied in directionality.
+For fall events, only two field studies found no relationship (and ref Fig 2c-d; CITE Rosique-Esplugas 2021, Fraxinus excelsior from garden LETTER, which was in the UK and Petkova et al., 2017 Fagus sylvatica from garden LETTER, which was in Bulgaria, ), while another study was the only greenhouse experiment included (Rehfeldt 1994, also included Aitken and Bammals (2015)), which uniquely used the fall event of `the day by which seedling elongation had finished.'  Spring event clines were always weak, ...
+ %7July2023: Do any of the spring event clines for individual species x garden actually exclude 0 in their 90% interval? If not, add here:
+ % (all species x garden clines included 0 in their 90\% intervals).
+ % If so do not then add:
+% ... were weak. Only XX and YY showed clines (give values with intervals) whereas the rest did not (all remaining species x garden clines included 0 in their 90\% intervals).
 
@@
 % emwJun4 -- can you reference the figures in the supp that have the latitude results?
+%7July2023 -- supp figures? Please add ref to them. 
 Effects of provenance latitude on both spring and fall events were similar across angiosperms and gymnosperms.
-Spring events slightly delayed \Sexpr{format(round(abs(mean(deciduous_springpost_fitC_spring_lat)),digits = 2), nsmall=2)} (\Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_lat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_lat, 0.95)), digits = 2),nsmall=2)}) days per degree north in angiosperms and \Sexpr{format(round(abs(mean(evergreens_springpost_fitC_spring_lat)),digits = 2), nsmall=2)} (\Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_lat, 0.05)), digits = 2),nsmall=2)}) days per degree north in gymnosperms.
+Spring events slightly delayed, at a rate of \Sexpr{format(round(abs(mean(deciduous_springpost_fitC_spring_lat)),digits = 2), nsmall=2)} (\Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_lat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_lat, 0.95)), digits = 2),nsmall=2)}) days per degree north in angiosperms and \Sexpr{format(round(abs(mean(evergreens_springpost_fitC_spring_lat)),digits = 2), nsmall=2)} (\Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_lat, 0.05)), digits = 2),nsmall=2)}) days per degree north in gymnosperms.
 Fall events advanced \Sexpr{format(round(abs(mean(deciduouspost_fitC_fall_lat)),digits = 2), nsmall=2)} (\Sexpr{format(round(abs(quantile(deciduouspost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(deciduouspost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree north in angiosperms and \Sexpr{format(round(abs(mean(evergreenspost_fitC_fall_lat)),digits = 2), nsmall=2)} (\Sexpr{format(round(abs(quantile(evergreenspost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)}-\Sexpr{format(round(abs(quantile(evergreenspost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree north in gymnosperms.
@@
 % emwJun4 -- give more explanation of what we mean by  'in months that matter for the events' ... is this in the methods? 
-While we believe that coarse metrics such as latitude and MAT would ultimately represent how similar the climates are between the provenances and gardens, we also estimated climate overlap in months that matter for the events to further test how much climate similarity between provenances and gardens predicts local adaptation.
+While we expected that coarse metrics, such as latitude and MAT, would generally represent how similar the climates are between the provenances and gardens, we also estimated climate overlap in months much closer to the events to further test how much climate similarity between provenances and gardens predicts local adaptation.
 For spring events, we considered overlap across March to May.
@@
 
+Opening (no subheader) 
 \begin{enumerate}
-\item Opening: 1-2 paragraph overview of our findings (spring phenology show no real clines, some variation by continent and gymnosperm/angiosperm but hard to fully disentangle).
-Suggests some small clines of spring phenology in some studies, but no consistent effects
-% The inconsistent and weak clines in spring events that we found suggest high plasticity in spring phenology across continents and species.
-Fall events, on the other hand, exhibit stronger clines which suggest more local adaptation, especially in North America.
-Overall, our results predict that warming springs will continue to be tracked more closely phenologically by trees than warming fall temperatures.
-The weak relationship between spring event dates and provenance latitude and MAT that we find in European studies might be explained by the higher extent of climate overlap in those studies.
+\item Overall, our results show inconsistent and weak clines in spring events across continents, while fall events generally showed much stronger clines, especially in North America
+\begin{enumerate}
+\item For spring phenology there were weak diverging effects by species groupings -- which may be important, and suggest different trends that mute an overall response. 
+\item Our results suggest that spring events are highly plastic, and thus predict that warming springs will continue to be tracked more closely phenologically by trees than warming fall temperatures.
+%7July2023 -- careful when talking about plasticity and adaptation; our results can be suggestive but do not test either of these things directly
+\end{enumerate}
+\item  These interesting trends by species grouping for spring phenology and continent for fall phenology, suggests data on more species and greater information on important factors, such as their geographic location in relation to their origins and elevation, would aid forecasting.
+\begin{enumerate}
+\item We found MAT and latitude still good predictors, better than climate overlap or other metrics. 
+\item While coarse metrics such as latitude and MAT appear to represent how similar the climates are between the provenances and gardens in times that matter for the events, at least in the narrow range of northern temperature and boreal forests we studied
+\end{enumerate}
+\end{enumerate}
+%7July2023 -- I don't understand the below: where do we show these continent effects for spring phenology?
+% The weak relationship between spring event dates and provenance latitude and MAT that we find in European studies might be explained by the higher extent of climate overlap in those studies.
 The more similar the climate is between provenances and gardens, the less difference between spring event dates.
-In contrast to spring events, we found strong latitudinal clines in fall events across both continents, with local adaptation appearing much stronger in North America than in Europe.
-Our results show that spring events are highly plastic, and thus may shift with warming, but data on more species and greater information on important factors, such as their geographic location in relation to their origins and elevation, are needed for forecasting.
+
+Subheader: Variation across continents and species groupings
+\begin{enumerate}
+\item Species differences
+\begin{enumerate}
+\item Angiosperm vs gymnosperm showed some effects, which makes sense as timing of leafout here is very different (gymnosperem photosynthesize without leafout and thus generally leafout later than angiosperms, after much risk of spring frost)
+\item Suggests there may be additional differences if we have more species and more information on leafout timing and frost risk approach ... 
+\item discuss xylem (ring versus diffuse porous spp) and embolism  (see Salk article)
+\item Maybe also here: Utilizing the early portion of the growing season can be especially critical for species in colder regions \citep{morin07, dantec15} and the ones with suboptimal shade tolerance \citep{richardson09}.
+\end{enumerate}
+\item Continental differences
+\begin{enumerate}
+\item No strong evidence of continental differences, could make sense based on long-term climate history perhaps ... 
+\item Maybe also discuss ice sheet dynamics in Europe/North America, which could impact rates of adaptation.
+\end{enumerate}
+\end{enumerate}
+
+Subheader: Adaptation in spring vs fall phenology: Implications for climate change responses
+\begin{enumerate}
+\item Our results suggest that current advances in spring phenology may continue, but delays in fall phenology could depend on speed of adaptation
+\begin{enumerate}
+\item Spring events appear highly plastic from these studies, which are good tests. 
+\item Some previous chamber studies finding potential adaptation (cited in Ettinger et al.
+2020 Nature Climate Change, I believe -- check the paper and supp) but these could be influenced by chilling at the provenance site, whereas gardens do not have this effect
+\item Previous work has suggested strong clines in fall phenology and our findings back this up... 
+\item These findings have important implications for growing discussion of how shifting phenology and tree growth relate as our results suggest there are strong limits on shifts in fall phenology (Zohner2023 and other papers such as Zani 2020 -- need a couple sentences here) % D.
+Zani, T.
+W.
+Crowther, L.
+Mo, S.
+S.
+Renner, C.
+M.
+Zohner, Increased growing-season productivity drives earlier autumn leaf senescence in temperate trees.
+Science 370, 1066–1071 (2020).
+\end{enumerate}
+\item These trends likely vary by species and depend on frost tolerance strategy
+\begin{enumerate}
+\item If current advances in spring phenology continues it could mean shifts in spring frost etc., but differences across species suggest this will not impact all species the same. 
+\item In temperate and boreal environments, the optimal timing for spring growth initiation is determined by a balance of avoiding late spring frost and having a long growing season \citep{alberto11, lenz16, allevato19} ...
+further Plants that budburst earlier can take advantage of a longer growing season and avoid competition  \citep{guo22}....
+and species vary in how the balance this tradeoff.
+\end{enumerate}
+\item Species may also vary in whether we predict co- or counter-gradient relationships with spring phenology
+\begin{enumerate}
+\item Fall events were always co-gradient
+\item Spring events were more varied -- some appeared to be co or counter ... 
+\item Add some discussion of what this might mean (one idea is that spring leafout sets up flowering and thus some counter-gradient could aid gene flow over larger areas?
+Other ideas you have?) and cite some of the relevant papers.
+\end{enumerate}
+\item All of this critical to predicting not just phenological change, but range expansions with climate change: An accurate understanding of how environmental conditions might influence species distribution at a continental scale is critical for future range shift prediction.
+\begin{enumerate}
+\item Implications of small differences in spring phenology could mean different gene flow etc. 
+\item the degree of adaptation dictated by climate change also differs across different geographical areas \citep{Loarie09}. 
+\item Mention elevational results here? 
+\end{enumerate}
+\end{enumerate}
 
@@
 Therefore, climate warming and the associated changes in spring phenology timing, such as earlier leafout or insect emergence due to climate change, can disrupt this synchronization and negatively impact animal populations \citep{wann19,renner18, chu17}.
- \item Subheader: Adaptation in spring vs fall phenology: Implications for climate change responses
- \begin{enumerate}
- \item Mostly unorganized list of things we may want to say in this section
- \begin{enumerate}
- \item Current advances in spring phenology may continue, but delays in fall phenology will depend on speed of adaptation
- \item If current advances in spring phenology continues it could mean shifts in spring frost etc.
-  \item Spring phenology, specifically the timing of budburst and leaf flush, plays a significant role in determining plant fitness and has been widely adopted as a useful tool to investigate plant adaptation mechanisms \citep{guo22, chuine01}.
-In temperate and boreal environments, the optimal timing for spring growth initiation is determined by a balance of avoiding late spring frost and having a long growing season \citep{alberto11, lenz16, allevato19}.
-Utilizing the early portion of the growing season can be especially critical for species in colder regions \citep{morin07, dantec15} and the ones with suboptimal shade tolerance \citep{richardson09}.
-Plants that budburst earlier can take advantage of a longer growing season and avoid competition  \citep{guo22}.
- \item Clines in fall phenology are stronger and more consistent while clines in spring phenology are weaker and can go either way  (highlight that some studies and we should include WHICH ones do show clines by XX days).
- \item Mention some of of the co and counter-gradient studies here
-  \item  An accurate understanding of how environmental conditions might influence species distribution at a continental scale is critical for future range shift prediction.
- \item Implications of small differences in spring phenology could mean different gene flow etc. 
-  \item the degree of adaptation dictated by climate change differs across different geographical areas \citep{Loarie09}. 
-  \item Need to cite elevation studies and discuss them somewhere here.
-  \item Somewhere mention results from chamber studies and local adaptation (which are hard to interpret given issues with chilling) 
- \item What else?
- % AlinaJune10: Maybe also discuss ice sheet dynamics in Europe/North America, which could impact rates of local adaptation.
- \end{enumerate}
-  \end{enumerate}
- \item Subheader: Variation across continents and species groupings
- \begin{enumerate}
- \item  Unorganized list of things we may want to say in this section
- \begin{enumerate}
- \item No strong evidence of continental differences, could make sense based on long-term climate history perhaps ... 
-\item discuss xylem (ring versus diffuse porous spp) and embolism but might be too in the weeds (see Salk article)
-\item Angiosperm vs gymnosperm
- \end{enumerate}
- \end{enumerate}
-\end{enumerate}
-
-%emw2Mar -- there's a delicate balance of how much rationale you can give for your methods in the methods (and similar for the results). I would move the below to the discussion section. 
-While coarse metrics such as latitude and MAT may represent how similar the climates are between the provenances and gardens in times that matter for the events.
-If climates are very similar, then we would expect similar timings [add more here].
-
-%emw2Mar -- save the below for now; may fit in the discussion or could add if reveiwers complain.
-For now, I added the events above -- it's quicker and IMHO we don't currently need to justify why we pooled these.
-We pooled the timing of budburst and leaf flush into a single category of ‘spring events’ and the timing of bud set, leaf senescence, growth cessation, and leaf abscission into ‘fall events.’ Such pooling is justified because of the shared pressures from natural selection that govern these events \citep{Gill15}.
 

### 2023-07-20 Docs/manuscript.Rnw
addressed comments

@@
 \pagestyle{fancy}
-\fancyhead[LO]{March 2023}
+\fancyhead[LO]{July 2023}
 \fancyhead[RO]{Manuscript}
@@
     \caption{Distribution of common gardens (denoted as stars) and provenances (denoted as circles) included in this meta-analysis.
-The distribution was skewed toward North America (number of studies = 12).
-See Table.\ref{supplement-table:all_studies} in Supplements for information on selected studies.}  % emwJun4 -- add to caption how many studies in europe (7?)
+The distribution was skewed toward North America (12 North American studies versus 5 European studies).
+See Table.\ref{supplement-table:all_studies} in Supplements for information on selected studies.}
     \label{figure:map_gardens}
@@
 Using these data we estimated how much the daily temperatures overlapped between garden and provenance locations, which we call `climate overlap.' For this we used the 'overlap' package in R to calculate the percentage overlap of the daily temperature of each provenance and their corresponding gardens in spring months (March to May) from 2011 to 2020.
-Finally, using the daily temperature data, we also calculated Growing degree days (GDD), a commonly used heat accumulation measure to forecast phenological development in plants (Miller et al., 2001), based on the accumulation of mean daily temperatures ($T_{m}$)from 2011 to 2020 above a baseline of 0$^{\circ}$C, from January 1 until budburst and leaf flush with the following formula: GDD = $\sum$($T_{m}$ - $0^{\circ}$C for $T_{m}$ $\ge$ $0^{\circ}$C; 0 for $T_{m}$ $\le$ $0^{\circ}$C).
-% emwJun4 -- can you clarify in the text how you did GDD ...
-did you calculate it specific per tree or garden or ...
-?
-And what about data before 2011?
-(I know we discussed, but I have forgotten))
+Finally, using the daily temperature data, we also calculated growing degree days (GDD), a commonly used heat accumulation measure to forecast phenological development in plants \citep{miller01}, for each provenance and garden, based on the accumulation of mean daily temperatures ($T_{m}$)from 2011 to 2020 above a baseline of 0$^{\circ}$C, from January 1 until budburst and leaf flush with the following formula: GDD = $\sum$($T_{m}$ - $0^{\circ}$C for $T_{m}$ $\ge$ $0^{\circ}$C; 0 for $T_{m}$ $\le$ $0^{\circ}$C).
 
@@
 We present estimates as mean $\pm$ 90\% uncertainty intervals given parenthetically, unless otherwise stated.
-%emw2Mar -- It would be good if you have tidy code you can post with the data so folks can fit the models themselves
-% emwJun4 -- do you need help with ADD here ... if so, ping me in the old issue and we can work on it. It should be easy if you have cleared up all divergent transitions!
+
+% alinaJuly20 I need help with ADD here (ping'd you in the old issue. I have removed all divergent transitions already)
+
 % We pooled the timing of budburst and leaf flush into a single category of ‘spring events’ and the timing of bud set, leaf senescence, growth cessation, and leaf abscission into ‘fall events.’ Such pooling is justified because of the shared pressures from natural selection that govern these events \citep{Gill15}.
@@
 \ref{figure:springfall_latmat}, Table.
-\ref{supplement-table:model_spring_lat} \& \ref{supplement-table:model_spring_mat} in Supplements), nor within North America (latitude: \Sexpr{format(round(mean(northamerican_springpost_fitC_spring_lat),digits = 2), nsmall=2)} days/degree [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.05), digits = 1)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(mean(northamerican_springpost_fitC_spring_mat),digits = 1)} days/$^{\circ}$C [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.95), digits = 2)}]) or Europe (latitude: \Sexpr{format(round(mean(european_springpost_fitC_spring_lat),digits = 2),nsmall=2)} days/degree [\Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT:\Sexpr{format(round(mean(european_springpost_fitC_spring_mat),digits = 2),nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.95), digits = 2)}]).
+\ref{supplement-table:model_spring_lat} \& \ref{supplement-table:model_spring_mat} in Supplements), nor within North America (latitude: \Sexpr{format(round(mean(northamerican_springpost_fitC_spring_lat),digits = 2), nsmall=2)} days/degree [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.05), digits = 1)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(mean(northamerican_springpost_fitC_spring_mat),digits = 1)} days/$^{\circ}$C [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.95), digits = 2)}]) or Europe (latitude: \Sexpr{format(round(mean(european_springpost_fitC_spring_lat),digits = 2),nsmall=2)} days/degree [\Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT:\Sexpr{format(round(mean(european_springpost_fitC_spring_mat),digits = 2),nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.95), digits = 2)}]) (Fig.\ref{figure:continental_spptype_lat}A \& \ref{figure:continental_spptype_effect}A).
 Results were similar using other distance metrics in lieu of latitude (see Supplement Fig.\ref{supplement-figure:lat_distance} for results using the difference between provenance and garden latitude, and the spherical distance between provenance and garden).
@@
 \ref{figure:springfall_latmat}, Table.\ref{supplement-table:model_fall_lat} \& \ref{supplement-table:model_fall_mat} in Supplements ).
-This relationship, however, was observed mostly in North America where fall events advanced \Sexpr{format(round(abs(mean(northamericanpost_fitC_fall_lat)),digits = 2),nsmall=2)} (\Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree northward, or \Sexpr{format(round(mean(northamericanpost_fitC_fall_mat),digits = 2),nsmall=2)} days (\Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_mat, 0.95)), digits = 2),nsmall=2)}) per degree decline in MAT ($^{\circ}$C), whereas in Europe these relationships were weaker: advance of \Sexpr{format(round(abs(mean(europeanpost_fitC_fall_lat)),digits = 2),nsmall=2)} (\Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree northward, or \Sexpr{format(round(mean(europeanpost_fitC_fall_mat),digits = 2),nsmall=2)} days (\Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_mat, 0.95)), digits = 2),nsmall=2)}) per degree decline in MAT ($^{\circ}$C) (Fig.
+This relationship, however, was observed mostly in North America where fall events advanced \Sexpr{format(round(abs(mean(northamericanpost_fitC_fall_lat)),digits = 2),nsmall=2)} (\Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree northward, or \Sexpr{format(round(mean(northamericanpost_fitC_fall_mat),digits = 2),nsmall=2)} days (\Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_mat, 0.95)), digits = 2),nsmall=2)}) per degree decline in MAT ($^{\circ}$C), whereas in Europe such relationship was weaker: advance of \Sexpr{format(round(abs(mean(europeanpost_fitC_fall_lat)),digits = 2),nsmall=2)} (\Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree northward, or \Sexpr{format(round(mean(europeanpost_fitC_fall_mat),digits = 2),nsmall=2)} days (\Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_mat, 0.95)), digits = 2),nsmall=2)}) per degree decline in MAT ($^{\circ}$C) (Fig.
 \ref{figure:continental_spptype_effect}A).
@@
 % Note to us, see also https://github.com/lizzieinvancouver/localadaptclim/issues/25
-%7July2023: Add to this please! Add refs to figure and add the correct bibtex refs.
- \item Clines in fall phenology were stronger and more consistent whereas clines in spring phenology are weaker and somewhat varied in directionality.
-For fall events, only two field studies found no relationship (and ref Fig 2c-d; CITE Rosique-Esplugas 2021, Fraxinus excelsior from garden LETTER, which was in the UK and Petkova et al., 2017 Fagus sylvatica from garden LETTER, which was in Bulgaria, ), while another study was the only greenhouse experiment included (Rehfeldt 1994, also included Aitken and Bammals (2015)), which uniquely used the fall event of `the day by which seedling elongation had finished.'  Spring event clines were always weak, ...
- %7July2023: Do any of the spring event clines for individual species x garden actually exclude 0 in their 90% interval? If not, add here:
- % (all species x garden clines included 0 in their 90\% intervals).
- % If so do not then add:
-% ... were weak. Only XX and YY showed clines (give values with intervals) whereas the rest did not (all remaining species x garden clines included 0 in their 90\% intervals).
+
+Clines in fall phenology were stronger and more consistent whereas clines in spring phenology are weaker and somewhat varied in directionality.
+For fall events, only two field studies found no relationship (Fig.
+\ref{figure:springfall_latmat}): \emph{Fraxinus excelsior} from Garden Q* in the UK \citep{rosique22} and \emph{Fagus sylvatica} from Garden R* in Bulgaria \citep{petkova17}.
+Another study that found no relationship was the only greenhouse experiment included (\emph{Picea engelmannii} from Garden B in the USA, also included in \citealp{AitkenBemmels16}), which uniquely used the fall event of `the day by which seedling elongation had finished' \citep{rehfeldt94}.
+Spring event clines were always weak: all species x garden clines included 0 in their 90\% intervals.
+
+\begin{figure}[!h] 
+    \centering
+ \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/continental_spptype_lat.jpg}
+    \caption{Posterior distributions for the effect of provenance latitude across different continents and species types.
+Zero -- no effect -- is shown with a dashed line.
+Solid line and shading in posteriors represent mean and 50 percent interval.
+(A) Effect of latitude and continent on spring and fall event date (DOY).
+Fall events advanced strongly per degree northward, particularly notably in North America.
+(B) Effect of latitude and leaf type on spring and fall event date (DOY).
+As we move north, spring events slightly advanced in gymnosperms and delayed in angiosperms.
+Fall events advanced per degree northward for both species types.}
+    \label{figure:continental_spptype_lat}
+\end{figure}
 
@@
 Fall events advanced strongly with decreasing MAT, particularly notably in North America.
-Effect of MAT and leaf type on spring and fall event date (DOY) weakly diverged.
 (B) Effect of MAT and leaf type on spring and fall event date (DOY).
@@
 
-% emwJun4 -- can you reference the figures in the supp that have the latitude results?
-%7July2023 -- supp figures? Please add ref to them. 
-Effects of provenance latitude on both spring and fall events were similar across angiosperms and gymnosperms.
-Spring events slightly delayed, at a rate of \Sexpr{format(round(abs(mean(deciduous_springpost_fitC_spring_lat)),digits = 2), nsmall=2)} (\Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_lat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_lat, 0.95)), digits = 2),nsmall=2)}) days per degree north in angiosperms and \Sexpr{format(round(abs(mean(evergreens_springpost_fitC_spring_lat)),digits = 2), nsmall=2)} (\Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_lat, 0.05)), digits = 2),nsmall=2)}) days per degree north in gymnosperms.
+Effects of provenance latitude on both spring and fall events were similar across angiosperms and gymnosperms (Fig.\ref{figure:continental_spptype_lat}B).
+Spring events weakly diverged, delaying at a rate of \Sexpr{format(round(abs(mean(deciduous_springpost_fitC_spring_lat)),digits = 2), nsmall=2)} (\Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_lat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_lat, 0.95)), digits = 2),nsmall=2)}) days per degree north in angiosperms and advancing \Sexpr{format(round(abs(mean(evergreens_springpost_fitC_spring_lat)),digits = 2), nsmall=2)} (\Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_lat, 0.05)), digits = 2),nsmall=2)}) days per degree north in gymnosperms.
 Fall events advanced \Sexpr{format(round(abs(mean(deciduouspost_fitC_fall_lat)),digits = 2), nsmall=2)} (\Sexpr{format(round(abs(quantile(deciduouspost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(deciduouspost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree north in angiosperms and \Sexpr{format(round(abs(mean(evergreenspost_fitC_fall_lat)),digits = 2), nsmall=2)} (\Sexpr{format(round(abs(quantile(evergreenspost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)}-\Sexpr{format(round(abs(quantile(evergreenspost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree north in gymnosperms.
-However, effects of MAT on spring events weakly diverged (Fig.\ref{figure:continental_spptype_effect}B).
+Effects of MAT on spring events also weakly diverged (Fig.\ref{figure:continental_spptype_effect}B).
 Spring events advanced \Sexpr{format(round(abs(mean(deciduous_springpost_fitC_spring_mat)),digits = 2), nsmall=2)} (\Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_mat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_mat, 0.05)), digits = 2),nsmall=2)}) days/$^{\circ}$C as MAT increased in angiosperms and delayed \Sexpr{format(round(abs(mean(evergreens_springpost_fitC_spring_mat)),digits = 2), nsmall=2)} (\Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_mat, 0.95)), digits = 2),nsmall=2)}) days/$^{\circ}$C as MAT increased in gymnosperms.
@@
 
-%EMWApr20 -- need to refer to some supp figures or tables or such below.
-Once you have the supp so I can see better what we're referring to here in first sentence we should likely edit this paragraph a little.
-% emwJun4 -- give more explanation of what we mean by  'in months that matter for the events' ... is this in the methods? 
 While we expected that coarse metrics, such as latitude and MAT, would generally represent how similar the climates are between the provenances and gardens, we also estimated climate overlap in months much closer to the events to further test how much climate similarity between provenances and gardens predicts local adaptation.
@@
 \item We found MAT and latitude still good predictors, better than climate overlap or other metrics. 
-\item While coarse metrics such as latitude and MAT appear to represent how similar the climates are between the provenances and gardens in times that matter for the events, at least in the narrow range of northern temperature and boreal forests we studied
+\item While coarse metrics such as latitude and MAT appear to represent how similar the climates are between the provenances and gardens in months closer to the events, at least in the narrow range of northern temperature and boreal forests we studied
 \end{enumerate}
 \end{enumerate}
-%7July2023 -- I don't understand the below: where do we show these continent effects for spring phenology?
-% The weak relationship between spring event dates and provenance latitude and MAT that we find in European studies might be explained by the higher extent of climate overlap in those studies.
-The more similar the climate is between provenances and gardens, the less difference between spring event dates.
 

### 2023-07-20 Docs/manuscript.tex
addressed comments

@@
 \ref{figure:springfall_latmat}, Table.
-\ref{supplement-table:model_spring_lat} \& \ref{supplement-table:model_spring_mat} in Supplements), nor in North America (latitude: 0.10 days/degree [-0.06 - 0.26]; MAT: -0.09 days/$^{\circ}$C [-0.36 - 0.18]) or Europe (latitude: 0.10 days/degree [-0.23 - 0.42]; MAT:-0.16 days/$^{\circ}$C [-0.55 - 0.23]).
+\ref{supplement-table:model_spring_lat} \& \ref{supplement-table:model_spring_mat} in Supplements), nor within North America (latitude: 0.10 days/degree [-0.06 - 0.26]; MAT: -0.09 days/$^{\circ}$C [-0.36 - 0.18]) or Europe (latitude: 0.10 days/degree [-0.23 - 0.42]; MAT:-0.16 days/$^{\circ}$C [-0.55 - 0.23]) (Fig.\ref{figure:continental_spptype_lat}A \& \ref{figure:continental_spptype_effect}A).
 Results were similar using other distance metrics in lieu of latitude (see Supplement Fig.\ref{supplement-figure:lat_distance} for results using the difference between provenance and garden latitude, and the spherical distance between provenance and garden).
@@
 
-In contrast, fall events (budset, leaf senescence, leaf abscission) advanced strongly with provenance latitude (3.16 days/degree [2.87-3.45]) and the decrease in MAT (4.78 days/$^{\circ}$C [4.1 - 5.4]), meaning fall events Were earlier where provenance MAT was lower (Fig.
+In contrast, fall events (e.g., budset, leaf senescence, leaf abscission) advanced strongly with provenance latitude (3.16 days/degree [2.87-3.45]) and the decrease in MAT (4.78 days/$^{\circ}$C [4.1 - 5.4]), meaning fall events were earlier where provenance MAT was lower (Fig.
 \ref{figure:springfall_latmat}, Table.\ref{supplement-table:model_fall_lat} \& \ref{supplement-table:model_fall_mat} in Supplements ).
@@
 \ref{figure:continental_spptype_effect}A).
+
+% Note to us, see also https://github.com/lizzieinvancouver/localadaptclim/issues/25
+
+Clines in fall phenology were stronger and more consistent whereas clines in spring phenology are weaker and somewhat varied in directionality.
+For fall events, only two field studies found no relationship (Fig.
+\ref{figure:springfall_latmat}): \emph{Fraxinus excelsior} from Garden Q* in the UK \citep{rosique22} and \emph{Fagus sylvatica} from Garden R* in Bulgaria \citep{petkova17}.
+Another study that found no relationship was the only greenhouse experiment included (\emph{Picea engelmannii} from Garden B in the USA, also included in \citealp{AitkenBemmels16}), which uniquely used the fall event of `the day by which seedling elongation had finished' \citep{rehfeldt94}.
+Spring event clines were always weak: all species x garden clines included 0 in their 90\% intervals.
+
+\begin{figure}[!h] 
+    \centering
+ \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/continental_spptype_lat.jpg}
+    \caption{Posterior distributions for the effect of provenance latitude across different continents and species types.
+Zero -- no effect -- is shown with a dashed line.
+Solid line and shading in posteriors represent mean and 50 percent interval.
+(A) Effect of latitude and continent on spring and fall event date (DOY).
+Fall events advanced strongly per degree northward, particularly notably in North America.
+(B) Effect of latitude and leaf type on spring and fall event date (DOY).
+As we move north, spring events slightly advanced in gymnosperms and delayed in angiosperms.
+Fall events advanced per degree northward for both species types.}
+    \label{figure:continental_spptype_lat}
+\end{figure}
 
@@
 % emwJun4 -- can you reference the figures in the supp that have the latitude results?
-Effects of provenance latitude on both spring and fall events were similar across angiosperms and gymnosperms.
-Spring events slightly delayed 0.37 (0.15 - 0.59) days per degree north in angiosperms and 0.23 (0.00 - 0.46) days per degree north in gymnosperms.
+
+Effects of provenance latitude on both spring and fall events were similar across angiosperms and gymnosperms ().
+Spring events weakly diverged, delaying at a rate of 0.37 (0.15 - 0.59) days per degree north in angiosperms and advancing 0.23 (0.00 - 0.46) days per degree north in gymnosperms.
 Fall events advanced 3.18 (2.76 - 3.62) days per degree north in angiosperms and 3.14 (2.81-3.47) days per degree north in gymnosperms.
-However, effects of MAT on spring events weakly diverged (Fig.\ref{figure:continental_spptype_effect}B).
+Effects of MAT on spring events also weakly diverged (Fig.\ref{figure:continental_spptype_effect}B).
 Spring events advanced 0.82 (0.54 - 1.11) days/$^{\circ}$C as MAT increased in angiosperms and delayed 0.76 (0.37 - 1.14) days/$^{\circ}$C as MAT increased in gymnosperms.
@@
 
-%EMWApr20 -- need to refer to some supp figures or tables or such below.
-Once you have the supp so I can see better what we're referring to here in first sentence we should likely edit this paragraph a little.
 % emwJun4 -- give more explanation of what we mean by  'in months that matter for the events' ... is this in the methods? 
-While we believe that coarse metrics such as latitude and MAT would ultimately represent how similar the climates are between the provenances and gardens, we also estimated climate overlap in months that matter for the events to further test how much climate similarity between provenances and gardens predicts local adaptation.
+While we expected that coarse metrics, such as latitude and MAT, would generally represent how similar the climates are between the provenances and gardens, we also estimated climate overlap in months much closer to the events to further test how much climate similarity between provenances and gardens predicts local adaptation.
 For spring events, we considered overlap across March to May.

### 2023-07-20 Analyses/source_fall_models_for_sweave.R

@@ -208,12 +208,12 @@ png(filename="leaf_effect_lat_fall.png",
     res=300)
 leaftypeplot_fitC_fall_lat <- as.matrix(cbind(evergreenspost_fitC_fall_lat, deciduouspost_fitC_fall_lat))
 plot_title <- ggtitle("Leaf Type Effects on Fall DOY in Relation to Provenance Latitude")
-color_scheme_set("viridis") 
+color_scheme_set("yellow")
 new_labels_leaftype <- c("Angiosperm","Gymnosperm")
 mcmc_areas(leaftypeplot_fitC_fall_lat)+ 
   plot_title +
-  theme(axis.text.x = element_text(size = 15))+             # x-axis text size
-  theme(axis.text.y = element_text(size = 15))   +          # y-axis text size
+  theme(axis.text.x = element_text(size = 35))+             # x-axis text size
+  theme(axis.text.y = element_text(size = 35))   +          # y-axis text size
   theme(plot.title = element_text(size = 21))  +            # plot title
   theme(text=element_text(family="sans"))+
   scale_y_discrete(labels = rev(new_labels_leaftype))
@@ -249,12 +249,12 @@ png(filename="continent_effect_lat_fall.png",
 
 continentplot_fitC_fall_lat <- as.matrix(cbind(northamericanpost_fitC_fall_lat, europeanpost_fitC_fall_lat ))
 plot_title <- ggtitle("Continental Effects on Fall DOY in Relation to Provenance Latitude")
-color_scheme_set("viridis") 
+color_scheme_set("yellow")
 new_labels_continent <- c("Europe","North America")
 mcmc_areas(continentplot_fitC_fall_lat)+ 
   plot_title +
-  theme(axis.text.x = element_text(size = 15))+             # x-axis text size
-  theme(axis.text.y = element_text(size = 15))   +          # y-axis text size
+  theme(axis.text.x = element_text(size = 35))+             # x-axis text size
+  theme(axis.text.y = element_text(size = 35))   +          # y-axis text size
   theme(plot.title = element_text(size = 21))  +            # plot title
   theme(text=element_text(family="sans"))+
   scale_y_discrete(labels = rev(new_labels_continent))

### 2023-07-20 Analyses/source_spring_models_for_sweave.R

@@ -250,12 +250,12 @@ png(filename="leaf_effect_lat_spring.png",
     res=300)
 leaftypeplot_fitC_spring_lat <- as.matrix(cbind(evergreens_springpost_fitC_spring_lat, deciduous_springpost_fitC_spring_lat))
 plot_title <- ggtitle("Leaf Type Effects on spring DOY in Relation to Provenance Latitude")
-color_scheme_set("viridis") 
+color_scheme_set("mix-teal-pink")
 new_labels_leaftype <- c("Angiosperm","Gymnosperm")
 mcmc_areas(leaftypeplot_fitC_spring_lat)+ 
   plot_title +
-  theme(axis.text.x = element_text(size = 15))+             # x-axis text size
-  theme(axis.text.y = element_text(size = 15))   +          # y-axis text size
+  theme(axis.text.x = element_text(size = 35))+             # x-axis text size
+  theme(axis.text.y = element_text(size = 35))   +          # y-axis text size
   theme(plot.title = element_text(size = 21))  +            # plot title
   theme(text=element_text(family="sans"))+
   scale_y_discrete(labels = rev(new_labels_leaftype))
@@ -300,12 +300,12 @@ png(filename="continent_effect_lat_spring.png",
 
 continentplot_fitC_spring_lat <- as.matrix(cbind(northamerican_springpost_fitC_spring_lat, european_springpost_fitC_spring_lat ))
 plot_title <- ggtitle("Continental Effects on spring DOY in Relation to Provenance Latitude")
-color_scheme_set("viridis") 
+color_scheme_set("mix-teal-pink")
 new_labels_continent <- c("Europe","North America")
 mcmc_areas(continentplot_fitC_spring_lat)+ 
   plot_title +
-  theme(axis.text.x = element_text(size = 15))+             # x-axis text size
-  theme(axis.text.y = element_text(size = 15))   +          # y-axis text size
+  theme(axis.text.x = element_text(size = 35))+             # x-axis text size
+  theme(axis.text.y = element_text(size = 35))   +          # y-axis text size
   theme(plot.title = element_text(size = 21))  +            # plot title
   theme(text=element_text(family="sans"))+
   scale_y_discrete(labels = rev(new_labels_continent))

### 2023-07-25 Docs/manuscript.Rnw
Updated discussion

@@
 
-%7July2023:
-% Zohner2023 throughout is: 
-% https://www.science.org/doi/10.1126/science.adf5098
-
 \section{Introduction}
@@
 Predicting the biological impacts of climate change has made understanding how organisms cope with environmental variation more urgent \citep{botero15}.
-In particular, the relative importance of genetic adaptation versus plasticity is of paramount importance for prediction \citep{chevin10}, with plasticity expected to allow species to shift more rapidly with climate change than environmental responses based on local adaptations \citep{chevin102, snell18}.
+In particular, the relative importance of plasticity versus genetic adaptation is vital for prediction \citep{chevin10}, with plasticity expected to allow species to shift more rapidly with climate change than environmental responses based on local adaptation \citep{chevin102, snell18}.
 
@@
 Shifted phenology in recent decades---with many events moving several days per decade \citep{vita21,khar18,Menzel06}---has led to concerns about fitness consequences, and the limits of possible future shifts.
-Future phenological shifts will depend on how much phenology is determined by adaptation versus plasticity; yet our understanding of the balance of these two approaches to variation is limited, even for species groups that are critical to both forecasting and have been well studied, such as trees.
+Future phenological shifts will depend on how much phenology is determined by plasticity versus adaptation; yet our understanding of the balance of these two approaches to variation is limited, even for species groups that are critical to both forecasting and have been well studied, such as trees.
 
@@
 The timing of budburst and senescence can impact plant competition, plant invasions and community assembly \citep{fridley12}.
-These effects often occur as phenology effects growth \citep{myneni97}, scaling up to impact ecosystem-level carbon sequestration \citep{Barichivich12}, and thus is integrated into forecasts of climate change itself.
-Growing evidence, however, suggests links between growth and phenology are not as consistent as previously predicted---or currently modelled,
-% cite https://www.nature.com/articles/s41586-022-05092-3
-, with recent work suggesting how much spring versus fall events shift may determine impacts on tree growth. % cite Zohner2023
-Such connections between tree phenology and global forecasts make understanding how plastic---or not--it may be important.
+These effects often occur as phenology affects growth \citep{myneni97}, scaling up to impact ecosystem-level carbon sequestration \citep{Barichivich12}, and thus is integrated into forecasts of climate change itself.
+Growing evidence, however, suggests links between growth and phenology are not as consistent as previously predicted---or currently modelled \citep{dow22}, with recent work suggesting how much spring versus fall events shift may determine impacts on tree growth \citep{zohner23}.
 
-Studies of adaptation versus plasticity in tree phenology have been conducted for centuries \citep{Cleland:2007or}, through common garden studies \citep[i.e., studies---often for forestry purposes---where researchers grow trees of different geographical origins under the same environmental conditions to disentangle the effects of environmental and genetic variation on trees’ phenology and phenotype][]{AitkenBemmels16, Alberto13}.
+Studies of adaptation versus plasticity in tree phenology have been conducted for centuries \citep{Cleland:2007or}, through common garden studies (i.e., studies---often for forestry purposes---where researchers grow trees of different geographical origins under the same environmental conditions to disentangle the effects of environmental and genetic variation on trees’ phenology and phenotype) \citep{AitkenBemmels16, Alberto13}.
 This work has established common clines in fall phenology, as source locations with shorter growing season (poleward and higher elevations) exhibit earlier stopping of growth (e.g., budset).
 Research has connected these clines to an underlying proximate mechanism of changing photoperiod cues (i.e., shifts in the photoperiod threshold that trigger budset), driven by local adaptation to the local growing season \citep{Alberto13,Savolainen07}.
-In contrast, spring phenology appears far more plastic \citep{AitkenBemmels16} and determined more strongly by temperature \citep{flynn18}.
+In contrast, spring phenology appears far more plastic \citep{AitkenBemmels16} and is determined more strongly by temperature \citep{flynn18}.
 Many studies, however, have argued that spring phenology shows important levels of adaptation.
+% alinaJuly20: im not sure what important levels of adaptation means here
 
@@
 To test for evidence of adaptation in spring phenology and what factors may underlie differences observed across studies, we comprehensively examined clines for spring and fall events.
-We test for evidence of adaptation via provenance trends with latitude and climate and examine possible factors that underlie these clines, asking:
+We tested for evidence of adaptation via provenance trends with latitude and climate and examined possible factors that underlie these clines, asking:
 \begin{itemize}
@@
 gymnosperm).
-All models were fit in rstanarm package (version 2.21.3; \citealp{brilleman2018}) using default priors, with 4 chains and 1000 sampling iterations per chain for a total of 4000 samples.
+All models were fit in `rstanarm' package (version 2.21.3; \citealp{brilleman2018}) using default priors, with 4 chains and 1000 sampling iterations per chain for a total of 4000 samples.
 We checked for model fit by ...
@@
 
-Effects of provenance latitude on both spring and fall events were similar across angiosperms and gymnosperms (Fig.\ref{figure:continental_spptype_lat}B).
+Effects of provenance latitude on fall events were similar across angiosperms and gymnosperms (Fig.\ref{figure:continental_spptype_lat}B).
 Spring events weakly diverged, delaying at a rate of \Sexpr{format(round(abs(mean(deciduous_springpost_fitC_spring_lat)),digits = 2), nsmall=2)} (\Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_lat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_lat, 0.95)), digits = 2),nsmall=2)}) days per degree north in angiosperms and advancing \Sexpr{format(round(abs(mean(evergreens_springpost_fitC_spring_lat)),digits = 2), nsmall=2)} (\Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_lat, 0.05)), digits = 2),nsmall=2)}) days per degree north in gymnosperms.
@@
 \ref{supplement-figure:overlap_scatterplot} in Supplements).
-We observed very weak effects of climate overlap on spring events (\Sexpr{format(round(abs(mean(all_species_springpost_fitC_spring_overlap)),digits = 2), nsmall=2)} [\Sexpr{format(round(abs(quantile(all_species_springpost_fitC_spring_overlap, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(all_species_springpost_fitC_spring_overlap, 0.95)), digits = 2),nsmall=2)}] days per one per cent increase in climate overlap), nearly identical across angiosperms (\Sexpr{format(round(abs(mean(deciduous_springpost_fitC_spring_overlap)),digits = 2), nsmall=2)} [\Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_overlap, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_overlap, 0.05)), digits = 2),nsmall=2)}]) and gymnosperms (\Sexpr{format(round(abs(mean(evergreens_springpost_fitC_spring_overlap)),digits = 2), nsmall=2)} [\Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_overlap, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_overlap, 0.95)), digits = 2),nsmall=2)}]).
-Fall events advanced as climate overlap declined, but slightly more strongly for gymnosperms (advancing \Sexpr{format(round(abs(mean(evergreenspost_fitC_fall_overlap)),digits = 2), nsmall=2)} [\Sexpr{format(round(abs(quantile(evergreenspost_fitC_fall_overlap, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(evergreenspost_fitC_fall_overlap, 0.95)), digits = 2),nsmall=2)}] days per one per cent decline in climate overlap) (Fig.\ref{supplement-figure:overlap_posterior} in Supplements).
+We observed very weak effects of climate overlap on spring events (\Sexpr{format(round(abs(mean(all_species_springpost_fitC_spring_overlap)),digits = 2), nsmall=2)} [\Sexpr{format(round(abs(quantile(all_species_springpost_fitC_spring_overlap, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(all_species_springpost_fitC_spring_overlap, 0.95)), digits = 2),nsmall=2)}] days per one percent increase in climate overlap), nearly identical across angiosperms (\Sexpr{format(round(abs(mean(deciduous_springpost_fitC_spring_overlap)),digits = 2), nsmall=2)} [\Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_overlap, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_overlap, 0.05)), digits = 2),nsmall=2)}]) and gymnosperms (\Sexpr{format(round(abs(mean(evergreens_springpost_fitC_spring_overlap)),digits = 2), nsmall=2)} [\Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_overlap, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_overlap, 0.95)), digits = 2),nsmall=2)}]).
+Fall events advanced as climate overlap declined, but slightly more strongly for gymnosperms (advancing \Sexpr{format(round(abs(mean(evergreenspost_fitC_fall_overlap)),digits = 2), nsmall=2)} [\Sexpr{format(round(abs(quantile(evergreenspost_fitC_fall_overlap, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(evergreenspost_fitC_fall_overlap, 0.95)), digits = 2),nsmall=2)}] days per one percent decline in climate overlap) (Fig.\ref{supplement-figure:overlap_posterior} in Supplements).
 
@@
 
-Opening (no subheader) 
+Overall, our results demonstrated inconsistent and weak clines in spring events across North America and Europe.
+Specifically, there were weak diverging effects by species x garden groupings, which might be important and suggest different trends that mute an overall response.
+Fall events generally showed much stronger clines, suggesting a higher degree of local adaptation, especially in North America.
+While not directly testing plasticity or adaptation, our study suggests that spring events are highly plastic, and thus we predict that warming springs will continue to be tracked more closely phenologically by trees than warming fall temperatures.
+These interesting trends by species grouping for spring events and continent for fall events suggest that data on more species and greater information on important factors, such as their geographic location in relation to their origins and elevation, would aid forecasting.
+We found coarse metrics such as provenance latitude and MAT to be better predictors of timing than climate overlap and other metrics.
+Latitude and MAT appeared to well represent how similar the climates are between the provenances and gardens in months close to the events, at least in the narrow range of northern temperature and boreal forests we studied.
+
+\subsection{Variation across continents and species groupings}
+
+Our results indicated that the effects of provenance latitude and MAT on fall events were similar across angiosperm and gymnosperm species: Fall events delay as MAT increases for both leaf types, slightly more so for gymnosperms, explained by sensitivity to day-length signals.
+However, the two leaf types exhibited slightly diverging trends in spring events:  advancing in angiosperms and delaying in most gymnosperms as MAT increases.
+Such differences make sense as timing of budburst and leafout varies by leaf type.
+Gymnosperms photosynthesize without leafout and thus generally leafout later than angiosperms, after much risk of spring frost \citep{panchen14}.
+Utilizing the early portion of the growing season can be especially critical for species in colder regions \citep{morin07, dantec15} and the ones with suboptimal shade tolerance \citep{richardson09}.
+Meanwhile, xylem anatomy also plays a part in determining spring timing due to the importance of vessel diameter to embolism risks induced by frost in spring.
+According to \citet{salk20}, species with narrow xylem such as coniferous gymnosperms and diffuse-porous angiosperms did not exhibit the consistent phenological trends shown in ring-porous species.
+Therefore, there might be additional differences if we have more species and more information on leafout timing and frost risk approach.
+
 \begin{enumerate}
-\item Overall, our results show inconsistent and weak clines in spring events across continents, while fall events generally showed much stronger clines, especially in North America
-\begin{enumerate}
-\item For spring phenology there were weak diverging effects by species groupings -- which may be important, and suggest different trends that mute an overall response. 
-\item Our results suggest that spring events are highly plastic, and thus predict that warming springs will continue to be tracked more closely phenologically by trees than warming fall temperatures.
-%7July2023 -- careful when talking about plasticity and adaptation; our results can be suggestive but do not test either of these things directly
-\end{enumerate}
-\item  These interesting trends by species grouping for spring phenology and continent for fall phenology, suggests data on more species and greater information on important factors, such as their geographic location in relation to their origins and elevation, would aid forecasting.
-\begin{enumerate}
-\item We found MAT and latitude still good predictors, better than climate overlap or other metrics. 
-\item While coarse metrics such as latitude and MAT appear to represent how similar the climates are between the provenances and gardens in months closer to the events, at least in the narrow range of northern temperature and boreal forests we studied
-\end{enumerate}
-\end{enumerate}
-
-Subheader: Variation across continents and species groupings
-\begin{enumerate}
-\item Species differences
-\begin{enumerate}
-\item Angiosperm vs gymnosperm showed some effects, which makes sense as timing of leafout here is very different (gymnosperem photosynthesize without leafout and thus generally leafout later than angiosperms, after much risk of spring frost)
-\item Suggests there may be additional differences if we have more species and more information on leafout timing and frost risk approach ... 
-\item discuss xylem (ring versus diffuse porous spp) and embolism  (see Salk article)
-\item Maybe also here: Utilizing the early portion of the growing season can be especially critical for species in colder regions \citep{morin07, dantec15} and the ones with suboptimal shade tolerance \citep{richardson09}.
-\end{enumerate}
-\item Continental differences
-\begin{enumerate}
-\item No strong evidence of continental differences, could make sense based on long-term climate history perhaps ... 
 \item Maybe also discuss ice sheet dynamics in Europe/North America, which could impact rates of adaptation.
 \end{enumerate}
-\end{enumerate}
+While we noticed more prominent clines in fall events in North America, there was no strong evidence of overall continental differences in spring events, which might again suggest a higher degree of plasticity in spring events.
+Meanwhile, most North American gardens and provenances included in this study were along the Pacific Northwest region \ref{}, which might mask the kind of climate pattern we were hoping to see from the entirety of North America.
+% alinaJuly20: not sure about how to link to long-term climate history.
 
-Subheader: Adaptation in spring vs fall phenology: Implications for climate change responses
+\subsection{Adaptation in spring vs fall phenology: Implications for climate change responses}
 \begin{enumerate}

### 2023-07-25 Docs/manuscript.tex
Updated discussion

@@
 \ref{figure:springfall_latmat}, Table.\ref{supplement-table:model_fall_lat} \& \ref{supplement-table:model_fall_mat} in Supplements ).
-This relationship, however, was observed mostly in North America where fall events advanced 4.24 (3.95 - 4.53) days per degree northward, or 6.41 days (5.78 - 7.04) per degree decline in MAT ($^{\circ}$C), whereas in Europe these relationships were weaker: advance of 0.47 (0.21 - 1.17) days per degree northward, or 0.70 days (1.04 - 2.42) per degree decline in MAT ($^{\circ}$C) (Fig.
+This relationship, however, was observed mostly in North America where fall events advanced 4.24 (3.95 - 4.53) days per degree northward, or 6.41 days (5.78 - 7.04) per degree decline in MAT ($^{\circ}$C), whereas in Europe such relationship was weaker: advance of 0.47 (0.21 - 1.17) days per degree northward, or 0.70 days (1.04 - 2.42) per degree decline in MAT ($^{\circ}$C) (Fig.
 \ref{figure:continental_spptype_effect}A).
@@
 
-% emwJun4 -- can you reference the figures in the supp that have the latitude results?
-
-Effects of provenance latitude on both spring and fall events were similar across angiosperms and gymnosperms ().
+Effects of provenance latitude on fall events were similar across angiosperms and gymnosperms (Fig.\ref{figure:continental_spptype_lat}B).
 Spring events weakly diverged, delaying at a rate of 0.37 (0.15 - 0.59) days per degree north in angiosperms and advancing 0.23 (0.00 - 0.46) days per degree north in gymnosperms.
@@
 
-% emwJun4 -- give more explanation of what we mean by  'in months that matter for the events' ... is this in the methods? 
 While we expected that coarse metrics, such as latitude and MAT, would generally represent how similar the climates are between the provenances and gardens, we also estimated climate overlap in months much closer to the events to further test how much climate similarity between provenances and gardens predicts local adaptation.
@@
 \ref{supplement-figure:overlap_scatterplot} in Supplements).
-We observed very weak effects of climate overlap on spring events (0.01 [0.02 - 0.03] days per one per cent increase in climate overlap), nearly identical across angiosperms (0.02 [0.00 - 0.05]) and gymnosperms (0.04 [0.00 - 0.09]).
-Fall events advanced as climate overlap declined, but slightly more strongly for gymnosperms (advancing 0.72 [0.51 - 0.92] days per one per cent decline in climate overlap) (Fig.\ref{supplement-figure:overlap_posterior} in Supplements).
+We observed very weak effects of climate overlap on spring events (0.01 [0.02 - 0.03] days per one percent increase in climate overlap), nearly identical across angiosperms (0.02 [0.00 - 0.05]) and gymnosperms (0.04 [0.00 - 0.09]).
+Fall events advanced as climate overlap declined, but slightly more strongly for gymnosperms (advancing 0.72 [0.51 - 0.92] days per one percent decline in climate overlap) (Fig.\ref{supplement-figure:overlap_posterior} in Supplements).
 
@@
 
-Opening (no subheader) 
+Overall, our results demonstrated inconsistent and weak clines in spring events across North America and Europe.
+Specifically, there were weak diverging effects by species x garden groupings, which might be important and suggest different trends that mute an overall response.
+Fall events generally showed much stronger clines, suggesting a higher degree of local adaptation, especially in North America.
+While not directly testing plasticity or adaptation, our study suggests that spring events are highly plastic, and thus we predict that warming springs will continue to be tracked more closely phenologically by trees than warming fall temperatures.
+These interesting trends by species grouping for spring events and continent for fall events suggest that data on more species and greater information on important factors, such as their geographic location in relation to their origins and elevation, would aid forecasting.
+We found coarse metrics such as provenance latitude and MAT to be better predictors of timing than climate overlap and other metrics.
+Latitude and MAT appeared to well represent how similar the climates are between the provenances and gardens in months close to the events, at least in the narrow range of northern temperature and boreal forests we studied.
+
+\subsection{Variation across continents and species groupings}
+
+Our results indicated that the effects of provenance latitude and MAT on fall events were similar across angiosperm and gymnosperm species: Fall events delay as MAT increases for both leaf types, slightly more so for gymnosperms, explained by sensitivity to day-length signals.
+However, the two leaf types exhibited slightly diverging trends in spring events:  advancing in angiosperms and delaying in most gymnosperms as MAT increases.
+Such differences make sense as timing of budburst and leafout varies by leaf type.
+Gymnosperms photosynthesize without leafout and thus generally leafout later than angiosperms, after much risk of spring frost \citep{panchen14}.
+Utilizing the early portion of the growing season can be especially critical for species in colder regions \citep{morin07, dantec15} and the ones with suboptimal shade tolerance \citep{richardson09}.
+Meanwhile, xylem anatomy also plays a part in determining spring timing due to the importance of vessel diameter to embolism risks induced by frost in spring.
+According to \citet{salk20}, species with narrow xylem such as coniferous gymnosperms and diffuse-porous angiosperms did not exhibit the consistent phenological trends shown in ring-porous species.
+Therefore, there might be additional differences if we have more species and more information on leafout timing and frost risk approach.
+
 \begin{enumerate}
-\item Overall, our results show inconsistent and weak clines in spring events across continents, while fall events generally showed much stronger clines, especially in North America
-\begin{enumerate}
-\item For spring phenology there were weak diverging effects by species groupings -- which may be important, and suggest different trends that mute an overall response. 
-\item Our results suggest that spring events are highly plastic, and thus predict that warming springs will continue to be tracked more closely phenologically by trees than warming fall temperatures.
-%7July2023 -- careful when talking about plasticity and adaptation; our results can be suggestive but do not test either of these things directly
-\end{enumerate}
-\item  These interesting trends by species grouping for spring phenology and continent for fall phenology, suggests data on more species and greater information on important factors, such as their geographic location in relation to their origins and elevation, would aid forecasting.
-\begin{enumerate}
-\item We found MAT and latitude still good predictors, better than climate overlap or other metrics. 
-\item While coarse metrics such as latitude and MAT appear to represent how similar the climates are between the provenances and gardens in times that matter for the events, at least in the narrow range of northern temperature and boreal forests we studied
-\end{enumerate}
-\end{enumerate}
-%7July2023 -- I don't understand the below: where do we show these continent effects for spring phenology?
-% The weak relationship between spring event dates and provenance latitude and MAT that we find in European studies might be explained by the higher extent of climate overlap in those studies.
-The more similar the climate is between provenances and gardens, the less difference between spring event dates.
-
-Subheader: Variation across continents and species groupings
-\begin{enumerate}
-\item Species differences
-\begin{enumerate}
-\item Angiosperm vs gymnosperm showed some effects, which makes sense as timing of leafout here is very different (gymnosperem photosynthesize without leafout and thus generally leafout later than angiosperms, after much risk of spring frost)
-\item Suggests there may be additional differences if we have more species and more information on leafout timing and frost risk approach ... 
-\item discuss xylem (ring versus diffuse porous spp) and embolism  (see Salk article)
-\item Maybe also here: Utilizing the early portion of the growing season can be especially critical for species in colder regions \citep{morin07, dantec15} and the ones with suboptimal shade tolerance \citep{richardson09}.
-\end{enumerate}
-\item Continental differences
-\begin{enumerate}
-\item No strong evidence of continental differences, could make sense based on long-term climate history perhaps ... 
 \item Maybe also discuss ice sheet dynamics in Europe/North America, which could impact rates of adaptation.
 \end{enumerate}
-\end{enumerate}
+While we noticed more prominent clines in fall events in North America, there was no strong evidence of overall continental differences in spring events, which might again suggest a higher degree of plasticity in spring events.
+Meanwhile, most North American gardens and provenances included in this study were along the Pacific Northwest region \ref{}, which might mask the kind of climate pattern we were hoping to see from the entirety of North America.
+% alinaJuly20: not sure about how to link to long-term climate history.
 
-Subheader: Adaptation in spring vs fall phenology: Implications for climate change responses
+\subsection{Adaptation in spring vs fall phenology: Implications for climate change responses}
 \begin{enumerate}

### 2023-08-01 Docs/manuscript.Rnw
Tiny edits and some work on opening of discussion

@@
 This is important because...
-\end{abstract}
+\end{abstract} % seven angiosperm and eight gymnosperm species from 19 common gardens, encompassing 384 North American provenances and 101 European provenances, with observations from 1962 to 2019
 
@@
 These effects often occur as phenology affects growth \citep{myneni97}, scaling up to impact ecosystem-level carbon sequestration \citep{Barichivich12}, and thus is integrated into forecasts of climate change itself.
-Growing evidence, however, suggests links between growth and phenology are not as consistent as previously predicted---or currently modelled \citep{dow22}, with recent work suggesting how much spring versus fall events shift may determine impacts on tree growth \citep{zohner23}.
+Growing evidence, however, suggests links between growth and phenology are not as consistent as previously predicted---or currently modelled \citep{dow22}---with recent work suggesting how much spring versus fall events shift may determine impacts on tree growth \citep{zohner23}.
 
-Studies of adaptation versus plasticity in tree phenology have been conducted for centuries \citep{Cleland:2007or}, through common garden studies (i.e., studies---often for forestry purposes---where researchers grow trees of different geographical origins under the same environmental conditions to disentangle the effects of environmental and genetic variation on trees’ phenology and phenotype) \citep{AitkenBemmels16, Alberto13}.
+Studies of adaptation versus plasticity in tree phenology have been conducted for centuries \citep{Cleland:2007or}, through common garden studies  \citep[i.e., studies---often for forestry purposes---where researchers grow trees of different geographical origins under the same environmental conditions to disentangle the effects of environmental and genetic variation on trees’ phenology and phenotype][]{AitkenBemmels16, Alberto13}.
 This work has established common clines in fall phenology, as source locations with shorter growing season (poleward and higher elevations) exhibit earlier stopping of growth (e.g., budset).
@@
 
-To test for evidence of adaptation in spring phenology and what factors may underlie differences observed across studies, we comprehensively examined clines for spring and fall events.
+To test for evidence of adaptation in spring phenology and what factors may underlie differences observed across studies, we comprehensively examined clines for spring events, including fall events when possible.
 We tested for evidence of adaptation via provenance trends with latitude and climate and examined possible factors that underlie these clines, asking:
@@
 The distribution was skewed toward North America (12 North American studies versus 5 European studies).
-See Table.\ref{supplement-table:all_studies} in Supplements for information on selected studies.}
+See Table.\ref{supplement-table:all_studies} in Supplement for information on selected studies.} %emwaug1: 12+5=17 but below we say 19 studies, which is correct?
     \label{figure:map_gardens}
@@
 To examine more explicitly climate near spring events, we used gridded daily temperature data for 2011 to 2020 for all European and North American provenances and gardens from E-OBS and the `daymetr' package in R respectively \citep{cornes2018,hufkens2018}.
-Using these data we estimated how much the daily temperatures overlapped between garden and provenance locations, which we call `climate overlap.' For this we used the 'overlap' package in R to calculate the percentage overlap of the daily temperature of each provenance and their corresponding gardens in spring months (March to May) from 2011 to 2020.
+Using these data we estimated how much the daily temperatures overlapped between garden and provenance locations, which we call `climate overlap.' For this we used the `overlap' package in R to calculate the percentage overlap of the daily temperature of each provenance and their corresponding gardens in spring months (March to May) from 2011 to 2020.
 Finally, using the daily temperature data, we also calculated growing degree days (GDD), a commonly used heat accumulation measure to forecast phenological development in plants \citep{miller01}, for each provenance and garden, based on the accumulation of mean daily temperatures ($T_{m}$)from 2011 to 2020 above a baseline of 0$^{\circ}$C, from January 1 until budburst and leaf flush with the following formula: GDD = $\sum$($T_{m}$ - $0^{\circ}$C for $T_{m}$ $\ge$ $0^{\circ}$C; 0 for $T_{m}$ $\le$ $0^{\circ}$C).
@@
 \ref{figure:springfall_latmat}, Table.\ref{supplement-table:model_fall_lat} \& \ref{supplement-table:model_fall_mat} in Supplements ).
-This relationship, however, was observed mostly in North America where fall events advanced \Sexpr{format(round(abs(mean(northamericanpost_fitC_fall_lat)),digits = 2),nsmall=2)} (\Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree northward, or \Sexpr{format(round(mean(northamericanpost_fitC_fall_mat),digits = 2),nsmall=2)} days (\Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_mat, 0.95)), digits = 2),nsmall=2)}) per degree decline in MAT ($^{\circ}$C), whereas in Europe such relationship was weaker: advance of \Sexpr{format(round(abs(mean(europeanpost_fitC_fall_lat)),digits = 2),nsmall=2)} (\Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree northward, or \Sexpr{format(round(mean(europeanpost_fitC_fall_mat),digits = 2),nsmall=2)} days (\Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_mat, 0.95)), digits = 2),nsmall=2)}) per degree decline in MAT ($^{\circ}$C) (Fig.
+This relationship, however, was observed mostly in North America where fall events advanced \Sexpr{format(round(abs(mean(northamericanpost_fitC_fall_lat)),digits = 2),nsmall=2)} (\Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree northward, or \Sexpr{format(round(mean(northamericanpost_fitC_fall_mat),digits = 2),nsmall=2)} days (\Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_mat, 0.95)), digits = 2),nsmall=2)}) per degree decline in MAT ($^{\circ}$C), whereas in Europe these relationships were weaker: advance of \Sexpr{format(round(abs(mean(europeanpost_fitC_fall_lat)),digits = 2),nsmall=2)} (\Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree northward, or \Sexpr{format(round(mean(europeanpost_fitC_fall_mat),digits = 2),nsmall=2)} days (\Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_mat, 0.95)), digits = 2),nsmall=2)}) per degree decline in MAT ($^{\circ}$C) (Fig.
 \ref{figure:continental_spptype_effect}A).
@@
 
-Clines in fall phenology were stronger and more consistent whereas clines in spring phenology are weaker and somewhat varied in directionality.
+Clines in fall phenology were stronger and more consistent whereas clines in spring phenology were weaker and somewhat varied in directionality.
 For fall events, only two field studies found no relationship (Fig.
@@
 Another study that found no relationship was the only greenhouse experiment included (\emph{Picea engelmannii} from Garden B in the USA, also included in \citealp{AitkenBemmels16}), which uniquely used the fall event of `the day by which seedling elongation had finished' \citep{rehfeldt94}.
-Spring event clines were always weak: all species x garden clines included 0 in their 90\% intervals.
+In contrast, spring event clines were always weak: all species x garden clines included 0 in their 90\% intervals.
 
@@
     \caption{Posterior distributions for the effect of provenance latitude across different continents and species types.
-Zero -- no effect -- is shown with a dashed line.
-Solid line and shading in posteriors represent mean and 50 percent interval.
+Zero---no effect---is shown with a dashed line.
+Solid line and shading in posteriors represent mean and 50 percent interval (90\% intervals given in text).
 (A) Effect of latitude and continent on spring and fall event date (DOY).
@@
     \caption{Posterior distributions for the effect of MAT across different continents and species types.
-Zero -- no effect -- is shown with a dashed line.
-Solid line and shading in posteriors represent mean and 50 percent interval.
+Zero---no effect---is shown with a dashed line.
+Solid line and shading in posteriors represent mean and 50 percent interval (90\% intervals given in text).
 (A) Effect of MAT and continent on spring and fall event date (DOY).
@@
 Overall, our results demonstrated inconsistent and weak clines in spring events across North America and Europe.
-Specifically, there were weak diverging effects by species x garden groupings, which might be important and suggest different trends that mute an overall response.
-Fall events generally showed much stronger clines, suggesting a higher degree of local adaptation, especially in North America.
-While not directly testing plasticity or adaptation, our study suggests that spring events are highly plastic, and thus we predict that warming springs will continue to be tracked more closely phenologically by trees than warming fall temperatures.
-These interesting trends by species grouping for spring events and continent for fall events suggest that data on more species and greater information on important factors, such as their geographic location in relation to their origins and elevation, would aid forecasting.
-We found coarse metrics such as provenance latitude and MAT to be better predictors of timing than climate overlap and other metrics.
+In contrast, fall events generally showed much stronger clines, especially in North America, and in support of many previous studies (CITES).
+While previous studies have suggested spring events are far more plastic compared to fall events (CITES) our study provides the first major test of this across continents and species and suggests overall no general trend for clines in spring phenology.
+While not a direct tests, our results suggest that spring events are highly plastic; they thus predict that warming springs will continue to be tracked more closely phenologically by trees than warming fall temperatures, in line with most current climate change impact studies (CITES, could cite IPCC here I think -- check).
+
+Trends between spring phenology and latitude or MAT were weak, but suggested the possibility of diverging results that could mute an overall trend (albeit a much weaker one than for fall phenology).
+Clines with MAT diverged between angiosperms and gymnosperms.
+Combined with our finding of stronger clines for fall phenology in North America, our results suggest that data on more species tested across continents could yield important insights into plasticity versus adaptation limits for spring and fall phenology, which would greatly aid forecasting.
+
+We found the coarse metrics of provenance latitude and MAT were generally good predictors of phenology, performing better than climate overlap.
 Latitude and MAT appeared to well represent how similar the climates are between the provenances and gardens in months close to the events, at least in the narrow range of northern temperature and boreal forests we studied.

### 2023-08-16 Docs/manuscript.Rnw
A couple more notes
discussion edits

@@
 \usepackage{amsmath}
+\usepackage[T1]{fontenc}
 \usepackage[semicolon]{natbib}
@@
 This is important because...
-\end{abstract} % seven angiosperm and eight gymnosperm species from 19 common gardens, encompassing 384 North American provenances and 101 European provenances, with observations from 1962 to 2019
+\end{abstract} % seven angiosperm and eight gymnosperm species from 17 studies, encompassing 384 North American provenances and 101 European provenances, with observations from 1962 to 2019
 
@@
 
-Studies of adaptation versus plasticity in tree phenology have been conducted for centuries \citep{Cleland:2007or}, through common garden studies  \citep[i.e., studies---often for forestry purposes---where researchers grow trees of different geographical origins under the same environmental conditions to disentangle the effects of environmental and genetic variation on trees’ phenology and phenotype][]{AitkenBemmels16, Alberto13}.
+Studies of adaptation versus plasticity in tree phenology have been conducted for centuries \citep{Cleland:2007or}, through common garden studies  (i.e., studies---often for forestry purposes---where researchers grow trees of different geographical origins under the same environmental conditions to disentangle the effects of environmental and genetic variation on trees’ phenology and phenotype) \citep{AitkenBemmels16, Alberto13}.
 This work has established common clines in fall phenology, as source locations with shorter growing season (poleward and higher elevations) exhibit earlier stopping of growth (e.g., budset).
@@
 In contrast, spring phenology appears far more plastic \citep{AitkenBemmels16} and is determined more strongly by temperature \citep{flynn18}.
-Many studies, however, have argued that spring phenology shows important levels of adaptation.
-% alinaJuly20: im not sure what important levels of adaptation means here
+Many studies, however, have argued that spring phenology shows high enough levels of adaptation to matter biologically.
+% alinaJuly20: im not sure what important levels of adaptation means here, %emwaug1: I meant that the 'amount' of local adaptation is high enough to matter (contrasting with the reality that if you have a big enough sample you can often find a small difference that is 'significant' but it might not matter biologically).
 
@@
 The distribution was skewed toward North America (12 North American studies versus 5 European studies).
-See Table.\ref{supplement-table:all_studies} in Supplement for information on selected studies.} %emwaug1: 12+5=17 but below we say 19 studies, which is correct?
+See Table.\ref{supplement-table:all_studies} in Supplement for information on selected studies.}
     \label{figure:map_gardens}
@@
 All models were fit in `rstanarm' package (version 2.21.3; \citealp{brilleman2018}) using default priors, with 4 chains and 1000 sampling iterations per chain for a total of 4000 samples.
-We checked for model fit by ...
-[ADD here].
+We checked for model fit by confirming no divergent transitions (which required setting \texttt{adapt\_delta} to 0.99 for some models), \^{R} values close to 1, and sufficient effective sample sizes.
 We present estimates as mean $\pm$ 90\% uncertainty intervals given parenthetically, unless otherwise stated.
-
-% alinaJuly20 I need help with ADD here (ping'd you in the old issue. I have removed all divergent transitions already)
 
@@
 \section{Results}
-Our final dataset included seven angiosperm and eight gymnosperm species from 19 common gardens, encompassing 384 North American provenances and 101 European provenances, with observations from 1962 to 2019.
+Our final dataset included seven angiosperm and eight gymnosperm species from 17 studies, encompassing 384 North American provenances and 101 European provenances, with observations from 1962 to 2019.
 Seven species also had fall event information available.
@@
 Overall, our results demonstrated inconsistent and weak clines in spring events across North America and Europe.
-In contrast, fall events generally showed much stronger clines, especially in North America, and in support of many previous studies (CITES).
-While previous studies have suggested spring events are far more plastic compared to fall events (CITES) our study provides the first major test of this across continents and species and suggests overall no general trend for clines in spring phenology.
-While not a direct tests, our results suggest that spring events are highly plastic; they thus predict that warming springs will continue to be tracked more closely phenologically by trees than warming fall temperatures, in line with most current climate change impact studies (CITES, could cite IPCC here I think -- check).
+In contrast, fall events generally showed much stronger clines, especially in North America, and in support of many previous studies \citep{AitkenBemmels16, Alberto13}.
+While previous studies have suggested spring events are far more plastic compared to fall events \citep{Li97,farmer93,mimura07}, our study provides the first major test of this across continents and species and suggests overall no general trend for clines in spring phenology.
+While not a direct test, our results suggest that spring events are highly plastic; they thus predict that warming springs will continue to be tracked more closely phenologically by trees than warming fall temperatures, in line with most current climate change impact studies \citep{IPCC22}
 
@@
 Gymnosperms photosynthesize without leafout and thus generally leafout later than angiosperms, after much risk of spring frost \citep{panchen14}.
-Utilizing the early portion of the growing season can be especially critical for species in colder regions \citep{morin07, dantec15} and the ones with suboptimal shade tolerance \citep{richardson09}.
 Meanwhile, xylem anatomy also plays a part in determining spring timing due to the importance of vessel diameter to embolism risks induced by frost in spring.
@@
 
-\begin{enumerate}
-\item Maybe also discuss ice sheet dynamics in Europe/North America, which could impact rates of adaptation.
-\end{enumerate}
 While we noticed more prominent clines in fall events in North America, there was no strong evidence of overall continental differences in spring events, which might again suggest a higher degree of plasticity in spring events.
-Meanwhile, most North American gardens and provenances included in this study were along the Pacific Northwest region \ref{}, which might mask the kind of climate pattern we were hoping to see from the entirety of North America.
-% alinaJuly20: not sure about how to link to long-term climate history.
+Meanwhile, most North American gardens and provenances included in this study were limited to the Pacific Northwest region, which might mask the kind of climate pattern we were hoping to see from the entirety of North America (Fig.
+\ref{figure:map_gardens}).
 
 \subsection{Adaptation in spring vs fall phenology: Implications for climate change responses}
-\begin{enumerate}
-\item Our results suggest that current advances in spring phenology may continue, but delays in fall phenology could depend on speed of adaptation
-\begin{enumerate}
-\item Spring events appear highly plastic from these studies, which are good tests. 
-\item Some previous chamber studies finding potential adaptation (cited in Ettinger et al.
-2020 Nature Climate Change, I believe -- check the paper and supp) but these could be influenced by chilling at the provenance site, whereas gardens do not have this effect
-\item Previous work has suggested strong clines in fall phenology and our findings back this up... 
-\item These findings have important implications for growing discussion of how shifting phenology and tree growth relate as our results suggest there are strong limits on shifts in fall phenology (Zohner2023 and other papers such as Zani 2020 -- need a couple sentences here) % D.
-Zani, T.
-W.
-Crowther, L.
-Mo, S.
-S.
-Renner, C.
-M.
-Zohner, Increased growing-season productivity drives earlier autumn leaf senescence in temperate trees.
-Science 370, 1066–1071 (2020).
-\end{enumerate}
-\item These trends likely vary by species and depend on frost tolerance strategy
-\begin{enumerate}
-\item If current advances in spring phenology continues it could mean shifts in spring frost etc., but differences across species suggest this will not impact all species the same. 
-\item In temperate and boreal environments, the optimal timing for spring growth initiation is determined by a balance of avoiding late spring frost and having a long growing season \citep{alberto11, lenz16, allevato19} ...
-further Plants that budburst earlier can take advantage of a longer growing season and avoid competition  \citep{guo22}....
-and species vary in how the balance this tradeoff.
-\end{enumerate}
-\item Species may also vary in whether we predict co- or counter-gradient relationships with spring phenology
-\begin{enumerate}
-\item Fall events were always co-gradient
-\item Spring events were more varied -- some appeared to be co or counter ... 
-\item Add some discussion of what this might mean (one idea is that spring leafout sets up flowering and thus some counter-gradient could aid gene flow over larger areas?
-Other ideas you have?) and cite some of the relevant papers.
-\end{enumerate}
-\item All of this critical to predicting not just phenological change, but range expansions with climate change: An accurate understanding of how environmental conditions might influence species distribution at a continental scale is critical for future range shift prediction.
-\begin{enumerate}
-\item Implications of small differences in spring phenology could mean different gene flow etc. 
-\item the degree of adaptation dictated by climate change also differs across different geographical areas \citep{Loarie09}. 
-\item Mention elevational results here? 
-\end{enumerate}
-\end{enumerate}
+
+Our results suggest that current advances in spring phenology may continue, but delays in fall phenology could depend on the speed of adaptation.
+While some previous chamber studies have found potential adaptation, these could have been influenced by chilling at the provenance site \citep{ettinger20}.
+However, field studies and common garden experiences do not have this effect.
+Spring events from the studies we looked at appeared to exhibit high plasticity, while fall events showed strong clines, which is consistent with previous work \citep{AitkenBemmels16, Alberto13}.
+These findings have important implications for growing discussion of how shifting phenology and tree growth relate as our results suggest there are strong limits on shifts in fall phenology \citep{zohner23, zani20}.
+
+It is worth noting that these trends likely vary by species and depend on frost tolerance strategy.
+If current advances in spring phenology continues, it could suggest shifts in spring frost.
+In temperate and boreal environments, the optimal timing for spring growth initiation is determined by a balance of avoiding late spring frost and having a long growing season \citep{alberto11, lenz16, allevato19}.
+Utilizing the early portion of the growing season can be especially critical for species in colder regions \citep{morin07, dantec15} and the ones with suboptimal shade tolerance \citep{richardson09}.
+Further, plants that budburst earlier can take advantage of a longer growing season and avoid competition \citep{guo22}.
+However, differences across species means that this will not impact all species the same.
+While fall event always exhibited co-gradient relationships (i.e.
+events advanced as we moved northward or up in elevation), spring events were more varied among species.
+The implication of such variations could be that spring leafout sets up flowering, and thus, some counter-gradient could better aid gene glow over larger geographical regions.
+
+All of above are critical to predicting not just phenological change, but range expansions with climate change: An accurate understanding of how environmental conditions might influence species distribution at a continental scale is critical for future range shift prediction.
+Implications of small differences in spring phenology could mean different levels of gene flow.
+
+Our observations reveal that although temperature by itself could serve as a reasonable indicator for predicting when fall events occur across various locations, it's crucial to acknowledge that the degree of adaptation dictated by climate change differs across different geographical areas \citep{Loarie09}.
+High-latitude sites, for example, are likely restricted by the impact of photoperiod \citep{Gill15}.
+Diverse mechanisms that govern the timing of leaf senescence at the ecosystem level hold the potential to exert significant influence over not only the interactions occurring within plant communities but also the intricate process of carbon storage within ecosystems.
+With the anticipated escalation of global temperatures in the forthcoming century, these mechanisms are poised to play a pivotal role in shaping the dynamics of plant communities and the overall carbon balance of ecosystems.
 

### 2023-08-16 Docs/manuscript.tex
discussion edits

@@
 This is important because...
-\end{abstract}
+\end{abstract} % seven angiosperm and eight gymnosperm species from 17 studies, encompassing 384 North American provenances and 101 European provenances, with observations from 1962 to 2019
 
@@
 These effects often occur as phenology affects growth \citep{myneni97}, scaling up to impact ecosystem-level carbon sequestration \citep{Barichivich12}, and thus is integrated into forecasts of climate change itself.
-Growing evidence, however, suggests links between growth and phenology are not as consistent as previously predicted---or currently modelled \citep{dow22}, with recent work suggesting how much spring versus fall events shift may determine impacts on tree growth \citep{zohner23}.
+Growing evidence, however, suggests links between growth and phenology are not as consistent as previously predicted---or currently modelled \citep{dow22}---with recent work suggesting how much spring versus fall events shift may determine impacts on tree growth \citep{zohner23}.
 
@@
 The distribution was skewed toward North America (12 North American studies versus 5 European studies).
-See Table.\ref{supplement-table:all_studies} in Supplements for information on selected studies.}
+See Table.\ref{supplement-table:all_studies} in Supplement for information on selected studies.}
     \label{figure:map_gardens}
@@
 \ref{figure:springfall_latmat}, Table.\ref{supplement-table:model_fall_lat} \& \ref{supplement-table:model_fall_mat} in Supplements ).
-This relationship, however, was observed mostly in North America where fall events advanced 4.24 (3.95 - 4.53) days per degree northward, or 6.41 days (5.78 - 7.04) per degree decline in MAT ($^{\circ}$C), whereas in Europe such relationship was weaker: advance of 0.47 (0.21 - 1.17) days per degree northward, or 0.70 days (1.04 - 2.42) per degree decline in MAT ($^{\circ}$C) (Fig.
+This relationship, however, was observed mostly in North America where fall events advanced 4.24 (3.95 - 4.53) days per degree northward, or 6.41 days (5.78 - 7.04) per degree decline in MAT ($^{\circ}$C), whereas in Europe these relationships were weaker: advance of 0.47 (0.21 - 1.17) days per degree northward, or 0.70 days (1.04 - 2.42) per degree decline in MAT ($^{\circ}$C) (Fig.
 \ref{figure:continental_spptype_effect}A).
@@
 Overall, our results demonstrated inconsistent and weak clines in spring events across North America and Europe.
-Specifically, there were weak diverging effects by species x garden groupings, which might be important and suggest different trends that mute an overall response.
-Fall events generally showed much stronger clines, suggesting a higher degree of local adaptation, especially in North America.
-While not directly testing plasticity or adaptation, our study suggests that spring events are highly plastic, and thus we predict that warming springs will continue to be tracked more closely phenologically by trees than warming fall temperatures.
-These interesting trends by species grouping for spring events and continent for fall events suggest that data on more species and greater information on important factors, such as their geographic location in relation to their origins and elevation, would aid forecasting.
-We found coarse metrics such as provenance latitude and MAT to be better predictors of timing than climate overlap and other metrics.
+In contrast, fall events generally showed much stronger clines, especially in North America, and in support of many previous studies \citep{AitkenBemmels16, Alberto13}.
+While previous studies have suggested spring events are far more plastic compared to fall events \citep{Li97,farmer93,mimura07}, our study provides the first major test of this across continents and species and suggests overall no general trend for clines in spring phenology.
+While not a direct test, our results suggest that spring events are highly plastic; they thus predict that warming springs will continue to be tracked more closely phenologically by trees than warming fall temperatures, in line with most current climate change impact studies \citep{IPCC22}
+
+Trends between spring phenology and latitude or MAT were weak, but suggested the possibility of diverging results that could mute an overall trend (albeit a much weaker one than for fall phenology).
+Clines with MAT diverged between angiosperms and gymnosperms.
+Combined with our finding of stronger clines for fall phenology in North America, our results suggest that data on more species tested across continents could yield important insights into plasticity versus adaptation limits for spring and fall phenology, which would greatly aid forecasting.
+
+We found the coarse metrics of provenance latitude and MAT were generally good predictors of phenology, performing better than climate overlap.
 Latitude and MAT appeared to well represent how similar the climates are between the provenances and gardens in months close to the events, at least in the narrow range of northern temperature and boreal forests we studied.

### 2023-09-01 Docs/manuscript.Rnw
Transcribing some edits from my paper copy -- not done yet

@@
 \begin{abstract}
-Climate change forecasting ....
-plasticity....
-adaptation.
-Spring tree phenology is critical to forecasting ...
+Forecasting the biological impacts of climate change requires understanding how species respond to warmer temperatures through both inter-annual flexible variation versus through adaptation to local conditions.
+Yet we often lack this information entirely, or find conflicting evidence across studies.
+This is the case for one of the most reported and consistent impacts of anthropogenic climate change---earlier spring phenology, which is also one of the most critical to forecasting given its role in carbon sequestration.
+Evidence from some common garden studies suggest important provenance effects, which suggest local adaptation in the underlying cues of spring phenology and supports findings for fall events, while other studies find no evidence.
+Here we synthesize common garden studies across Europe and North America that reported spring event dates for a mix of angiosperm and gymnosperm tree species in the northern hemisphere, capturing data from 384 North American provenances and 101 European provenances with observations from 1962 to 2019, alongside fall event data when provided.
+Across continents we find no evidence of provenance effects in spring phenology, but strong clines with latitude and mean annual temperature (MAT) for fall events.
+These effects, however, appear to diverge by continent and species type (gymnosperm versus angiosperm), especially for fall events where clines with latitude and MAT are much stronger in North America.
+Our results suggest flexible, likely plastic responses, in spring phenology with warming, and potential limits---at least in the short term---due to provenance effects for fall phenology.
+They also highlight that, after over 50 years of common garden studies on tree phenology we may still lack a holistic predictive model of clines across species and phenological events.
+\end{abstract} %Spring tree phenology is critical to forecasting ...
 Despite a long history of study, however, the balance of local adaptation versus plasticity in determining environmental responses is not well understood, especially for spring events.
-We conducted the first cross-continental meta-analysis of published studies from the peer-reviewed literature that reported spring event dates for a mix of angiosperm and gymnosperm tree species in the northern hemisphere, capturing data from 384 North American provenances and 101 European provenances with observations from 1962 to 2019.
+We conducted the first cross-continental meta-analysis of published studies from the peer-reviewed literature that reported spring event dates for a mix of angiosperm and gymnosperm tree species in the northern hemisphere, capturing data  for  seven angiosperm and eight gymnosperm species from 17 studies, from 384 North American provenances and 101 European provenances with observations from 1962 to 2019.
 We found ...
 This is important because...
-\end{abstract} % seven angiosperm and eight gymnosperm species from 17 studies, encompassing 384 North American provenances and 101 European provenances, with observations from 1962 to 2019
 
@@
 To understand how climatic differences, in addition to geographical differences, shape local adaptation in spring events we extracted several types of climate data using information about provenance latitude, longitude, and elevation from original publications.
-We estimated the mean annual temperature (MAT) for each provenance using the Climate Information Tool by Food and Agriculture Organization of the United Nations \citep{FAO2022}and ClimateWNA \citep{wang2016}.
+We estimated the mean annual temperature (MAT) from 1960-1991 for each provenance using the Climate Information Tool by Food and Agriculture Organization of the United Nations \citep{FAO2022}.
+We verified our calculated MAT was was similar to MAT calculated using ClimateWNA \citep{wang2016}, a source used in previous analyses.
 
@@
 Using these data we estimated how much the daily temperatures overlapped between garden and provenance locations, which we call `climate overlap.' For this we used the `overlap' package in R to calculate the percentage overlap of the daily temperature of each provenance and their corresponding gardens in spring months (March to May) from 2011 to 2020.
-Finally, using the daily temperature data, we also calculated growing degree days (GDD), a commonly used heat accumulation measure to forecast phenological development in plants \citep{miller01}, for each provenance and garden, based on the accumulation of mean daily temperatures ($T_{m}$)from 2011 to 2020 above a baseline of 0$^{\circ}$C, from January 1 until budburst and leaf flush with the following formula: GDD = $\sum$($T_{m}$ - $0^{\circ}$C for $T_{m}$ $\ge$ $0^{\circ}$C; 0 for $T_{m}$ $\le$ $0^{\circ}$C).
+% Finally, using the daily temperature data, we also calculated growing degree days (GDD), a commonly used heat accumulation measure to forecast phenological development in plants \citep{miller01}, for each provenance and garden, based on the accumulation of mean daily temperatures ($T_{m}$)from 2011 to 2020 above a baseline of 0$^{\circ}$C, from January 1 until budburst and leaf flush with the following formula: GDD = $\sum$($T_{m}$ - $0^{\circ}$C for $T_{m}$ $\ge$ $0^{\circ}$C; 0 for $T_{m}$ $\le$ $0^{\circ}$C).
+% See notes on 31 Aug 2023 by Lizzie at https://github.com/lizzieinvancouver/localadaptclim/issues/26 for more methods on GDD.
 

### 2023-09-06 Docs/manuscript.Rnw
Edits to current draft, esp. discussion

@@
 Predicting the biological impacts of climate change has made understanding how organisms cope with environmental variation more urgent \citep{botero15}.
-In particular, the relative importance of plasticity versus genetic adaptation is vital for prediction \citep{chevin10}, with plasticity expected to allow species to shift more rapidly with climate change than environmental responses based on local adaptation \citep{chevin102, snell18}.
+In particular, the relative importance of plasticity versus genetic adaptation is vital for prediction \citep{chevin10}, with plasticity expected to allow species to shift more rapidly with climate change than environmental responses based on local adaptation, but possibly limiting responses after the limits of plasticity are reached \citep{chevin102,snell18}.
 
@@
 Thus I slimmed this way down and kept many references, but less of the examples.
-Many of the currently observed responses to climate change appear to be mostly plastic \citep{burton22,zettle21,bonamour19, king17}, including the most reported biological response to climate change---shifting phenology.
-Phenology (the timing of recurring seasonal events) governs the timing of transitions between dormancy and active growth for many organisms, allowing them to time reproduction and exploit the resources of each growing season \citep{chuine10,hanninen11,rytteri21,posle18}.
+Many of the currently observed responses to climate change appear to be mainly plastic \citep{burton22,zettle21,bonamour19, king17}, including the most reported biological response to climate change---shifting phenology.
+Phenology---the timing of recurring seasonal events---governs the timing of transitions between dormancy and active growth for many organisms, allowing them to time reproduction and exploit the resources of each growing season \citep{chuine10,hanninen11,rytteri21,posle18}.
 As such, phenology plays a significant role in determining fitness for both plants \citep{guo22,chuine01} and animals \citep{wann19,renner18,chu17}.
@@
 Shifted phenology in recent decades---with many events moving several days per decade \citep{vita21,khar18,Menzel06}---has led to concerns about fitness consequences, and the limits of possible future shifts.
-Future phenological shifts will depend on how much phenology is determined by plasticity versus adaptation; yet our understanding of the balance of these two approaches to variation is limited, even for species groups that are critical to both forecasting and have been well studied, such as trees.
+While future phenological shifts will depend on how much phenology is determined by plasticity versus adaptation, our understanding of the balance of these two approaches to variation is limited.
+This is the case even for species groups that are critical to both forecasting and have been well studied, such as trees.
 
@@
 
-Studies of adaptation versus plasticity in tree phenology have been conducted for centuries \citep{Cleland:2007or}, through common garden studies  (i.e., studies---often for forestry purposes---where researchers grow trees of different geographical origins under the same environmental conditions to disentangle the effects of environmental and genetic variation on trees’ phenology and phenotype) \citep{AitkenBemmels16, Alberto13}.
-This work has established common clines in fall phenology, as source locations with shorter growing season (poleward and higher elevations) exhibit earlier stopping of growth (e.g., budset).
-Research has connected these clines to an underlying proximate mechanism of changing photoperiod cues (i.e., shifts in the photoperiod threshold that trigger budset), driven by local adaptation to the local growing season \citep{Alberto13,Savolainen07}.
-In contrast, spring phenology appears far more plastic \citep{AitkenBemmels16} and is determined more strongly by temperature \citep{flynn18}.
-Many studies, however, have argued that spring phenology shows high enough levels of adaptation to matter biologically.
-% alinaJuly20: im not sure what important levels of adaptation means here, %emwaug1: I meant that the 'amount' of local adaptation is high enough to matter (contrasting with the reality that if you have a big enough sample you can often find a small difference that is 'significant' but it might not matter biologically).
+Studies of adaptation versus plasticity in tree phenology have been conducted for centuries \citep{Cleland:2007or}, through common garden studies.
+In these studies---conducted often for forestry purposes---researchers grow trees of different geographical origins under the same environmental conditions to disentangle the effects of environmental and genetic variation on trees’ phenology and phenotype \citep{AitkenBemmels16, Alberto13}.
+Such work has established common clines in fall phenology, as source locations with shorter growing seasons (poleward and higher elevations) exhibit earlier stopping of growth (such as budset).
+Research has connected these clines to an underlying proximate mechanism of changing photoperiod cues (i.e., shifts in the photoperiod threshold required to trigger budset), driven by adaptation to the local growing season \citep{Alberto13,Savolainen07}.
+In contrast, spring phenology appears more plastic \citep{AitkenBemmels16} and is determined more strongly by temperature \citep{flynn18}.
+Many studies, however, have argued that spring phenology shows levels of adaptation that may be critical to forecasting and mitigation \citep{vitasse2009,Basler:2012}.
 
-Results, however, have not always been consistent.
-Studies have documented provenance differences of 2-4 days per degree latitude in spring phenology for some species (\emph{Picea abies} in \citealp{sog08} and \emph{Quercus petraea} in \citealp{deans96}) while others have not found any trends in budburst along latitudinal gradients (for example, \emph{Picea sitchensis} in \citealp{mimura07}, \emph{Picea glauca} in \citealp{Li97}, and \emph{Populus balsamifera} in \citealp{farmer93}).
-This has led to some debate over the prevalence and importance of adaptation in spring tree phenology.
-While clines of spring phenology have been found in studies in both Europe \citep{sog08,deans96,von95} and North America \citep{rossi15, soo13, hannerz99}, they could vary by continent.
+These contrasting studies highlight how inconsistent evidence for adaptation in tree spring phenology has been.
+Studies have documented provenance differences of 2-4 days per degree latitude in spring phenology for some species (\emph{Picea abies} in \citealp{sog08} and \emph{Quercus petraea} in \citealp{deans96}) while others have failed to find similar trends along latitudinal gradients (for example, \emph{Picea sitchensis} in \citealp{mimura07}, \emph{Picea glauca} in \citealp{Li97}, and \emph{Populus balsamifera} in \citealp{farmer93}).
+This has led to debate over the prevalence and importance of adaptation in spring tree phenology.
+While clines of spring phenology have been found in studies in both Europe \citep{sog08,deans96,von95} and North America \citep{rossi15, soo13, hannerz99}, they may be more studies and debated in Europe \citep{deans96,vitasse2009,Basler:2012}, raising the possibility that they could vary by continent.
+%EMWSep4: Check deans96 is accurate ref I added.
+Other refs I added to 'more' bib in case you do not have them.
 
-Continental differences in patterns of adaptation versus plasticity could be driven by climatic differences, as North American springs are more variable across years than European ones \citep{tward21,zohner2017,schwartz00}.
+Continental differences in patterns of adaptation versus plasticity could be driven by climatic differences, especially as North American springs are more variable across years than European ones \citep{tward21,zohner2017,schwartz00}.
 Such high temporal variability means that distant sites can effectively experience the same spring climate, but in different years.
-Studies of spring phenology in arboreta suggest cues for budburst may vary depending on continental climate \citep{zohner2017}, but are poorly controlled compared to traditional provenance trials, making them difficult to use for tests of plasticity versus adaptation \citep{gauzere2020}.
-Even for more carefully designed common garden differences in study design---especially studied species---may complicate understanding what underlies potential trends across continents, with European studies focusing more on deciduous angiosperms and North American studies focusing often on evergreen gymnosperms.
+Studies of spring phenology in arboreta suggest cues for budburst may vary depending on continental climate \citep{zohner2017}, but are poorly controlled compared to traditional common garden studies, making them difficult to use for tests of plasticity versus adaptation \citep{gauzere2020}.
+Even for more carefully designed common gardens, differences in species studied or other differences in design may complicate understanding what underlies potential trends across continents.
+%EMWSep4: Moved last point to discussion.
 
@@
 \section{Results}
-Our final dataset included seven angiosperm and eight gymnosperm species from 17 studies, encompassing 384 North American provenances and 101 European provenances, with observations from 1962 to 2019.
-Seven species also had fall event information available.
+Our final dataset included seven deciduous angiosperm and eight evergreen gymnosperm species from 17 studies and 19 gardens, encompassing 384 North American provenances and 101 European provenances, with observations from 1962 to 2019.
+Seven species (five in North America and two in Europe) also had fall event information available.
+%EMWSep4 -- check my edits here; also if we don't have it somewhere in the results yet, please add gymnosperms/angiosperm species # in North America versus Europe here, something like: Most species in North American gardens were gymosperms (7/11 species) while most species were angiosperms in Europe (3/4 species).
 <<label=source_spring_models, echo=FALSE>>=
@@
 
-In contrast, fall events (e.g., budset, leaf senescence, leaf abscission) advanced strongly with provenance latitude (\Sexpr{format(round(abs(mean(all_species_fallpost_fitC_fall_lat)),digits = 2), nsmall=2)} days/degree [\Sexpr{format(round(abs(quantile(all_species_fallpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)}-\Sexpr{format(round(abs(quantile(all_species_fallpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}]) and the decrease in MAT (\Sexpr{format(round(abs(mean(all_species_fallpost_fitC_fall_mat)),digits = 2), nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(all_species_fallpost_fitC_fall_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(all_species_fallpost_fitC_fall_mat, 0.95), digits = 2)}]), meaning fall events were earlier where provenance MAT was lower (Fig.
+%EMWSep4 -- small edits here, hopefully I did not screw up the formatting
+In contrast, fall events (e.g., budset, leaf senescence, leaf abscission) were earlier at more norther, cooler MAT sites (that is, they advanced strongly with provenance latitude: \Sexpr{format(round(abs(mean(all_species_fallpost_fitC_fall_lat)),digits = 2), nsmall=2)} days/degree [\Sexpr{format(round(abs(quantile(all_species_fallpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)}-\Sexpr{format(round(abs(quantile(all_species_fallpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}], and with decreasing MAT:\Sexpr{format(round(abs(mean(all_species_fallpost_fitC_fall_mat)),digits = 2), nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(all_species_fallpost_fitC_fall_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(all_species_fallpost_fitC_fall_mat, 0.95), digits = 2)}], Fig.
 \ref{figure:springfall_latmat}, Table.\ref{supplement-table:model_fall_lat} \& \ref{supplement-table:model_fall_mat} in Supplements ).
@@
 
-While we expected that coarse metrics, such as latitude and MAT, would generally represent how similar the climates are between the provenances and gardens, we also estimated climate overlap in months much closer to the events to further test how much climate similarity between provenances and gardens predicts local adaptation.
+While we expected that coarse metrics, such as latitude and MAT, would generally represent how similar the climates are between the provenances and gardens, we also estimated climate overlap in months much closer to the events to further test how much climate similarity between provenances and gardens predicts provenance effects.
 For spring events, we considered overlap across March to May.
@@
 In contrast, fall events generally showed much stronger clines, especially in North America, and in support of many previous studies \citep{AitkenBemmels16, Alberto13}.
-While previous studies have suggested spring events are far more plastic compared to fall events \citep{Li97,farmer93,mimura07}, our study provides the first major test of this across continents and species and suggests overall no general trend for clines in spring phenology.
-While not a direct test, our results suggest that spring events are highly plastic; they thus predict that warming springs will continue to be tracked more closely phenologically by trees than warming fall temperatures, in line with most current climate change impact studies \citep{IPCC22}
+While previous studies have suggested spring events are far more plastic compared to fall events \citep{Li97,farmer93,mimura07}, our study provides the first major test of this across continents and species and suggests no general trend for clines in spring phenology.
+Our results thus predict that warming springs will continue to be tracked more closely phenologically by trees than warming fall temperatures \citep{IPCC22}
 
-Trends between spring phenology and latitude or MAT were weak, but suggested the possibility of diverging results that could mute an overall trend (albeit a much weaker one than for fall phenology).
-Clines with MAT diverged between angiosperms and gymnosperms.
-Combined with our finding of stronger clines for fall phenology in North America, our results suggest that data on more species tested across continents could yield important insights into plasticity versus adaptation limits for spring and fall phenology, which would greatly aid forecasting.
+Trends between spring phenology and latitude or MAT were weak, but suggested the possibility of diverging results that could mute an overall trend---albeit a much weaker one than for fall phenology.
+We found angiosperm versus gymnosperm species diverged in their clines with MAT.
+Combined with our finding of much stronger clines for fall phenology in North America, these results support the idea of potential variation across continents and/or species type that may underlie the debate in whether spring events show important clinal variation.
+As clines with spring events were very weak, however, and gardens almost always focus on only one species, understanding these diverging results well enough to aid forecasting would take significant additional investment in common garden studies.
+%EMWSep4 -- we need to be careful in how many times we say 'we need more data' -- it's a common need, so it's better to focus on specifics of what the data suggests, or doesn't and say we need more data -- at most once in a paper.
+More on this below.
 
-We found the coarse metrics of provenance latitude and MAT were generally good predictors of phenology, performing better than climate overlap.
-Latitude and MAT appeared to well represent how similar the climates are between the provenances and gardens in months close to the events, at least in the narrow range of northern temperature and boreal forests we studied.
+We found the coarse metrics of provenance latitude and MAT were generally good predictors of phenology, performing better than our more complex and data-rich metric of climate overlap.
+Latitude and MAT appeared to well represent how similar the climates are between the provenances and gardens in  the temperate and boreal forest species we studied, yielding similar results metrics calculated specifically in the spring with daily climate data.
+%EMWSep4 -- also avoid repeating how narrow the data is etc.
+-- it's good to point out limitations, but repeating them too much is not helpful -- this is the biggest dataset for this question on earth, we need to focus on what we learned from it, not so much on the limitations.
+Also, Europeans would not call all of the forests northern.
 
-\subsection{Variation across continents and species groupings}
+\subsection{Variation across continents and species types}
 
-Our results indicated that the effects of provenance latitude and MAT on fall events were similar across angiosperm and gymnosperm species: Fall events delay as MAT increases for both leaf types, slightly more so for gymnosperms, explained by sensitivity to day-length signals.
-However, the two leaf types exhibited slightly diverging trends in spring events:  advancing in angiosperms and delaying in most gymnosperms as MAT increases.
-Such differences make sense as timing of budburst and leafout varies by leaf type.
-Gymnosperms photosynthesize without leafout and thus generally leafout later than angiosperms, after much risk of spring frost \citep{panchen14}.
-Meanwhile, xylem anatomy also plays a part in determining spring timing due to the importance of vessel diameter to embolism risks induced by frost in spring.
+Our results highlight that spring events show much weaker clines than fall events in tree species, but suggest important variation between species types in spring events.
+Angiosperms tended to budburst earlier in provenances that were warmer and more southerly, while gymnosperms trended in the opposite direction.
+Such differences could be driven by their varying investment strategies.
+As gymnosperms photosynthesize without leafout they generally leafout much later than angiosperms, after most risk of spring frost \citep{panchen14}, and thus may avoid frost risk more than angiosperms.
+In contrast, angiosperms may tend towards earlier leafout in warmer climates to compete best for access to light and other resources \citep{cat2019}.
+Testing these hypotheses would require More information on frost risk and forest community assembly from across the provenance locations.
+%EMWSep4 -- I like the below (and I think it may have been my idea), but we'd likely be asked to test for it ...
+and I am not 100% sure it makes sense -- both angiosperms and conifers need to activate their xylem to photosynthesize I believe.
+% Meanwhile, xylem anatomy also plays a part in determining spring timing due to the importance of vessel diameter to embolism risks induced by frost in spring.
 According to \citet{salk20}, species with narrow xylem such as coniferous gymnosperms and diffuse-porous angiosperms did not exhibit the consistent phenological trends shown in ring-porous species.
@@
 
-While we noticed more prominent clines in fall events in North America, there was no strong evidence of overall continental differences in spring events, which might again suggest a higher degree of plasticity in spring events.
-Meanwhile, most North American gardens and provenances included in this study were limited to the Pacific Northwest region, which might mask the kind of climate pattern we were hoping to see from the entirety of North America (Fig.
+Fall event clines were more similar across species types, but diverged by continent with the common clines of later events in warmer, more southerly latitudes found only in North American studies.
+However, as we were focused on spring events, we had limited data on fall events, which likely explains our results which contrast with meta-analyses of fall events in Europe (citeAlbero?).
+In our dataset, more data on fall events were reported in North America, while only two of the six European gardens reported leaf senescence-related metrics, one studying \emph{Fagus sylvatica} and one studying \emph{Fraxinus excelsior}, and neither showed strong clines.
+In North America studies of both gymnosperm and angiosperm species reported strong clines, and most often focused on budset, while the two European gardens reported data on `leaf senescence' (EXPLAIN what their metric was?, cite Table S1).
+Budset is most closely related to growth cessation and tied to photoperiod signals, making it generally considered a more robust indicator of when plants stop investing in growth and more genetically controlled than leaf senescence \citep{Alberto13}.
+%EMWSep4 -- I am actually not sure if `our results which contrast with meta-analyses of fall events in Europe' is true -- is it?
+I could not tell from reviewing Alberto very quickly.
+If it is true from that paper or another, then I think this is the best explanation of our results.
+If it's not true, I can adjust the text.
+% Also, can you say how the two gardens defined `leaf senescence' inside (EXPLAIN)?
+
+%EMWSep4 -- I like to have one paragraph in a paper (usually) focused on biases and limitations, so the below is that paragraph.
+I also aimed for it to have more of a message specific to this work beyond 'we need more data' (as we always need more data, you can write that in any paper).
+Taken together, these results could explain some of the existing debate on the strength of spring clinal variation, but also highlight how data biases make disentangling the drivers of variation difficult.
+Angiosperm species showed a weak trend towards earlier budburst in populations from warmer locations.
+In contrast fall event clines in Europe were weak, but their fall event type (leaf senescence) is less likely to show clinal variation.
+Given European studies were biased towards studying angiosperms (3/4 species) that are often canopy-species (e.g., \emph{Fagus}, \emph{Quercus}) these results together could explain a greater focus on clinal variation in spring events across European studies (CITESa).
+Most North American gardens and provenances included in this study were limited to the Pacific Northwest region, and thus tended to focus on species from forests where gymnosperms are almost always the canopy species (\emph{Pseudotsuga, Tsuga, Picea, Pinus}) and angiosperms are much smaller, more opportunistic species (\emph{Alnus, Populus}).
+This might suggest a far different pattern if gardens included more evenly sampled North American tree species (Fig.
 \ref{figure:map_gardens}).
+Given the tremendous investment required for such gardens, however, it is not surprising they are often limited to one species---most often those relevant for forestry (CITES).
+Our results, however, highlight the limited inference we may gain from such an approach to understand trends across species more broadly.
+%EMWSep4 -- Great point you made about the bias in North American study locations, I built on it. 
+% CITESa -- could be Basler & Korner, Gauzere 2020 paper .. you likely have more ideas to add.
 
-\subsection{Adaptation in spring vs fall phenology: Implications for climate change responses}
+\subsection{Adaptation in tree phenology: Implications for climate change responses}
 
-Our results suggest that current advances in spring phenology may continue, but delays in fall phenology could depend on the speed of adaptation.
-While some previous chamber studies have found potential adaptation, these could have been influenced by chilling at the provenance site \citep{ettinger20}.
+%EMWSep4 -- good stuff here!
+I tried to refocus on spring events some as our data is focused there.
+And I limited the amount of hand-wavey stuff we discussed as reviewers only usually want so much extrapolation of your results (so I cut some of the ending stuff).
+Our results suggest that current advances in spring phenology will likely continue, but predicting any shifts due to provenance effects will require new data across more species.
+Plastic responses to warming, as our results suggest for spring events, mean species can adjust to the shifting environments of climate change---up to some point.
+Beyond that point theory suggests plastic traits may limit adaptation to climate change (CITE chevin I think?).
+This may mean species will be replaced when their plastic spring phenological responses are no longer adequate, or shifting reaction norms could lead to adaptation (cite Gauzere 2020?).
+
+Predicting this will require better understanding how different species strategies covary with early-season risks, and how such risks may shift with continued climate change.
+For temperate zones, many species strategies appear designed to cope with frost, either via avoidance or tolerance \citep{alberto11, lenz16, allevato19}, as utilizing the early portion of the growing season can be especially critical for species in colder regions \citep{morin07, dantec15}.
+Climate change at once shifts spring phenology---thus potentially changing each species frost risk---but also appears to reshape the drivers of frost climatically \citep{cat2021pep}.
+Layered onto this are possible shifts in early-season herbivory with warming \citep{meineke2019}, and trade-offs in early-season risks with access to a long growing season for growth and reproduction, with some species avoiding competition through being very early \citep{guo22}.
+Common garden studies that track and report frost and herbivory damage, alongside timing of flowering and fruiting, could help to understand the complex fitness landscape of spring phenology.
+
+In contrast to spring events which were weak and variable, fall events exhibited strong clines, which appeared co-gradient with the environment (i.e.
+late-season events advanced earlier with northward latitudes).
+This supports previous results and the general theory that fall events are consistently designed to avoid tissue loss at the end of the season, when the cost of such losses could be particularly high \citep{AitkenBemmels16, Alberto13}.
+Spring events showed no clear trends, but Increasing research into counter-gradient variation for flowering events could be relevant when spring vegetative and flowering events are linked \citep{dan2021nph}.
+
+Understanding environmental clines in phenology will ultimately be critical to predicting how species shift their ranges as well.
+Implications of small differences in spring phenology could mean different levels of gene flow, while co- and counter-gradient variation have important implications for gene glow over larger geographical regions (CITES), and thus could shape future range shifts.
+Much like frost risk, climate change likely shifts multiple factors at once, as climate change shifts pollinators and wind patterns (CITESb).
+With the anticipated escalation of global temperatures in the forthcoming century, these mechanisms are poised to play a pivotal role in shaping the dynamics of plant communities and the overall carbon balance of ecosystems.
+% CITESb Ackerly has a paper on shifting wind patterns in PNAS in the last couple years to cite here.
+
+% All of above are critical to predicting not just phenological change, but range expansions with climate change ...
+An accurate understanding of how environmental conditions might influence species distribution at a continental scale is critical for future range shift prediction.
+
+%EMWSep4 -- I know I wanted this included, but not sure it works. Let's leave out for now. 
+%  While some previous chamber studies have found potential adaptation, these could have been influenced by chilling at the provenance site \citep{ettinger20}.
 However, field studies and common garden experiences do not have this effect.
@@
 
-It is worth noting that these trends likely vary by species and depend on frost tolerance strategy.
-If current advances in spring phenology continues, it could suggest shifts in spring frost.
-In temperate and boreal environments, the optimal timing for spring growth initiation is determined by a balance of avoiding late spring frost and having a long growing season \citep{alberto11, lenz16, allevato19}.
-Utilizing the early portion of the growing season can be especially critical for species in colder regions \citep{morin07, dantec15} and the ones with suboptimal shade tolerance \citep{richardson09}.
-Further, plants that budburst earlier can take advantage of a longer growing season and avoid competition \citep{guo22}.
-However, differences across species means that this will not impact all species the same.
-While fall event always exhibited co-gradient relationships (i.e.
-events advanced as we moved northward or up in elevation), spring events were more varied among species.
-The implication of such variations could be that spring leafout sets up flowering, and thus, some counter-gradient could better aid gene glow over larger geographical regions.
-
-All of above are critical to predicting not just phenological change, but range expansions with climate change: An accurate understanding of how environmental conditions might influence species distribution at a continental scale is critical for future range shift prediction.
-Implications of small differences in spring phenology could mean different levels of gene flow.
-
-Our observations reveal that although temperature by itself could serve as a reasonable indicator for predicting when fall events occur across various locations, it's crucial to acknowledge that the degree of adaptation dictated by climate change differs across different geographical areas \citep{Loarie09}.
+% Our observations reveal that although temperature by itself could serve as a reasonable indicator for predicting when fall events occur across various locations, it's crucial to acknowledge that the degree of adaptation dictated by climate change differs across different geographical areas \citep{Loarie09}.
 High-latitude sites, for example, are likely restricted by the impact of photoperiod \citep{Gill15}.
 Diverse mechanisms that govern the timing of leaf senescence at the ecosystem level hold the potential to exert significant influence over not only the interactions occurring within plant communities but also the intricate process of carbon storage within ecosystems.
-With the anticipated escalation of global temperatures in the forthcoming century, these mechanisms are poised to play a pivotal role in shaping the dynamics of plant communities and the overall carbon balance of ecosystems.
 

### 2023-09-24 Docs/manuscript.Rnw
more edits, fixed plots

@@
 \pagestyle{fancy}
-\fancyhead[LO]{July 2023}
+\fancyhead[LO]{2023}
 \fancyhead[RO]{Manuscript}
+\usepackage{lineno} %alinaSep22: this is to number each line (GCB requires it).
+However, I dont know how not to number the title page, and I also would like the number to appear bigger.
+Can lizzie fix this please?
+\linenumbers
+\usepackage{setspace}
+% Using \doublespacing in the preamble changes the text to double-line spacing
+\doublespacing
 % more optionals! %
@@
 
-% Just jotting down and putting some bits you have for abstract for now. 
 \begin{abstract}
@@
 \section{Introduction}
-%emw: FYI this is a first draft -- meaning it still has lots of words, sentences and other bits to fix; but best to do that as we go. 
+
 %emw: I noticed a lot of papers on local adaptation don't always use the term 'local adaptation' so we may want to avoid using it too much?
@@
 %EMWSep4: Check deans96 is accurate ref I added.
-Other refs I added to 'more' bib in case you do not have them.
+%alinaSep22: Hi lizzie I am not sure how you are trying to phrase "they may be more studies and debated in Europe" but I think it's slightly off.
+could you fix it?
 
@@
 The distribution was skewed toward North America (12 North American studies versus 5 European studies).
-See Table.\ref{supplement-table:all_studies} in Supplement for information on selected studies.}
+See Table.\ref{supplement-table:all_studies} in Supplement for information on selected studies.
+Note: map lines delineate study regions and do not necessarily depict accepted national boundaries}
     \label{figure:map_gardens}
@@
 Seven species (five in North America and two in Europe) also had fall event information available.
-%EMWSep4 -- check my edits here; also if we don't have it somewhere in the results yet, please add gymnosperms/angiosperm species # in North America versus Europe here, something like: Most species in North American gardens were gymosperms (7/11 species) while most species were angiosperms in Europe (3/4 species).
+Most species in North American gardens were gymosperms (7/11 species) while most species in European gardens were angiosperms (3/4 species).
+
 <<label=source_spring_models, echo=FALSE>>=
@@
 \ref{figure:springfall_latmat}, Table.
-\ref{supplement-table:model_spring_lat} \& \ref{supplement-table:model_spring_mat} in Supplements), nor within North America (latitude: \Sexpr{format(round(mean(northamerican_springpost_fitC_spring_lat),digits = 2), nsmall=2)} days/degree [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.05), digits = 1)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(mean(northamerican_springpost_fitC_spring_mat),digits = 1)} days/$^{\circ}$C [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.95), digits = 2)}]) or Europe (latitude: \Sexpr{format(round(mean(european_springpost_fitC_spring_lat),digits = 2),nsmall=2)} days/degree [\Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT:\Sexpr{format(round(mean(european_springpost_fitC_spring_mat),digits = 2),nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.95), digits = 2)}]) (Fig.\ref{figure:continental_spptype_lat}A \& \ref{figure:continental_spptype_effect}A).
+\ref{supplement-table:model_spring_lat} \& \ref{supplement-table:model_spring_mat} in Supplements), nor within North America (latitude: \Sexpr{format(round(mean(northamerican_springpost_fitC_spring_lat),digits = 2), nsmall=2)} days/degree [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.05), digits = 1)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(mean(northamerican_springpost_fitC_spring_mat),digits = 1)} days/$^{\circ}$C [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.95), digits = 2)}]) or Europe (latitude: \Sexpr{format(round(mean(european_springpost_fitC_spring_lat),digits = 2),nsmall=2)} days/degree [\Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(round(mean(european_springpost_fitC_spring_mat),digits = 2),nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.95), digits = 2)}]) (Fig.\ref{figure:continental_spptype_lat}A \& \ref{figure:continental_spptype_effect}A).
 Results were similar using other distance metrics in lieu of latitude (see Supplement Fig.\ref{supplement-figure:lat_distance} for results using the difference between provenance and garden latitude, and the spherical distance between provenance and garden).
@@
 
-%EMWSep4 -- small edits here, hopefully I did not screw up the formatting
-In contrast, fall events (e.g., budset, leaf senescence, leaf abscission) were earlier at more norther, cooler MAT sites (that is, they advanced strongly with provenance latitude: \Sexpr{format(round(abs(mean(all_species_fallpost_fitC_fall_lat)),digits = 2), nsmall=2)} days/degree [\Sexpr{format(round(abs(quantile(all_species_fallpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)}-\Sexpr{format(round(abs(quantile(all_species_fallpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}], and with decreasing MAT:\Sexpr{format(round(abs(mean(all_species_fallpost_fitC_fall_mat)),digits = 2), nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(all_species_fallpost_fitC_fall_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(all_species_fallpost_fitC_fall_mat, 0.95), digits = 2)}], Fig.
+In contrast, fall events (e.g., budset, leaf senescence, leaf abscission) were earlier at more northern, cooler MAT sites (that is, they advanced strongly with provenance latitude: \Sexpr{format(round(abs(mean(all_species_fallpost_fitC_fall_lat)),digits = 2), nsmall=2)} days/degree [\Sexpr{format(round(abs(quantile(all_species_fallpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)}-\Sexpr{format(round(abs(quantile(all_species_fallpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}], and with decreasing MAT:\Sexpr{format(round(abs(mean(all_species_fallpost_fitC_fall_mat)),digits = 2), nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(all_species_fallpost_fitC_fall_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(all_species_fallpost_fitC_fall_mat, 0.95), digits = 2)}], Fig.
 \ref{figure:springfall_latmat}, Table.\ref{supplement-table:model_fall_lat} \& \ref{supplement-table:model_fall_mat} in Supplements ).
@@
  \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/continental_spptype_lat.jpg}
-    \caption{Posterior distributions for the effect of provenance latitude across different continents and species types.
+    \caption{Posterior distributions (99\% percentile) for the effect of provenance latitude across different continents and species types.
 Zero---no effect---is shown with a dashed line.
 Solid line and shading in posteriors represent mean and 50 percent interval (90\% intervals given in text).
-(A) Effect of latitude and continent on spring and fall event date (DOY).
+(a \& b) Effect of latitude and continent on spring and fall event date (DOY).
 Fall events advanced strongly per degree northward, particularly notably in North America.
-(B) Effect of latitude and leaf type on spring and fall event date (DOY).
+(c \& d) Effect of latitude and leaf type on spring and fall event date (DOY).
 As we move north, spring events slightly advanced in gymnosperms and delayed in angiosperms.
@@
  \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/continental_spptype_effect.jpg}
-    \caption{Posterior distributions for the effect of MAT across different continents and species types.
+    \caption{Posterior distributions (99\% percentile) for the effect of MAT across different continents and species types.
 Zero---no effect---is shown with a dashed line.
 Solid line and shading in posteriors represent mean and 50 percent interval (90\% intervals given in text).
-(A) Effect of MAT and continent on spring and fall event date (DOY).
+(a \& b) Effect of MAT and continent on spring and fall event date (DOY).
 Fall events advanced strongly with decreasing MAT, particularly notably in North America.
-(B) Effect of MAT and leaf type on spring and fall event date (DOY).
+(c \& d) Effect of MAT and leaf type on spring and fall event date (DOY).
 As MAT increased, spring events slightly advanced in angiosperms and delayed in gymnosperms.
@@
 
-Fall event clines were more similar across species types, but diverged by continent with the common clines of later events in warmer, more southerly latitudes found only in North American studies.
-However, as we were focused on spring events, we had limited data on fall events, which likely explains our results which contrast with meta-analyses of fall events in Europe (citeAlbero?).
-In our dataset, more data on fall events were reported in North America, while only two of the six European gardens reported leaf senescence-related metrics, one studying \emph{Fagus sylvatica} and one studying \emph{Fraxinus excelsior}, and neither showed strong clines.
-In North America studies of both gymnosperm and angiosperm species reported strong clines, and most often focused on budset, while the two European gardens reported data on `leaf senescence' (EXPLAIN what their metric was?, cite Table S1).
-Budset is most closely related to growth cessation and tied to photoperiod signals, making it generally considered a more robust indicator of when plants stop investing in growth and more genetically controlled than leaf senescence \citep{Alberto13}.
-%EMWSep4 -- I am actually not sure if `our results which contrast with meta-analyses of fall events in Europe' is true -- is it?
-I could not tell from reviewing Alberto very quickly.
-If it is true from that paper or another, then I think this is the best explanation of our results.
-If it's not true, I can adjust the text.
-% Also, can you say how the two gardens defined `leaf senescence' inside (EXPLAIN)?
+In our dataset, more data on fall events were reported in North America.
+In Europe, only two gardens reported fall event metrics, one studying \emph{Fagus sylvatica} and one studying \emph{Fraxinus excelsior}.
+Fall events were earlier in colder, more northern latitudes in both continents, which is consistent with previous literature \citep{AitkenBemmels16, Alberto13}.
+North American studies of both gymnosperm and angiosperm species reported strong clines, and most often focused on budset, while the two European gardens reported data on `leaf senescence' (Supplement Table.\ref{supplement-table:all_studies}).
+Leaf senescence was visually inspected based on the percentage of the tree crown that had changed color when the assessor stood on the south-facing side of each tree \citep{rosique22}.
+We note that budset is most closely related to growth cessation and tied to photoperiod signals, making it generally considered a more robust indicator of when plants stop investing in growth and more genetically controlled than leaf senescence \citep{Alberto13}.
 
@@
 In contrast fall event clines in Europe were weak, but their fall event type (leaf senescence) is less likely to show clinal variation.
-Given European studies were biased towards studying angiosperms (3/4 species) that are often canopy-species (e.g., \emph{Fagus}, \emph{Quercus}) these results together could explain a greater focus on clinal variation in spring events across European studies (CITESa).
+Given European studies were biased towards studying angiosperms (3/4 species) that are often canopy-species (e.g., \emph{Fagus}, \emph{Quercus}) these results together could explain a greater focus on clinal variation in spring events across European studies \citep{Basler:2012,gauzere2020,sog08,deans96,von95}.
 Most North American gardens and provenances included in this study were limited to the Pacific Northwest region, and thus tended to focus on species from forests where gymnosperms are almost always the canopy species (\emph{Pseudotsuga, Tsuga, Picea, Pinus}) and angiosperms are much smaller, more opportunistic species (\emph{Alnus, Populus}).
@@
 \ref{figure:map_gardens}).
-Given the tremendous investment required for such gardens, however, it is not surprising they are often limited to one species---most often those relevant for forestry (CITES).
+Given the tremendous investment required for such gardens, however, it is not surprising they are often limited to one species---most often those relevant for forestry (e.g.,\emph{Pseudotsuga menziesii, Tsuja plicata, Tsuga heterophylla, Picea sitchensis, Pinus albicaulis}).
 Our results, however, highlight the limited inference we may gain from such an approach to understand trends across species more broadly.
-%EMWSep4 -- Great point you made about the bias in North American study locations, I built on it. 
-% CITESa -- could be Basler & Korner, Gauzere 2020 paper .. you likely have more ideas to add.
 
@@
 Plastic responses to warming, as our results suggest for spring events, mean species can adjust to the shifting environments of climate change---up to some point.
-Beyond that point theory suggests plastic traits may limit adaptation to climate change (CITE chevin I think?).
-This may mean species will be replaced when their plastic spring phenological responses are no longer adequate, or shifting reaction norms could lead to adaptation (cite Gauzere 2020?).
+Beyond that point theory suggests plastic traits may limit adaptation to climate change \citep{chevin10}.
+This may mean species will be replaced when their plastic spring phenological responses are no longer adequate, or shifting reaction norms could lead to adaptation \citep{gauzere2020}.
 
@@
 Understanding environmental clines in phenology will ultimately be critical to predicting how species shift their ranges as well.
-Implications of small differences in spring phenology could mean different levels of gene flow, while co- and counter-gradient variation have important implications for gene glow over larger geographical regions (CITES), and thus could shape future range shifts.
-Much like frost risk, climate change likely shifts multiple factors at once, as climate change shifts pollinators and wind patterns (CITESb).
+Implications of small differences in spring phenology could mean different levels of gene flow, while co- and counter-gradient variation have important implications for gene glow over larger geographical regions \citep{bach20}, and thus could shape future range shifts.
+Much like frost risk, climate change likely shifts multiple factors at once, as climate change shifts pollinators and wind patterns \citep{kling21}.
 With the anticipated escalation of global temperatures in the forthcoming century, these mechanisms are poised to play a pivotal role in shaping the dynamics of plant communities and the overall carbon balance of ecosystems.
-% CITESb Ackerly has a paper on shifting wind patterns in PNAS in the last couple years to cite here.
 
@@
 \section{Figures}
+%alinaSep22: i dont know how to force all figures to be in this section instead of in between texts
 
+\section{References}
 \bibliography{bibliography_local_adaptation.bib}

### 2023-09-24 Docs/manuscript.tex
more edits, fixed plots

@@
 
-% Just jotting down and putting some bits you have for abstract for now. 
 \begin{abstract}
-Climate change forecasting ....
-plasticity....
-adaptation.
-Spring tree phenology is critical to forecasting ...
+Forecasting the biological impacts of climate change requires understanding how species respond to warmer temperatures through both inter-annual flexible variation versus through adaptation to local conditions.
+Yet we often lack this information entirely, or find conflicting evidence across studies.
+This is the case for one of the most reported and consistent impacts of anthropogenic climate change---earlier spring phenology, which is also one of the most critical to forecasting given its role in carbon sequestration.
+Evidence from some common garden studies suggest important provenance effects, which suggest local adaptation in the underlying cues of spring phenology and supports findings for fall events, while other studies find no evidence.
+Here we synthesize common garden studies across Europe and North America that reported spring event dates for a mix of angiosperm and gymnosperm tree species in the northern hemisphere, capturing data from 384 North American provenances and 101 European provenances with observations from 1962 to 2019, alongside fall event data when provided.
+Across continents we find no evidence of provenance effects in spring phenology, but strong clines with latitude and mean annual temperature (MAT) for fall events.
+These effects, however, appear to diverge by continent and species type (gymnosperm versus angiosperm), especially for fall events where clines with latitude and MAT are much stronger in North America.
+Our results suggest flexible, likely plastic responses, in spring phenology with warming, and potential limits---at least in the short term---due to provenance effects for fall phenology.
+They also highlight that, after over 50 years of common garden studies on tree phenology we may still lack a holistic predictive model of clines across species and phenological events.
+\end{abstract} %Spring tree phenology is critical to forecasting ...
 Despite a long history of study, however, the balance of local adaptation versus plasticity in determining environmental responses is not well understood, especially for spring events.
-We conducted the first cross-continental meta-analysis of published studies from the peer-reviewed literature that reported spring event dates for a mix of angiosperm and gymnosperm tree species in the northern hemisphere, capturing data from 384 North American provenances and 101 European provenances with observations from 1962 to 2019.
+We conducted the first cross-continental meta-analysis of published studies from the peer-reviewed literature that reported spring event dates for a mix of angiosperm and gymnosperm tree species in the northern hemisphere, capturing data  for  seven angiosperm and eight gymnosperm species from 17 studies, from 384 North American provenances and 101 European provenances with observations from 1962 to 2019.
 We found ...
 This is important because...
-\end{abstract} % seven angiosperm and eight gymnosperm species from 17 studies, encompassing 384 North American provenances and 101 European provenances, with observations from 1962 to 2019
 
 \section{Introduction}
-%emw: FYI this is a first draft -- meaning it still has lots of words, sentences and other bits to fix; but best to do that as we go. 
+
 %emw: I noticed a lot of papers on local adaptation don't always use the term 'local adaptation' so we may want to avoid using it too much?
@@
 
-Studies of adaptation versus plasticity in tree phenology have been conducted for centuries \citep{Cleland:2007or}, through common garden studies  (i.e., studies---often for forestry purposes---where researchers grow trees of different geographical origins under the same environmental conditions to disentangle the effects of environmental and genetic variation on trees’ phenology and phenotype) \citep{AitkenBemmels16, Alberto13}.
-This work has established common clines in fall phenology, as source locations with shorter growing season (poleward and higher elevations) exhibit earlier stopping of growth (e.g., budset).
-Research has connected these clines to an underlying proximate mechanism of changing photoperiod cues (i.e., shifts in the photoperiod threshold that trigger budset), driven by local adaptation to the local growing season \citep{Alberto13,Savolainen07}.
-In contrast, spring phenology appears far more plastic \citep{AitkenBemmels16} and is determined more strongly by temperature \citep{flynn18}.
-Many studies, however, have argued that spring phenology shows high enough levels of adaptation to matter biologically.
-% alinaJuly20: im not sure what important levels of adaptation means here, %emwaug1: I meant that the 'amount' of local adaptation is high enough to matter (contrasting with the reality that if you have a big enough sample you can often find a small difference that is 'significant' but it might not matter biologically).
+Studies of adaptation versus plasticity in tree phenology have been conducted for centuries \citep{Cleland:2007or}, through common garden studies.
+In these studies---conducted often for forestry purposes---researchers grow trees of different geographical origins under the same environmental conditions to disentangle the effects of environmental and genetic variation on trees’ phenology and phenotype \citep{AitkenBemmels16, Alberto13}.
+Such work has established common clines in fall phenology, as source locations with shorter growing seasons (poleward and higher elevations) exhibit earlier stopping of growth (such as budset).
+Research has connected these clines to an underlying proximate mechanism of changing photoperiod cues (i.e., shifts in the photoperiod threshold required to trigger budset), driven by adaptation to the local growing season \citep{Alberto13,Savolainen07}.
+In contrast, spring phenology appears more plastic \citep{AitkenBemmels16} and is determined more strongly by temperature \citep{flynn18}.
+Many studies, however, have argued that spring phenology shows levels of adaptation that may be critical to forecasting and mitigation \citep{vitasse2009,Basler:2012}.
 
-Results, however, have not always been consistent.
-Studies have documented provenance differences of 2-4 days per degree latitude in spring phenology for some species (\emph{Picea abies} in \citealp{sog08} and \emph{Quercus petraea} in \citealp{deans96}) while others have not found any trends in budburst along latitudinal gradients (for example, \emph{Picea sitchensis} in \citealp{mimura07}, \emph{Picea glauca} in \citealp{Li97}, and \emph{Populus balsamifera} in \citealp{farmer93}).
-This has led to some debate over the prevalence and importance of adaptation in spring tree phenology.
-While clines of spring phenology have been found in studies in both Europe \citep{sog08,deans96,von95} and North America \citep{rossi15, soo13, hannerz99}, they could vary by continent.
+These contrasting studies highlight how inconsistent evidence for adaptation in tree spring phenology has been.
+Studies have documented provenance differences of 2-4 days per degree latitude in spring phenology for some species (\emph{Picea abies} in \citealp{sog08} and \emph{Quercus petraea} in \citealp{deans96}) while others have failed to find similar trends along latitudinal gradients (for example, \emph{Picea sitchensis} in \citealp{mimura07}, \emph{Picea glauca} in \citealp{Li97}, and \emph{Populus balsamifera} in \citealp{farmer93}).
+This has led to debate over the prevalence and importance of adaptation in spring tree phenology.
+While clines of spring phenology have been found in studies in both Europe \citep{sog08,deans96,von95} and North America \citep{rossi15, soo13, hannerz99}, they may be more studies and debated in Europe \citep{deans96,vitasse2009,Basler:2012}, raising the possibility that they could vary by continent.
+%EMWSep4: Check deans96 is accurate ref I added.
+%alinaSep22: Hi lizzie I am not sure how you are trying to phrase "they may be more studies and debated in Europe" but I think it's slightly off.
+could you fix it?
 
-Continental differences in patterns of adaptation versus plasticity could be driven by climatic differences, as North American springs are more variable across years than European ones \citep{tward21,zohner2017,schwartz00}.
+Continental differences in patterns of adaptation versus plasticity could be driven by climatic differences, especially as North American springs are more variable across years than European ones \citep{tward21,zohner2017,schwartz00}.
 Such high temporal variability means that distant sites can effectively experience the same spring climate, but in different years.
-Studies of spring phenology in arboreta suggest cues for budburst may vary depending on continental climate \citep{zohner2017}, but are poorly controlled compared to traditional provenance trials, making them difficult to use for tests of plasticity versus adaptation \citep{gauzere2020}.
-Even for more carefully designed common garden differences in study design---especially studied species---may complicate understanding what underlies potential trends across continents, with European studies focusing more on deciduous angiosperms and North American studies focusing often on evergreen gymnosperms.
+Studies of spring phenology in arboreta suggest cues for budburst may vary depending on continental climate \citep{zohner2017}, but are poorly controlled compared to traditional common garden studies, making them difficult to use for tests of plasticity versus adaptation \citep{gauzere2020}.
+Even for more carefully designed common gardens, differences in species studied or other differences in design may complicate understanding what underlies potential trends across continents.
+%EMWSep4: Moved last point to discussion.
 
@@
 \section{Results}
-Our final dataset included seven angiosperm and eight gymnosperm species from 17 studies, encompassing 384 North American provenances and 101 European provenances, with observations from 1962 to 2019.
-Seven species also had fall event information available.
+Our final dataset included seven deciduous angiosperm and eight evergreen gymnosperm species from 17 studies and 19 gardens, encompassing 384 North American provenances and 101 European provenances, with observations from 1962 to 2019.
+Seven species (five in North America and two in Europe) also had fall event information available.
+Most species in North American gardens were gymosperms (7/11 species) while most species in European gardens were angiosperms (3/4 species).
 
@@
 \ref{figure:springfall_latmat}, Table.
-\ref{supplement-table:model_spring_lat} \& \ref{supplement-table:model_spring_mat} in Supplements), nor within North America (latitude: 0.10 days/degree [-0.06 - 0.26]; MAT: -0.09 days/$^{\circ}$C [-0.36 - 0.18]) or Europe (latitude: 0.10 days/degree [-0.23 - 0.42]; MAT:-0.16 days/$^{\circ}$C [-0.55 - 0.23]) (Fig.\ref{figure:continental_spptype_lat}A \& \ref{figure:continental_spptype_effect}A).
+\ref{supplement-table:model_spring_lat} \& \ref{supplement-table:model_spring_mat} in Supplements), nor within North America (latitude: 0.10 days/degree [-0.06 - 0.26]; MAT: -0.09 days/$^{\circ}$C [-0.36 - 0.18]) or Europe (latitude: 0.10 days/degree [-0.23 - 0.42]; MAT: -0.16 days/$^{\circ}$C [-0.55 - 0.23]) (Fig.\ref{figure:continental_spptype_lat}A \& \ref{figure:continental_spptype_effect}A).
 Results were similar using other distance metrics in lieu of latitude (see Supplement Fig.\ref{supplement-figure:lat_distance} for results using the difference between provenance and garden latitude, and the spherical distance between provenance and garden).
@@
 
-In contrast, fall events (e.g., budset, leaf senescence, leaf abscission) advanced strongly with provenance latitude (3.16 days/degree [2.87-3.45]) and the decrease in MAT (4.78 days/$^{\circ}$C [4.1 - 5.4]), meaning fall events were earlier where provenance MAT was lower (Fig.
+In contrast, fall events (e.g., budset, leaf senescence, leaf abscission) were earlier at more northern, cooler MAT sites (that is, they advanced strongly with provenance latitude: 3.16 days/degree [2.87-3.45], and with decreasing MAT:4.78 days/$^{\circ}$C [4.1 - 5.4], Fig.
 \ref{figure:springfall_latmat}, Table.\ref{supplement-table:model_fall_lat} \& \ref{supplement-table:model_fall_mat} in Supplements ).
@@
 In contrast, fall events generally showed much stronger clines, especially in North America, and in support of many previous studies \citep{AitkenBemmels16, Alberto13}.
-While previous studies have suggested spring events are far more plastic compared to fall events \citep{Li97,farmer93,mimura07}, our study provides the first major test of this across continents and species and suggests overall no general trend for clines in spring phenology.
-While not a direct test, our results suggest that spring events are highly plastic; they thus predict that warming springs will continue to be tracked more closely phenologically by trees than warming fall temperatures, in line with most current climate change impact studies \citep{IPCC22}
+While previous studies have suggested spring events are far more plastic compared to fall events \citep{Li97,farmer93,mimura07}, our study provides the first major test of this across continents and species and suggests no general trend for clines in spring phenology.
+Our results thus predict that warming springs will continue to be tracked more closely phenologically by trees than warming fall temperatures \citep{IPCC22}
 
-Trends between spring phenology and latitude or MAT were weak, but suggested the possibility of diverging results that could mute an overall trend (albeit a much weaker one than for fall phenology).
-Clines with MAT diverged between angiosperms and gymnosperms.
-Combined with our finding of stronger clines for fall phenology in North America, our results suggest that data on more species tested across continents could yield important insights into plasticity versus adaptation limits for spring and fall phenology, which would greatly aid forecasting.
+Trends between spring phenology and latitude or MAT were weak, but suggested the possibility of diverging results that could mute an overall trend---albeit a much weaker one than for fall phenology.
+We found angiosperm versus gymnosperm species diverged in their clines with MAT.
+Combined with our finding of much stronger clines for fall phenology in North America, these results support the idea of potential variation across continents and/or species type that may underlie the debate in whether spring events show important clinal variation.
+As clines with spring events were very weak, however, and gardens almost always focus on only one species, understanding these diverging results well enough to aid forecasting would take significant additional investment in common garden studies.
+%EMWSep4 -- we need to be careful in how many times we say 'we need more data' -- it's a common need, so it's better to focus on specifics of what the data suggests, or doesn't and say we need more data -- at most once in a paper.
+More on this below.
 
-We found the coarse metrics of provenance latitude and MAT were generally good predictors of phenology, performing better than climate overlap.
-Latitude and MAT appeared to well represent how similar the climates are between the provenances and gardens in months close to the events, at least in the narrow range of northern temperature and boreal forests we studied.
+We found the coarse metrics of provenance latitude and MAT were generally good predictors of phenology, performing better than our more complex and data-rich metric of climate overlap.
+Latitude and MAT appeared to well represent how similar the climates are between the provenances and gardens in the temperate and boreal forest species we studied, yielding similar results metrics calculated specifically in the spring with daily climate data.
+%EMWSep4 -- also avoid repeating how narrow the data is etc.
+-- it's good to point out limitations, but repeating them too much is not helpful -- this is the biggest dataset for this question on earth, we need to focus on what we learned from it, not so much on the limitations.
+Also, Europeans would not call all of the forests northern.
 
-\subsection{Variation across continents and species groupings}
+\subsection{Variation across continents and species types}
 
-Our results indicated that the effects of provenance latitude and MAT on fall events were similar across angiosperm and gymnosperm species: Fall events delay as MAT increases for both leaf types, slightly more so for gymnosperms, explained by sensitivity to day-length signals.
-However, the two leaf types exhibited slightly diverging trends in spring events:  advancing in angiosperms and delaying in most gymnosperms as MAT increases.
-Such differences make sense as timing of budburst and leafout varies by leaf type.
-Gymnosperms photosynthesize without leafout and thus generally leafout later than angiosperms, after much risk of spring frost \citep{panchen14}.
-Meanwhile, xylem anatomy also plays a part in determining spring timing due to the importance of vessel diameter to embolism risks induced by frost in spring.
+Our results highlight that spring events show much weaker clines than fall events in tree species, but suggest important variation between species types in spring events.
+Angiosperms tended to budburst earlier in provenances that were warmer and more southerly, while gymnosperms trended in the opposite direction.
+Such differences could be driven by their varying investment strategies.
+As gymnosperms photosynthesize without leafout they generally leafout much later than angiosperms, after most risk of spring frost \citep{panchen14}, and thus may avoid frost risk more than angiosperms.
+In contrast, angiosperms may tend towards earlier leafout in warmer climates to compete best for access to light and other resources \citep{cat2019}.
+Testing these hypotheses would require More information on frost risk and forest community assembly from across the provenance locations.
+%EMWSep4 -- I like the below (and I think it may have been my idea), but we'd likely be asked to test for it ...
+and I am not 100% sure it makes sense -- both angiosperms and conifers need to activate their xylem to photosynthesize I believe.
+% Meanwhile, xylem anatomy also plays a part in determining spring timing due to the importance of vessel diameter to embolism risks induced by frost in spring.
 According to \citet{salk20}, species with narrow xylem such as coniferous gymnosperms and diffuse-porous angiosperms did not exhibit the consistent phenological trends shown in ring-porous species.
@@
 
-While we noticed more prominent clines in fall events in North America, there was no strong evidence of overall continental differences in spring events, which might again suggest a higher degree of plasticity in spring events.
-Meanwhile, most North American gardens and provenances included in this study were limited to the Pacific Northwest region, which might mask the kind of climate pattern we were hoping to see from the entirety of North America (Fig.
+In our dataset, more data on fall events were reported in North America.
+In Europe, only two gardens reported fall event metrics, one studying \emph{Fagus sylvatica} and one studying \emph{Fraxinus excelsior}.
+Fall events were earlier in colder, more northern latitudes in both continents, which is consistent with previous literature \citep{AitkenBemmels16, Alberto13}.
+North American studies of both gymnosperm and angiosperm species reported strong clines, and most often focused on budset, while the two European gardens reported data on `leaf senescence' (Supplement Table.\ref{supplement-table:all_studies}).
+Leaf senescence was visually inspected based on the percentage of the tree crown that had changed color when the assessor stood on the south-facing side of each tree \citep{rosique22}.
+We note that budset is most closely related to growth cessation and tied to photoperiod signals, making it generally considered a more robust indicator of when plants stop investing in growth and more genetically controlled than leaf senescence \citep{Alberto13}.
+
+%EMWSep4 -- I like to have one paragraph in a paper (usually) focused on biases and limitations, so the below is that paragraph.
+I also aimed for it to have more of a message specific to this work beyond 'we need more data' (as we always need more data, you can write that in any paper).
+Taken together, these results could explain some of the existing debate on the strength of spring clinal variation, but also highlight how data biases make disentangling the drivers of variation difficult.
+Angiosperm species showed a weak trend towards earlier budburst in populations from warmer locations.
+In contrast fall event clines in Europe were weak, but their fall event type (leaf senescence) is less likely to show clinal variation.
+Given European studies were biased towards studying angiosperms (3/4 species) that are often canopy-species (e.g., \emph{Fagus}, \emph{Quercus}) these results together could explain a greater focus on clinal variation in spring events across European studies \citep{Basler:2012,gauzere2020,sog08,deans96,von95}.
+Most North American gardens and provenances included in this study were limited to the Pacific Northwest region, and thus tended to focus on species from forests where gymnosperms are almost always the canopy species (\emph{Pseudotsuga, Tsuga, Picea, Pinus}) and angiosperms are much smaller, more opportunistic species (\emph{Alnus, Populus}).
+This might suggest a far different pattern if gardens included more evenly sampled North American tree species (Fig.
 \ref{figure:map_gardens}).
+Given the tremendous investment required for such gardens, however, it is not surprising they are often limited to one species---most often those relevant for forestry (e.g.,\emph{Pseudotsuga menziesii, Tsuja plicata, Tsuga heterophylla, Picea sitchensis, Pinus albicaulis}).
+Our results, however, highlight the limited inference we may gain from such an approach to understand trends across species more broadly.
 
-\subsection{Adaptation in spring vs fall phenology: Implications for climate change responses}
+\subsection{Adaptation in tree phenology: Implications for climate change responses}
 
-Our results suggest that current advances in spring phenology may continue, but delays in fall phenology could depend on the speed of adaptation.
-While some previous chamber studies have found potential adaptation, these could have been influenced by chilling at the provenance site \citep{ettinger20}.
+%EMWSep4 -- good stuff here!
+I tried to refocus on spring events some as our data is focused there.
+And I limited the amount of hand-wavey stuff we discussed as reviewers only usually want so much extrapolation of your results (so I cut some of the ending stuff).
+Our results suggest that current advances in spring phenology will likely continue, but predicting any shifts due to provenance effects will require new data across more species.
+Plastic responses to warming, as our results suggest for spring events, mean species can adjust to the shifting environments of climate change---up to some point.
+Beyond that point theory suggests plastic traits may limit adaptation to climate change \citep{chevin10}.
+This may mean species will be replaced when their plastic spring phenological responses are no longer adequate, or shifting reaction norms could lead to adaptation \citep{gauzere2020}.
+
+Predicting this will require better understanding how different species strategies covary with early-season risks, and how such risks may shift with continued climate change.
+For temperate zones, many species strategies appear designed to cope with frost, either via avoidance or tolerance \citep{alberto11, lenz16, allevato19}, as utilizing the early portion of the growing season can be especially critical for species in colder regions \citep{morin07, dantec15}.
+Climate change at once shifts spring phenology---thus potentially changing each species frost risk---but also appears to reshape the drivers of frost climatically \citep{cat2021pep}.
+Layered onto this are possible shifts in early-season herbivory with warming \citep{meineke2019}, and trade-offs in early-season risks with access to a long growing season for growth and reproduction, with some species avoiding competition through being very early \citep{guo22}.
+Common garden studies that track and report frost and herbivory damage, alongside timing of flowering and fruiting, could help to understand the complex fitness landscape of spring phenology.
+
+In contrast to spring events which were weak and variable, fall events exhibited strong clines, which appeared co-gradient with the environment (i.e.
+late-season events advanced earlier with northward latitudes).
+This supports previous results and the general theory that fall events are consistently designed to avoid tissue loss at the end of the season, when the cost of such losses could be particularly high \citep{AitkenBemmels16, Alberto13}.
+Spring events showed no clear trends, but Increasing research into counter-gradient variation for flowering events could be relevant when spring vegetative and flowering events are linked \citep{dan2021nph}.
+
+Understanding environmental clines in phenology will ultimately be critical to predicting how species shift their ranges as well.
+Implications of small differences in spring phenology could mean different levels of gene flow, while co- and counter-gradient variation have important implications for gene glow over larger geographical regions \citep{bach20}, and thus could shape future range shifts.
+Much like frost risk, climate change likely shifts multiple factors at once, as climate change shifts pollinators and wind patterns \citep{kling21}.
+With the anticipated escalation of global temperatures in the forthcoming century, these mechanisms are poised to play a pivotal role in shaping the dynamics of plant communities and the overall carbon balance of ecosystems.
+
+% All of above are critical to predicting not just phenological change, but range expansions with climate change ...
+An accurate understanding of how environmental conditions might influence species distribution at a continental scale is critical for future range shift prediction.
+
+%EMWSep4 -- I know I wanted this included, but not sure it works. Let's leave out for now. 
+%  While some previous chamber studies have found potential adaptation, these could have been influenced by chilling at the provenance site \citep{ettinger20}.
 However, field studies and common garden experiences do not have this effect.

### 2023-09-24 Docs/supplement.Rnw
more edits, fixed plots

@@
 \fancyhead[LO]{2023}
-\fancyhead[RO]{Manuscript}
+\fancyhead[RO]{Supplement}
+\usepackage{setspace}
+% Using \doublespacing in the preamble changes the text to double-line spacing
+\doublespacing
 % more optionals! %
@@
 
-\begin{figure}[!h] 
-    \centering
- \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/gdd_ms.png}
-    \caption{Growing Degree Days (GDD)on each day of spring event in relation to provenance latitude and MAT, coded by symbol for species and color for garden with linear fits from hierarchical Bayesian models.}
-    \label{figure:gdd}
-\end{figure}
+% \begin{figure}[!h] 
+%     \centering
+%  \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/gdd_ms.png}
+%     \caption{Growing Degree Days (GDD)on each day of spring event in relation to provenance latitude and MAT, coded by symbol for species and color for garden with linear fits from hierarchical Bayesian models.}
+%     \label{figure:gdd}
+% \end{figure}
 
-\begin{figure}[!h] 
-    \centering
- \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_supp/climate_overlap_continent_comparison.png}
-    \caption{The closer a garden is to a provenance, the more overlap in temperature. 
-Higher extend of climate overlap in European studies.}
-    \label{figure:overlap_continent}
-\end{figure}
-
-\begin{figure}[!h] 
-    \centering
- \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_supp/prov_lat_mat_continent.png}
-    \caption{Placeholder.}
-    \label{figure:lat_mat_continent}
-\end{figure}
-
-\begin{figure}[!h] 
-    \centering
- \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_supp/spp_type_continent.png}
-    \caption{Placeholder}
-    \label{figure:spp_continent}
-\end{figure}
-
+% 
+% \begin{figure}[!h] 
+%     \centering
+%  \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_supp/climate_overlap_continent_comparison.png}
+%     \caption{The closer a garden is to a provenance, the more overlap in temperature. 
+% Higher extend of climate overlap in European studies.}
+%     \label{figure:overlap_continent}
+% \end{figure}
+% 
+% \begin{figure}[!h] 
+%     \centering
+%  \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_supp/prov_lat_mat_continent.png}
+%     \caption{Placeholder.}
+%     \label{figure:lat_mat_continent}
+% \end{figure}
+% 
+% 
+% \begin{figure}[!h] 
+%     \centering
+%  \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_supp/spp_type_continent.png}
+%     \caption{Placeholder}
+%     \label{figure:spp_continent}
+% \end{figure}
+% 
  \end{document}

### 2023-09-24 Analyses/source_fall_models_for_sweave.R

@@ -101,6 +101,7 @@ color_scheme_set("yellow")
 new_labels_leaftype <- c("Angiosperm","Gymnosperm")
 mcmc_areas(leaftypeplot_fitC_fall_mat)+
   plot_title +
+  lims(x= c(-2,8))+
   theme(axis.text.x = element_text(size = 35))+             # x-axis text size
   theme(axis.text.y = element_text(size = 35))   +          # y-axis text size
   theme(plot.title = element_text(size = 21))  +            # plot title
@@ -151,6 +152,7 @@ color_scheme_set("yellow")
 new_labels_continent <- c("Europe","North America")
 mcmc_areas(continentplot_fitC_fall_mat)+ 
   plot_title +
+  lims(x= c(-2,8))+
   theme(axis.text.x = element_text(size = 35))+             # x-axis text size
   theme(axis.text.y = element_text(size = 35))   +          # y-axis text size
   theme(plot.title = element_text(size = 21))  +            # plot title
@@ -212,6 +214,7 @@ color_scheme_set("yellow")
 new_labels_leaftype <- c("Angiosperm","Gymnosperm")
 mcmc_areas(leaftypeplot_fitC_fall_lat)+ 
   plot_title +
+  lims(x= c(-5,1))+
   theme(axis.text.x = element_text(size = 35))+             # x-axis text size
   theme(axis.text.y = element_text(size = 35))   +          # y-axis text size
   theme(plot.title = element_text(size = 21))  +            # plot title
@@ -253,6 +256,7 @@ color_scheme_set("yellow")
 new_labels_continent <- c("Europe","North America")
 mcmc_areas(continentplot_fitC_fall_lat)+ 
   plot_title +
+  lims(x= c(-5,1))+
   theme(axis.text.x = element_text(size = 35))+             # x-axis text size
   theme(axis.text.y = element_text(size = 35))   +          # y-axis text size
   theme(plot.title = element_text(size = 21))  +            # plot title
@@ -319,6 +323,7 @@ color_scheme_set("yellow")
 new_labels_leaftype <- c("Angiosperm","Gymnosperm")
 mcmc_areas(leaftypeplot_fitC_fall_overlap)+ 
   plot_title +
+  lims(x= c(-0.5,1.5))+
   theme(axis.text.x = element_text(size = 35))+             # x-axis text size
   theme(axis.text.y = element_text(size = 35))   +          # y-axis text size
   theme(plot.title = element_text(size = 21))  +            # plot title

### 2023-09-24 Analyses/source_spring_models_for_sweave.R

@@ -123,6 +123,7 @@ color_scheme_set("viridis")
 new_labels_leaftype <- c("Angiosperm","Gymnosperm")
 mcmc_areas(leaftypeplot_fitC_spring_mat)+ 
   plot_title +
+  lims(x= c(-2,8))+
   theme(axis.text.x = element_text(size = 35))+             # x-axis text size
   theme(axis.text.y = element_text(size = 35))   +          # y-axis text size
   theme(plot.title = element_text(size = 21))  +            # plot title
@@ -145,6 +146,7 @@ color_scheme_set("mix-teal-pink")
 new_labels_leaftype <- c("Angiosperm","Gymnosperm")
 mcmc_areas(leaftypeplot_fitC_spring_mat)+ 
   plot_title +
+  lims(x= c(-2,8))+
   theme(axis.text.x = element_text(size = 35))+             # x-axis text size
   theme(axis.text.y = element_text(size = 35))   +          # y-axis text size
   theme(plot.title = element_text(size = 21))  +            # plot title
@@ -195,6 +197,7 @@ color_scheme_set("mix-teal-pink")
 new_labels_continent <- c("Europe","North America")
 mcmc_areas(continentplot_fitC_spring_mat)+ 
   plot_title +
+  lims(x= c(-2,8))+
   theme(axis.text.x = element_text(size = 35))+             # x-axis text size
   theme(axis.text.y = element_text(size = 35))   +          # y-axis text size
   theme(plot.title = element_text(size = 21))  +            # plot title
@@ -254,6 +257,7 @@ color_scheme_set("mix-teal-pink")
 new_labels_leaftype <- c("Angiosperm","Gymnosperm")
 mcmc_areas(leaftypeplot_fitC_spring_lat)+ 
   plot_title +
+  lims(x= c(-5,1))+
   theme(axis.text.x = element_text(size = 35))+             # x-axis text size
   theme(axis.text.y = element_text(size = 35))   +          # y-axis text size
   theme(plot.title = element_text(size = 21))  +            # plot title
@@ -304,6 +308,7 @@ color_scheme_set("mix-teal-pink")
 new_labels_continent <- c("Europe","North America")
 mcmc_areas(continentplot_fitC_spring_lat)+ 
   plot_title +
+  lims(x= c(-5,1))+
   theme(axis.text.x = element_text(size = 35))+             # x-axis text size
   theme(axis.text.y = element_text(size = 35))   +          # y-axis text size
   theme(plot.title = element_text(size = 21))  +            # plot title

### 2023-09-30 Docs/manuscript.Rnw
Edits to ms, seems pretty close!

@@
 Can lizzie fix this please?
-\linenumbers
 \usepackage{setspace}
@@
 \SweaveOpts{concordance=TRUE} % For RStudio hiccups
-\title{{\huge Changes and trends in budburst and leaf flush across Europe and North America} \\A meta-analysis of local adaptation in spring phenology studies}
+\title{{\huge Weak evidence of provenance effects in spring phenology across Europe and North America}
 \author{Ziyun Zeng \& E. M. Wolkovich}
@@
 maybe New Phytologist but I think they could be picky.
-
+\linenumbers
 \begin{abstract}
@@
 Yet we often lack this information entirely, or find conflicting evidence across studies.
-This is the case for one of the most reported and consistent impacts of anthropogenic climate change---earlier spring phenology, which is also one of the most critical to forecasting given its role in carbon sequestration.
-Evidence from some common garden studies suggest important provenance effects, which suggest local adaptation in the underlying cues of spring phenology and supports findings for fall events, while other studies find no evidence.
+The latter is the case for shifts in spring phenology---one of the most reported and consistent impacts of anthropogenic climate change, and also one of the most critical to forecasting, given its role in carbon sequestration.
+Some common garden studies have found evidence of important provenance effects, which suggest there may be local adaptation in the underlying cues of spring phenology and mirrors findings for fall events, while other studies find no evidence.
 Here we synthesize common garden studies across Europe and North America that reported spring event dates for a mix of angiosperm and gymnosperm tree species in the northern hemisphere, capturing data from 384 North American provenances and 101 European provenances with observations from 1962 to 2019, alongside fall event data when provided.
@@
 Our results suggest flexible, likely plastic responses, in spring phenology with warming, and potential limits---at least in the short term---due to provenance effects for fall phenology.
-They also highlight that, after over 50 years of common garden studies on tree phenology we may still lack a holistic predictive model of clines across species and phenological events.
+They also highlight that, after over 50 years of common garden studies on tree phenology, we still lack a holistic predictive model of clines across species and phenological events.
 \end{abstract} %Spring tree phenology is critical to forecasting ...
@@
 Predicting the biological impacts of climate change has made understanding how organisms cope with environmental variation more urgent \citep{botero15}.
-In particular, the relative importance of plasticity versus genetic adaptation is vital for prediction \citep{chevin10}, with plasticity expected to allow species to shift more rapidly with climate change than environmental responses based on local adaptation, but possibly limiting responses after the limits of plasticity are reached \citep{chevin102,snell18}.
+In particular, the relative importance of plasticity versus genetic adaptation is vital for prediction \citep{chevin10}, with plasticity expected to allow species to shift more rapidly with climate change than environmental responses based on local adaptation, but possibly stalling responses after the limits of plasticity are reached \citep{chevin102,snell18}.
 
@@
 
-Tree phenology is critically important to climate change forecasting at both the community and ecosystem levels.
+Tree phenology is important to climate change forecasting at both the community and ecosystem levels.
 The timing of budburst and senescence can impact plant competition, plant invasions and community assembly \citep{fridley12}.
-These effects often occur as phenology affects growth \citep{myneni97}, scaling up to impact ecosystem-level carbon sequestration \citep{Barichivich12}, and thus is integrated into forecasts of climate change itself.
+Shifts in phenology can affect tree growth \citep{myneni97}, scaling up to impact ecosystem-level carbon sequestration \citep{Barichivich12}, and thus forecasts of climate change.
 Growing evidence, however, suggests links between growth and phenology are not as consistent as previously predicted---or currently modelled \citep{dow22}---with recent work suggesting how much spring versus fall events shift may determine impacts on tree growth \citep{zohner23}.
@@
 Studies of adaptation versus plasticity in tree phenology have been conducted for centuries \citep{Cleland:2007or}, through common garden studies.
-In these studies---conducted often for forestry purposes---researchers grow trees of different geographical origins under the same environmental conditions to disentangle the effects of environmental and genetic variation on trees’ phenology and phenotype \citep{AitkenBemmels16, Alberto13}.
+In these studies---conducted often for forestry purposes---researchers grow trees of different geographical origins under the same environmental conditions to disentangle the effects of environmental and genetic variation on trees’ phenotypes \citep{AitkenBemmels16, Alberto13}.
 Such work has established common clines in fall phenology, as source locations with shorter growing seasons (poleward and higher elevations) exhibit earlier stopping of growth (such as budset).
 Research has connected these clines to an underlying proximate mechanism of changing photoperiod cues (i.e., shifts in the photoperiod threshold required to trigger budset), driven by adaptation to the local growing season \citep{Alberto13,Savolainen07}.
-In contrast, spring phenology appears more plastic \citep{AitkenBemmels16} and is determined more strongly by temperature \citep{flynn18}.
+In contrast, spring phenology appears more plastic \citep{AitkenBemmels16} and determined more strongly by temperature \citep{flynn18}.
 Many studies, however, have argued that spring phenology shows levels of adaptation that may be critical to forecasting and mitigation \citep{vitasse2009,Basler:2012}.
@@
 This has led to debate over the prevalence and importance of adaptation in spring tree phenology.
-While clines of spring phenology have been found in studies in both Europe \citep{sog08,deans96,von95} and North America \citep{rossi15, soo13, hannerz99}, they may be more studies and debated in Europe \citep{deans96,vitasse2009,Basler:2012}, raising the possibility that they could vary by continent.
+Though clines of spring phenology have been found in both Europe \citep{sog08,deans96,von95} and North America \citep{rossi15, soo13, hannerz99}, there is continuing debate, especially in Europe \citep{deans96,vitasse2009,Basler:2012}, raising the possibility that they could vary by continent.
 %EMWSep4: Check deans96 is accurate ref I added.
@@
 could you fix it?
+%EMWSep30: Yes, I have been struggling with this sentence .... it's tricky, but I think this works now.  Let me know what you think.
 
@@
 Such high temporal variability means that distant sites can effectively experience the same spring climate, but in different years.
-Studies of spring phenology in arboreta suggest cues for budburst may vary depending on continental climate \citep{zohner2017}, but are poorly controlled compared to traditional common garden studies, making them difficult to use for tests of plasticity versus adaptation \citep{gauzere2020}.
+Studies of spring phenology in arboreta suggest cues for budburst may vary depending on continental climate \citep{zohner2017}, but are poorly controlled compared to traditional common garden studies, making them difficult to use for inference of plasticity versus adaptation \citep{gauzere2020}.
 Even for more carefully designed common gardens, differences in species studied or other differences in design may complicate understanding what underlies potential trends across continents.
@@
 To test for evidence of adaptation in spring phenology and what factors may underlie differences observed across studies, we comprehensively examined clines for spring events, including fall events when possible.
-We tested for evidence of adaptation via provenance trends with latitude and climate and examined possible factors that underlie these clines, asking:
-\begin{itemize}
-\item How different are clines between spring and fall phenology? 
-\item How different are clines between Europe and North America?
-\item How different are clines between angiosperm and gymnosperm species?
-\end{itemize}
+We tested for evidence of adaptation via provenance trends with latitude and climate and examined possible factors that underlie these clines, including for differences between: (1) spring and fall phenology, (2) studies in Europe and North America, (3) angiosperm and gymnosperm species.
 To address these questions, we combined Bayesian hierarchical models with a new meta-analysis of all common garden experiments in temperate tree species across Europe and North America reporting spring phenology.
+%EMWSep30: Questions were so similar, so I simplified into once sentence/
 
@@
 We also contacted authors of previous review papers \citep{AitkenBemmels16, Alberto13}, to help further search the literature.
-We then reviewed the methods and results of all publications to refine to only studies that met the following criteria: (a) focused on woody plants originating from either Europe or North America, (b) had provenance trials/common gardens on the same continent, (c) reported latitude and longitude of provenances and gardens, and (d) reported spring events in units of calendar days (day of year or DOY) or could be converted into DOY (see \nameref{supplement-section:addmethods} in the Supplements).
+We then reviewed the methods and results of all publications to refine to only studies that met the following criteria: (a) focused on woody plants originating from either Europe or North America (also the locations of most studies), (b) had provenance trials/common gardens on the same continent, (c) reported latitude and longitude of provenances and gardens, and (d) reported spring events in units of calendar days (day of year or DOY) or could be converted into DOY (see \nameref{supplement-section:addmethods} in the Supplements).
 
@@
 \citealp{Rehfeldt1994}), we converted such dates to DOY using the `lubridate' package in R \citep{Grolemund11}.
-% emwJun4 -- you need to use ` to open single quotes in Latex.
-\begin{figure}[!h] 
-    \centering
- \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/map_gardens.jpg}
-    \caption{Distribution of common gardens (denoted as stars) and provenances (denoted as circles) included in this meta-analysis.
-The distribution was skewed toward North America (12 North American studies versus 5 European studies).
-See Table.\ref{supplement-table:all_studies} in Supplement for information on selected studies.
-Note: map lines delineate study regions and do not necessarily depict accepted national boundaries}
-    \label{figure:map_gardens}
-\end{figure}
-\newline
+
 To understand how climatic differences, in addition to geographical differences, shape local adaptation in spring events we extracted several types of climate data using information about provenance latitude, longitude, and elevation from original publications.
@@
 
-To examine more explicitly climate near spring events, we used gridded daily temperature data for 2011 to 2020 for all European and North American provenances and gardens from E-OBS and the `daymetr' package in R respectively \citep{cornes2018,hufkens2018}.
-Using these data we estimated how much the daily temperatures overlapped between garden and provenance locations, which we call `climate overlap.' For this we used the `overlap' package in R to calculate the percentage overlap of the daily temperature of each provenance and their corresponding gardens in spring months (March to May) from 2011 to 2020.
-% Finally, using the daily temperature data, we also calculated growing degree days (GDD), a commonly used heat accumulation measure to forecast phenological development in plants \citep{miller01}, for each provenance and garden, based on the accumulation of mean daily temperatures ($T_{m}$)from 2011 to 2020 above a baseline of 0$^{\circ}$C, from January 1 until budburst and leaf flush with the following formula: GDD = $\sum$($T_{m}$ - $0^{\circ}$C for $T_{m}$ $\ge$ $0^{\circ}$C; 0 for $T_{m}$ $\le$ $0^{\circ}$C).
+To examine climate near spring events more explicitly than MAT allows, we used gridded daily temperature data for March-May for 2011 to 2020 for all provenances and gardens.
+We extracted data from E-OBS for European locations, and using the `daymetr' in R for North American locations \citep{cornes2018,hufkens2018}.
+Then, using these data and the `overlap' package in R, we estimated how much the daily temperatures overlapped between each provenance location and their corresponding gardens across the three months for our 10 year period, which we call `climate overlap.'  % Finally, using the daily temperature data, we also calculated growing degree days (GDD), a commonly used heat accumulation measure to forecast phenological development in plants \citep{miller01}, for each provenance and garden, based on the accumulation of mean daily temperatures ($T_{m}$)from 2011 to 2020 above a baseline of 0$^{\circ}$C, from January 1 until budburst and leaf flush with the following formula: GDD = $\sum$($T_{m}$ - $0^{\circ}$C for $T_{m}$ $\ge$ $0^{\circ}$C; 0 for $T_{m}$ $\le$ $0^{\circ}$C).
 % See notes on 31 Aug 2023 by Lizzie at https://github.com/lizzieinvancouver/localadaptclim/issues/26 for more methods on GDD.
@@
 
-\begin{figure}[!h] 
-    \centering
- \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/springfall_latmat.jpg}
-    \caption{Event day of year (DOY) in relation to provenance latitude and MAT, coded by symbol for species and color for garden with linear fits from hierarchical Bayesian models.
-Spring events shown on top and fall events at the bottom.
-European gardens and species are bolded and denoted by an asterisk (*).
-}
-    \label{figure:springfall_latmat}
-\end{figure}
-
 <<label=source_fall_models, echo=FALSE>>=
@@
 In contrast, fall events (e.g., budset, leaf senescence, leaf abscission) were earlier at more northern, cooler MAT sites (that is, they advanced strongly with provenance latitude: \Sexpr{format(round(abs(mean(all_species_fallpost_fitC_fall_lat)),digits = 2), nsmall=2)} days/degree [\Sexpr{format(round(abs(quantile(all_species_fallpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)}-\Sexpr{format(round(abs(quantile(all_species_fallpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}], and with decreasing MAT:\Sexpr{format(round(abs(mean(all_species_fallpost_fitC_fall_mat)),digits = 2), nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(all_species_fallpost_fitC_fall_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(all_species_fallpost_fitC_fall_mat, 0.95), digits = 2)}], Fig.
-\ref{figure:springfall_latmat}, Table.\ref{supplement-table:model_fall_lat} \& \ref{supplement-table:model_fall_mat} in Supplements ).
+\ref{figure:springfall_latmat}, Table.\ref{supplement-table:model_fall_lat} \& \ref{supplement-table:model_fall_mat} in Supplements).
 This relationship, however, was observed mostly in North America where fall events advanced \Sexpr{format(round(abs(mean(northamericanpost_fitC_fall_lat)),digits = 2),nsmall=2)} (\Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree northward, or \Sexpr{format(round(mean(northamericanpost_fitC_fall_mat),digits = 2),nsmall=2)} days (\Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_mat, 0.95)), digits = 2),nsmall=2)}) per degree decline in MAT ($^{\circ}$C), whereas in Europe these relationships were weaker: advance of \Sexpr{format(round(abs(mean(europeanpost_fitC_fall_lat)),digits = 2),nsmall=2)} (\Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree northward, or \Sexpr{format(round(mean(europeanpost_fitC_fall_mat),digits = 2),nsmall=2)} days (\Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_mat, 0.95)), digits = 2),nsmall=2)}) per degree decline in MAT ($^{\circ}$C) (Fig.
@@
 In contrast, spring event clines were always weak: all species x garden clines included 0 in their 90\% intervals.
-
-\begin{figure}[!h] 
-    \centering
- \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/continental_spptype_lat.jpg}
-    \caption{Posterior distributions (99\% percentile) for the effect of provenance latitude across different continents and species types.
-Zero---no effect---is shown with a dashed line.
-Solid line and shading in posteriors represent mean and 50 percent interval (90\% intervals given in text).
-(a \& b) Effect of latitude and continent on spring and fall event date (DOY).
-Fall events advanced strongly per degree northward, particularly notably in North America.
-(c \& d) Effect of latitude and leaf type on spring and fall event date (DOY).
-As we move north, spring events slightly advanced in gymnosperms and delayed in angiosperms.
-Fall events advanced per degree northward for both species types.}
-    \label{figure:continental_spptype_lat}
-\end{figure}
-
-\begin{figure}[!h] 
-    \centering
- \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/continental_spptype_effect.jpg}
-    \caption{Posterior distributions (99\% percentile) for the effect of MAT across different continents and species types.
-Zero---no effect---is shown with a dashed line.
-Solid line and shading in posteriors represent mean and 50 percent interval (90\% intervals given in text).
-(a \& b) Effect of MAT and continent on spring and fall event date (DOY).
-Fall events advanced strongly with decreasing MAT, particularly notably in North America.
-(c \& d) Effect of MAT and leaf type on spring and fall event date (DOY).
-As MAT increased, spring events slightly advanced in angiosperms and delayed in gymnosperms.
-Fall events delayed in warmer locations for both species types.}
-    \label{figure:continental_spptype_effect}
-\end{figure}
 
@@
 We found the coarse metrics of provenance latitude and MAT were generally good predictors of phenology, performing better than our more complex and data-rich metric of climate overlap.
-Latitude and MAT appeared to well represent how similar the climates are between the provenances and gardens in the temperate and boreal forest species we studied, yielding similar results metrics calculated specifically in the spring with daily climate data.
+Latitude and MAT appeared to well represent how similar the climates are between the provenances and gardens in the temperate and boreal forest species we studied, yielding similar results to metrics calculated specifically in the spring with daily climate data.
 %EMWSep4 -- also avoid repeating how narrow the data is etc.
@@
 
+%EMWSep30: Please double-check we're explaining trends correctly before you submit; I think we are -- but please check closely.
 Our results highlight that spring events show much weaker clines than fall events in tree species, but suggest important variation between species types in spring events.
@@
 In contrast, angiosperms may tend towards earlier leafout in warmer climates to compete best for access to light and other resources \citep{cat2019}.
-Testing these hypotheses would require More information on frost risk and forest community assembly from across the provenance locations.
+Testing these hypotheses would require more information on frost risk and forest community assembly from across the provenance locations.
 %EMWSep4 -- I like the below (and I think it may have been my idea), but we'd likely be asked to test for it ...
@@
 Fall events were earlier in colder, more northern latitudes in both continents, which is consistent with previous literature \citep{AitkenBemmels16, Alberto13}.
-North American studies of both gymnosperm and angiosperm species reported strong clines, and most often focused on budset, while the two European gardens reported data on `leaf senescence' (Supplement Table.\ref{supplement-table:all_studies}).
-Leaf senescence was visually inspected based on the percentage of the tree crown that had changed color when the assessor stood on the south-facing side of each tree \citep{rosique22}.
-We note that budset is most closely related to growth cessation and tied to photoperiod signals, making it generally considered a more robust indicator of when plants stop investing in growth and more genetically controlled than leaf senescence \citep{Alberto13}.
+North American studies of both gymnosperm and angiosperm species reported strong clines, and most often focused on budset, while the two European gardens reported data on `leaf senescence' \citep{rosique22}, measured as percentage of tree crown that had changed color (Supplement Table.\ref{supplement-table:all_studies}).
+These different metrics could explain the different results.
+Research has shown that budset is more closely related to growth cessation and thus considered a more robust indicator of when plants stop investing in growth than leaf senescence; it is also more genetically controlled \citep[with different photoperiod thresholds required to trigger budset in different provenances,][]{Alberto13}.
+%EMWSep30: I believe you never need 'note that' -- cutting it is almost ALWAYS a good way to simplify. 
+%Leaf senescence was visually inspected based on the percentage of the tree crown that had changed color when the assessor stood on the south-facing side of each tree.
+%EMWSep30: Good info, just too much detail; I edited.
 
@@
 \ref{figure:map_gardens}).
-Given the tremendous investment required for such gardens, however, it is not surprising they are often limited to one species---most often those relevant for forestry (e.g.,\emph{Pseudotsuga menziesii, Tsuja plicata, Tsuga heterophylla, Picea sitchensis, Pinus albicaulis}).
-Our results, however, highlight the limited inference we may gain from such an approach to understand trends across species more broadly.
+Given the tremendous investment required for such gardens, however, it is not surprising they are often limited to one species---most often those relevant for forestry (e.g., \emph{Pseudotsuga menziesii, Tsuja plicata, Tsuga heterophylla, Picea sitchensis, Pinus albicaulis}).
+Our results, however, highlight the limited inference we may gain from such an approach to understand trends across species more broadly, a critical need for climate change forecasting.
 
@@
 This supports previous results and the general theory that fall events are consistently designed to avoid tissue loss at the end of the season, when the cost of such losses could be particularly high \citep{AitkenBemmels16, Alberto13}.
-Spring events showed no clear trends, but Increasing research into counter-gradient variation for flowering events could be relevant when spring vegetative and flowering events are linked \citep{dan2021nph}.
+Spring events showed no clear trends, but increasing research into counter-gradient variation for flowering events could be relevant when spring vegetative and flowering events are linked \citep{dan2021nph}.
 
@@
 
+ \section{Acknowledgement}
+%EMWSep30 -- check umlaut on Korner works
+We thank S. Aitken,  I. Chuine, R. Guy, C K\"{o}rner and Y. Vitasse for reviewing our list of papers for possible additional common garden studies.
+
+\section{References} %EMWSep30: Check requested order by GCB ... I think it's usually acknowledgements, references then figures so I moved these section up. 
+\bibliography{bibliography_local_adaptation.bib}
+
+\clearpage
 \section{Figures}
 %alinaSep22: i dont know how to force all figures to be in this section instead of in between texts
+%EMWSep30 -- you just move them here! Please check I put them in the right order; I also added a \clearpage which can help with formatting looking clean (see if it works, you can remove if not). 
+\begin{figure}[!h] 
+    \centering
+ \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/map_gardens.jpg}
+    \caption{Distribution of common gardens (denoted as stars) and provenances (denoted as circles) included in this meta-analysis.
+The distribution was skewed toward North America (12 North American studies versus 5 European studies).
+See Table.\ref{supplement-table:all_studies} in Supplement for information on selected studies.
+} %EMWSep30: These lines don't actually delineate study regions ...
+so I suggest you delete the whole sentence unless the journal requires it or you really want it; if you keep it in, it should be: Note: map lines do not necessarily depict accepted national boundaries.
+    \label{figure:map_gardens}
+\end{figure}
 
-\section{References}
-\bibliography{bibliography_local_adaptation.bib}
+\begin{figure}[!h] 
+    \centering
+ \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/springfall_latmat.jpg}
+    \caption{Event day of year (DOY) in relation to provenance latitude and MAT, coded by symbol for species and color for garden with linear fits from hierarchical Bayesian models.
+Spring events shown on top and fall events at the bottom.
+European gardens and species are bolded and denoted by an asterisk (*).
+}
+    \label{figure:springfall_latmat}
+\end{figure}
 
-\section{Acknowledgement}
-We thank S. Aitken,  I. Chuine, R. Guy, C Korner and Y. Vitasse for reviewing our list of papers for possible additional common garden studies.
+%EMWSep30: Updated figures are nice!
+I shortened the descriptions in both and re-arranged a little.
+Try to keep phrasing more consistent here (don't mix 'as we move north' with 'per degree northward' -- pick one).
+Check my edits closely to make sure they're correct.
+\begin{figure}[!h] 
+    \centering
+ \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/continental_spptype_lat.jpg}
+    \caption{Effects of latitude on spring and fall event date (DOY) depending on (a \& b) continent, and (c \& d) species leaf type shown as posterior distributions (full distribution represents 99\% percentile, solid line and shading in posteriors represent mean and 50 percent interval, 90\% intervals given in text).
+Zero---no effect---is shown with a dashed line.
+Spring events did not shift with latitude by continent, but fall events advanced strongly per degree northward, particularly in North America.
+Spring events slightly advanced in gymnosperms and delayed in angiosperms per degree northward.
+Fall events advanced per degree northward for both species types.}
+    \label{figure:continental_spptype_lat}
+\end{figure}
+
+\begin{figure}[!h] 
+    \centering
+ \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/continental_spptype_effect.jpg}
+    \caption{Effects of MAT on spring and fall event date (DOY) depending on (a \& b) continent, and (c \& d) species leaf type shown as posterior distributions (full distribution represents 99\% percentile, solid line and shading in posteriors represent mean and 50 percent interval, 90\% intervals given in text).
+Zero---no effect---is shown with a dashed line.
+Spring events did not shift with MAT by continent, but fall events advanced strongly with decreasing MAT, particularly notably in North America.
+Spring events slightly advanced in angiosperms and delayed in gymnosperms with increasing MAT.
+Fall events delayed with increasing MAT for both species types.}
+    \label{figure:continental_spptype_effect}
+\end{figure}
 

### 2023-10-04 Docs/manuscript.Rnw
made small edits

@@
 \usepackage{Sweave}
+\usepackage{xurl}
 \usepackage{graphicx}
@@
 \SweaveOpts{concordance=TRUE} % For RStudio hiccups
-\title{{\huge Weak evidence of provenance effects in spring phenology across Europe and North America}
-\author{Ziyun Zeng \& E. M. Wolkovich}
-\date{2023}
-\maketitle
+\begin{flushright}
+Version dated: \today
+\end{flushright}
+
+\bigskip
+\medskip
+\begin{center}
+
+% Insert your title:
+\noindent{\Large {\bf Weak evidence of provenance effects in spring phenology across Europe and North America}}\\ 
+\bigskip
+
+\noindent {\normalsize \sc
+Z. A. Zeng $^{1}$ \& E. M. Wolkovich$^{2}$}\\
+
+\noindent {\small \it
+$^1$ Forest Resources Management, Faculty of Forestry, University of British Columbia, 2424 Main Mall, Vancouver, BC V6T 1Z4\\
+$^2$ Forest \& Conservation Sciences, Faculty of Forestry, University of British Columbia, 2424 Main Mall, Vancouver, BC V6T 1Z4}
+\end{center}
+\medskip
+\noindent{\bf Corresponding author:} Z. A. Zeng, see $^{1}$ above ; E-mail: alinazengziyun@yahoo.com\\
+
+\noindent Additional information: Zeng (0009-0000-4647-9457); Wolkovich (0000-0001-7653-893X)\\
+
+\noindent Running head: Clines in spring phenology
 
@@
 \begin{abstract}
-Forecasting the biological impacts of climate change requires understanding how species respond to warmer temperatures through both inter-annual flexible variation versus through adaptation to local conditions.
-Yet we often lack this information entirely, or find conflicting evidence across studies.
+Forecasting the biological impacts of climate change requires understanding how species respond to warmer temperatures through inter-annual flexible variation versus through adaptation to local conditions.
+Yet, we often lack this information entirely or find conflicting evidence across studies.
 The latter is the case for shifts in spring phenology---one of the most reported and consistent impacts of anthropogenic climate change, and also one of the most critical to forecasting, given its role in carbon sequestration.
 Some common garden studies have found evidence of important provenance effects, which suggest there may be local adaptation in the underlying cues of spring phenology and mirrors findings for fall events, while other studies find no evidence.
-Here we synthesize common garden studies across Europe and North America that reported spring event dates for a mix of angiosperm and gymnosperm tree species in the northern hemisphere, capturing data from 384 North American provenances and 101 European provenances with observations from 1962 to 2019, alongside fall event data when provided.
-Across continents we find no evidence of provenance effects in spring phenology, but strong clines with latitude and mean annual temperature (MAT) for fall events.
+Here, we synthesize common garden studies across Europe and North America that reported spring event dates for a mix of angiosperm and gymnosperm tree species in the northern hemisphere, capturing data from 384 North American provenances and 101 European provenances with observations from 1962 to 2019, alongside fall event data when provided.
+Across continents, we find no evidence of provenance effects in spring phenology, but strong clines with latitude and mean annual temperature (MAT) for fall events.
 These effects, however, appear to diverge by continent and species type (gymnosperm versus angiosperm), especially for fall events where clines with latitude and MAT are much stronger in North America.
 Our results suggest flexible, likely plastic responses, in spring phenology with warming, and potential limits---at least in the short term---due to provenance effects for fall phenology.
-They also highlight that, after over 50 years of common garden studies on tree phenology, we still lack a holistic predictive model of clines across species and phenological events.
+They also highlight that, after over 250 years of common garden studies on tree phenology, we still lack a holistic predictive model of clines across species and phenological events.
 \end{abstract} %Spring tree phenology is critical to forecasting ...
@@
 This is important because...
+
+\noindent \emph{Keywords:} spring phenology, provenance, common gardens, budburst,
+leafout, budset, forests, climate change\\
 
@@
 Tree phenology is important to climate change forecasting at both the community and ecosystem levels.
-The timing of budburst and senescence can impact plant competition, plant invasions and community assembly \citep{fridley12}.
+The timing of budburst and senescence can impact plant competition, plant invasions, and community assembly \citep{fridley12}.
 Shifts in phenology can affect tree growth \citep{myneni97}, scaling up to impact ecosystem-level carbon sequestration \citep{Barichivich12}, and thus forecasts of climate change.
@@
 Though clines of spring phenology have been found in both Europe \citep{sog08,deans96,von95} and North America \citep{rossi15, soo13, hannerz99}, there is continuing debate, especially in Europe \citep{deans96,vitasse2009,Basler:2012}, raising the possibility that they could vary by continent.
-%EMWSep4: Check deans96 is accurate ref I added.
-%alinaSep22: Hi lizzie I am not sure how you are trying to phrase "they may be more studies and debated in Europe" but I think it's slightly off.
-could you fix it?
-%EMWSep30: Yes, I have been struggling with this sentence .... it's tricky, but I think this works now.  Let me know what you think.
 
@@
 To address these questions, we combined Bayesian hierarchical models with a new meta-analysis of all common garden experiments in temperate tree species across Europe and North America reporting spring phenology.
-%EMWSep30: Questions were so similar, so I simplified into once sentence/
 
@@
 To understand how climatic differences, in addition to geographical differences, shape local adaptation in spring events we extracted several types of climate data using information about provenance latitude, longitude, and elevation from original publications.
-We estimated the mean annual temperature (MAT) from 1960-1991 for each provenance using the Climate Information Tool by Food and Agriculture Organization of the United Nations \citep{FAO2022}.
-We verified our calculated MAT was was similar to MAT calculated using ClimateWNA \citep{wang2016}, a source used in previous analyses.
+We estimated each provenance's mean annual temperature (MAT) from 1960 to 1991 using the monthly temperature data in the Climate Information Tool by Food and Agriculture Organization of the United Nations \citep{FAO2022}.
+We verified our estimated MAT was similar to MAT calculated using ClimateWNA \citep{wang2016}, a source used in previous analyses.
 
-To examine climate near spring events more explicitly than MAT allows, we used gridded daily temperature data for March-May for 2011 to 2020 for all provenances and gardens.
-We extracted data from E-OBS for European locations, and using the `daymetr' in R for North American locations \citep{cornes2018,hufkens2018}.
-Then, using these data and the `overlap' package in R, we estimated how much the daily temperatures overlapped between each provenance location and their corresponding gardens across the three months for our 10 year period, which we call `climate overlap.'  % Finally, using the daily temperature data, we also calculated growing degree days (GDD), a commonly used heat accumulation measure to forecast phenological development in plants \citep{miller01}, for each provenance and garden, based on the accumulation of mean daily temperatures ($T_{m}$)from 2011 to 2020 above a baseline of 0$^{\circ}$C, from January 1 until budburst and leaf flush with the following formula: GDD = $\sum$($T_{m}$ - $0^{\circ}$C for $T_{m}$ $\ge$ $0^{\circ}$C; 0 for $T_{m}$ $\le$ $0^{\circ}$C).
+To examine climate near spring events more explicitly than MAT allows, we used gridded daily temperature data for March-May from 2011 to 2020 for all provenances and gardens.
+We extracted data from E-OBS for European locations and used the `daymetr' in R for North American locations \citep{cornes2018,hufkens2018}.
+Then, using these data and the `overlap' package in R, we estimated how much the daily temperatures overlapped between each provenance location and their corresponding gardens across the three months for our 10 year period, which we call `climate overlap.' Dataset containing event dates, geographic information, and climatic information of all provenances are archived in  Knowledge Network for Biocomplexity (KNB) \citep{zeng23}.
+% Finally, using the daily temperature data, we also calculated growing degree days (GDD), a commonly used heat accumulation measure to forecast phenological development in plants \citep{miller01}, for each provenance and garden, based on the accumulation of mean daily temperatures ($T_{m}$)from 2011 to 2020 above a baseline of 0$^{\circ}$C, from January 1 until budburst and leaf flush with the following formula: GDD = $\sum$($T_{m}$ - $0^{\circ}$C for $T_{m}$ $\ge$ $0^{\circ}$C; 0 for $T_{m}$ $\le$ $0^{\circ}$C).
 % See notes on 31 Aug 2023 by Lizzie at https://github.com/lizzieinvancouver/localadaptclim/issues/26 for more methods on GDD.
@@
 To estimate clines in spring and fall phenological events across species we used Bayesian hierarchical models.
-We regressed DOY of events against geographical and climate predictors with partial pooling (sometimes called `random effects') on the intercept and slope for each species within each garden.
+We regressed DOY of events against geographical and climatic predictors with partial pooling (sometimes called `random effects') on the intercept and slope for each species within each garden.
 Because most tree species were present in only one common garden in our dataset, it was impossible to fit garden and species separately, thus we treat each species within a garden as a unique group.
@@
 Research has shown that budset is more closely related to growth cessation and thus considered a more robust indicator of when plants stop investing in growth than leaf senescence; it is also more genetically controlled \citep[with different photoperiod thresholds required to trigger budset in different provenances,][]{Alberto13}.
-%EMWSep30: I believe you never need 'note that' -- cutting it is almost ALWAYS a good way to simplify. 
 %Leaf senescence was visually inspected based on the percentage of the tree crown that had changed color when the assessor stood on the south-facing side of each tree.
-%EMWSep30: Good info, just too much detail; I edited.
 
@@
 \section{Acknowledgement}
-%EMWSep30 -- check umlaut on Korner works
 We thank S. Aitken,  I. Chuine, R. Guy, C K\"{o}rner and Y. Vitasse for reviewing our list of papers for possible additional common garden studies.
+
+\section{Data Availability Statement}
+The data that support the findings of this study are openly available in the Knowledge Network for Biocomplexity (KNB) repository at \url{https://knb.ecoinformatics.org/view/urn\%3Auuid\%3Aa37258b9-23e7-4b0c-a20f-9185cbc27194}.
 
@@
 See Table.\ref{supplement-table:all_studies} in Supplement for information on selected studies.
-} %EMWSep30: These lines don't actually delineate study regions ...
-so I suggest you delete the whole sentence unless the journal requires it or you really want it; if you keep it in, it should be: Note: map lines do not necessarily depict accepted national boundaries.
+Note: map lines do not necessarily depict accepted national boundaries.}
     \label{figure:map_gardens}
@@
 
-% helping with abstract language
-% Alberto16
-% "On the basis of the patterns of quantitative variation for 19 adaptation-related traits studied in 59 tree species (mostly temperate and boreal species from the Northern hemisphere), we found that genetic differentiation between populations and clinal variation along environmental gradients were very common (respectively, 90% and 78% of cases)."
+% A data availability statement which provides information about where the research data and other artifacts supporting the results reported in the paper can be found must be submitted.
+Links to the repository where the dataset(s) are publicly archived and DOIs must be included.
+A list of standard templates for the text for the ‘Data Availability Statement’ is available here.
+% Data must be cited within the text in the in the Materials and Methods section.
+% Data must be included as a formal citation in the reference section. More detail on how to do this can be found here.
+% Datasets containing population means from all studies are archived in Dryad
+
+% Data Archiving Statement
+% Data available from the Dryad Digital Repository: http://dx.doi.org/10.5061/dryad.rj15h
 

### 2023-10-04 Docs/manuscript.tex
made small edits

@@
 \input{/Users/alina/Documents/git/localadaptclim/Docs/figure_ms/demoFig-concordance} % For RStudio hiccups
-\title{{\huge Changes and trends in budburst and leaf flush across Europe and North America} \\A meta-analysis of local adaptation in spring phenology studies}
-\author{Ziyun Zeng \& E. M. Wolkovich}
-\date{2023}
-\maketitle
+\begin{flushright}
+Version dated: \today
+\end{flushright}
+
+\bigskip
+\medskip
+\begin{center}
+
+% Insert your title:
+\noindent{\Large {\bf Weak evidence of provenance effects in spring phenology across Europe and North America}}\\ 
+\bigskip
+
+\noindent {\normalsize \sc
+Z. A. Zeng $^{1}$ \& E. M. Wolkovich$^{2}$}\\
+
+\noindent {\small \it
+$^1$ Forest Resources Management, Faculty of Forestry, University of British Columbia, 2424 Main Mall, Vancouver, BC V6T 1Z4\\
+$^2$ Forest \& Conservation Sciences, Faculty of Forestry, University of British Columbia, 2424 Main Mall, Vancouver, BC V6T 1Z4}
+\end{center}
+\medskip
+\noindent{\bf Corresponding author:} Z. A. Zeng, see $^{1}$ above ; E-mail: alinazengziyun@yahoo.com\\
+
+\noindent Additional information: Zeng (0009-0000-4647-9457); Wolkovich (0000-0001-7653-893X)\\
+
+\noindent Running head: Clines in spring phenology
 
@@
 maybe New Phytologist but I think they could be picky.
-
+\linenumbers
 \begin{abstract}
-Forecasting the biological impacts of climate change requires understanding how species respond to warmer temperatures through both inter-annual flexible variation versus through adaptation to local conditions.
-Yet we often lack this information entirely, or find conflicting evidence across studies.
-This is the case for one of the most reported and consistent impacts of anthropogenic climate change---earlier spring phenology, which is also one of the most critical to forecasting given its role in carbon sequestration.
-Evidence from some common garden studies suggest important provenance effects, which suggest local adaptation in the underlying cues of spring phenology and supports findings for fall events, while other studies find no evidence.
-Here we synthesize common garden studies across Europe and North America that reported spring event dates for a mix of angiosperm and gymnosperm tree species in the northern hemisphere, capturing data from 384 North American provenances and 101 European provenances with observations from 1962 to 2019, alongside fall event data when provided.
-Across continents we find no evidence of provenance effects in spring phenology, but strong clines with latitude and mean annual temperature (MAT) for fall events.
+Forecasting the biological impacts of climate change requires understanding how species respond to warmer temperatures through inter-annual flexible variation versus through adaptation to local conditions.
+Yet, we often lack this information entirely or find conflicting evidence across studies.
+The latter is the case for shifts in spring phenology---one of the most reported and consistent impacts of anthropogenic climate change, and also one of the most critical to forecasting, given its role in carbon sequestration.
+Some common garden studies have found evidence of important provenance effects, which suggest there may be local adaptation in the underlying cues of spring phenology and mirrors findings for fall events, while other studies find no evidence.
+Here, we synthesize common garden studies across Europe and North America that reported spring event dates for a mix of angiosperm and gymnosperm tree species in the northern hemisphere, capturing data from 384 North American provenances and 101 European provenances with observations from 1962 to 2019, alongside fall event data when provided.
+Across continents, we find no evidence of provenance effects in spring phenology, but strong clines with latitude and mean annual temperature (MAT) for fall events.
 These effects, however, appear to diverge by continent and species type (gymnosperm versus angiosperm), especially for fall events where clines with latitude and MAT are much stronger in North America.
 Our results suggest flexible, likely plastic responses, in spring phenology with warming, and potential limits---at least in the short term---due to provenance effects for fall phenology.
-They also highlight that, after over 50 years of common garden studies on tree phenology we may still lack a holistic predictive model of clines across species and phenological events.
+They also highlight that, after over 250 years of common garden studies on tree phenology, we still lack a holistic predictive model of clines across species and phenological events.
 \end{abstract} %Spring tree phenology is critical to forecasting ...
@@
 
-Tree phenology is critically important to climate change forecasting at both the community and ecosystem levels.
-The timing of budburst and senescence can impact plant competition, plant invasions and community assembly \citep{fridley12}.
-These effects often occur as phenology affects growth \citep{myneni97}, scaling up to impact ecosystem-level carbon sequestration \citep{Barichivich12}, and thus is integrated into forecasts of climate change itself.
+Tree phenology is important to climate change forecasting at both the community and ecosystem levels.
+The timing of budburst and senescence can impact plant competition, plant invasions, and community assembly \citep{fridley12}.
+Shifts in phenology can affect tree growth \citep{myneni97}, scaling up to impact ecosystem-level carbon sequestration \citep{Barichivich12}, and thus forecasts of climate change.
 Growing evidence, however, suggests links between growth and phenology are not as consistent as previously predicted---or currently modelled \citep{dow22}---with recent work suggesting how much spring versus fall events shift may determine impacts on tree growth \citep{zohner23}.
@@
 This has led to debate over the prevalence and importance of adaptation in spring tree phenology.
-While clines of spring phenology have been found in studies in both Europe \citep{sog08,deans96,von95} and North America \citep{rossi15, soo13, hannerz99}, they may be more studies and debated in Europe \citep{deans96,vitasse2009,Basler:2012}, raising the possibility that they could vary by continent.
-%EMWSep4: Check deans96 is accurate ref I added.
-%alinaSep22: Hi lizzie I am not sure how you are trying to phrase "they may be more studies and debated in Europe" but I think it's slightly off.
-could you fix it?
+Though clines of spring phenology have been found in both Europe \citep{sog08,deans96,von95} and North America \citep{rossi15, soo13, hannerz99}, there is continuing debate, especially in Europe \citep{deans96,vitasse2009,Basler:2012}, raising the possibility that they could vary by continent.
 
@@
 To test for evidence of adaptation in spring phenology and what factors may underlie differences observed across studies, we comprehensively examined clines for spring events, including fall events when possible.
-We tested for evidence of adaptation via provenance trends with latitude and climate and examined possible factors that underlie these clines, asking:
-\begin{itemize}
-\item How different are clines between spring and fall phenology? 
-\item How different are clines between Europe and North America?
-\item How different are clines between angiosperm and gymnosperm species?
-\end{itemize}
+We tested for evidence of adaptation via provenance trends with latitude and climate and examined possible factors that underlie these clines, including for differences between: (1) spring and fall phenology, (2) studies in Europe and North America, (3) angiosperm and gymnosperm species.
 To address these questions, we combined Bayesian hierarchical models with a new meta-analysis of all common garden experiments in temperate tree species across Europe and North America reporting spring phenology.
@@
 \citealp{Rehfeldt1994}), we converted such dates to DOY using the `lubridate' package in R \citep{Grolemund11}.
-% emwJun4 -- you need to use ` to open single quotes in Latex.
-\begin{figure}[!h] 
-    \centering
- \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/map_gardens.jpg}
-    \caption{Distribution of common gardens (denoted as stars) and provenances (denoted as circles) included in this meta-analysis.
-The distribution was skewed toward North America (12 North American studies versus 5 European studies).
-See Table.\ref{supplement-table:all_studies} in Supplement for information on selected studies.
-Note: map lines delineate study regions and do not necessarily depict accepted national boundaries}
-    \label{figure:map_gardens}
-\end{figure}
-\newline
+
 To understand how climatic differences, in addition to geographical differences, shape local adaptation in spring events we extracted several types of climate data using information about provenance latitude, longitude, and elevation from original publications.
-We estimated the mean annual temperature (MAT) from 1960-1991 for each provenance using the Climate Information Tool by Food and Agriculture Organization of the United Nations \citep{FAO2022}.
-We verified our calculated MAT was was similar to MAT calculated using ClimateWNA \citep{wang2016}, a source used in previous analyses.
+We estimated each provenance's mean annual temperature (MAT) from 1960 to 1991 using the monthly temperature data in the Climate Information Tool by Food and Agriculture Organization of the United Nations \citep{FAO2022}.
+We verified our estimated MAT was similar to MAT calculated using ClimateWNA \citep{wang2016}, a source used in previous analyses.
 
-To examine more explicitly climate near spring events, we used gridded daily temperature data for 2011 to 2020 for all European and North American provenances and gardens from E-OBS and the `daymetr' package in R respectively \citep{cornes2018,hufkens2018}.
-Using these data we estimated how much the daily temperatures overlapped between garden and provenance locations, which we call `climate overlap.' For this we used the `overlap' package in R to calculate the percentage overlap of the daily temperature of each provenance and their corresponding gardens in spring months (March to May) from 2011 to 2020.
+To examine climate near spring events more explicitly than MAT allows, we used gridded daily temperature data for March-May from 2011 to 2020 for all provenances and gardens.
+We extracted data from E-OBS for European locations and used the `daymetr' in R for North American locations \citep{cornes2018,hufkens2018}.
+Then, using these data and the `overlap' package in R, we estimated how much the daily temperatures overlapped between each provenance location and their corresponding gardens across the three months for our 10 year period, which we call `climate overlap.' Dataset containing event dates, geographic information, and climatic information of all provenances are archived in  Knowledge Network for Biocomplexity (KNB) \citep{zeng23}.
 % Finally, using the daily temperature data, we also calculated growing degree days (GDD), a commonly used heat accumulation measure to forecast phenological development in plants \citep{miller01}, for each provenance and garden, based on the accumulation of mean daily temperatures ($T_{m}$)from 2011 to 2020 above a baseline of 0$^{\circ}$C, from January 1 until budburst and leaf flush with the following formula: GDD = $\sum$($T_{m}$ - $0^{\circ}$C for $T_{m}$ $\ge$ $0^{\circ}$C; 0 for $T_{m}$ $\le$ $0^{\circ}$C).
@@
 
-\begin{figure}[!h] 
-    \centering
- \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/springfall_latmat.jpg}
-    \caption{Event day of year (DOY) in relation to provenance latitude and MAT, coded by symbol for species and color for garden with linear fits from hierarchical Bayesian models.
-Spring events shown on top and fall events at the bottom.
-European gardens and species are bolded and denoted by an asterisk (*).
-}
-    \label{figure:springfall_latmat}
-\end{figure}
-
 In contrast, fall events (e.g., budset, leaf senescence, leaf abscission) were earlier at more northern, cooler MAT sites (that is, they advanced strongly with provenance latitude: 3.16 days/degree [2.87-3.45], and with decreasing MAT:4.78 days/$^{\circ}$C [4.1 - 5.4], Fig.
-\ref{figure:springfall_latmat}, Table.\ref{supplement-table:model_fall_lat} \& \ref{supplement-table:model_fall_mat} in Supplements ).
+\ref{figure:springfall_latmat}, Table.\ref{supplement-table:model_fall_lat} \& \ref{supplement-table:model_fall_mat} in Supplements).
 This relationship, however, was observed mostly in North America where fall events advanced 4.24 (3.95 - 4.53) days per degree northward, or 6.41 days (5.78 - 7.04) per degree decline in MAT ($^{\circ}$C), whereas in Europe these relationships were weaker: advance of 0.47 (0.21 - 1.17) days per degree northward, or 0.70 days (1.04 - 2.42) per degree decline in MAT ($^{\circ}$C) (Fig.
@@
 Fall events were earlier in colder, more northern latitudes in both continents, which is consistent with previous literature \citep{AitkenBemmels16, Alberto13}.
-North American studies of both gymnosperm and angiosperm species reported strong clines, and most often focused on budset, while the two European gardens reported data on `leaf senescence' (Supplement Table.\ref{supplement-table:all_studies}).
-Leaf senescence was visually inspected based on the percentage of the tree crown that had changed color when the assessor stood on the south-facing side of each tree \citep{rosique22}.
-We note that budset is most closely related to growth cessation and tied to photoperiod signals, making it generally considered a more robust indicator of when plants stop investing in growth and more genetically controlled than leaf senescence \citep{Alberto13}.
+North American studies of both gymnosperm and angiosperm species reported strong clines, and most often focused on budset, while the two European gardens reported data on `leaf senescence' \citep{rosique22}, measured as percentage of tree crown that had changed color (Supplement Table.\ref{supplement-table:all_studies}).
+These different metrics could explain the different results.
+Research has shown that budset is more closely related to growth cessation and thus considered a more robust indicator of when plants stop investing in growth than leaf senescence; it is also more genetically controlled \citep[with different photoperiod thresholds required to trigger budset in different provenances,][]{Alberto13}.
+%Leaf senescence was visually inspected based on the percentage of the tree crown that had changed color when the assessor stood on the south-facing side of each tree.
 
@@
 
+\section{Acknowledgement}
+We thank S. Aitken,  I. Chuine, R. Guy, C K\"{o}rner and Y. Vitasse for reviewing our list of papers for possible additional common garden studies.
+
+\section{Data Availability Statement}
+The data that support the findings of this study are openly available in the Knowledge Network for Biocomplexity (KNB) repository at \url{https://knb.ecoinformatics.org/view/urn\%3Auuid\%3Aa37258b9-23e7-4b0c-a20f-9185cbc27194}.
+
+\section{References} %EMWSep30: Check requested order by GCB ... I think it's usually acknowledgements, references then figures so I moved these section up. 
+\bibliography{bibliography_local_adaptation.bib}
+
+\clearpage
 \section{Figures}
 %alinaSep22: i dont know how to force all figures to be in this section instead of in between texts
+%EMWSep30 -- you just move them here! Please check I put them in the right order; I also added a \clearpage which can help with formatting looking clean (see if it works, you can remove if not). 
+\begin{figure}[!h] 
+    \centering
+ \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/map_gardens.jpg}
+    \caption{Distribution of common gardens (denoted as stars) and provenances (denoted as circles) included in this meta-analysis.
+The distribution was skewed toward North America (12 North American studies versus 5 European studies).
+See Table.\ref{supplement-table:all_studies} in Supplement for information on selected studies.
+Note: map lines do not necessarily depict accepted national boundaries.}
+    \label{figure:map_gardens}
+\end{figure}
 
-\section{References}
-\bibliography{bibliography_local_adaptation.bib}
+\begin{figure}[!h] 
+    \centering
+ \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/springfall_latmat.jpg}
+    \caption{Event day of year (DOY) in relation to provenance latitude and MAT, coded by symbol for species and color for garden with linear fits from hierarchical Bayesian models.
+Spring events shown on top and fall events at the bottom.
+European gardens and species are bolded and denoted by an asterisk (*).
+}
+    \label{figure:springfall_latmat}
+\end{figure}
 
-\section{Acknowledgement}
-We thank S. Aitken,  I. Chuine, R. Guy, C Korner and Y. Vitasse for reviewing our list of papers for possible additional common garden studies.
+%EMWSep30: Updated figures are nice!
+I shortened the descriptions in both and re-arranged a little.
+Try to keep phrasing more consistent here (don't mix 'as we move north' with 'per degree northward' -- pick one).
+Check my edits closely to make sure they're correct.
+\begin{figure}[!h] 
+    \centering
+ \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/continental_spptype_lat.jpg}
+    \caption{Effects of latitude on spring and fall event date (DOY) depending on (a \& b) continent, and (c \& d) species leaf type shown as posterior distributions (full distribution represents 99\% percentile, solid line and shading in posteriors represent mean and 50 percent interval, 90\% intervals given in text).
+Zero---no effect---is shown with a dashed line.
+Spring events did not shift with latitude by continent, but fall events advanced strongly per degree northward, particularly in North America.
+Spring events slightly advanced in gymnosperms and delayed in angiosperms per degree northward.
+Fall events advanced per degree northward for both species types.}
+    \label{figure:continental_spptype_lat}
+\end{figure}
+
+\begin{figure}[!h] 
+    \centering
+ \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/continental_spptype_effect.jpg}
+    \caption{Effects of MAT on spring and fall event date (DOY) depending on (a \& b) continent, and (c \& d) species leaf type shown as posterior distributions (full distribution represents 99\% percentile, solid line and shading in posteriors represent mean and 50 percent interval, 90\% intervals given in text).
+Zero---no effect---is shown with a dashed line.
+Spring events did not shift with MAT by continent, but fall events advanced strongly with decreasing MAT, particularly notably in North America.
+Spring events slightly advanced in angiosperms and delayed in gymnosperms with increasing MAT.
+Fall events delayed with increasing MAT for both species types.}
+    \label{figure:continental_spptype_effect}
+\end{figure}
 

### 2023-10-04 Docs/supplement.Rnw
made small edits

@@
 \SweaveOpts{concordance=TRUE} % For RStudio hiccups
-\title{{\huge Supplements for: Changes and trends in budburst and leaf flush across Europe and North America} \\A meta-analysis of local adaptation in spring phenology studies}
-\author{Ziyun Zeng \& E. M. Wolkovich}
-\date{2023}
-\maketitle
+\begin{flushright}
+Version dated: \today
+\end{flushright}
 
+\bigskip
+\medskip
+\begin{center}
+
+% Insert your title:
+\noindent{\Large {\bf Supporting Information: Weak evidence of provenance effects in spring phenology across Europe and North America}}\\ 
+\bigskip
+
+\noindent {\normalsize \sc
+Z. A. Zeng $^{1}$ \& E. M. Wolkovich$^{2}$}\\
+
+\noindent {\small \it
+$^1$ Forest Resources Management, Faculty of Forestry, University of British Columbia, 2424 Main Mall, Vancouver, BC V6T 1Z4\\
+$^2$ Forest \& Conservation Sciences, Faculty of Forestry, University of British Columbia, 2424 Main Mall, Vancouver, BC V6T 1Z4}
+\end{center}
+\medskip
+\noindent{\bf Corresponding author:} Z. A. Zeng, see $^{1}$ above ; E-mail: alinazengziyun@yahoo.com\\
+
+% \noindent Additional information: Zeng (0009-0000-4647-9457); Wolkovich (0000-0001-7653-893X)\\
+
+% \title{{\huge Supplements for: Changes and trends in budburst and leaf flush across Europe and North America} \\A meta-analysis of local adaptation in spring phenology studies}
+% \author{Ziyun Zeng \& E. M. Wolkovich}
+% \date{2023}
+% \maketitle
+
+\newpage
 \section{Additional Methods}

### 2023-10-04 Docs/supplement.tex
made small edits

@@
 \input{/Users/alina/Documents/git/localadaptclim/Docs/figure_ms/demoFig-concordance} % For RStudio hiccups
-\title{{\huge Supplements for: Changes and trends in budburst and leaf flush across Europe and North America} \\A meta-analysis of local adaptation in spring phenology studies}
-\author{Ziyun Zeng \& E. M. Wolkovich}
-\date{2023}
-\maketitle
+\begin{flushright}
+Version dated: \today
+\end{flushright}
 
+\bigskip
+\medskip
+\begin{center}
+
+% Insert your title:
+\noindent{\Large {\bf Supporting Information: Weak evidence of provenance effects in spring phenology across Europe and North America}}\\ 
+\bigskip
+
+\noindent {\normalsize \sc
+Z. A. Zeng $^{1}$ \& E. M. Wolkovich$^{2}$}\\
+
+\noindent {\small \it
+$^1$ Forest Resources Management, Faculty of Forestry, University of British Columbia, 2424 Main Mall, Vancouver, BC V6T 1Z4\\
+$^2$ Forest \& Conservation Sciences, Faculty of Forestry, University of British Columbia, 2424 Main Mall, Vancouver, BC V6T 1Z4}
+\end{center}
+\medskip
+\noindent{\bf Corresponding author:} Z. A. Zeng, see $^{1}$ above ; E-mail: alinazengziyun@yahoo.com\\
+
+% \noindent Additional information: Zeng (0009-0000-4647-9457); Wolkovich (0000-0001-7653-893X)\\
+
+% \title{{\huge Supplements for: Changes and trends in budburst and leaf flush across Europe and North America} \\A meta-analysis of local adaptation in spring phenology studies}
+% \author{Ziyun Zeng \& E. M. Wolkovich}
+% \date{2023}
+% \maketitle
+
+\newpage
 \section{Additional Methods}

### 2023-10-31 Docs/manuscript.Rnw
submitted to GCB

@@
 \usepackage[T1]{fontenc}
-\usepackage[semicolon]{natbib}
-\bibliographystyle{plainnat} %{unsrtnat}
+% \usepackage[semicolon]{natbib} %apacite takes care of everything now
+% \bibliographystyle{plainnat} %{unsrtnat}
+\usepackage[natbibapa]{apacite}
+\bibliographystyle{apacite}
 \usepackage{xr-hyper}
@@
 
-To test for evidence of adaptation in spring phenology and what factors may underlie differences observed across studies, we comprehensively examined clines for spring events, including fall events when possible.
+To test for evidence of adaptation in spring phenology and what factors may underlies differences observed across studies, we comprehensively examined clines for spring events, including fall events when possible.
 We tested for evidence of adaptation via provenance trends with latitude and climate and examined possible factors that underlie these clines, including for differences between: (1) spring and fall phenology, (2) studies in Europe and North America, (3) angiosperm and gymnosperm species.
@@
 \section{References} %EMWSep30: Check requested order by GCB ... I think it's usually acknowledgements, references then figures so I moved these section up.
-\bibliography{bibliography_local_adaptation.bib}
+
+\bibliography{..//..//localadaptclim/Docs/bibliography_local_adaptation.bib}
 

### 2023-11-26 Docs/manuscript.Rnw
updated manuscript for resubmission

@@
 add \usepackage{parskip} to preamble, delete \\ and see how it looks
+\UseRawInputEncoding % helps with "LaTeX Error: Invalid UTF-8 byte sequence"
 \usepackage{parskip}
@@
 \usepackage[T1]{fontenc}
-% \usepackage[semicolon]{natbib} %apacite takes care of everything now
-% \bibliographystyle{plainnat} %{unsrtnat}
-\usepackage[natbibapa]{apacite}
-\bibliographystyle{apacite}
+% \usepackage[natbibapa]{apacite}  % this is apa
+% \bibliographystyle{apacite} % this is apa
+\usepackage[comma]{natbib}
+\bibliographystyle{agsm} %Harvard referencing style
 \usepackage{xr-hyper}
@@
 % Using \doublespacing in the preamble changes the text to double-line spacing
-\doublespacing
+% \doublespacing
+\setstretch{1.5} %1.5 spacing
 % more optionals! %
@@
 \linenumbers
+
+\section*{Summary}
+
 \begin{abstract}
@@
 
-\noindent \emph{Keywords:} spring phenology, provenance, common gardens, budburst,
-leafout, budset, forests, climate change\\
+\noindent \emph{Keywords: budburst, budset, climate change, common gardens, forests, leafout, provenance, spring phenology}\\
 
@@
 
-\section{Methods}
-\subsection{Data Collection}
+\section{Materials and Methods}
+\subsection{Data collection}
 To locate common garden studies that reported the timing of spring events of woody plant species we searched and reviewed the peer-reviewed literature.
@@
 We also contacted authors of previous review papers \citep{AitkenBemmels16, Alberto13}, to help further search the literature.
-We then reviewed the methods and results of all publications to refine to only studies that met the following criteria: (a) focused on woody plants originating from either Europe or North America (also the locations of most studies), (b) had provenance trials/common gardens on the same continent, (c) reported latitude and longitude of provenances and gardens, and (d) reported spring events in units of calendar days (day of year or DOY) or could be converted into DOY (see \nameref{supplement-section:addmethods} in the Supplements).
+We then reviewed the methods and results of all publications to refine to only studies that met the following criteria: (a) focused on woody plants originating from either Europe or North America (also the locations of most studies), (b) had provenance trials/common gardens on the same continent, (c) reported latitude and longitude of provenances and gardens, and (d) reported spring events in units of calendar days (day of year or DOY) or could be converted into DOY (see \nameref{supplement-section:addmethods} in Supporting Information).
 
-Based on these criteria we found 19 common gardens distributed throughout North America and Europe, with the majority of data concentrated in North America (Fig.
-\ref{figure:map_gardens} \& Supplement Table.\ref{supplement-table:all_studies}).
+Based on these criteria we found 19 common gardens distributed throughout North America and Europe, with the majority of data concentrated in North America (Fig.\ref{figure:map_gardens} \& Table.S\ref{supplement-table:all_studies} in Supporting Information).
 From each common garden study we extracted phenological data on spring events (budburst and leaf flush) in DOY and, when present in the same paper, fall events (bud set, leaf senescence, growth cessation, and leaf abscission) by species and the geographic information of provenances and gardens.
@@
 
-Overall, spring events were not related to provenance latitude or MAT, neither across continents (latitude: \Sexpr{format(round(mean(all_species_springpost_fitC_spring_lat),digits = 2), nsmall=2)} days/degree [\Sexpr{format(quantile(all_species_springpost_fitC_spring_lat, 0.05), digits = 1)} - \Sexpr{format(quantile(all_species_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(round(mean(all_species_springpost_fitC_spring_mat),digits = 2), nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(all_species_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(all_species_springpost_fitC_spring_mat, 0.95), digits = 2)}]) (Fig.
-\ref{figure:springfall_latmat}, Table.
-\ref{supplement-table:model_spring_lat} \& \ref{supplement-table:model_spring_mat} in Supplements), nor within North America (latitude: \Sexpr{format(round(mean(northamerican_springpost_fitC_spring_lat),digits = 2), nsmall=2)} days/degree [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.05), digits = 1)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(mean(northamerican_springpost_fitC_spring_mat),digits = 1)} days/$^{\circ}$C [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.95), digits = 2)}]) or Europe (latitude: \Sexpr{format(round(mean(european_springpost_fitC_spring_lat),digits = 2),nsmall=2)} days/degree [\Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(round(mean(european_springpost_fitC_spring_mat),digits = 2),nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.95), digits = 2)}]) (Fig.\ref{figure:continental_spptype_lat}A \& \ref{figure:continental_spptype_effect}A).
-Results were similar using other distance metrics in lieu of latitude (see Supplement Fig.\ref{supplement-figure:lat_distance} for results using the difference between provenance and garden latitude, and the spherical distance between provenance and garden).
+Overall, spring events were not related to provenance latitude or MAT, neither across continents (latitude: \Sexpr{format(round(mean(all_species_springpost_fitC_spring_lat),digits = 2), nsmall=2)} days/degree [\Sexpr{format(quantile(all_species_springpost_fitC_spring_lat, 0.05), digits = 1)} - \Sexpr{format(quantile(all_species_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(round(mean(all_species_springpost_fitC_spring_mat),digits = 2), nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(all_species_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(all_species_springpost_fitC_spring_mat, 0.95), digits = 2)}]) (Fig.\ref{figure:springfall_latmat}, Table.S\ref{supplement-table:model_spring_lat} \& S\ref{supplement-table:model_spring_mat} in Supporting Information), nor within North America (latitude: \Sexpr{format(round(mean(northamerican_springpost_fitC_spring_lat),digits = 2), nsmall=2)} days/degree [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.05), digits = 1)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(mean(northamerican_springpost_fitC_spring_mat),digits = 1)} days/$^{\circ}$C [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.95), digits = 2)}]) or Europe (latitude: \Sexpr{format(round(mean(european_springpost_fitC_spring_lat),digits = 2),nsmall=2)} days/degree [\Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(round(mean(european_springpost_fitC_spring_mat),digits = 2),nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.95), digits = 2)}]) (Fig.\ref{figure:continental_spptype_lat}A \& \ref{figure:continental_spptype_effect}A).
+Results were similar using other distance metrics in lieu of latitude (see Fig.S\ref{supplement-figure:lat_distance} for results using the difference between provenance and garden latitude, and the spherical distance between provenance and garden).
 % Spring events advanced slightly with provenance latitude in Europe spring events are slightly earlier where provenance MAT is lower (i.e., higher, more northern latitudes), though uncertainty intervals strongly overlapped zero.
@@
 
-In contrast, fall events (e.g., budset, leaf senescence, leaf abscission) were earlier at more northern, cooler MAT sites (that is, they advanced strongly with provenance latitude: \Sexpr{format(round(abs(mean(all_species_fallpost_fitC_fall_lat)),digits = 2), nsmall=2)} days/degree [\Sexpr{format(round(abs(quantile(all_species_fallpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)}-\Sexpr{format(round(abs(quantile(all_species_fallpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}], and with decreasing MAT:\Sexpr{format(round(abs(mean(all_species_fallpost_fitC_fall_mat)),digits = 2), nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(all_species_fallpost_fitC_fall_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(all_species_fallpost_fitC_fall_mat, 0.95), digits = 2)}], Fig.
-\ref{figure:springfall_latmat}, Table.\ref{supplement-table:model_fall_lat} \& \ref{supplement-table:model_fall_mat} in Supplements).
-This relationship, however, was observed mostly in North America where fall events advanced \Sexpr{format(round(abs(mean(northamericanpost_fitC_fall_lat)),digits = 2),nsmall=2)} (\Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree northward, or \Sexpr{format(round(mean(northamericanpost_fitC_fall_mat),digits = 2),nsmall=2)} days (\Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_mat, 0.95)), digits = 2),nsmall=2)}) per degree decline in MAT ($^{\circ}$C), whereas in Europe these relationships were weaker: advance of \Sexpr{format(round(abs(mean(europeanpost_fitC_fall_lat)),digits = 2),nsmall=2)} (\Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree northward, or \Sexpr{format(round(mean(europeanpost_fitC_fall_mat),digits = 2),nsmall=2)} days (\Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_mat, 0.95)), digits = 2),nsmall=2)}) per degree decline in MAT ($^{\circ}$C) (Fig.
-\ref{figure:continental_spptype_effect}A).
+In contrast, fall events (e.g., budset, leaf senescence, leaf abscission) were earlier at more northern, cooler MAT sites (that is, they advanced strongly with provenance latitude: \Sexpr{format(round(abs(mean(all_species_fallpost_fitC_fall_lat)),digits = 2), nsmall=2)} days/degree [\Sexpr{format(round(abs(quantile(all_species_fallpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)}-\Sexpr{format(round(abs(quantile(all_species_fallpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}], and with decreasing MAT:\Sexpr{format(round(abs(mean(all_species_fallpost_fitC_fall_mat)),digits = 2), nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(all_species_fallpost_fitC_fall_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(all_species_fallpost_fitC_fall_mat, 0.95), digits = 2)}], Fig.\ref{figure:springfall_latmat}, Table.S\ref{supplement-table:model_fall_lat} \& S\ref{supplement-table:model_fall_mat} in Supporting Information).
+This relationship, however, was observed mostly in North America where fall events advanced \Sexpr{format(round(abs(mean(northamericanpost_fitC_fall_lat)),digits = 2),nsmall=2)} (\Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree northward, or \Sexpr{format(round(mean(northamericanpost_fitC_fall_mat),digits = 2),nsmall=2)} days (\Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_mat, 0.95)), digits = 2),nsmall=2)}) per degree decline in MAT ($^{\circ}$C), whereas in Europe these relationships were weaker: advance of \Sexpr{format(round(abs(mean(europeanpost_fitC_fall_lat)),digits = 2),nsmall=2)} (\Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree northward, or \Sexpr{format(round(mean(europeanpost_fitC_fall_mat),digits = 2),nsmall=2)} days (\Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_mat, 0.95)), digits = 2),nsmall=2)}) per degree decline in MAT ($^{\circ}$C) (Fig.\ref{figure:continental_spptype_effect}A).
 
@@
 Clines in fall phenology were stronger and more consistent whereas clines in spring phenology were weaker and somewhat varied in directionality.
-For fall events, only two field studies found no relationship (Fig.
-\ref{figure:springfall_latmat}): \emph{Fraxinus excelsior} from Garden Q* in the UK \citep{rosique22} and \emph{Fagus sylvatica} from Garden R* in Bulgaria \citep{petkova17}.
+For fall events, only two field studies found no relationship (Fig.\ref{figure:springfall_latmat}): \emph{Fraxinus excelsior} from Garden Q* in the UK \citep{rosique22} and \emph{Fagus sylvatica} from Garden R* in Bulgaria \citep{petkova17}.
 Another study that found no relationship was the only greenhouse experiment included (\emph{Picea engelmannii} from Garden B in the USA, also included in \citealp{AitkenBemmels16}), which uniquely used the fall event of `the day by which seedling elongation had finished' \citep{rehfeldt94}.
@@
 Spring events advanced \Sexpr{format(round(abs(mean(deciduous_springpost_fitC_spring_mat)),digits = 2), nsmall=2)} (\Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_mat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_mat, 0.05)), digits = 2),nsmall=2)}) days/$^{\circ}$C as MAT increased in angiosperms and delayed \Sexpr{format(round(abs(mean(evergreens_springpost_fitC_spring_mat)),digits = 2), nsmall=2)} (\Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_mat, 0.95)), digits = 2),nsmall=2)}) days/$^{\circ}$C as MAT increased in gymnosperms.
-Fall events delayed in warmer locations for both species types, but slightly more so for gymnosperms (\Sexpr{format(round(mean(evergreenspost_fitC_fall_mat),digits = 2),nsmall=2)} days) than angiosperms (\Sexpr{format(round(mean(deciduouspost_fitC_fall_mat),digits = 2),nsmall=2)} days) (Fig.
-\ref{figure:continental_spptype_effect}B).
+Fall events delayed in warmer locations for both species types, but slightly more so for gymnosperms (\Sexpr{format(round(mean(evergreenspost_fitC_fall_mat),digits = 2),nsmall=2)} days) than angiosperms (\Sexpr{format(round(mean(deciduouspost_fitC_fall_mat),digits = 2),nsmall=2)} days) (Fig.\ref{figure:continental_spptype_effect}B).
 
@@
 For spring events, we considered overlap across March to May.
-However, results were not qualitatively different than using MAT (See Fig.
-\ref{supplement-figure:overlap_scatterplot} in Supplements).
+However, results were not qualitatively different than using MAT (See Fig.S\ref{supplement-figure:overlap_scatterplot} in Supporting Information).
 We observed very weak effects of climate overlap on spring events (\Sexpr{format(round(abs(mean(all_species_springpost_fitC_spring_overlap)),digits = 2), nsmall=2)} [\Sexpr{format(round(abs(quantile(all_species_springpost_fitC_spring_overlap, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(all_species_springpost_fitC_spring_overlap, 0.95)), digits = 2),nsmall=2)}] days per one percent increase in climate overlap), nearly identical across angiosperms (\Sexpr{format(round(abs(mean(deciduous_springpost_fitC_spring_overlap)),digits = 2), nsmall=2)} [\Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_overlap, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_overlap, 0.05)), digits = 2),nsmall=2)}]) and gymnosperms (\Sexpr{format(round(abs(mean(evergreens_springpost_fitC_spring_overlap)),digits = 2), nsmall=2)} [\Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_overlap, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_overlap, 0.95)), digits = 2),nsmall=2)}]).
-Fall events advanced as climate overlap declined, but slightly more strongly for gymnosperms (advancing \Sexpr{format(round(abs(mean(evergreenspost_fitC_fall_overlap)),digits = 2), nsmall=2)} [\Sexpr{format(round(abs(quantile(evergreenspost_fitC_fall_overlap, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(evergreenspost_fitC_fall_overlap, 0.95)), digits = 2),nsmall=2)}] days per one percent decline in climate overlap) (Fig.\ref{supplement-figure:overlap_posterior} in Supplements).
+Fall events advanced as climate overlap declined, but slightly more strongly for gymnosperms (advancing \Sexpr{format(round(abs(mean(evergreenspost_fitC_fall_overlap)),digits = 2), nsmall=2)} [\Sexpr{format(round(abs(quantile(evergreenspost_fitC_fall_overlap, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(evergreenspost_fitC_fall_overlap, 0.95)), digits = 2),nsmall=2)}] days per one percent decline in climate overlap) (Fig.S\ref{supplement-figure:overlap_posterior} in Supporting Information).
 
@@
 Fall events were earlier in colder, more northern latitudes in both continents, which is consistent with previous literature \citep{AitkenBemmels16, Alberto13}.
-North American studies of both gymnosperm and angiosperm species reported strong clines, and most often focused on budset, while the two European gardens reported data on `leaf senescence' \citep{rosique22}, measured as percentage of tree crown that had changed color (Supplement Table.\ref{supplement-table:all_studies}).
+North American studies of both gymnosperm and angiosperm species reported strong clines, and most often focused on budset, while the two European gardens reported data on `leaf senescence' \citep{rosique22}, measured as percentage of tree crown that had changed color (Table.S\ref{supplement-table:all_studies} in Supporting Information).
 These different metrics could explain the different results.
@@
 Most North American gardens and provenances included in this study were limited to the Pacific Northwest region, and thus tended to focus on species from forests where gymnosperms are almost always the canopy species (\emph{Pseudotsuga, Tsuga, Picea, Pinus}) and angiosperms are much smaller, more opportunistic species (\emph{Alnus, Populus}).
-This might suggest a far different pattern if gardens included more evenly sampled North American tree species (Fig.
-\ref{figure:map_gardens}).
+This might suggest a far different pattern if gardens included more evenly sampled North American tree species (Fig.\ref{figure:map_gardens}).
 Given the tremendous investment required for such gardens, however, it is not surprising they are often limited to one species---most often those relevant for forestry (e.g., \emph{Pseudotsuga menziesii, Tsuja plicata, Tsuga heterophylla, Picea sitchensis, Pinus albicaulis}).
@@
 
-\section{Data Availability Statement}
+\section{Competing interests}
+We state that we do not possess any recognized conflicting financial interests or personal connections that might have seemed to impact the findings presented in this paper.
+
+\section{Author contributions}
+ZAZ collected and cleaned data, performed computations, and analyzed data in an iterative process.
+EMW conceived of the presented idea, designed the computational framework, and verified the analytical methods.
+Both authors discussed results and contributed to the final manuscript.
+
+\section{Data availability}
 The data that support the findings of this study are openly available in the Knowledge Network for Biocomplexity (KNB) repository at \url{https://knb.ecoinformatics.org/view/urn\%3Auuid\%3Aa37258b9-23e7-4b0c-a20f-9185cbc27194}.
@@
 The distribution was skewed toward North America (12 North American studies versus 5 European studies).
-See Table.\ref{supplement-table:all_studies} in Supplement for information on selected studies.
+See Table.S\ref{supplement-table:all_studies} in Supporting Information for sourcing information on selected studies.
 Note: map lines do not necessarily depict accepted national boundaries.}
@@
 
+\section{Supporting Information (brief legends)}
+
+1. Additional methods on exact procedures we followed for meta-analysis
+and reasons for the exclusion of a small number of studies.
+2. Table S1: showing all publications included in meta-analysis.
+3. Tables S2-5: showing model estimates.
+4. Figures S1-3: supporting figures.
+
 \end{document}

### 2023-11-26 Docs/manuscript.tex
updated manuscript for resubmission

@@
 
-Overall, spring events were not related to provenance latitude or MAT, neither across continents (latitude: 0.10 days/degree [-0.05 - 0.25]; MAT: -0.11 days/$^{\circ}$C [-0.34 - 0.12]) (Fig.
-\ref{figure:springfall_latmat}, Table.
-\ref{supplement-table:model_spring_lat} \& \ref{supplement-table:model_spring_mat} in Supplements), nor within North America (latitude: 0.10 days/degree [-0.06 - 0.26]; MAT: -0.09 days/$^{\circ}$C [-0.36 - 0.18]) or Europe (latitude: 0.10 days/degree [-0.23 - 0.42]; MAT: -0.16 days/$^{\circ}$C [-0.55 - 0.23]) (Fig.\ref{figure:continental_spptype_lat}A \& \ref{figure:continental_spptype_effect}A).
-Results were similar using other distance metrics in lieu of latitude (see Supplement Fig.\ref{supplement-figure:lat_distance} for results using the difference between provenance and garden latitude, and the spherical distance between provenance and garden).
+Overall, spring events were not related to provenance latitude or MAT, neither across continents (latitude: 0.10 days/degree [-0.05 - 0.25]; MAT: -0.11 days/$^{\circ}$C [-0.34 - 0.12]) (Fig.\ref{figure:springfall_latmat}, Table.S\ref{supplement-table:model_spring_lat} \& S\ref{supplement-table:model_spring_mat} in Supporting Information), nor within North America (latitude: 0.10 days/degree [-0.06 - 0.26]; MAT: -0.09 days/$^{\circ}$C [-0.36 - 0.18]) or Europe (latitude: 0.10 days/degree [-0.23 - 0.42]; MAT: -0.16 days/$^{\circ}$C [-0.55 - 0.23]) (Fig.\ref{figure:continental_spptype_lat}A \& \ref{figure:continental_spptype_effect}A).
+Results were similar using other distance metrics in lieu of latitude (see Fig.S\ref{supplement-figure:lat_distance} for results using the difference between provenance and garden latitude, and the spherical distance between provenance and garden).
 % Spring events advanced slightly with provenance latitude in Europe spring events are slightly earlier where provenance MAT is lower (i.e., higher, more northern latitudes), though uncertainty intervals strongly overlapped zero.
@@
 
-In contrast, fall events (e.g., budset, leaf senescence, leaf abscission) were earlier at more northern, cooler MAT sites (that is, they advanced strongly with provenance latitude: 3.16 days/degree [2.87-3.45], and with decreasing MAT:4.78 days/$^{\circ}$C [4.1 - 5.4], Fig.
-\ref{figure:springfall_latmat}, Table.\ref{supplement-table:model_fall_lat} \& \ref{supplement-table:model_fall_mat} in Supplements).
-This relationship, however, was observed mostly in North America where fall events advanced 4.24 (3.95 - 4.53) days per degree northward, or 6.41 days (5.78 - 7.04) per degree decline in MAT ($^{\circ}$C), whereas in Europe these relationships were weaker: advance of 0.47 (0.21 - 1.17) days per degree northward, or 0.70 days (1.04 - 2.42) per degree decline in MAT ($^{\circ}$C) (Fig.
-\ref{figure:continental_spptype_effect}A).
+In contrast, fall events (e.g., budset, leaf senescence, leaf abscission) were earlier at more northern, cooler MAT sites (that is, they advanced strongly with provenance latitude: 3.16 days/degree [2.87-3.45], and with decreasing MAT:4.78 days/$^{\circ}$C [4.1 - 5.4], Fig.\ref{figure:springfall_latmat}, Table.S\ref{supplement-table:model_fall_lat} \& S\ref{supplement-table:model_fall_mat} in Supporting Information).
+This relationship, however, was observed mostly in North America where fall events advanced 4.24 (3.95 - 4.53) days per degree northward, or 6.41 days (5.78 - 7.04) per degree decline in MAT ($^{\circ}$C), whereas in Europe these relationships were weaker: advance of 0.47 (0.21 - 1.17) days per degree northward, or 0.70 days (1.04 - 2.42) per degree decline in MAT ($^{\circ}$C) (Fig.\ref{figure:continental_spptype_effect}A).
 
@@
 Spring events advanced 0.82 (0.54 - 1.11) days/$^{\circ}$C as MAT increased in angiosperms and delayed 0.76 (0.37 - 1.14) days/$^{\circ}$C as MAT increased in gymnosperms.
-Fall events delayed in warmer locations for both species types, but slightly more so for gymnosperms (6.23 days) than angiosperms (3.69 days) (Fig.
-\ref{figure:continental_spptype_effect}B).
+Fall events delayed in warmer locations for both species types, but slightly more so for gymnosperms (6.23 days) than angiosperms (3.69 days) (Fig.\ref{figure:continental_spptype_effect}B).
 
@@
 For spring events, we considered overlap across March to May.
-However, results were not qualitatively different than using MAT (See Fig.
-\ref{supplement-figure:overlap_scatterplot} in Supplements).
+However, results were not qualitatively different than using MAT (See Fig.S\ref{supplement-figure:overlap_scatterplot} in Supporting Information).
 We observed very weak effects of climate overlap on spring events (0.01 [0.02 - 0.03] days per one percent increase in climate overlap), nearly identical across angiosperms (0.02 [0.00 - 0.05]) and gymnosperms (0.04 [0.00 - 0.09]).
-Fall events advanced as climate overlap declined, but slightly more strongly for gymnosperms (advancing 0.72 [0.51 - 0.92] days per one percent decline in climate overlap) (Fig.\ref{supplement-figure:overlap_posterior} in Supplements).
+Fall events advanced as climate overlap declined, but slightly more strongly for gymnosperms (advancing 0.72 [0.51 - 0.92] days per one percent decline in climate overlap) (Fig.S\ref{supplement-figure:overlap_posterior} in Supporting Information).
 

### 2024-02-05 Docs/manuscriptRev.Rnw
Edits to address some of the reviewer concerns.

--- 2023-11-26 Docs/manuscript.Rnw
@@
 
-\begin{abstract}
+\begin{abstract} %EWMFeb5 -- edited to define provenance differences 
 Forecasting the biological impacts of climate change requires understanding how species respond to warmer temperatures through inter-annual flexible variation versus through adaptation to local conditions.
@@
 The latter is the case for shifts in spring phenology---one of the most reported and consistent impacts of anthropogenic climate change, and also one of the most critical to forecasting, given its role in carbon sequestration.
-Some common garden studies have found evidence of important provenance effects, which suggest there may be local adaptation in the underlying cues of spring phenology and mirrors findings for fall events, while other studies find no evidence.
+Some common garden studies have found evidence of important provenance effects (i.e.
+population differences), which suggest there may be local adaptation in the underlying cues of spring phenology and mirrors findings for fall events, while other studies find no evidence.
 Here, we synthesize common garden studies across Europe and North America that reported spring event dates for a mix of angiosperm and gymnosperm tree species in the northern hemisphere, capturing data from 384 North American provenances and 101 European provenances with observations from 1962 to 2019, alongside fall event data when provided.
@@
 
-\noindent \emph{Keywords: budburst, budset, climate change, common gardens, forests, leafout, provenance, spring phenology}\\
+\noindent \emph{Keywords: budburst, leafout, budset, senescence, climate change, common gardens, deciduous and evergreen trees, provenance, spring phenology}\\ %EWMFeb5 -- if you have an abundance of keywords, remove ones that are redundant with title/abstract
 
@@
 
+%EWMFeb5 -- edits below to define provenance; and added a reference to 'local adaptation'
 Studies of adaptation versus plasticity in tree phenology have been conducted for centuries \citep{Cleland:2007or}, through common garden studies.
-In these studies---conducted often for forestry purposes---researchers grow trees of different geographical origins under the same environmental conditions to disentangle the effects of environmental and genetic variation on trees’ phenotypes \citep{AitkenBemmels16, Alberto13}.
-Such work has established common clines in fall phenology, as source locations with shorter growing seasons (poleward and higher elevations) exhibit earlier stopping of growth (such as budset).
+In these studies---conducted often for forestry purposes---researchers grow trees of different geographical origins (called `provenances' often in forestry) under the same environmental conditions to disentangle the effects of environmental and genetic variation on trees’ phenotypes \citep{AitkenBemmels16, Alberto13}.
+Such work has established common clines in fall phenology suggestive of local adaptation, as source locations with shorter growing seasons (poleward and higher elevations) exhibit earlier stopping of growth (such as budset).
 Research has connected these clines to an underlying proximate mechanism of changing photoperiod cues (i.e., shifts in the photoperiod threshold required to trigger budset), driven by adaptation to the local growing season \citep{Alberto13,Savolainen07}.
@@
 To test for evidence of adaptation in spring phenology and what factors may underlies differences observed across studies, we comprehensively examined clines for spring events, including fall events when possible.
-We tested for evidence of adaptation via provenance trends with latitude and climate and examined possible factors that underlie these clines, including for differences between: (1) spring and fall phenology, (2) studies in Europe and North America, (3) angiosperm and gymnosperm species.
+We tested for evidence of adaptation via provenance trends with latitude and climate and examined possible factors that underlie these clines, including for differences between: (1) spring and fall phenology, (2) studies in Europe and North America, (3) angiosperm and gymnosperm species, which represent a deep evolutionary split in the plant tree of life.
 To address these questions, we combined Bayesian hierarchical models with a new meta-analysis of all common garden experiments in temperate tree species across Europe and North America reporting spring phenology.
+%EWMFeb5 -- edits above to deal with angio/gymno
 
@@
 \section{Results}
+%EWMFeb5 -- we do a nice job here of mentioning the angiosperms are deciduous and the gymnosperms are evergreen. Could highlight this in response letter. 
 Our final dataset included seven deciduous angiosperm and eight evergreen gymnosperm species from 17 studies and 19 gardens, encompassing 384 North American provenances and 101 European provenances, with observations from 1962 to 2019.
@@
 
+%EWMFeb5 -- small edits below about angio/gymno
 Trends between spring phenology and latitude or MAT were weak, but suggested the possibility of diverging results that could mute an overall trend---albeit a much weaker one than for fall phenology.
-We found angiosperm versus gymnosperm species diverged in their clines with MAT.
+We found angiosperm (all deciduous) versus gymnosperm (all evergreen) species diverged in their clines with MAT.
 Combined with our finding of much stronger clines for fall phenology in North America, these results support the idea of potential variation across continents and/or species type that may underlie the debate in whether spring events show important clinal variation.
@@
 %EMWSep30: Please double-check we're explaining trends correctly before you submit; I think we are -- but please check closely.
+%EWMFeb5 -- more edits here about angio/gymno
 Our results highlight that spring events show much weaker clines than fall events in tree species, but suggest important variation between species types in spring events.
 Angiosperms tended to budburst earlier in provenances that were warmer and more southerly, while gymnosperms trended in the opposite direction.
-Such differences could be driven by their varying investment strategies.
-As gymnosperms photosynthesize without leafout they generally leafout much later than angiosperms, after most risk of spring frost \citep{panchen14}, and thus may avoid frost risk more than angiosperms.
-In contrast, angiosperms may tend towards earlier leafout in warmer climates to compete best for access to light and other resources \citep{cat2019}.
-Testing these hypotheses would require more information on frost risk and forest community assembly from across the provenance locations.
+Such differences could be driven by the varying investment strategies, given that all our studied gymnosperms were evergreen and all angiosperms were deciduous.
+As evergreen species photosynthesize without leafout they generally leafout much later than deciduous species, after most risk of spring frost \citep{panchen14}, and thus may avoid frost risk.
+In contrast, deciduous species may tend towards earlier leafout in warmer climates to compete best for access to light and other resources \citep{cat2019}.
+Testing these hypotheses would require more information on frost risk and forest community assembly from across the provenance locations, but seems an important step towards understanding the drivers of this variation.
+Without a clear mechanism, extrapolating these results to other species or across Europe and North America may be difficult, especially given biases in the data.
+%EWMFeb5 -- trying to address request by R2 about extrapolating while also keeping ms flow.
 %EMWSep4 -- I like the below (and I think it may have been my idea), but we'd likely be asked to test for it ...
@@
 Spring events did not shift with latitude by continent, but fall events advanced strongly per degree northward, particularly in North America.
-Spring events slightly advanced in gymnosperms and delayed in angiosperms per degree northward.
-Fall events advanced per degree northward for both species types.}
+Spring events slightly advanced in gymnosperms (all evergreen species) and delayed in angiosperms (all deciduous species) per degree northward.
+Fall events advanced per degree northward for both species types.} %EWMFeb5 -- more edits for angio/gymno
     \label{figure:continental_spptype_lat}

### 2024-02-17 Docs/manuscriptRev.Rnw
revised ms

@@
 %%% start preambling . . .  %%%
-\documentclass{article}
+\documentclass[12pt]{article}
+% \setmainfont{Times New Roman}
 % required
@@
 \pagestyle{fancy}
-\fancyhead[LO]{2023}
+\fancyhead[LO]{2024}
 \fancyhead[RO]{Manuscript}
@@
 \noindent{\bf Corresponding author:} Z. A. Zeng, see $^{1}$ above ; E-mail: alinazengziyun@yahoo.com\\
-
-\noindent Additional information: Zeng (0009-0000-4647-9457); Wolkovich (0000-0001-7653-893X)\\
-
-\noindent Running head: Clines in spring phenology
+\noindent{\bf Additional information:} Zeng (0009-0000-4647-9457); Wolkovich (0000-0001-7653-893X)\\
+\noindent{\bf Running head:} Clines in spring phenology\\
+\noindent{\bf Word count:} Summary - 198, Introduction - 1077, Materials and Methods - 821, Results - 1012, Discussion - 1400\\
+\noindent{\bf Number of figures:} 4 (Among these, Figure 1 and 2 should be published in color)\\
+1. Figure 1: Map showing the distribution of common gardens and provenances.\\
+2. Figure 2: Event day of year in relation to provenance latitude and MAT.\\
+3. Figure 3: Effects of latitude on spring and fall event day of year depending on continent and species leaf type.\\
+4. Figure 4: Effects of MAT on spring and fall event day of year depending on continent and species leaf type.\\
+\noindent{\bf Supporting Information (brief legends):}\\
+1. Methods S1: Additional methods.\\
+2. Table S1: Showing all publications included in meta-analysis.\\
+3. Table S2-5: Showing summary of model estimates.\\
+4. Figures S1-3: Supporting figures.
 
@@
 
-\begin{abstract} %EWMFeb5 -- edited to define provenance differences 
-Forecasting the biological impacts of climate change requires understanding how species respond to warmer temperatures through inter-annual flexible variation versus through adaptation to local conditions.
+% \begin{abstract} %EWMFeb5 -- edited to define provenance differences
+
+\begin{itemize}
+  \item Forecasting the biological impacts of climate change requires understanding how species respond to warmer temperatures through inter-annual flexible variation vs.
+through adaptation to local conditions.
+Yet, we often lack this information entirely or find conflicting evidence across previous studies.
+  \item We synthesize common garden studies across Europe and North America that reported spring event dates for a mix of angiosperm and gymnosperm tree species in the northern hemisphere, capturing data from 384 North American and 101 European provenances (i.e.
+populations) with observations from 1962 to 2019, alongside fall event data when provided.
+  \item Across continents, we find no evidence of provenance effects (i.e.
+population differences) in spring phenology, but strong clines with latitude and mean annual temperature (MAT) in fall.
+These effects, however, appear to diverge by continent and species type (gymnosperm vs.
+angiosperm), with particularly pronounced clines in North America in fall.
+  \item Our results suggest flexible, likely plastic responses, in spring phenology with warming, and potential limits---at least in the short term---due to provenance effects for fall phenology.
+They also highlight that, after over 250 years of common garden studies on tree phenology, we still lack a holistic predictive model of clines across species and phenological events.
+\end{itemize}
+% Forecasting the biological impacts of climate change requires understanding how species respond to warmer temperatures through inter-annual flexible variation versus through adaptation to local conditions.
 Yet, we often lack this information entirely or find conflicting evidence across studies.
@@
 They also highlight that, after over 250 years of common garden studies on tree phenology, we still lack a holistic predictive model of clines across species and phenological events.
-\end{abstract} %Spring tree phenology is critical to forecasting ...
-Despite a long history of study, however, the balance of local adaptation versus plasticity in determining environmental responses is not well understood, especially for spring events.
-We conducted the first cross-continental meta-analysis of published studies from the peer-reviewed literature that reported spring event dates for a mix of angiosperm and gymnosperm tree species in the northern hemisphere, capturing data  for  seven angiosperm and eight gymnosperm species from 17 studies, from 384 North American provenances and 101 European provenances with observations from 1962 to 2019.
-We found ...
-This is important because...
+% \end{abstract}
 
-\noindent \emph{Keywords: budburst, leafout, budset, senescence, climate change, common gardens, deciduous and evergreen trees, provenance, spring phenology}\\ %EWMFeb5 -- if you have an abundance of keywords, remove ones that are redundant with title/abstract
+\noindent \emph{Keywords: budburst, budset, climate change, common gardens, deciduous and evergreen trees, leafout, senescence, spring phenology}\\ %EWMFeb5 -- if you have an abundance of keywords, remove ones that are redundant with title/abstract
 
@@
 Studies of adaptation versus plasticity in tree phenology have been conducted for centuries \citep{Cleland:2007or}, through common garden studies.
-In these studies---conducted often for forestry purposes---researchers grow trees of different geographical origins (called `provenances' often in forestry) under the same environmental conditions to disentangle the effects of environmental and genetic variation on trees’ phenotypes \citep{AitkenBemmels16, Alberto13}.
-Such work has established common clines in fall phenology suggestive of local adaptation, as source locations with shorter growing seasons (poleward and higher elevations) exhibit earlier stopping of growth (such as budset).
+In these studies---conducted often for forestry purposes---researchers grow trees of different geographical origins (called `provenances' often in forestry) under the same environmental conditions to disentangle the effects of environmental and genetic variation on trees' phenotypes \citep{AitkenBemmels16, Alberto13}.
+Such work has established common clines in fall phenology suggestive of local adaptation, as source locations with shorter growing seasons (poleward and higher elevations) exhibit earlier growth cessation (such as budset).
 Research has connected these clines to an underlying proximate mechanism of changing photoperiod cues (i.e., shifts in the photoperiod threshold required to trigger budset), driven by adaptation to the local growing season \citep{Alberto13,Savolainen07}.
@@
 
-Based on these criteria we found 19 common gardens distributed throughout North America and Europe, with the majority of data concentrated in North America (Fig.\ref{figure:map_gardens} \& Table.S\ref{supplement-table:all_studies} in Supporting Information).
+Based on these criteria we found 19 common gardens distributed throughout North America and Europe, with the majority of data concentrated in western North America (Fig.\ref{figure:map_gardens} \& Table.S\ref{supplement-table:all_studies} in Supporting Information).
 From each common garden study we extracted phenological data on spring events (budburst and leaf flush) in DOY and, when present in the same paper, fall events (bud set, leaf senescence, growth cessation, and leaf abscission) by species and the geographic information of provenances and gardens.
@@
 We extracted data from E-OBS for European locations and used the `daymetr' in R for North American locations \citep{cornes2018,hufkens2018}.
-Then, using these data and the `overlap' package in R, we estimated how much the daily temperatures overlapped between each provenance location and their corresponding gardens across the three months for our 10 year period, which we call `climate overlap.' Dataset containing event dates, geographic information, and climatic information of all provenances are archived in  Knowledge Network for Biocomplexity (KNB) \citep{zeng23}.
+Then, using these data and the `overlap' package in R, we estimated how much the daily temperatures overlapped between each provenance location and their corresponding gardens across the three months from 2011 to 2020, which we call `climate overlap.' Dataset containing event dates, geographic information, and climatic information of all provenances are archived in  Knowledge Network for Biocomplexity (KNB) \citep{zeng23}.
 % Finally, using the daily temperature data, we also calculated growing degree days (GDD), a commonly used heat accumulation measure to forecast phenological development in plants \citep{miller01}, for each provenance and garden, based on the accumulation of mean daily temperatures ($T_{m}$)from 2011 to 2020 above a baseline of 0$^{\circ}$C, from January 1 until budburst and leaf flush with the following formula: GDD = $\sum$($T_{m}$ - $0^{\circ}$C for $T_{m}$ $\ge$ $0^{\circ}$C; 0 for $T_{m}$ $\le$ $0^{\circ}$C).
@@
 
-Overall, spring events were not related to provenance latitude or MAT, neither across continents (latitude: \Sexpr{format(round(mean(all_species_springpost_fitC_spring_lat),digits = 2), nsmall=2)} days/degree [\Sexpr{format(quantile(all_species_springpost_fitC_spring_lat, 0.05), digits = 1)} - \Sexpr{format(quantile(all_species_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(round(mean(all_species_springpost_fitC_spring_mat),digits = 2), nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(all_species_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(all_species_springpost_fitC_spring_mat, 0.95), digits = 2)}]) (Fig.\ref{figure:springfall_latmat}, Table.S\ref{supplement-table:model_spring_lat} \& S\ref{supplement-table:model_spring_mat} in Supporting Information), nor within North America (latitude: \Sexpr{format(round(mean(northamerican_springpost_fitC_spring_lat),digits = 2), nsmall=2)} days/degree [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.05), digits = 1)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(mean(northamerican_springpost_fitC_spring_mat),digits = 1)} days/$^{\circ}$C [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.95), digits = 2)}]) or Europe (latitude: \Sexpr{format(round(mean(european_springpost_fitC_spring_lat),digits = 2),nsmall=2)} days/degree [\Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(round(mean(european_springpost_fitC_spring_mat),digits = 2),nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.95), digits = 2)}]) (Fig.\ref{figure:continental_spptype_lat}A \& \ref{figure:continental_spptype_effect}A).
+Overall, spring events such as budburst and leaf flush were not related to provenance latitude or MAT, neither across continents (latitude: \Sexpr{format(round(mean(all_species_springpost_fitC_spring_lat),digits = 2), nsmall=2)} days/degree [\Sexpr{format(quantile(all_species_springpost_fitC_spring_lat, 0.05), digits = 1)} - \Sexpr{format(quantile(all_species_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(round(mean(all_species_springpost_fitC_spring_mat),digits = 2), nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(all_species_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(all_species_springpost_fitC_spring_mat, 0.95), digits = 2)}]) (Fig.\ref{figure:springfall_latmat}, Table.S\ref{supplement-table:model_spring_lat} \& S\ref{supplement-table:model_spring_mat} in Supporting Information), nor within North America (latitude: \Sexpr{format(round(mean(northamerican_springpost_fitC_spring_lat),digits = 2), nsmall=2)} days/degree [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.05), digits = 1)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(mean(northamerican_springpost_fitC_spring_mat),digits = 1)} days/$^{\circ}$C [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.95), digits = 2)}]) or Europe (latitude: \Sexpr{format(round(mean(european_springpost_fitC_spring_lat),digits = 2),nsmall=2)} days/degree [\Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(round(mean(european_springpost_fitC_spring_mat),digits = 2),nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.95), digits = 2)}]) (Fig.\ref{figure:continental_spptype_lat}A \& \ref{figure:continental_spptype_effect}A).
 Results were similar using other distance metrics in lieu of latitude (see Fig.S\ref{supplement-figure:lat_distance} for results using the difference between provenance and garden latitude, and the spherical distance between provenance and garden).
@@
 
-While we expected that coarse metrics, such as latitude and MAT, would generally represent how similar the climates are between the provenances and gardens, we also estimated climate overlap in months much closer to the events to further test how much climate similarity between provenances and gardens predicts provenance effects.
+While we expected that coarse metrics, such as latitude and MAT, would generally represent how similar the climates are between the provenances and gardens, we also estimated climate overlap in months much closer to the events to further test how much climate similarity between provenances and gardens predicts provenance effects (i.e.
+differential responses observed among plant populations from different geographical origins).
 For spring events, we considered overlap across March to May.
@@
 Testing these hypotheses would require more information on frost risk and forest community assembly from across the provenance locations, but seems an important step towards understanding the drivers of this variation.
-Without a clear mechanism, extrapolating these results to other species or across Europe and North America may be difficult, especially given biases in the data.
+Without a clear mechanism, extrapolating these results to other species or across Europe and North America may be difficult, especially given biases in the data, and the distinct climatic, geographical, and ecological contexts of Europe and North America.
+Without a clear and universal mechanism driving the observed phenological responses, the direct applicability of our findings to other contexts must be approached with caution.
 %EWMFeb5 -- trying to address request by R2 about extrapolating while also keeping ms flow.
@@
 Angiosperm species showed a weak trend towards earlier budburst in populations from warmer locations.
-In contrast fall event clines in Europe were weak, but their fall event type (leaf senescence) is less likely to show clinal variation.
+In contrast, fall event clines in Europe were weak, but their fall event type (leaf senescence) is less likely to show clinal variation.
 Given European studies were biased towards studying angiosperms (3/4 species) that are often canopy-species (e.g., \emph{Fagus}, \emph{Quercus}) these results together could explain a greater focus on clinal variation in spring events across European studies \citep{Basler:2012,gauzere2020,sog08,deans96,von95}.
@@
 See Table.S\ref{supplement-table:all_studies} in Supporting Information for sourcing information on selected studies.
-Note: map lines do not necessarily depict accepted national boundaries.}
+Note: map lines do not necessarily depict accepted national boundaries.
+European sites are made bold and denoted by an asterisk (*).
+}
     \label{figure:map_gardens}
@@
 Spring events shown on top and fall events at the bottom.
-European gardens and species are bolded and denoted by an asterisk (*).
+European gardens and species are made bold and denoted by an asterisk (*).
 }
@@
 
-\section{Supporting Information (brief legends)}
-
-1. Additional methods on exact procedures we followed for meta-analysis
-and reasons for the exclusion of a small number of studies.
-2. Table S1: showing all publications included in meta-analysis.
-3. Tables S2-5: showing model estimates.
-4. Figures S1-3: supporting figures.
-
 \end{document}

### 2024-02-17 Docs/manuscriptRev.tex
revised ms

--- 2024-02-17 Docs/manuscriptRev.Rnw
@@
 \begin{document}
-\SweaveOpts{concordance=TRUE} % For RStudio hiccups
+\input{/Users/alina/Documents/git/localadaptclim/Docs/figure_ms/demoFig-concordance} % For RStudio hiccups
 \begin{flushright}
@@
 
-<<label=source_spring_models, echo=FALSE>>=
-#load models and data to report numbers of things
-# "C:/Users/alina/Documents/git/localadaptclim"
-# setwd("C:/Users/alina/Documents/git/localadaptclim/Docs")
-#getwd()
-source("../Analyses/source_spring_models_for_sweave.R")
-@
-
-Overall, spring events such as budburst and leaf flush were not related to provenance latitude or MAT, neither across continents (latitude: \Sexpr{format(round(mean(all_species_springpost_fitC_spring_lat),digits = 2), nsmall=2)} days/degree [\Sexpr{format(quantile(all_species_springpost_fitC_spring_lat, 0.05), digits = 1)} - \Sexpr{format(quantile(all_species_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(round(mean(all_species_springpost_fitC_spring_mat),digits = 2), nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(all_species_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(all_species_springpost_fitC_spring_mat, 0.95), digits = 2)}]) (Fig.\ref{figure:springfall_latmat}, Table.S\ref{supplement-table:model_spring_lat} \& S\ref{supplement-table:model_spring_mat} in Supporting Information), nor within North America (latitude: \Sexpr{format(round(mean(northamerican_springpost_fitC_spring_lat),digits = 2), nsmall=2)} days/degree [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.05), digits = 1)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(mean(northamerican_springpost_fitC_spring_mat),digits = 1)} days/$^{\circ}$C [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.95), digits = 2)}]) or Europe (latitude: \Sexpr{format(round(mean(european_springpost_fitC_spring_lat),digits = 2),nsmall=2)} days/degree [\Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(round(mean(european_springpost_fitC_spring_mat),digits = 2),nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.95), digits = 2)}]) (Fig.\ref{figure:continental_spptype_lat}A \& \ref{figure:continental_spptype_effect}A).
+Overall, spring events such as budburst and leaf flush were not related to provenance latitude or MAT, neither across continents (latitude: 0.10 days/degree [-0.05 - 0.25]; MAT: -0.11 days/$^{\circ}$C [-0.34 - 0.12]) (Fig.\ref{figure:springfall_latmat}, Table.S\ref{supplement-table:model_spring_lat} \& S\ref{supplement-table:model_spring_mat} in Supporting Information), nor within North America (latitude: 0.10 days/degree [-0.06 - 0.26]; MAT: -0.09 days/$^{\circ}$C [-0.36 - 0.18]) or Europe (latitude: 0.10 days/degree [-0.23 - 0.42]; MAT: -0.16 days/$^{\circ}$C [-0.55 - 0.23]) (Fig.\ref{figure:continental_spptype_lat}A \& \ref{figure:continental_spptype_effect}A).
 Results were similar using other distance metrics in lieu of latitude (see Fig.S\ref{supplement-figure:lat_distance} for results using the difference between provenance and garden latitude, and the spherical distance between provenance and garden).
@@
 
-<<label=source_fall_models, echo=FALSE>>=
-#load models and data to report numbers of things
-# "C:/Users/alina/Documents/git/localadaptclim"
-# setwd("C:/Users/alina/Documents/git/localadaptclim/Docs")Aitken and Bammals (2015
-#getwd()
-source("../Analyses/source_fall_models_for_sweave.R")
-@
-
-In contrast, fall events (e.g., budset, leaf senescence, leaf abscission) were earlier at more northern, cooler MAT sites (that is, they advanced strongly with provenance latitude: \Sexpr{format(round(abs(mean(all_species_fallpost_fitC_fall_lat)),digits = 2), nsmall=2)} days/degree [\Sexpr{format(round(abs(quantile(all_species_fallpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)}-\Sexpr{format(round(abs(quantile(all_species_fallpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}], and with decreasing MAT:\Sexpr{format(round(abs(mean(all_species_fallpost_fitC_fall_mat)),digits = 2), nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(all_species_fallpost_fitC_fall_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(all_species_fallpost_fitC_fall_mat, 0.95), digits = 2)}], Fig.\ref{figure:springfall_latmat}, Table.S\ref{supplement-table:model_fall_lat} \& S\ref{supplement-table:model_fall_mat} in Supporting Information).
-This relationship, however, was observed mostly in North America where fall events advanced \Sexpr{format(round(abs(mean(northamericanpost_fitC_fall_lat)),digits = 2),nsmall=2)} (\Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree northward, or \Sexpr{format(round(mean(northamericanpost_fitC_fall_mat),digits = 2),nsmall=2)} days (\Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_mat, 0.95)), digits = 2),nsmall=2)}) per degree decline in MAT ($^{\circ}$C), whereas in Europe these relationships were weaker: advance of \Sexpr{format(round(abs(mean(europeanpost_fitC_fall_lat)),digits = 2),nsmall=2)} (\Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree northward, or \Sexpr{format(round(mean(europeanpost_fitC_fall_mat),digits = 2),nsmall=2)} days (\Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_mat, 0.95)), digits = 2),nsmall=2)}) per degree decline in MAT ($^{\circ}$C) (Fig.\ref{figure:continental_spptype_effect}A).
+In contrast, fall events (e.g., budset, leaf senescence, leaf abscission) were earlier at more northern, cooler MAT sites (that is, they advanced strongly with provenance latitude: 3.16 days/degree [2.87-3.45], and with decreasing MAT:4.78 days/$^{\circ}$C [4.1 - 5.4], Fig.\ref{figure:springfall_latmat}, Table.S\ref{supplement-table:model_fall_lat} \& S\ref{supplement-table:model_fall_mat} in Supporting Information).
+This relationship, however, was observed mostly in North America where fall events advanced 4.24 (3.95 - 4.53) days per degree northward, or 6.41 days (5.78 - 7.04) per degree decline in MAT ($^{\circ}$C), whereas in Europe these relationships were weaker: advance of 0.47 (0.21 - 1.17) days per degree northward, or 0.70 days (1.04 - 2.42) per degree decline in MAT ($^{\circ}$C) (Fig.\ref{figure:continental_spptype_effect}A).
 
@@
 Effects of provenance latitude on fall events were similar across angiosperms and gymnosperms (Fig.\ref{figure:continental_spptype_lat}B).
-Spring events weakly diverged, delaying at a rate of \Sexpr{format(round(abs(mean(deciduous_springpost_fitC_spring_lat)),digits = 2), nsmall=2)} (\Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_lat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_lat, 0.95)), digits = 2),nsmall=2)}) days per degree north in angiosperms and advancing \Sexpr{format(round(abs(mean(evergreens_springpost_fitC_spring_lat)),digits = 2), nsmall=2)} (\Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_lat, 0.05)), digits = 2),nsmall=2)}) days per degree north in gymnosperms.
-Fall events advanced \Sexpr{format(round(abs(mean(deciduouspost_fitC_fall_lat)),digits = 2), nsmall=2)} (\Sexpr{format(round(abs(quantile(deciduouspost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(deciduouspost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree north in angiosperms and \Sexpr{format(round(abs(mean(evergreenspost_fitC_fall_lat)),digits = 2), nsmall=2)} (\Sexpr{format(round(abs(quantile(evergreenspost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)}-\Sexpr{format(round(abs(quantile(evergreenspost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree north in gymnosperms.
+Spring events weakly diverged, delaying at a rate of 0.37 (0.15 - 0.59) days per degree north in angiosperms and advancing 0.23 (0.00 - 0.46) days per degree north in gymnosperms.
+Fall events advanced 3.18 (2.76 - 3.62) days per degree north in angiosperms and 3.14 (2.81-3.47) days per degree north in gymnosperms.
 Effects of MAT on spring events also weakly diverged (Fig.\ref{figure:continental_spptype_effect}B).
-Spring events advanced \Sexpr{format(round(abs(mean(deciduous_springpost_fitC_spring_mat)),digits = 2), nsmall=2)} (\Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_mat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_mat, 0.05)), digits = 2),nsmall=2)}) days/$^{\circ}$C as MAT increased in angiosperms and delayed \Sexpr{format(round(abs(mean(evergreens_springpost_fitC_spring_mat)),digits = 2), nsmall=2)} (\Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_mat, 0.95)), digits = 2),nsmall=2)}) days/$^{\circ}$C as MAT increased in gymnosperms.
-Fall events delayed in warmer locations for both species types, but slightly more so for gymnosperms (\Sexpr{format(round(mean(evergreenspost_fitC_fall_mat),digits = 2),nsmall=2)} days) than angiosperms (\Sexpr{format(round(mean(deciduouspost_fitC_fall_mat),digits = 2),nsmall=2)} days) (Fig.\ref{figure:continental_spptype_effect}B).
+Spring events advanced 0.82 (0.54 - 1.11) days/$^{\circ}$C as MAT increased in angiosperms and delayed 0.76 (0.37 - 1.14) days/$^{\circ}$C as MAT increased in gymnosperms.
+Fall events delayed in warmer locations for both species types, but slightly more so for gymnosperms (6.23 days) than angiosperms (3.69 days) (Fig.\ref{figure:continental_spptype_effect}B).
 
@@
 However, results were not qualitatively different than using MAT (See Fig.S\ref{supplement-figure:overlap_scatterplot} in Supporting Information).
-We observed very weak effects of climate overlap on spring events (\Sexpr{format(round(abs(mean(all_species_springpost_fitC_spring_overlap)),digits = 2), nsmall=2)} [\Sexpr{format(round(abs(quantile(all_species_springpost_fitC_spring_overlap, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(all_species_springpost_fitC_spring_overlap, 0.95)), digits = 2),nsmall=2)}] days per one percent increase in climate overlap), nearly identical across angiosperms (\Sexpr{format(round(abs(mean(deciduous_springpost_fitC_spring_overlap)),digits = 2), nsmall=2)} [\Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_overlap, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_overlap, 0.05)), digits = 2),nsmall=2)}]) and gymnosperms (\Sexpr{format(round(abs(mean(evergreens_springpost_fitC_spring_overlap)),digits = 2), nsmall=2)} [\Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_overlap, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_overlap, 0.95)), digits = 2),nsmall=2)}]).
-Fall events advanced as climate overlap declined, but slightly more strongly for gymnosperms (advancing \Sexpr{format(round(abs(mean(evergreenspost_fitC_fall_overlap)),digits = 2), nsmall=2)} [\Sexpr{format(round(abs(quantile(evergreenspost_fitC_fall_overlap, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(evergreenspost_fitC_fall_overlap, 0.95)), digits = 2),nsmall=2)}] days per one percent decline in climate overlap) (Fig.S\ref{supplement-figure:overlap_posterior} in Supporting Information).
+We observed very weak effects of climate overlap on spring events (0.01 [0.02 - 0.03] days per one percent increase in climate overlap), nearly identical across angiosperms (0.02 [0.00 - 0.05]) and gymnosperms (0.04 [0.00 - 0.09]).
+Fall events advanced as climate overlap declined, but slightly more strongly for gymnosperms (advancing 0.72 [0.51 - 0.92] days per one percent decline in climate overlap) (Fig.S\ref{supplement-figure:overlap_posterior} in Supporting Information).
 

### 2024-02-21 Docs/manuscriptRev.Rnw
tracking changes -> diff file
A couple small tweaks throughout for flow, clarity

@@
 
-\begin{itemize}
+\begin{itemize} %EMWFeb20 -- small tweaks to improve flow and we do not need to say 'population' so much, methinks. 
   \item Forecasting the biological impacts of climate change requires understanding how species respond to warmer temperatures through inter-annual flexible variation vs.
 through adaptation to local conditions.
-Yet, we often lack this information entirely or find conflicting evidence across previous studies.
+Yet, we often lack this information entirely or find conflicting evidence across studies, which is the case for spring phenology.
   \item We synthesize common garden studies across Europe and North America that reported spring event dates for a mix of angiosperm and gymnosperm tree species in the northern hemisphere, capturing data from 384 North American and 101 European provenances (i.e.
 populations) with observations from 1962 to 2019, alongside fall event data when provided.
-  \item Across continents, we find no evidence of provenance effects (i.e.
-population differences) in spring phenology, but strong clines with latitude and mean annual temperature (MAT) in fall.
+  \item Across continents, we find no evidence of provenance effects in spring phenology, but strong clines with latitude and mean annual temperature (MAT) in fall.
 These effects, however, appear to diverge by continent and species type (gymnosperm vs.
-angiosperm), with particularly pronounced clines in North America in fall.
+angiosperm), with particularly pronounced clines in North America in fall events.
+  %EWMFeb5 -- if you need to save words, can cut 'due to provenance effects' below. 
   \item Our results suggest flexible, likely plastic responses, in spring phenology with warming, and potential limits---at least in the short term---due to provenance effects for fall phenology.
@@
 
-Overall, spring events such as budburst and leaf flush were not related to provenance latitude or MAT, neither across continents (latitude: \Sexpr{format(round(mean(all_species_springpost_fitC_spring_lat),digits = 2), nsmall=2)} days/degree [\Sexpr{format(quantile(all_species_springpost_fitC_spring_lat, 0.05), digits = 1)} - \Sexpr{format(quantile(all_species_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(round(mean(all_species_springpost_fitC_spring_mat),digits = 2), nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(all_species_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(all_species_springpost_fitC_spring_mat, 0.95), digits = 2)}]) (Fig.\ref{figure:springfall_latmat}, Table.S\ref{supplement-table:model_spring_lat} \& S\ref{supplement-table:model_spring_mat} in Supporting Information), nor within North America (latitude: \Sexpr{format(round(mean(northamerican_springpost_fitC_spring_lat),digits = 2), nsmall=2)} days/degree [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.05), digits = 1)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(mean(northamerican_springpost_fitC_spring_mat),digits = 1)} days/$^{\circ}$C [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.95), digits = 2)}]) or Europe (latitude: \Sexpr{format(round(mean(european_springpost_fitC_spring_lat),digits = 2),nsmall=2)} days/degree [\Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(round(mean(european_springpost_fitC_spring_mat),digits = 2),nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.95), digits = 2)}]) (Fig.\ref{figure:continental_spptype_lat}A \& \ref{figure:continental_spptype_effect}A).
+Overall, spring events such as budburst and leaf flush, were not related to provenance latitude or MAT, neither across continents (latitude: \Sexpr{format(round(mean(all_species_springpost_fitC_spring_lat),digits = 2), nsmall=2)} days/degree [\Sexpr{format(quantile(all_species_springpost_fitC_spring_lat, 0.05), digits = 1)} - \Sexpr{format(quantile(all_species_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(round(mean(all_species_springpost_fitC_spring_mat),digits = 2), nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(all_species_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(all_species_springpost_fitC_spring_mat, 0.95), digits = 2)}]) (Fig.\ref{figure:springfall_latmat}, Table.S\ref{supplement-table:model_spring_lat} \& S\ref{supplement-table:model_spring_mat} in Supporting Information), nor within North America (latitude: \Sexpr{format(round(mean(northamerican_springpost_fitC_spring_lat),digits = 2), nsmall=2)} days/degree [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.05), digits = 1)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(mean(northamerican_springpost_fitC_spring_mat),digits = 1)} days/$^{\circ}$C [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.95), digits = 2)}]) or Europe (latitude: \Sexpr{format(round(mean(european_springpost_fitC_spring_lat),digits = 2),nsmall=2)} days/degree [\Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(round(mean(european_springpost_fitC_spring_mat),digits = 2),nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.95), digits = 2)}]) (Fig.\ref{figure:continental_spptype_lat}A \& \ref{figure:continental_spptype_effect}A).
 Results were similar using other distance metrics in lieu of latitude (see Fig.S\ref{supplement-figure:lat_distance} for results using the difference between provenance and garden latitude, and the spherical distance between provenance and garden).
@@
 
-We found the coarse metrics of provenance latitude and MAT were generally good predictors of phenology, performing better than our more complex and data-rich metric of climate overlap.
+%EWMFeb20 -- tweak below, could cite these lines also for the response to reviewer about MAT
+We found the coarse metrics of provenance latitude and MAT were generally good predictors of phenology, performing better than our more complex, data-rich, and season-specific metric of climate overlap.
 Latitude and MAT appeared to well represent how similar the climates are between the provenances and gardens in the temperate and boreal forest species we studied, yielding similar results to metrics calculated specifically in the spring with daily climate data.
@@
 Without a clear mechanism, extrapolating these results to other species or across Europe and North America may be difficult, especially given biases in the data, and the distinct climatic, geographical, and ecological contexts of Europe and North America.
-Without a clear and universal mechanism driving the observed phenological responses, the direct applicability of our findings to other contexts must be approached with caution.
+%EWMFeb20 -- deleted last sentence here, update letter
 %EWMFeb5 -- trying to address request by R2 about extrapolating while also keeping ms flow. 
@@
 Note: map lines do not necessarily depict accepted national boundaries.
-European sites are made bold and denoted by an asterisk (*).
+European sites are shown in bold and denoted by an asterisk (*).
 }
@@
 Spring events shown on top and fall events at the bottom.
-European gardens and species are made bold and denoted by an asterisk (*).
+European gardens and species are shown in bold and denoted by an asterisk (*).
 }

### 2024-02-21 Docs/manuscriptRev.tex
tracking changes -> diff file
submitting revision to new phytologist

@@
 
-Overall, spring events such as budburst and leaf flush were not related to provenance latitude or MAT, neither across continents (latitude: 0.10 days/degree [-0.05 - 0.25]; MAT: -0.11 days/$^{\circ}$C [-0.34 - 0.12]) (Fig.\ref{figure:springfall_latmat}, Table.S\ref{supplement-table:model_spring_lat} \& S\ref{supplement-table:model_spring_mat} in Supporting Information), nor within North America (latitude: 0.10 days/degree [-0.06 - 0.26]; MAT: -0.09 days/$^{\circ}$C [-0.36 - 0.18]) or Europe (latitude: 0.10 days/degree [-0.23 - 0.42]; MAT: -0.16 days/$^{\circ}$C [-0.55 - 0.23]) (Fig.\ref{figure:continental_spptype_lat}A \& \ref{figure:continental_spptype_effect}A).
+Overall, spring events such as budburst and leaf flush, were not related to provenance latitude or MAT, neither across continents (latitude: 0.10 days/degree [-0.05 - 0.25]; MAT: -0.11 days/$^{\circ}$C [-0.34 - 0.12]) (Fig.\ref{figure:springfall_latmat}, Table.S\ref{supplement-table:model_spring_lat} \& S\ref{supplement-table:model_spring_mat} in Supporting Information), nor within North America (latitude: 0.10 days/degree [-0.06 - 0.26]; MAT: -0.09 days/$^{\circ}$C [-0.36 - 0.18]) or Europe (latitude: 0.10 days/degree [-0.23 - 0.42]; MAT: -0.16 days/$^{\circ}$C [-0.55 - 0.23]) (Fig.\ref{figure:continental_spptype_lat}A \& \ref{figure:continental_spptype_effect}A).
 Results were similar using other distance metrics in lieu of latitude (see Fig.S\ref{supplement-figure:lat_distance} for results using the difference between provenance and garden latitude, and the spherical distance between provenance and garden).

### 2024-02-25 Docs/manuscriptRev.Rnw
Removed text from two captions

@@
  \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/continental_spptype_lat.jpg}
-    \caption{Effects of latitude on spring and fall event date (DOY) depending on (a \& b) continent, and (c \& d) species leaf type shown as posterior distributions (full distribution represents 99\% percentile, solid line and shading in posteriors represent mean and 50 percent interval, 90\% intervals given in text).
+    \caption{Effects of latitude on spring and fall event date (DOY) depending on (a \& b) continent, and (c \& d) species leaf type (gymnosperms, which were all evergreen species, versus angiosperms, which were all deciduous species) shown as posterior distributions (full distribution represents 99\% percentile, solid line and shading in posteriors represent mean and 50 percent interval, 90\% intervals given in text).
 Zero---no effect---is shown with a dashed line.
-Spring events did not shift with latitude by continent, but fall events advanced strongly per degree northward, particularly in North America.
+} %RM for editor: Spring events did not shift with latitude by continent, but fall events advanced strongly per degree northward, particularly in North America.
 Spring events slightly advanced in gymnosperms (all evergreen species) and delayed in angiosperms (all deciduous species) per degree northward.
-Fall events advanced per degree northward for both species types.} %EWMFeb5 -- more edits for angio/gymno
+Fall events advanced per degree northward for both species types.
 \label{figure:continental_spptype_lat}
@@
  \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/continental_spptype_effect.jpg}
-    \caption{Effects of MAT on spring and fall event date (DOY) depending on (a \& b) continent, and (c \& d) species leaf type shown as posterior distributions (full distribution represents 99\% percentile, solid line and shading in posteriors represent mean and 50 percent interval, 90\% intervals given in text).
+    \caption{Effects of MAT on spring and fall event date (DOY) depending on (a \& b) continent, and (c \& d) species leaf type (gymnosperms, which were all evergreen species, versus angiosperms, which were all deciduous species)  shown as posterior distributions (full distribution represents 99\% percentile, solid line and shading in posteriors represent mean and 50 percent interval, 90\% intervals given in text).
 Zero---no effect---is shown with a dashed line.
-Spring events did not shift with MAT by continent, but fall events advanced strongly with decreasing MAT, particularly notably in North America.
+} % RM per editor request: Spring events did not shift with MAT by continent, but fall events advanced strongly with decreasing MAT, particularly notably in North America.
 Spring events slightly advanced in angiosperms and delayed in gymnosperms with increasing MAT.
-Fall events delayed with increasing MAT for both species types.}
+Fall events delayed with increasing MAT for both species types.
     \label{figure:continental_spptype_effect}

### 2024-02-26 Docs/manuscriptRev.Rnw
Revision bound 2

@@
     \caption{Effects of latitude on spring and fall event date (DOY) depending on (a \& b) continent, and (c \& d) species leaf type (gymnosperms, which were all evergreen species, versus angiosperms, which were all deciduous species) shown as posterior distributions (full distribution represents 99\% percentile, solid line and shading in posteriors represent mean and 50 percent interval, 90\% intervals given in text).
-Zero---no effect---is shown with a dashed line.
-} %RM for editor: Spring events did not shift with latitude by continent, but fall events advanced strongly per degree northward, particularly in North America.
+Zero---no effect---is shown with a dashed line.}
+    %RM for editor: Spring events did not shift with latitude by continent, but fall events advanced strongly per degree northward, particularly in North America.
 Spring events slightly advanced in gymnosperms (all evergreen species) and delayed in angiosperms (all deciduous species) per degree northward.
@@
 Zero---no effect---is shown with a dashed line.
-} % RM per editor request: Spring events did not shift with MAT by continent, but fall events advanced strongly with decreasing MAT, particularly notably in North America.
+}
+    % RM per editor request: Spring events did not shift with MAT by continent, but fall events advanced strongly with decreasing MAT, particularly notably in North America.
 Spring events slightly advanced in angiosperms and delayed in gymnosperms with increasing MAT.

### 2024-02-26 Docs/manuscriptRev.tex
Revision bound 2

@@
  \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/continental_spptype_lat.jpg}
-    \caption{Effects of latitude on spring and fall event date (DOY) depending on (a \& b) continent, and (c \& d) species leaf type shown as posterior distributions (full distribution represents 99\% percentile, solid line and shading in posteriors represent mean and 50 percent interval, 90\% intervals given in text).
-Zero---no effect---is shown with a dashed line.
-Spring events did not shift with latitude by continent, but fall events advanced strongly per degree northward, particularly in North America.
+    \caption{Effects of latitude on spring and fall event date (DOY) depending on (a \& b) continent, and (c \& d) species leaf type (gymnosperms, which were all evergreen species, versus angiosperms, which were all deciduous species) shown as posterior distributions (full distribution represents 99\% percentile, solid line and shading in posteriors represent mean and 50 percent interval, 90\% intervals given in text).
+Zero---no effect---is shown with a dashed line.}
+    %RM for editor: Spring events did not shift with latitude by continent, but fall events advanced strongly per degree northward, particularly in North America.
 Spring events slightly advanced in gymnosperms (all evergreen species) and delayed in angiosperms (all deciduous species) per degree northward.
-Fall events advanced per degree northward for both species types.} %EWMFeb5 -- more edits for angio/gymno
+Fall events advanced per degree northward for both species types.
     \label{figure:continental_spptype_lat}
@@
  \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/continental_spptype_effect.jpg}
-    \caption{Effects of MAT on spring and fall event date (DOY) depending on (a \& b) continent, and (c \& d) species leaf type shown as posterior distributions (full distribution represents 99\% percentile, solid line and shading in posteriors represent mean and 50 percent interval, 90\% intervals given in text).
+    \caption{Effects of MAT on spring and fall event date (DOY) depending on (a \& b) continent, and (c \& d) species leaf type (gymnosperms, which were all evergreen species, versus angiosperms, which were all deciduous species)  shown as posterior distributions (full distribution represents 99\% percentile, solid line and shading in posteriors represent mean and 50 percent interval, 90\% intervals given in text).
 Zero---no effect---is shown with a dashed line.
-Spring events did not shift with MAT by continent, but fall events advanced strongly with decreasing MAT, particularly notably in North America.
+}
+    % RM per editor request: Spring events did not shift with MAT by continent, but fall events advanced strongly with decreasing MAT, particularly notably in North America.
 Spring events slightly advanced in angiosperms and delayed in gymnosperms with increasing MAT.
-Fall events delayed with increasing MAT for both species types.}
+Fall events delayed with increasing MAT for both species types.
     \label{figure:continental_spptype_effect}
